### 2013-06-11 paper.tex
Stub of paper, and list of what to do

\input a4
\input xpmath
\input respnotes

\title{The instability of variational quantum mechanics with Gaussian approximants is (not) entirely due to the unstable least squares problem}

Coherent states have been considered as an approximant for variational quantum mechanics since 197?.  Superpositions of them are an obvious way to extend the approximant, so that the variational dynamics converge to Hamiltonian evolution.  However, these dynamics are numerically unstable, and it has not been possible to increase the size of the superposition to see this convergence.  The coherent states are overcomplete, and it has often been suggested that this instability is due to the variational problem being underdetermined.

A numerical experiment was carried out to test this hypothesis.  The dynamics of a Morse oscillator were computed by three different methods.  First, a harmonic potential was fit to the bottom of the Morse potential, the state and the Hamiltonian were expanded over its eigenstates, and Schrödinger's equation was integrated directly in Hilbert space.  Secondly, differential equations were derived for the variational parameters in the standard way, using the normal equations of the regularised variational problem, and these were solved in parameter space.  The third method was intermediate: Schrödinger's equation was solved in Hilbert space, but the state was approximated by a superposition of coherent states, updated by least squares at each timestep.  Finally, the eigenstates of this potential are known exactly, and the state was expanded over those.

In a sense, expanding over Fock states is cheating.  If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis would be unmanagably large.  In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.  However, this would raise the question whether those least-square solutions from the normal equations were stable.  By using an orthonormal basis, we can be more confident of this.

A few different ODE solving algorithms were used, with different tolerance for stiff ODEs.  Adaptive 4th/5th order Runge-Kutta was used.  This is the algorithm that everyone will try by default, and it would be useful to know if it performs much worse than other algorithms.  ? and ? were used for comparison.  These discretisations have very different regions of stability ...

The solutions from method 3 provide a test of the stiffness of the ODEs defined in parameter space by the least squares problem.  The step in Hilbert space is very nearly the same from one time step to the next.  If the step in parameter space changes dramatically, this would suggest that the variational solution is very sensitive to the state, and thus to some of the parameters.  This would indicate that the ODEs are stiff.

The initial state was the ground state of the Morse oscillator potential, given various momentum kicks.  The number of components in the superposition representing this was varied.  The initial superposition was found by the AMPL optimisation program, with the constraint that no component would have a weight less that ?.

I want to measure the numerical stability of the variational dynamics derived from Schrödinger's equation.  I can do that by measuring how the discretisation error converges as the timestep is reduced.  The second hypothesis is that the instability is due to stiffness, which can be tested by discretising time in ways with a variety of stability regions, and seeing whether the stability differs.

\bye

### 2013-06-11 paper.tex
task

@@
 
+To do this, we need some bracket matrices.  The representation used is
+$$|ψ(φ,α)〉=e^{φ+αa†}|0〉=e^{φ+½|α|²}|α〉,$$
+whence
+$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
+The bracket with the number states is easily evaluated:
+$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
+as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
+$$〈n|H|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
+
 \bye

### 2013-06-11 paper.tex
task

--- 2013-06-11 paper.tex
@@
 
+The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian $$H=\hbar a^{2\dagger}a²$.  It will do so by the midpoint method, discretising the differential equation
+$$Du(t)=f(t)$$
+to the difference equation
+$$v^{n+1}=v^{n-1}+2τfⁿ,$$
+where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  The results will be compared for three difference choices of $v$ and $f$.
+
+In the first choice, the quantum state at time $t$ is expanded over a truncated Fock basis $|N〉$, as 
+$$|ν(c(t))〉=|N〉c(t).$$
+Schrödinger's equation expands to
+$$|N〉Dc(t)=|D(ν\circ c)(t)〉≈{H\over i\hbar}|ν(c(t))〉≈|N〉〈N|{H\over i\hbar}|N〉c(t),$$
+whence
+$$Dc(t)≈〈N|{H\over i\hbar}|N〉c(t).$$
+In this method, we set
+$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=〈N|{H\over i\hbar}|N〉c(t),$$
+solving Schrödinger's equation quite conventionally.
+
+In the second choice, the state is approximated as a superposition of coherent states,
+$$|ψ(z(t))〉=∑_{φ,α∈z(t)}e^{φ+αa†}|0〉=∑_{φ,α∈z(t)}e^{φ+½|α|²}|α〉.$$
+Time is discretised immediately, setting
+$$vⁿ=|ψ(zⁿ)〉\qquad{\rm and}\qquad fⁿ={H\over i\hbar}|ψ(zⁿ)〉,$$
+so that the midpoint formula becomes
+$$|ψ(z^{n+1})〉≈|ψ(z^{n-1})〉+{2τH\over i\hbar}|ψ(z^ⁿ)〉.$$
+The approximation is due to basis truncation, as always in a least squares sense.  Writing the unknown $|ψ(z^{n+1})〉$ in terms of $|Dψ(zⁿ)〉$ and expanding over a Fock basis gives
+$$ 〈N|Dψ(zⁿ)〉\bigl(z^{n+1}-zⁿ\bigr)≈〈N|ψ(z^{n-1})〉-〈N|ψ(zⁿ)〉-2iτ〈N|H|ψ(zⁿ)〉.$$
+This matrix equation can be solved for $z^{n+1}-zⁿ$, which can be used to update $z$.
+
+The third choice, the conventional variational one, is to discretise the state first, setting
+$$vⁿ=zⁿ\qquad{\rm and}\qquad |Dψ(zⁿ)〉fⁿ≈{H\over i\hbar}|ψ(zⁿ)〉.$$
+The midpoint formula is used exactly, and $fⁿ$ satisfies the least squares problem
+$$〈N|Dψ(zⁿ)〉fⁿ≈-i〈N|Dψ(zⁿ)〉{H\over \hbar}|ψ(zⁿ)〉.$$
+Usually, this would be expanded over $〈Dψ(zⁿ)|$ to give a set of normal equations.
+However, when $|Dψ(zⁿ)〉$ is near singular, the numerical stability of the normal equations is suspect.
+Expanding over an orthonormal basis removes this confounding instability.
+
+To do this, we need some bracket matrices.  We have
+$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
+The bracket with the number states is easily evaluated:
+$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
+as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
+$$〈n|H|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
+
 \bye

### 2013-06-12 paper.tex
task

@@
 
-The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian $$H=\hbar a^{2\dagger}a²$.  It will do so by the midpoint method, discretising the differential equation
+The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian 
+$$H=\hbar a^{2\dagger}a².$$
+It will do so by the midpoint method, discretising the differential equation
 $$Du(t)=f(t)$$
@@
 so that the midpoint formula becomes
-$$|ψ(z^{n+1})〉≈|ψ(z^{n-1})〉+{2τH\over i\hbar}|ψ(z^ⁿ)〉.$$
+$$|ψ(z^{n+1})〉≈|ψ(z^{n-1})〉+{2τH\over i\hbar}|ψ(zⁿ)〉.$$
 The approximation is due to basis truncation, as always in a least squares sense.  Writing the unknown $|ψ(z^{n+1})〉$ in terms of $|Dψ(zⁿ)〉$ and expanding over a Fock basis gives
@@
 $$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
-The bracket with the number states is easily evaluated:
+The brackets with the number states are
+$$〈m|H/\hbar|n〉=n(n-1)δ_{mn}$$
+and
 $$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
 as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
-$$〈n|H|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
+$$〈n|H/\hbar|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
 

### 2013-06-12 brackets.py
Bracket code written

@@ -1,0 +1,47 @@
+"""Calculate the required brackets for the quartic oscillator.
+
+Throughout, the Fock basis comprises |0> through |N>.  The variable N must be defined before this is imported.
+
+The convention is that psi is a function of (psi0, psi1, ... psi_R, alpha0, alpha1, ..., alpha_R)
+"""
+
+N = 5
+
+from numpy import abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander, vstack, zeros_like
+from scipy.misc import factorial
+
+nhn = arange(N+1)*arange(-1,N)
+"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
+
+nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
+ndqr = matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
+"""sum(nq*evan(f,a), axis=1) gives the bracket <N|psi>.
+    hstack((nq*evan(f,a), ndqr*evan(f,a))) gives <N|Dpsi>
+"""
+
+def evan(f, a):
+	expf = exp(f)[newaxis,:]
+	va = vander(a, N+1)[:,::-1].T
+	return matrix(expf*va)
+
+	
+nhq = matrix(diag(sqrt(concatenate(([0, 0], [m*(m-1)/factorial(m-2) for m in range(2,N+1)])))))
+"""sum(nhq*evan(f,a), axis=1) gives the brackets of the Hamiltonian between number states and the superposition.
+	"""
+
+
+def check(f, a):
+	"""Test routine.  Answers should be on the order of epsilon times the infinity norms of f and a."""
+	R = len(f)
+	x = hstack((nq*evan(f,a), ndqr*evan(f,a)))
+	for m in range(N+1):
+		for i in range(R):
+			x[m,i] = x[m,i] - exp(f[i])*a[i]**m/sqrt(factorial(m))
+	for m in range(1,N+1):
+		for i in range(R):
+			x[m,i+R] = x[m,i+R] - exp(f[i])*a[i]**(m-1)*sqrt(m/factorial(m-1))
+	y = nhq*evan(f,a)
+	for m in range(2,N+1):
+		for i in range(R):
+			y[m,i] = y[m,i] - exp(f[i])*a[i]**m*sqrt(m*(m-1)/factorial(m-2))
+	return max(abs(x)), max(abs(y))

### 2013-06-12 paper.tex
Bracket code written
First integration script

@@
 In this method, we set
-$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=〈N|{H\over i\hbar}|N〉c(t),$$
-solving Schrödinger's equation quite conventionally.
+$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=-i〈N|{H\over\hbar}|N〉c(t),$$
+solving Schrödinger's equation quite conventionally.  Note that $|N〉$ diagonalises $H$, so this is especially simple.
 

### 2013-06-12 brackets.py

@@ -1,20 +1,23 @@
 """Calculate the required brackets for the quartic oscillator.
 
-Throughout, the Fock basis comprises |0> through |N>.  The variable N must be defined before this is imported.
+Throughout, the Fock basis comprises |0> through |N>.
 
 The convention is that psi is a function of (psi0, psi1, ... psi_R, alpha0, alpha1, ..., alpha_R)
+
+These definitions will generally be copied rather than imported, because of the difficulties of passing around N.
 """
 
-N = 5
-
-from numpy import abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander, vstack, zeros_like
+from numpy import abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander
 from scipy.misc import factorial
 
-nhn = arange(N+1)*arange(-1,N)
-"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
+def get_nhn(N): 
+	"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
+	return arange(N+1)*arange(-1,N)
 
-nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
-ndqr = matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
+def get_nq(N):
+	return matrix(diag(1./sqrt(factorial(arange(N+1)))))
+def get_ndqr(N):
+	return matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
 """sum(nq*evan(f,a), axis=1) gives the bracket <N|psi>.
     hstack((nq*evan(f,a), ndqr*evan(f,a))) gives <N|Dpsi>
 """
@@ -25,10 +28,14 @@ def evan(f, a):
 	return matrix(expf*va)
 
 	
-nhq = matrix(diag(sqrt(concatenate(([0, 0], [m*(m-1)/factorial(m-2) for m in range(2,N+1)])))))
+def get_nhq(N):
+	return matrix(diag(sqrt(concatenate(([0, 0], [m*(m-1)/factorial(m-2) for m in range(2,N+1)])))))
 """sum(nhq*evan(f,a), axis=1) gives the brackets of the Hamiltonian between number states and the superposition.
 	"""
 
+def get_aop(N):
+	"Lowering operator"
+	return matrix(diag(sqrt(range(1,N)), 1))
 
 def check(f, a):
[…]

### 2013-06-12 dry.py

@@ -1,0 +1,31 @@
+"""Dry run of the numerical experiment, integrating a quartic oscillator in a Fock basis by the explicit Euler formula."""
+
+from numpy import arange, pi, zeros, abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander
+from scipy.misc import factorial
+import matplotlib.pyplot as plt
+
+N = 20
+dt = 0.1
+ts = dt*arange(100)
+ats = zeros(ts.shape, dtype=complex)
+
+nhn = matrix(diag(arange(N+1)*arange(-1,N)))
+nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
+aop = matrix(diag(sqrt(range(1,N+1)), 1))
+
+def evan(f, a):
+	expf = exp(f)[newaxis,:]
+	va = vander(a, N+1)[:,::-1].T
+	return matrix(expf*va)
+
+a0 = 2
+c = nq*evan([-0.5*2**2], [2])
+
+for m in range(100):
+	ats[m] = (c.H*aop*c)[0,0]
+	c = c-1j*nhn*c*dt
+	
+plt.semilogy(ts, ats.real)
+plt.show()
+
+	

### 2013-06-13 brackets.py
Rewrote brackets to be far too clever for its own good

@@ -1,4 +1,4 @@
-"""Calculate the required brackets for the quartic oscillator.
+"""Initialise the required brackets for the quartic oscillator.
 
 Throughout, the Fock basis comprises |0> through |N>.
 
@@ -10,34 +10,37 @@ These definitions will generally be copied rather than imported, because of the 
 from numpy import abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander
 from scipy.misc import factorial
 
-def get_nhn(N): 
-	"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
-	return arange(N+1)*arange(-1,N)
-
-def get_nq(N):
-	return matrix(diag(1./sqrt(factorial(arange(N+1)))))
-def get_ndqr(N):
-	return matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
-"""sum(nq*evan(f,a), axis=1) gives the bracket <N|psi>.
-    hstack((nq*evan(f,a), ndqr*evan(f,a))) gives <N|Dpsi>
-"""
+_names = "N", "nhn", "nq", "ndqr", "nhq", "aop"
+for x in _names:
+	globals()[x] = None
 
 def evan(f, a):
 	expf = exp(f)[newaxis,:]
 	va = vander(a, N+1)[:,::-1].T
 	return matrix(expf*va)
 
-	
-def get_nhq(N):
-	return matrix(diag(sqrt(concatenate(([0, 0], [m*(m-1)/factorial(m-2) for m in range(2,N+1)])))))
-"""sum(nhq*evan(f,a), axis=1) gives the brackets of the Hamiltonian between number states and the superposition.
+def init(N):
+	global nhn, nq, ndqr, nhq, aop
+
+	globals()["N"] = N
+
+	"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
+	nhn = arange(N+1)*arange(-1,N)
+
+	"""sum(nq*evan(f,a), axis=1) gives the bracket <N|psi>.
+	    hstack((nq*evan(f,a), ndqr*evan(f,a))) gives <N|Dpsi>
 	"""
+	nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
+	ndqr = matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
 
-def get_aop(N):
[…]

### 2013-06-13 paper.tex
stiffness of Schrodinger equation demonstrated

@@
 
-The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian 
+The first experiment computed the dynamics of a quartic oscillator, with Hamiltonian 
 $$H=\hbar a^{2\dagger}a².$$
-It will do so by the midpoint method, discretising the differential equation
+It used the explicit Euler formula, which discretises the differential equation
 $$Du(t)=f(t)$$
@@
 $$v^{n+1}=v^{n-1}+2τfⁿ,$$
-where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  The results will be compared for three difference choices of $v$ and $f$.
+where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.
 
@@
 solving Schrödinger's equation quite conventionally.  Note that $|N〉$ diagonalises $H$, so this is especially simple.
+
+\centerline{\XeTeXpicfile dry.pdf width \hsize}
 

### 2013-06-13 brackets.py

@@ -7,17 +7,13 @@ The convention is that psi is a function of (psi0, psi1, ... psi_R, alpha0, alph
 These definitions will generally be copied rather than imported, because of the difficulties of passing around N.
 """
 
-from numpy import abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander
-from scipy.misc import factorial
 
-_names = "N", "nhn", "nq", "ndqr", "nhq", "aop"
-for x in _names:
-	globals()[x] = None
+import numpy as _np, scipy.misc as _sp
 
 def evan(f, a):
-	expf = exp(f)[newaxis,:]
-	va = vander(a, N+1)[:,::-1].T
-	return matrix(expf*va)
+	expf = _np.exp(f)[_np.newaxis,:]
+	va = _np.vander(a, N+1)[:,::-1].T
+	return _np.matrix(expf*va)
 
 def init(N):
 	global nhn, nq, ndqr, nhq, aop
@@ -25,20 +21,20 @@ def init(N):
 	globals()["N"] = N
 
 	"""Diagonal brackets of the quartic oscillator Hamiltonian between number states."""
-	nhn = arange(N+1)*arange(-1,N)
+	nhn = _np.arange(N+1)*_np.arange(-1,N)
 
 	"""sum(nq*evan(f,a), axis=1) gives the bracket <N|psi>.
 	    hstack((nq*evan(f,a), ndqr*evan(f,a))) gives <N|Dpsi>
 	"""
-	nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
-	ndqr = matrix(diag(sqrt(arange(1,N+1)/factorial(arange(N))), -1))
+	nq = _np.matrix(_np.diag(1./_np.sqrt(_sp.factorial(_np.arange(N+1)))))
+	ndqr = _np.matrix(_np.diag(_np.sqrt(_np.arange(1,N+1)/_sp.factorial(_np.arange(N))), -1))
 
 	"""sum(nhq*evan(f,a), axis=1) gives the brackets of the Hamiltonian between number states and the superposition.
 	"""
-	nhq = matrix(diag(sqrt(concatenate(([0, 0], [m*(m-1)/factorial(m-2) for m in range(2,N+1)])))))
+	nhq = _np.matrix(_np.diag(_np.sqrt(_np.concatenate(([0, 0], [m*(m-1)/_sp.factorial(m-2) for m in range(2,N+1)])))))
 
[…]

### 2013-06-13 dry.py

@@ -1,31 +1,32 @@
-"""Dry run of the numerical experiment, integrating a quartic oscillator in a Fock basis by the explicit Euler formula."""
+"""Dry run of the numerical experiment, integrating a quartic
+oscillator in a Fock basis by the explicit Euler formula."""
 
-from numpy import arange, pi, zeros, abs, arange, concatenate, diag, exp, hstack, matrix, max, newaxis, sqrt, sum, vander
-from scipy.misc import factorial
+import brackets; brackets.init(10); from brackets import *
+from numpy import array, abs, arange, ceil, exp, logspace, pi, zeros
 import matplotlib.pyplot as plt
 
-N = 20
-dt = 0.1
-ts = dt*arange(100)
-ats = zeros(ts.shape, dtype=complex)
-
-nhn = matrix(diag(arange(N+1)*arange(-1,N)))
-nq = matrix(diag(1./sqrt(factorial(arange(N+1)))))
-aop = matrix(diag(sqrt(range(1,N+1)), 1))
-
-def evan(f, a):
-	expf = exp(f)[newaxis,:]
-	va = vander(a, N+1)[:,::-1].T
-	return matrix(expf*va)
-
+T = 4*pi
 a0 = 2
-c = nq*evan([-0.5*2**2], [2])
+c0 = array(nq*evan([-0.5*2**2], [2])).flatten()
 
-for m in range(100):
-	ats[m] = (c.H*aop*c)[0,0]
-	c = c-1j*nhn*c*dt
+lbl = "$c_{%d}$"
+mrk = {2:"ok", 4:"pb", 7:"sg", N:"*r"}
+
+for dt in logspace(-4, 1, 20):
+
+	ts = dt*arange(ceil(T/dt))
+	c = c0.copy()
+	for m in range(ts.size):
+		c = c-1j*nhn*c*dt
 	
-plt.semilogy(ts, ats.real)
-plt.show()
+	# FIXME: matplotlib mixes up legend unless orders match 
+	for n in sorted(mrk.keys()):
+		plt.loglog(dt, abs(c[n]/c0[n]), mrk[n], label=lbl % n)
+
+plt.title("Stiffness of Schrodinger's equation with $H=\hbar a^{2\dagger}a^2$")
+plt.xlabel(r"Euler formula step $\tau$")
+plt.ylabel("Growth of Fock coefficients $c_n(4\pi)/c_n(0)$")
[…]

### 2013-06-13 dry.py
writeup

@@ -25,7 +25,7 @@ for dt in logspace(-4, 1, 20):
 
 plt.title("Stiffness of Schrodinger's equation with $H=\hbar a^{2\dagger}a^2$")
 plt.xlabel(r"Euler formula step $\tau$")
-plt.ylabel("Growth of Fock coefficients $c_n(4\pi)/c_n(0)$")
+plt.ylabel("Growth of Fock coefficients $|c_n(4\pi)|/c_n(0)$")
 plt.legend([lbl % n for n in sorted(mrk.keys())], loc="upper left")
 plt.savefig("dry.pdf")
 

### 2013-06-14 dry.py
neater code

@@ -9,8 +9,8 @@ T = 4*pi
 a0 = 2
 c0 = array(nq*evan([-0.5*2**2], [2])).flatten()
 
-lbl = "$c_{%d}$"
 mrk = {2:"ok", 4:"pb", 7:"sg", N:"*r"}
+lbl = "$c_{%d}$"
 
 for dt in logspace(-4, 1, 20):
 
@@ -19,7 +19,8 @@ for dt in logspace(-4, 1, 20):
 	for m in range(ts.size):
 		c = c-1j*nhn*c*dt
 	
-	# FIXME: matplotlib mixes up legend unless orders match 
+	# sorted is a workaround for
+	# https://github.com/matplotlib/matplotlib/issues/2127
 	for n in sorted(mrk.keys()):
 		plt.loglog(dt, abs(c[n]/c0[n]), mrk[n], label=lbl % n)
 

### 2013-06-14 paper.tex

--- 2013-06-12 paper.tex
@@
 
-The first attempt at this will compute the dynamics of a quartic oscillator, with Hamiltonian 
-$$H=\hbar a^{2\dagger}a².$$
-It will do so by the midpoint method, discretising the differential equation
-$$Du(t)=f(t)$$
-to the difference equation
-$$v^{n+1}=v^{n-1}+2τfⁿ,$$
-where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  The results will be compared for three difference choices of $v$ and $f$.
-
-In the first choice, the quantum state at time $t$ is expanded over a truncated Fock basis $|N〉$, as 
+The first experiment computed the dynamics of a quartic oscillator, with Hamiltonian 
+$$H=\hbar a^{2\dagger}a²,$$
+solving Schrödinger's equation quite conventionally.
+The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
 $$|ν(c(t))〉=|N〉c(t).$$
 Schrödinger's equation expands to
-$$|N〉Dc(t)=|D(ν\circ c)(t)〉≈{H\over i\hbar}|ν(c(t))〉≈|N〉〈N|{H\over i\hbar}|N〉c(t),$$
+$$|N〉Dc(t)=|D(ν\circ c)(t)〉={H\over i\hbar}|ν(c(t))〉=|N〉〈N|{H\over i\hbar}|N〉c(t),$$
 whence
-$$Dc(t)≈〈N|{H\over i\hbar}|N〉c(t).$$
-In this method, we set
-$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=〈N|{H\over i\hbar}|N〉c(t),$$
-solving Schrödinger's equation quite conventionally.
+$$Dc(t)=-i〈N|{H\over \hbar}|N〉c(t).$$
+Usually, these expansions would be approximate due to basis truncation.
+In this case, the Fock states diagonalise $H$, so they are exact.
+However, in general the initial state can only be approximately expanded over $|N\rangle$.
+Time was discretised with the explicit Euler formula, which replaces the differential equation
+$$Du(t)=f(u(t))$$
+with a difference equation
+$$v^{n+1}=vⁿ+τfⁿ,$$
+where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  Let
+$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=-i〈N|{H\over\hbar}|N〉c(t).$$
+Since $〈N|H|N〉$ is diagonal, the matrix product can be performed sparely as a dot product.
+The results are shown below.
+The Fock states are eigenstates of the Hamiltonian, so their amplitudes should be constant: in the exact solution of Schrödinger's equation, all the ratios plotted are exactly 1.
+
+%\centerline{\XeTeXpicfile dry.pdf width \hsize}
+
+These results are the textbook signs of stiffness.
+As the time step is reduced from values on the order of 1, the discretised solutions diverge hugely, before converging at very small time steps.
+In the standard theory of discrete approximations to $Du(t)=f(u(t))$, stability is determined by the eigenvalues of the linear transform $Df$.
+In Schrödinger's equation, $f$ is the Hamiltonian, a linear transform, so in our expansion, $df=f=-i〈N|H/\hbar|N〉$, and the relevant eigenvalues are $-iEⁿ/\hbar=-in(n-1)$.
+The stiffness of a differential equation is often measured by the ratio of the sizes of these eigenvalues.
+For the quartic oscillator, some eigenvalues are zero, so this is not a sensible measure.
+However, the largest eigenvalue, which often determines the timestep, increases quadratically with the size of the Fock basis as $N(N-1)$.
+
+The stability region of the explicit Euler formula is ?.  This means ?.
 

### 2013-06-17 paper.tex
semi-implicit win

@@
 
-The stability region of the explicit Euler formula is ?.  This means ?.
+The stability region of the explicit Euler formula is the disk $|\bar k-(-1)|≤1|$.
+So it is unstable for any Jacobian with complex eigenvalues, which is certainly the case for a Hamiltonian.
+The implicit Euler formula,
+$$v^{n+1}=vⁿ+τf^{n+1},$$
+has the opposite stability region.  For the oscillator, this becomes
+$$c^{n+1}=cⁿ-iτ〈N|{H\over\hbar}|N〉c^{n+1},$$
+so $c^{n+1}$ satisfies the linear equation
+$$(1+iτ〈N|{H\over\hbar}|N〉)c^{n+1}=cⁿ.$$
+This is still diagonal for the quartic oscillator.  To order $τ$, the implicitn formula is the same as the explicit one.  Obviously the higher powers of $τ$ make the difference when $τ$ is finite.
+
+The results of the backward Euler formula are the opposite of the Euler formula.
+Instead of increasing, the amplitudes of the larger number states decrease nearly to zero.
+This distorts the trajectory of $〈a(t)〉$, as show below.
+
+A third method is the semi-implicit one, with the formulae
+$$v^{\prime n}=vⁿ+½τf^{\prime n}\qquad v^{n+1}=vⁿ+τf^{\prime n}.$$
+With the Hamiltonian, this becomes
+$$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$
 

### 2013-06-17 dry.py

@@ -2,32 +2,70 @@
 oscillator in a Fock basis by the explicit Euler formula."""
 
 import brackets; brackets.init(10); from brackets import *
-from numpy import array, abs, arange, ceil, exp, logspace, pi, zeros
+from numpy import *
+from scipy.special import cbrt
 import matplotlib.pyplot as plt
 
-T = 4*pi
-a0 = 2
+T = pi
 c0 = array(nq*evan([-0.5*2**2], [2])).flatten()
+dts = logspace(-4, 0, 20)
+# dts = logspace(-3, 0, 3)
 
-mrk = {2:"ok", 4:"pb", 7:"sg", N:"*r"}
-lbl = "$c_{%d}$"
+mrk = {2:"o", 4:"p", 7:"s", N:"*"}
+sym = {}
 
-for dt in logspace(-4, 1, 20):
+rs = 0.5*T*(1+cbrt(linspace(-1, 1, 100)))
+ct = exp(-1j*outer(nhn,rs))*c0[:,newaxis]
+exact_ars = diag(matrix(ct).H*aop*matrix(ct))
 
-	ts = dt*arange(ceil(T/dt))
-	c = c0.copy()
-	for m in range(ts.size):
-		c = c-1j*nhn*c*dt
+splt = 0
+for dt in dts:
 	
-	# sorted is a workaround for
-	# https://github.com/matplotlib/matplotlib/issues/2127
-	for n in sorted(mrk.keys()):
-		plt.loglog(dt, abs(c[n]/c0[n]), mrk[n], label=lbl % n)
+	ts = dt*arange(ceil(T/dt))
+	ars = zeros(ts.shape, dtype=complex)
+	brs = zeros(ts.shape, dtype=complex)
+	j = 0
+	c = c0.copy()
+	d = c0.copy()
+	b = c0.copy()
 
+	hsi = 1-1j*dt*nhn/(1+0.5*1j*dt*nhn)
+	
+	for m in range(ts.size):
+		ars[m] = (matrix(d[:,newaxis]).H*aop*matrix(d[:,newaxis]))[0,0]
+		brs[m] = (matrix(b[:,newaxis]).H*aop*matrix(b[:,newaxis]))[0,0]
+		c = c-1j*nhn*c*dt	# Euler
+		d = d/(1+1j*nhn*dt)	# backward Euler
+		b = hsi*b			# semi-implicit
+	
+	plt.figure(1)
[…]

### 2013-06-17 dry.py
Going home

@@ -40,13 +40,12 @@ for dt in dts:
 	
 	plt.figure(1)
 	for n in mrk.keys():
-		plt.loglog(dt, abs(b[n]/c0[n]), mrk[n]+"b")
+		plt.loglog(dt, abs(b[n]/c0[n]), "+b")
 		plt.loglog(dt, abs(c[n]/c0[n]), mrk[n]+"r")
 		sym[n], = plt.loglog(dt, abs(d[n]/c0[n]), mrk[n]+"k")
 		
 	plt.figure(2)
 	plt.subplot(4,5,20-splt)
-	splt = splt + 1
 	plt.axis('off')
 	plt.axis('equal')
 	plt.plot(exact_ars.real, exact_ars.imag, 'k-', ars.real, ars.imag, 'r-', brs.real, brs.imag, 'b-')
@@ -54,6 +53,7 @@ for dt in dts:
 		plt.title(r"$\tau=%.4f$" % dt)
 	else:
 		plt.title("$%.4f$" % dt)
+	splt = splt + 1
 
 
 plt.figure(1)

### 2013-06-18 paper.tex
write up changes from original plan

@@
 
-\title{The instability of variational quantum mechanics with Gaussian approximants is (not) entirely due to the unstable least squares problem}
+\title{Stability of variational quantum dynamics with Gaussian approximants}
 
@@
 A numerical experiment was carried out to test this hypothesis.
-The dynamics of a Morse oscillator were computed by three different methods.
+The dynamics of a quartic oscillator were computed by three different methods.
 First, a harmonic potential was fit to the bottom of the Morse potential, the state and the Hamiltonian were expanded over its eigenstates, and Schrödinger's equation was integrated directly in Hilbert space.
@@
 The third method was intermediate: Schrödinger's equation was solved in Hilbert space, but the state was approximated by a superposition of coherent states, updated by least squares at each timestep.
-Finally, the eigenstates of this potential are known exactly, and the state was expanded over those.
+Finally, the eigenstates of this potential are known exactly: the state was expanded over them, and the exact dynamics calculated for comparison.
 
 In a sense, expanding over Fock states is cheating.
-If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis would be unmanagably large.
+If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis such as the Fock states would be unmanagably large.
 In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.
@@
 
-A few different ODE solving algorithms were used, with different tolerance for stiff ODEs.
-Adaptive 4th/5th order Runge-Kutta was used.
-This is the algorithm that everyone will try by default, and it would be useful to know if it performs much worse than other algorithms.
-?
-and ?
-were used for comparison.
-These discretisations have very different regions of stability ...
-
-The solutions from method 3 provide a test of the stiffness of the ODEs defined in parameter space by the least squares problem.
-The step in Hilbert space is very nearly the same from one time step to the next.
-If the step in parameter space changes dramatically, this would suggest that the variational solution is very sensitive to the state, and thus to some of the parameters.
-This would indicate that the ODEs are stiff.
-
-The initial state was the ground state of the Morse oscillator potential, given various momentum kicks.
-The number of components in the superposition representing this was varied.
-The initial superposition was found by the AMPL optimisation program, with the constraint that no component would have a weight less that ?.
-
-I want to measure the numerical stability of the variational dynamics derived from Schrödinger's equation.
-I can do that by measuring how the discretisation error converges as the timestep is reduced.
-The second hypothesis is that the instability is due to stiffness, which can be tested by discretising time in ways with a variety of stability regions, and seeing whether the stability differs.
-
-The first experiment computed the dynamics of a quartic oscillator, with Hamiltonian 
+I used three discretisation formulae for ODE, with different regions of stability.
+As a first step, the Hamiltonian and a coherent state with real amplitude 2 was expanded over Fock states $|0〉$ through $|10〉$, and the resulting equations for the components solved directly.
+The Hamiltonian is
 $$H=\hbar a^{2\dagger}a²,$$
-solving Schrödinger's equation quite conventionally.
+so the Fock states are eigenstates, although the spacing of their energies is uneven.
 The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
@@
 $$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$
+If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.
+The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.
+Different eigenstates circle the origin with different frequencies.
+
+The semi-implicit formula has a special property: when the trajectory of $u(t)$ is a circle, the $vⁿ$ lie exactly on that circle, and the discretisation error is equivalent to rescaling time by a factor that depends on the frequency and the discrete timestep.
+In Hilbert space, this means that the projections of the state onto eigenplanes still circle the origin exactly: only the frequencies are altered by discretisation.
+So the effect of discretising quantum dynamics using the semi-implicit formula is merely to shift the energies of the eigenstates.
+Of course, this can be a significant effect: it's the only difference between the harmonic and quartic oscillators.
+
+The solutions from method 3 provide a test of the stiffness of the ODEs defined in parameter space by the least squares problem.
+The step in Hilbert space is very nearly the same from one time step to the next.
+If the step in parameter space changes dramatically, this would suggest that the variational solution is very sensitive to the state, and thus to some of the parameters.
+This would indicate that the ODEs are stiff.
+
+The initial state was the ground state of the Morse oscillator potential, given various momentum kicks.
+The number of components in the superposition representing this was varied.
+The initial superposition was found by the AMPL optimisation program, with the constraint that no component would have a weight less that ?.
+
+I want to measure the numerical stability of the variational dynamics derived from Schrödinger's equation.
+I can do that by measuring how the discretisation error converges as the timestep is reduced.
+The second hypothesis is that the instability is due to stiffness, which can be tested by discretising time in ways with a variety of stability regions, and seeing whether the stability differs.
 

### 2013-06-20 brackets.py
switched to Matlab

@@ -15,6 +15,7 @@ def evan(f, a):
 	va = _np.vander(a, N+1)[:,::-1].T
 	return _np.matrix(expf*va)
 
+
 def init(N):
 	global nhn, nq, ndqr, nhq, aop
 
@@ -36,18 +37,19 @@ def init(N):
 	"Lowering operator"
 	aop = _np.matrix(_np.diag(_np.sqrt(_np.arange(1,N+1)), 1))
 
+
 def _check(f, a):
 	"""Test routine.  Answers should be on the order of epsilon times the infinity norms of f and a."""
 	R = len(f)
-	x = hstack((nq*evan(f,a), ndqr*evan(f,a)))
+	x = _np.hstack((nq*evan(f,a), ndqr*evan(f,a)))
 	for m in range(N+1):
 		for i in range(R):
-			x[m,i] = x[m,i] - exp(f[i])*a[i]**m/sqrt(factorial(m))
+			x[m,i] = x[m,i] - _np.exp(f[i])*a[i]**m/_np.sqrt(_sp.factorial(m))
 	for m in range(1,N+1):
 		for i in range(R):
-			x[m,i+R] = x[m,i+R] - exp(f[i])*a[i]**(m-1)*sqrt(m/factorial(m-1))
+			x[m,i+R] = x[m,i+R] - _np.exp(f[i])*a[i]**(m-1)*_np.sqrt(m/_sp.factorial(m-1))
 	y = nhq*evan(f,a)
 	for m in range(2,N+1):
 		for i in range(R):
-			y[m,i] = y[m,i] - exp(f[i])*a[i]**m*sqrt(m*(m-1)/factorial(m-2))
-	return max(abs(x)), max(abs(y))
+			y[m,i] = y[m,i] - _np.exp(f[i])*a[i]**m*_np.sqrt(m*(m-1)/_sp.factorial(m-2))
+	return _np.max(_np.abs(x)), _np.max(_np.abs(y))

### 2013-06-20 brackets.m
First Matlab version

@@ -1,0 +1,9 @@
+% The brackets <N|psi> are sum(nq*evan(f,a),2)
+%    <N|Dpsi> is [nq*evan(f,a), ndqr*evan(f,a)]
+
+global N aop nhn nq ndqr
+
+aop = diag(sqrt(1:N),1);
+nhn = ((0:N).*(-1:N-1)).';
+nq = diag(1./sqrt(factorial(0:N)));
+ndqr = diag(sqrt((1:N)./factorial(0:N-1)),-1);

### 2013-06-20 check.m

@@ -1,0 +1,25 @@
+function [dc, dd, npsi, ndpsi] = check(f,a)
+
+global N nq ndqr
+
+npsi = zeros(N+1,1);
+for m = 0:N
+    for r = 1:numel(a)
+        npsi(m+1) = npsi(m+1) + exp(f(r))*a(r)^m/sqrt(factorial(m));
+    end
+end
+dc = norm(npsi-sum(nq*evan(f,a),2));
+
+ndpsi = zeros(N+1,2*numel(a));
+for m = 0:N
+    for r = 1:numel(a)
+        ndpsi(m+1,r) = exp(f(r))*a(r)^m/sqrt(factorial(m));
+    end
+end
+for m = 1:N
+    for r = 1:numel(a)
+        ndpsi(m+1,r+numel(a)) = ...
+            exp(f(r))*a(r)^(m-1)*sqrt(m/factorial(m-1));
+    end
+end
+dd = norm(ndpsi-[nq*evan(f,a), ndqr*evan(f,a)]);

### 2013-06-20 cohfit.m

@@ -1,0 +1,32 @@
+function [z, rsdl] = cohfit(cs, z0)
+% fit a superposition of coherent states to the fock expansion cs,
+% starting from z.
+
+options = optimset('Display','iter','Algorithm','active-set');
+[z, rsdl] = fmincon(@(z)cost(z),z0,...
+                    [],[],[],[],[],[],@constraint,options);
+                
+                
+function y = cost(z)
+% Distance of z from state cs
+
+global nq
+
+R = 0.5*numel(z);
+y = norm(sum([nq*evan(z(1:R),z(R+1:2*R)),-cs],2));
+
+end
+
+function [y,yeq] = constraint(z)
+% positive when any component of z has norm less than 0.1
+
+global nq
+
+R = 0.5*numel(z);
+ncpts = sum(abs(nq*evan(z(1:R),z(R+1:2*R))).^2);
+y = 0.1^2 - min(ncpts);
+yeq = [];
+
+end
+
+end

### 2013-06-20 evan.m

@@ -1,0 +1,5 @@
+function X = evan(f,a)
+
+global N
+
+X = ((ones(N+1,1)*a).^((0:N)'*ones(size(a))))*diag(exp(f));

### 2013-06-20 projerr.m

@@ -1,0 +1,71 @@
+N = 10; brackets
+T = pi;
+dts = logspace(-3,-1,6);
+
+% amplitude 2 coherent state and 5-component approximation
+c0 = nq*evan(-2,2);
+a0 = 2+[0.5*1i.^(0:3), 0];  R = numel(a0);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+z = cohfit(c0, [f0,a0]); f0 = z(1:R); a0 = z(R+1:2*R);
+c0 = sum(nq*evan(f0,a0),2);
+
+% exact solution
+rs = 0.5*pi*(1+nthroot(linspace(-1, 1, 100), 3));
+crs = diag(c0)*exp(-1i*nhn*rs);
+ars = diag(crs'*aop*crs);
+
+
+h1 = figure;
+set(gcf,'DefaultAxesDataAspectRatio',[1 1 1], ...
+    'DefaultAxesVisible','off','DefaultAxesColor','none')
+h2 = figure;
+
+for p=1:numel(dts)
+    figure(h1); subplot(2,3,7-p);
+    dt = dts(p);
+    ts = (0:T/dt)*dt;
+    as = zeros(2, numel(ts)); as(:) = nan;
+    rsdl = zeros(2, numel(ts)); rsdl(:) = nan;
+    
+    f = f0; a = a0; sty1 = 'g-';
+    for q=1:numel(ts)
+        c = sum(nq*evan(f,a),2);
+        as(1,q) = c'*aop*c;
+        dc = expm1(-1i*nhn*dt).*exp(-1i*nhn*ts(q)).*c0;
+        ndq = [nq*evan(f,a), ndqr*evan(f,a)];
+        if any(isnan(ndq(:))) || abs(as(1,q)) > 4
+            as(1,q) = nan; sty1 = 'r-';
+            plot(real(as(1,q-1)), imag(as(1,q-1)), 'or'); hold on
+            break
+        end
+        dz = pinv(ndq, 0.1)*dc;
+        f = f + dz(1:R).'; a = a + dz(R+1:2*R).';
+        rsdl(1,q) = norm(dc-ndq*dz)/norm(dc);
+    end
+    
+    f = f0; a = a0; sty2 = 'b-';
+    for q=1:numel(ts)
+        c = sum(nq*evan(f,a),2);
+        as(2,q) = c'*aop*c;
[…]

### 2013-06-20 cohfit.m
refactored code works

@@ -5,6 +5,7 @@ function [z, rsdl] = cohfit(cs, z0)
 options = optimset('Display','iter','Algorithm','active-set');
 [z, rsdl] = fmincon(@(z)cost(z),z0,...
                     [],[],[],[],[],[],@constraint,options);
+z = z.';
                 
                 
 function y = cost(z)

### 2013-06-20 projerr.m

@@ -1,71 +1,46 @@
+global N
 N = 10; brackets
 T = pi;
 dts = logspace(-3,-1,6);
+h1 = figure; h2 = figure;
+set(h1,'DefaultAxesDataAspectRatio',[1 1 1], ...
+    'DefaultAxesVisible','off','DefaultAxesColor','none')
 
 % amplitude 2 coherent state and 5-component approximation
 c0 = nq*evan(-2,2);
-a0 = 2+[0.5*1i.^(0:3), 0];  R = numel(a0);
+a0 = 2+[0.5*1i.^(0:3), 0];
 f0 = -0.5*abs(a0).^2;
 f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
-z = cohfit(c0, [f0,a0]); f0 = z(1:R); a0 = z(R+1:2*R);
-c0 = sum(nq*evan(f0,a0),2);
+z0 = cohfit(c0, [f0,a0]);
 
-% exact solution
+% exact solutions
 rs = 0.5*pi*(1+nthroot(linspace(-1, 1, 100), 3));
 crs = diag(c0)*exp(-1i*nhn*rs);
 ars = diag(crs'*aop*crs);
 
-
-h1 = figure;
-set(gcf,'DefaultAxesDataAspectRatio',[1 1 1], ...
-    'DefaultAxesVisible','off','DefaultAxesColor','none')
-h2 = figure;
-
+% computing
+fmt = '%.3f';
 for p=1:numel(dts)
-    figure(h1); subplot(2,3,7-p);
     dt = dts(p);
-    ts = (0:T/dt)*dt;
-    as = zeros(2, numel(ts)); as(:) = nan;
-    rsdl = zeros(2, numel(ts)); rsdl(:) = nan;
-    
-    f = f0; a = a0; sty1 = 'g-';
-    for q=1:numel(ts)
-        c = sum(nq*evan(f,a),2);
-        as(1,q) = c'*aop*c;
-        dc = expm1(-1i*nhn*dt).*exp(-1i*nhn*ts(q)).*c0;
-        ndq = [nq*evan(f,a), ndqr*evan(f,a)];
-        if any(isnan(ndq(:))) || abs(as(1,q)) > 4
-            as(1,q) = nan; sty1 = 'r-';
-            plot(real(as(1,q-1)), imag(as(1,q-1)), 'or'); hold on
-            break
-        end
-        dz = pinv(ndq, 0.1)*dc;
[…]

### 2013-06-20 propagate.m

@@ -1,0 +1,28 @@
+function [ts, ms, rs] = propagate(dfun, mfun, z0, T, dt)
+% Propagate the superposition z0 by applying dfun(t,c) to its Fock
+% expansion, and projecting back to a superposition at each step.
+% Return the results of applying mfun to the fock expansion, at
+% each step.  Note that dfun is a differential, not a derivative.
+% The results are passed to the continuations sfun or ffun as the
+% integration succeeds or fails, and the result of that call returned.
+
+global N nq ndqr
+
+R = 0.5*numel(z0);
+ts = (0:T/dt+1)*dt;
+ms = zeros(numel(ts), numel(mfun(zeros(N+1,1))));
+rs = zeros(numel(ts), 1);
+f = z0(1:R).'; a = z0(R+1:2*R).';
+for q=1:numel(ts)
+    ndq = [nq*evan(f,a), ndqr*evan(f,a)];
+    if any(isnan(ndq(:)))
+        ts(q:end) = nan; ms(q:end) = nan; rs(q:end) = nan;
+        return 
+    end
+    c = sum(ndq(:,1:R),2);
+    ms(q,:) = mfun(c);
+    dc = dfun(ts(q), c);
+    dz = pinv(ndq, 0.1)*dc;
+    f = f + dz(1:R).'; a = a + dz(R+1:2*R).';
+    rs(q) = norm(dc-ndq*dz)/norm(dc);
+end

### 2013-06-24 paper.tex
writeup

@@
 By using an orthonormal basis, we can be more confident of this.
+
+Expanding the quartic oscillator over Fock states allows it to be solved very stably, and as precisely as desired by rotating the phases of the expansion coefficients.
+Also, it is easy to find the expansion over Fock states of a superpositon of coherent states.
+This means that the stability of time discretisation can be removed from the problem entirely.
+The state can be propagated exactly, and the coherent states adjusted to fit.
+The most direct way to do so is by nonlinear optimisation at every timestep; Matlab provides the necessary routines in a convenient package.
+Unfortunately, this is not at all stable.
+
+\centerline{\XeTeXpicfile flwc0n10.pdf width \hsize}
+
+\centerline{\XeTeXpicfile flwc0n5.pdf width \hsize}
+
+The above plots show my first attempt at this.
+The initial state was coherent, with real amplitude 2.
+A superposition of 5 coherent states was initialised with amplitudes normally distributed around 2, the norm of the superposition being 1, and the components having equal norms.
+This superposition was fit to the coherent state by minimising their distance in Hilbert space; for stability, the weights were constrained so that no component had a norm less than 0.1.
+This was done by the Matlab routine {\tt fmincon}.
+The initial state was truncated to Fock states $|0〉$ to $|10〉$, and propagated exactly in a Hamiltonian $\hbar a^{2\dagger}a²$ with timestep 0.03.
+After each step, the fitting of the superposition to the exact state was repeated.
+
+The results are shown in the first Figure.
+The left hand plot shows the complex amplitude of the exact solution in black, and the fitted solution in blue.
+The right hand plot shows the residual of the fit as a solid line, and the condition number of $|Dψ(z)〉$ at the fitted $z$ as a dotted line.
+The program stopped after a short time, when the fitted amplitudes caused overflow when expanded over Fock coefficients.
+
+The second Figure shows the same simulation, with the states expanded over Fock states $|0〉$ to $|5〉$.
+This has a significant effect on the truncated exact solution, adding an extra loop to the amplitude.
+The fitting works a bit longer, but still fails at a very short time.
+
+\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
+
+\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
+
+The next two plots show a different fitting strategy.
+Instead of constraining the weights, a cost term was added in quadrature to the Hilbert space distance, proportional to the logarithm of the condition number of $|Dψ(z)〉$ for the fitted $z$.
+The constant of proportionality was $10^{-5}$; this was adjusted until a good fit was obtained.
+The most apparent thing is that the fit is much more stable: this is being done with timestep 0.1 instead of 0.3, but the fit stays stable for the whole period.
+(I had to redraw the initial amplitudes a few times before that happened.)  It isn't a very good fit through the middle of the period: the residual increases to order 1.
+The really interesting thing is what happened when, by accident, I truncated the Fock basis at $N=5$.
+As before, this perturbs the trajectory in Hilbert space.
+However, the perturbed trajectory can be fit very well: the residual is between $10^{-2}$ and $10^{-5}$, and this is done with quite well conditioned superpositions.
+
+So even without stiffness in a discrete time propagation, there is something in the high energy components of $H|ψ〉$ that makes fitting the state with coherent states unstable.
+This isn't too surprising: these high energy components are rapidly oscillating, and will by magnified when the inverse problem is solved with fairly smooth, low amplitude coherent states.
+To fit the oscillating components, we need coherent states with large amplitudes, and low weights.
+But the low weights will increase the condition number of $|Dψ〉$, so the regularisation scheme will avoid these, unless there are many components in the superposition and all their weights are low.
+Let's investigate what happens as the number of components and the Fock truncation are varied together.
 
@@
 
+The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
+If a singular value decomposition $|Dψ〉=|U〉SV†$ is taken, the singularity is indicated by small singular values on the diagonal of $S$.
+It is often measured by the condition number of $|Dψ〉$, the ratio of the largest to the smallest singular value.
+Let $C(z)$ be the condition number of $|Dψ(z)〉$.
+As part of the regularisation process, it would be useful to avoid changing $z$ in directions that increase $C(z)$.
+For example, if two coherent states had real amplitudes, and the real part of the expected amplitude needed to increase, it would be done by shifting the right hand one away from the left hand one, not the left hand one closer to the right hand.
+Is there a way to make a computer do that?
+
+The least squares problem can be solved using the singular value decomposition: to approximate a state $|h〉≈|Dψ(z)〉dz$, we set $dz=VS^{-1}〈U|h〉$.
+The first problem is that some of the $σ_i$ on the diagonal of $S$ will be small, because the $|u_i〉$ that correspond to them change different components of the superposition $|ψ(z)〉$ in ways that nearly cancel out.
+So a error in the discrete representation of $|h〉$ in the direction $|u_i〉$ will cause a very large error in the least squares $dz$ in the direction $v_i$.
+The way to avoid such large errors is to shift the small singular values of $S$ up to a lower bound $ε$, and thus reduce the large singular values of $S^{-1}$ below an upper bound $ε^{-1}$.
+There are many and varied ways of doing that.
+
+To solving our differential equation, we need to solve an iterated least squares problem.
+We need to do so in a way that preserves the condition of $|Dψ〉$.
+When we shift $z$ in the direction $v_i$ as $dz=λv_i$, how does this condition number change?
+Without loss of generality, it suffices to consider the effect on the smallest and largest singular values, that determine $C(z)$.
+Note that $v_1$ and $v_R$ are eigenvectors of the matrix ${\cal V}(z)=〈Dψ(z)|Dψ(z)〉$, belonging to the eigenvalues $σ₁²$ and $σ_R²$; this is a familiar problem in pertubation theory.
+Assume for now that we have found $D{\cal V}(z)$—this won't be analytic, so we will need to use real and imaginary parts, or Wirtinger calculus.
+The change in $\cal V$ is of the form $d{\cal V}=∂_i{\cal V}(z)λ+∂*_i{\cal V}(z)λ*$, where the partial derivatives are taken in the directions $v_i$.
+
+As is well known in pertubation theory, the change in the eigenvalue $σ₁²$ corresponding to a small change $d{\cal V}$ is $v₁†d{\cal V}v₁$, and similarly for $σ_R²$.
+Therefore the change in $C(z)²$ is
+$$σ_R²v₁†d{\cal V}v₁-σ₁²v_R†d{\cal V}v_R \over σ_R².$$
+Given a proposed change $dz$, we can find the effect on the condition number.
+
+The question is how to apply this to condition the process of finding $dz$.
+Given $z$, we can find out which directions of $dz$ correspond to large changes in the condition number.
+However,  until we know $dz$, we don't know if these changes help or hurt.
+The process might run like this: first, truncate the singular values.
+Next, find the components $〈U_i|h〉$, and the corresponding components $v_iσ_i〈U_i|h〉$ of $dz$, given the truncation.
+Then, truncate the components that will cause $C(z)$ to increase to values that limits the increase to an acceptable amount.
+
+Coding this will be quite involved, and the result might not be very efficient.  However, if it allows physically chosen bases to be used in  variational simulations, it is likely to be worthwhile.
+
+\centerline{\XeTeXpicfile j1.pdf width \hsize}
+
+To test this idea, a five component superposition was fit to an amplitude 2 coherent state.
+The Hilbert space residual was $1.3×10^{-3}$.
+The amplitudes and norms of the components are shown above; the fitting routine constrained the norm of each component to be at least 0.1.
+Note that components 1 and 3 ended up with very similar amplitudes.
+
+Below is a plot of the condition number of $|Dψ(z)〉$ for this superposition, as the value of $z$ varies by $0.1$ in every direction.
+The top line shows the effects of changing $φ$, the bottom line $α$, components are shown left to right.
+Changes to real parts are black, imaginary parts red.
+There are a few things to note.
+The condition is very bad, with number $1.4×10^{11}$.
+This is entirely due to the similar amplitudes of components 1 and 3: those graphs are on a different scale to the others, and it's logarithmic.
+As they move closer and become identical, the condition number increases $10^{14}$.
+No doubt it would go higher if we sampled the graphs more finely, but this is huge—reciprocal floating point epsilon is $5×10^{15}$.
+If the separation between these components is doubled, the condition number improves by an order of magnitude.
+
+\centerline{\XeTeXpicfile j2.pdf width \hsize}
+
+The top line indicates that the condition number is independent of the phases of the components.
+I didn't forsee that, but it isn't too surprising: we can change the coefficients of the superposition in any direction we like, regardless of their current values, and the derivative is only sensitive to changes.
+
+The experiment with following the exact solution by nonlinear optimisation was repeated.
+Instead of constraining the weights of the components, the logarithm of the condition number of the derivative was added in quadrature to the residual of the fit to give a cost.
+This worked much better, getting stably to $t=0.4$ instead of under 0.1.
+At that point, the residuals increased to order 1, and the superposition stopped following the exact solution.
+This cold mean that 5 components weren't enough to represent those states, except that the same thing happened with 10 components.
+
+Note that the optimisation is a nonlinear least squares problem, and special solvers exist for this case.
+
+Some furt
+
 \bye

### 2013-06-24 cohfit.m

@@ -1,8 +1,8 @@
-function [z, rsdl] = cohfit(cs, z0)
+function [z, rsdl] = cohfit(cs, z0, ~)
 % fit a superposition of coherent states to the fock expansion cs,
 % starting from z.
 
-options = optimset('Display','iter','Algorithm','active-set');
+options = optimset('Display','off','Algorithm','active-set');
 [z, rsdl] = fmincon(@(z)cost(z),z0,...
                     [],[],[],[],[],[],@constraint,options);
 z = z.';

### 2013-06-24 wobble.m
breakthrough

@@ -1,0 +1,46 @@
+% Evolve a coherent state for t=0.1, then fit a superposition of five
+% coherent states to the result, truncated to an N=5 Fock basis.  Then,
+% for the expansions N=6:10, plot the components of vector from the 
+% superposition to the truncated state
+% along the singular directions of |DQ>.
+
+global N
+R = 5;
+w = 1e-5;
+t = 0.1;
+
+% amplitude 2 coherent state and 5-component approximation
+N = 5; brackets
+c = exp(-1i*nhn*t).*(nq*evan(-2,2));
+
+% 5-component approximation, randomly spread
+a0 = 2+0.3*randn(1,R)+0.3i*randn(1,R);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+
+figure(1)
+plot(real(a), imag(a), '*k')
+title(rsdl)
+
+[z, rsdl] = cohfit2(c, [f0,a0], w);
+f = z(1:R).'; a = z(R+1:2*R).';
+
+for N=5:2:11
+    brackets
+    c = exp(-1i*nhn*t).*(nq*evan(-2,2));
+    cprox = sum(nq*evan(f,a), 2);
+    [U,S,V] = svd([nq*evan(f,a), ndqr*evan(f,a)]);
+    incpt = U'*(c-cprox);
+    outcpt = S\incpt;
+    subplot(3,4,0.5*(N-3))
+    semilogy(1:numel(incpt), abs(incpt), '+k')
+    title(sprintf('N=%d', N))
+    if N==5; ylabel('\psi components'); end
+    subplot(3,4,0.5*(N-3)+4)
+    semilogy(1:numel(outcpt), abs(outcpt), '+r')
+    if N==5; ylabel('z components'); end
+    subplot(3,4,0.5*(N-3)+8)
+    semilogy(1:numel(diag(S)), diag(S), 'or')
+    if N==5; ylabel('singular values'); end
+end
+    

### 2013-06-28 paper.tex
figured out SVE of coherent state expansion.

@@
 By using an orthonormal basis, we can be more confident of this.
+
+The variational methods are based on the phase space mapping $G$ from square-integrable functions to Fock space, where
+$$Gf=∫f(α*)|α〉\,d²α.$$
+This is different from Glauber's use: my $f$ is his $fe^{-½|α|²}$.  (Check the half.)
+
+Hypothesis: The singular values of $G$ are 0 and $\sqrt π$, and the space belonging to $\sqrt π$ comprises the functions $f$ for which $fe^{½|α|²}$ is analytic.
+
+There are two ways one could set about proving this.
+Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
+Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
+ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
+
+This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
+My numerical experiment suggests otherwise; let's ignore that possibility for now.
+
+Let $g$ be a normalised entire function, in Glauber's sense.  Our expansion function is 
+$$f(α) = {1\over π}g(α)e^{-½|α|²}.$$
+If $g$ has the Taylor expansion
+$$g(z)=∑_n c_n{zⁿ\over(n!)^½},$$
+then 
+$$Gf = ∑_n c_n|n〉.$$
+The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
+$$f(α)={1\over π}e^{-½|α|²}{αⁿ\over(n!)^½}.$$
+We have 
+$$|f(α)|²={1\over π²n!}e^{-|α|²}|α|^{2n},$$
+so the squared norm is
+$$\|f\|²={1\over π²n!}∫₀^∞2πr^{2n+1}e^{-r²}\,dr={2\over πn!}∫₀^∞r^{2n+1}e^{-r²}\,dr.$$
+According to Wolfram, this is 
+$$\|f\|²={1\over π},$$
+whence
+$$\|f\|=π^{-½}.$$
+
+Note that the gaussian entire functions are a vector subspace of the square integrable functions.
+
+Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
+Or at least I think so.
+I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
 
@@
 Let's investigate what happens as the number of components and the Fock truncation are varied together.
+
+The things I've noticed so far are:
+\item{1.} When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.
+It isn't accurate, though.
+
+\item{2.} Measuring state differences in a truncated fock space makes the fit stable and accurate.
+Of course, the truncated state is different from the original one, so the original state is not fit precisely.
+When this was done, the least squares problem involved 6 fock coefficients, and the amplitude vector has 10 elements.
+So almost all the truncated problems were underdetermined: some singular values of the Jacobian are exactly zero.
+Matlab claims that backslash solves underdetermined problems by giving a solution with as many zero elements as possible.
+The documentation doesn't say how it chooses which elements to make zero.
+That's different from the other regularisation methods we've tried.
+Of course, the problems are being solved inside Matlab's optimisation routine, and I don't know what it's doing.
+
+Reducing the dimension of Fock space so that the problem is underdetermined means that, when the problem is solved by SVD, we only take components along $N+1$ directions in Fock space, and map them to $N+1$ directions in parameter space.
+The remaining $2R-N-1$ components of the solution are zero.
+As we add extra dimensions to Fock space, we are taking components in more Fock space directions, which include more oscillating components of the wavefunction.
+This is like sampling a function on a finer grid.
+With these directions, we can form some of the smaller singular vectors of the full problem.
+
+Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
+(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
+
+As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
+Presumably these converge to limits as $N$ increases.
+There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
+
+\item{3.} The difference between the truncated state and the original one fails to satisfy the Picard condition when with respect to $|Dψ〉$.  What about $H|ψ〉$?
+
+\item{4.} When the least squares problem is solved with a different distance function, which sets distances along the large $N$ directions to zero, the fit fails on the first step.
+Note that this isn't really a distance.
+Also, I did it by solving $P|h〉=P|Dψ〉dz$; when $P$ truncates the Fock space, the operator $P|Dψ〉$ is singular.
+Doing a weighted least squares fit, with the many particle directions having low weights, would be near-singular.
 

### 2013-06-28 brackets.m

@@ -1,9 +1,10 @@
 % The brackets <N|psi> are sum(nq*evan(f,a),2)
 %    <N|Dpsi> is [nq*evan(f,a), ndqr*evan(f,a)]
 
-global N aop nhn nq ndqr
+global N aop nhn nq ndqr nwgt
 
 aop = diag(sqrt(1:N),1);
 nhn = ((0:N).*(-1:N-1)).';
 nq = diag(1./sqrt(factorial(0:N)));
 ndqr = diag(sqrt((1:N)./factorial(0:N-1)),-1);
+nwgt = eye(N+1);

### 2013-06-28 paper.tex
Fitting states with randomly drawn initial guesses

@@
 \title{Stability of variational quantum dynamics with Gaussian approximants}
+
+When physicists began to study quantum mechanics, they could hardly have avoided discovering coherent states.
+When they grasped that observables were inherently uncertain and considered laws of probability, the normal distribution sprung to mind at once.
+Once they understood the dynamics of the free particle, and looked for a more interesting system to study, the obvious choice was the harmonic oscillator.
+The first question was its ground state; the second was why, in the classical limit, it oscillated.
+Coherent states were the answer to both.
+
+Variational mechanics was another obvious idea.
+The quantum equations of motion were hard to solve exactly, except for some very simple systems.
+It was known from very early on that the solutions were extema of certain properties: ground states had the lowest possible energy, trajectories the least possible action.
+So it was natural to consider a class of possible solutions that could be treated exactly, and find out which of them had the extreme value over the class.
+
+In the 1970s, Heller put the two ideas together, and proposed
 

### 2013-07-04 paper.tex
writing

@@
 It was known from very early on that the solutions were extema of certain properties: ground states had the lowest possible energy, trajectories the least possible action.
-So it was natural to consider a class of possible solutions that could be treated exactly, and find out which of them had the extreme value over the class.
+So it was natural to consider a class of approximants that could be treated exactly, and find out which of them had the extreme value over the class.
 
-In the 1970s, Heller put the two ideas together, and proposed
+In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using coherent states as the approximant.
+He eventually discovered that related methods had been used to find the dynamics of lasers, starting with Glauber's work in the 19?0s on expansions over coherent states, and continuing when ?
+derived the dynamics of these expansions, and solved for the quantum dyanmics of the laser.
 
-Coherent states have been considered as an approximant for variational quantum mechanics since 197?.
-Superpositions of them are an obvious way to extend the approximant, so that the variational dynamics converge to Hamiltonian evolution.
-However, these dynamics are numerically unstable, and it has not been possible to increase the size of the superposition to see this convergence.
-The coherent states are overcomplete, and it has often been suggested that this instability is due to the variational problem being underdetermined.
+When used on computers, the variational methods expand quantum states over finitely many coherent states.
+This is a discretisation of the expansion introduced by Glauber, a mapping $G$ from square-integrable functions to Fock space, where
+$$Gf={1\over π}∫g(α*)|α〉\,d²α.\eqno{(1)}$$
+This is different from Glauber's use: he uses $f=ge^{½|α|²}$.
+Is it easier to work Glauber's way, with the $e^{-½|α|²}$ in the mapping, so that the singular vectors are entire functions?
+You'd have to use a Bargmann inner product to normalise them; the theory of that is well developed.
 
-A numerical experiment was carried out to test this hypothesis.
-The dynamics of a quartic oscillator were computed by three different methods.
-First, a harmonic potential was fit to the bottom of the Morse potential, the state and the Hamiltonian were expanded over its eigenstates, and Schrödinger's equation was integrated directly in Hilbert space.
-Secondly, differential equations were derived for the variational parameters in the standard way, using the normal equations of the regularised variational problem, and these were solved in parameter space.
-The third method was intermediate: Schrödinger's equation was solved in Hilbert space, but the state was approximated by a superposition of coherent states, updated by least squares at each timestep.
-Finally, the eigenstates of this potential are known exactly: the state was expanded over them, and the exact dynamics calculated for comparison.
+Variational dynamics seeks an expansion $g$ to approximate a quantum state $|ψ〉≈Gg$.
+Given the form of Equation~1, this requires solving a Fredholm integral equation of the first kind.
+That is the canonical example of an ill-posed problem: rounding and discretisation errors in $|ψ〉$ are magnified enormously in the solution $g$.
+To solve these problems, it helps to understand the singular values and vectors of the operator $G$.
+
+The singular value expansion of the linear operator $G$ has the form
+$$Gf = ∑_{j=1}^∞ |g_j〉σ_j〈v_j,f〉,$$
+where the left singular vectors $|g_j〉$ are orthonormal kets, the singular values $σ_j$ are positive real numbers, the right singular vectors $v_j$ are normalised entire functions, and presumably $〈,〉$ is the Bargmann inner product.
+This can be written
+$$G = |G〉SV†,$$
+where $|G〉$ is a row vector of the left singular kets, $S$ is a matrix with the singular values along the diagonal, and $V$ maps complex coefficients to linear combinations of the functions $v_j$.
+
+Because the expansions are underdetermined, the set $|g_j〉$ will be complete, but the set $v_j$ will not be.
+Alternatively, some of the $v_j$ have zero singular values and no corresponding left singular vector.
+One can write a full SVE, where $V†$ is a full expansion, but $S$ truncates the vector of coefficients.
+
+The singular value expansion is useful for two reasons.
+Firstly, the discrete version, the singular value decomposition, is the most stable way to solve badly-posed least squares problems.
+Secondly, the sizes of the singular values, and the components of $H|ψ〉$ along the singular directions, determine whether the problem can be solved at all.
+
+The most important issue is the Picard condition.
+This restates a very basic observation: for the problem $G\dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $f$ over an orthonormal basis must have a finite norm.
+If the basis is chosen as the right singular vectors $v_j$, this means that the sum of the squared moduli of the components must be finite.
+For $\dot f$ to satisfy the equation, the component of $v_j$ is $〈g_j|H|ψ〉/σ_j$.
+So the Picard condition is that, for the equation to have a solution, the series
+$$∑_{j=1}^∞{\bigl|〈g_j|H|ψ〉\bigr|²\over σ_j²}$$
+must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.
+
+Hypothesis: The singular values of $G$ are 0 and $\sqrt π$, and the space belonging to $\sqrt π$ comprises the functions $f$ for which $fe^{½|α|²}$ is analytic.
+
+This claims that $G$ is a projection: if the inner product is normalised properly, $G$ keeps the component of $f$ in the space of entire functions, and discards the orthogonal component.
+In that case, we'd have a simple underdetermined problem, which could be regularised very easily by discarding the small singular values.
+But that isn't what I see with discrete versions: they seem to have a steady decay of singular values, not a sudden drop to zero.
+Is that an artefact of how I'm discretising?
+
+The theorems on how quickly singular values of convolution operators decay don't apply here.
+They apply to convolutions of real functions, which can only oscillate at a fixed rate.
+Complex functions have space in the Argand plane to oscillate arbitrarily rapidy; considered as wavefunctions, the coherent states do: the wavenumber is proportional to the imaginary part of $α$.
+
+If the Hypothesis is right, the Picard condition is very obviously satisfied.
+The operator $G$ is a projection.
+The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm.
+Functions in the null space of $G$ don't correspond to left singular vectors, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
+
+There are two ways one could set about proving the Hypothesis.
+Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
+Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
+ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
+
+This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
+My numerical experiment suggests otherwise; let's ignore that possibility for now.
+
+Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
+$$f(z)=∑c_n{zⁿ\over (n!)^½},$$
+for some normalised vector $c$.
+
+  Our expansion function is 
+$$g(α)=f(α)e^{-½|α|²}.$$
+If $f$ has the Taylor expansion
+$$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
+then 
+$$Gg = ∑_n c_n|n〉.$$
+The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
+$$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
+the Glauber preimage of $|n〉$.  We have 
+$$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
+so the squared norm is
+$$\|g\|²={1\over n!}∫₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}∫₀^∞r^{2n+1}e^{-r²}\,dr.$$
+According to Wolfram,
+$$∫₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
+so
+$$\|g\|=π^½.$$
+
+The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
+Numerical experiments in the script {\tt sve.m} support this.
+The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
+Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
+The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
+
+\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
+
+The following three figures show the right singular vectors, sampled versions of phase space functions.
+The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
+These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
+In the continuous limit there is only one singular value, and the vectors are degenerate.
+For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
+As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
+
+\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
+\XeTeXpicfile rsv3.pdf width 0.5\hsize}
+\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+
+Note that the gaussian entire functions are a vector subspace of the square integrable functions.
+
+Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
+Or at least I think so.
+I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
+
+To sum up, this analysis shows that $G$ is a projector.
+Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.
+This might depend somewhat on how evenly the amplitudes are spaced: I could start with a regular grid, and add Gaussian noise to see what changed.
+
+There are two types of ill-posed problems, rank-deficient and discrete.
+They are distingished by how the singular values decay to zero.
+In a rank-deficient problem, there is a sharp cut between large and small singular values.
+The large ones are large enough that components in their singular directions do not contribute excessively to the solution, the small ones would contribute huge components, but they are small enough to round to zero, and ignore the component along that singular direction.
+In discrete problems, the singular values decay continuously.
+The size of the components of the solution varies continuously: there is no point where they become huge, but a continuous minimum and a gradual rise as the singular values get smaller.
+
+I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.
+However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.
+This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
+$$f(t)=∫F(ω)e^{iωt}\,dω.$$
+This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.  (What is the SVD of the DFT matrix?)
+
+The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.
+I was working in a truncated Fock space; I wonder what happens as the truncation increases, while the grid stays a constant size?
+Or alternately, if the grid shrinks while the Fock space stays constant.
+
+I'm getting a picture of how the final algorithm will look.
+At each timestep, the directions belonging to large singular values of $|Dψ〉$ will be used for variational dynamics.
+This can be done by solving at Tychonov conditioned problem, for example.
+At the same time, the directions with small singular values will be used to keep the discretisation stable.
+I think this requires two things: that $|Dψ〉$ is reasonably well conditioned, and that the angle between $H|ψ〉$ and the span of $|Dψ〉$ is small.
+This raises a question: are there singular directions that have little effect on $|ψ〉$, but a large effect on $|Dψ〉$?
+This seems likely; for example, shuffling the amplitudes to move two components further apart in phase space, then adjusting the weights so that $|ψ〉$ stays the same.
+This could be done by a kind of Tychonov anti-regularisation: we improve the stability as much as we can, while constraining the change in $|ψ〉$.
+I might need to constrain the norm of $dz$ too, so that linear approximations hold.
+
+I've already figured out how to find the change in the condition number by pertubation theory.
+That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.
+How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?
+This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.
+Is there a better way?
+
+The final algorithm would propagate $|ψ〉$ in Hilbert space by a semi-implicit method, solving a variational problem to represent each state.
+This is a slight generalisation, because you need to project $H|ψ_{n+½}〉$ in to the span of $|Dψ_n〉$, and the ensembles at the two times are slightly different.
+It might help to propagate from $n$ to $n+½$ by changing only weights, leaving the same ensemble.
+On the other hand, the change is only that the $α*$ become $β*$, so it should be managable.
+
+There shouldn't be an need for fancy methods to fit the initial state: one can take a guess $z₀$, and variationally solve the equation $|Dψ(z)〉dz=|ψ₀〉-|ψ(z)〉$.
+One could even change the conditioning rules as the state propagates, so that $|Dψ〉$ is aligned with $|ψ₀〉-|ψ(z)〉$ to start with, then with $H|ψ(z)〉$ later on.
+That way, you start the actual simulation with a well conditioned approximant.
+
+Expanding the quartic oscillator over Fock states allows it to be solved very stably, and as precisely as desired by rotating the phases of the expansion coefficients.
+Also, it is easy to find the expansion over Fock states of a superpositon of coherent states.
+This means that the stability of time discretisation can be removed from the problem entirely.
+The state can be propagated exactly, and the coherent states adjusted to fit.
+The most direct way to do so is by nonlinear optimisation at every timestep; Matlab provides the necessary routines in a convenient package.
+Unfortunately, this is not at all stable.
 
@@
 
-The variational methods are based on the phase space mapping $G$ from square-integrable functions to Fock space, where
-$$Gf=∫f(α*)|α〉\,d²α.$$
-This is different from Glauber's use: my $f$ is his $fe^{-½|α|²}$.  (Check the half.)
+The following graphs show a coherent state, with real amplitude 2, propagating in a quartic oscillator potential.
+The propagation was done exactly, in a Fock space truncated at $|19〉$; at times $π/10, π/5, 3π/10,\ldots,π$, a 10 component superposition was fit to the state by the routine {\tt cohfit5}.
+The SVD of $|Dψ〉$ was computed for each of these superpositions, and the Picard condition investigated.
+It appears that this discretisation is inherently ill-posed.
 
-Hypothesis: The singular values of $G$ are 0 and $\sqrt π$, and the space belonging to $\sqrt π$ comprises the functions $f$ for which $fe^{½|α|²}$ is analytic.
+\centerline{\XeTeXpicfile spsns.pdf width \hsize}
+Amplitudes of components
 
-There are two ways one could set about proving this.
-Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
-Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
-ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
+\centerline{\XeTeXpicfile traj.pdf width \hsize}
+Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted)
 
-This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
-My numerical experiment suggests otherwise; let's ignore that possibility for now.
+\centerline{\XeTeXpicfile picard.pdf width \hsize}
+Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied.
 
-Let $g$ be a normalised entire function, in Glauber's sense.  Our expansion function is 
-$$f(α) = {1\over π}g(α)e^{-½|α|²}.$$
-If $g$ has the Taylor expansion
-$$g(z)=∑_n c_n{zⁿ\over(n!)^½},$$
-then 
-$$Gf = ∑_n c_n|n〉.$$
-The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
-$$f(α)={1\over π}e^{-½|α|²}{αⁿ\over(n!)^½}.$$
-We have 
-$$|f(α)|²={1\over π²n!}e^{-|α|²}|α|^{2n},$$
-so the squared norm is
-$$\|f\|²={1\over π²n!}∫₀^∞2πr^{2n+1}e^{-r²}\,dr={2\over πn!}∫₀^∞r^{2n+1}e^{-r²}\,dr.$$
-According to Wolfram, this is 
-$$\|f\|²={1\over π},$$
-whence
-$$\|f\|=π^{-½}.$$
-
-Note that the gaussian entire functions are a vector subspace of the square integrable functions.
-
-Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
-Or at least I think so.
-I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
-
-Expanding the quartic oscillator over Fock states allows it to be solved very stably, and as precisely as desired by rotating the phases of the expansion coefficients.
-Also, it is easy to find the expansion over Fock states of a superpositon of coherent states.
-This means that the stability of time discretisation can be removed from the problem entirely.
-The state can be propagated exactly, and the coherent states adjusted to fit.
-The most direct way to do so is by nonlinear optimisation at every timestep; Matlab provides the necessary routines in a convenient package.
-Unfortunately, this is not at all stable.
+\centerline{\XeTeXpicfile svnum.pdf width \hsize}
+Expected particle number for left singular vectors.
+The svs with small sws have large particle number.
+When $H$ is quartic, these have large components in $H|ψ〉$.
+In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$.
 

### 2013-07-04 evan.m

@@ -1,5 +1,22 @@
 function X = evan(f,a)
+% evan: Calculate the parts of Fock space brackets that depend on the
+% amplitudes of the superposition.
+% 
+% X = evan(z)
+% X = evan(f, a)
+%
+% npsi = sum(nq*evan(z),2)
+% ndpsi = [nq*evan(z), ndqr*evan(z)]
+%
+% The amplitudes can be provided as a pair of vectors f and a, or a vector
+% z = [f a].
 
 global N
+
+if nargin==1
+    R = 0.5*numel(f);
+    a = f(R+1:end); f = f(1:R);
+end
+a = a(:).';
 
 X = ((ones(N+1,1)*a).^((0:N)'*ones(size(a))))*diag(exp(f));

### 2013-07-06 paper.tex
writing

@@
+\nonstopmode
 \input a4
@@
 
-This claims that $G$ is a projection: if the inner product is normalised properly, $G$ keeps the component of $f$ in the space of entire functions, and discards the orthogonal component.
-In that case, we'd have a simple underdetermined problem, which could be regularised very easily by discarding the small singular values.
-But that isn't what I see with discrete versions: they seem to have a steady decay of singular values, not a sudden drop to zero.
-Is that an artefact of how I'm discretising?
+So $G$ is a projection: if the inner product is normalised properly, $G$ keeps the component of $f$ in the space of entire functions, and discards the orthogonal component.
+The continuous problem is simply underdetermined, and the inverse problem can be regularised very easily by excluding components with zero singular values from the solution.
+That isn't what I see with discrete versions: they seem to have a steady decay of singular values, not a sudden drop to zero.
+This must be entirely an artefact of the discretisation.
+
+This paper will talk about two related but distinct inverse problems.
+Both start with a given wavefunction $|ψ\rangle$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
+In systems with few modes, these points could be a grid, but in many-mode systems they will have to be more sparse than that: there isn't space to store grid coefficients.
+The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j\rangle$ approximates $|ψ\rangle$.
+This is simply finding a discrete inverse to $G$ sampled on the grid $α_j$; since $G$ is a projection, the pseudoinverse of $G$ should solve it well enough.
+The differential case, where we have an expansion for $|ψ\rangle$ which we want to perturb to represent $|ψ\rangle+|dψ\rangle$, is no harder.
+Because $G$ is linear, we can expand $|dψ\rangle$, and add it to the expansion of $|ψ\rangle$.
+
+The second problem, the variational one, is a generalisation.
+As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ\rangle$.
+Instead of expanding $|ψ\rangle$ over phase space, we want to expand it over the columns of $|Dψ(f,α)\rangle$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
+Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)\rangle$ to the solution of the inverse problem, and those of $|∂fψ(f,a)\rangle$ to zero.
+Is this the smallest norm solution?
+Probably not: if a coherent state was moving in an oscillator, it might be easier to just rotate the $α_j$.
+
+Since we have a totally stable solution to the problem, why would we try anything else?
+Because we're iterating the problem, and we need to preserve our ability to solve it at future steps.
+If the amplitudes form a grid, which covers all the energies the system could possibly access, we're set: we'll always be able to expand over them.
+However, we can't do that for many mode systems: there will always be accessible states of the system whose support in phase space is remote from the current grid.
+The set $|Dψ\rangle$ will have some small singular values, corresponding to directions where the amplitudes and samples are adjusted in a way that preserves the wavefunction.
+We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ\rangle$ in the future.
+
+So what are the critera for adjusting the grid?
+I think there are two obvious ones: firstly, that the inverse problem stays well-posed.
+Otherwise, we're wasting grid points.
+Secondly, when we're expanding a derivative over the tangent space $|Dψ\rangle$, we want to tilt the tangent space closer to the derivative.
+That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.
+The first criterion seems redundant: if we could better represent the goal by shifting two sample points apart, the second criterion will do that automatically.
+However, if we only care about the tangent space and not the derivative map, we might need very large adjustments of the amplitudes and weights to get to the point in the tangent space that we want.
+So a condition number criterion seems sensible.
+There might be other criteria that I've overlooked.
+
+I still haven't thought in detail about the results with truncated Fock spaces.
+Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
+Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
+So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
+And in fact, by solving the inverse problem, we can always get a nice solution in any case.
+Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
 
@@
 Complex functions have space in the Argand plane to oscillate arbitrarily rapidy; considered as wavefunctions, the coherent states do: the wavenumber is proportional to the imaginary part of $α$.
+
+It turns out that complex convolutions are not like real convolutions: the imaginary part of $t-t'$ can control the rate at which the kernel oscillates.
+Each line of constant imaginary part in the convolution is determined by a limited range of frequencies from the function, but that frequency range can vary with the imaginary part, so that all frequency components of the function contribute to the convolution somewhere.
+Is the Fourier transform a complex convolution?
+
+Not a
 

### 2013-07-08 paper.tex
pertubation of singular values

@@
 
-When physicists began to study quantum mechanics, they could hardly have avoided discovering coherent states.
-When they grasped that observables were inherently uncertain and considered laws of probability, the normal distribution sprung to mind at once.
-Once they understood the dynamics of the free particle, and looked for a more interesting system to study, the obvious choice was the harmonic oscillator.
-The first question was its ground state; the second was why, in the classical limit, it oscillated.
-Coherent states were the answer to both.
+When the first quantum physicists understood that measurements are essentially uncertain, they naturally wondered what conditions would produce the normal law of probability, and discovered gaussian wave packets.
+Once they got sick of the free particle and turned to the harmonic oscillator, they would have know to look for its ground state, and how it oscillated in the classical limit.
+Gaussian wave packets were the answer to both questions.
+Now, with their width fixed so that the action of a cycle was $\hbar$, they became coherent states.
 
-Variational mechanics was another obvious idea.
-The quantum equations of motion were hard to solve exactly, except for some very simple systems.
-It was known from very early on that the solutions were extema of certain properties: ground states had the lowest possible energy, trajectories the least possible action.
-So it was natural to consider a class of approximants that could be treated exactly, and find out which of them had the extreme value over the class.
+When quantum mechanics was applied to more complicated systems, the equations of motion could not be solved in closed form.
+Several approaches were used to approximate them: one of the most versatile was the variational method, in which a class of approximants was specified that could be treated exactly, and it was determined which of them most nearly solved the problem.
+The most obvious application was in finding ground states, where the best solution was the one with the lowest energy.
+Variational methods were also applied to quantum dynamics, by seeking approximants for which the difference between the two sides of Schrödinger's equation had the least norm in Hilbert space.
+Gaussian wavepackets were an obvious choice for the class of approximants.
+They are distinguished by their central position and momentum, and their scale: useful things to know about the solution to any problem.
+Also, the gaussian packets and their derivatives are easy to manipulate algebraically.
 
-In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using coherent states as the approximant.
-He eventually discovered that related methods had been used to find the dynamics of lasers, starting with Glauber's work in the 19?0s on expansions over coherent states, and continuing when ?
-derived the dynamics of these expansions, and solved for the quantum dyanmics of the laser.
+The laser was invented, and physicists needed a quantum theory of intense radiation.
+This obeyed simple laws of dynamics, very similar to harmonic oscillators.
+However, It involved up to $10^{18}$ photons, spread across thousands of modes, so the difficulty of the quantum problem came from its sheer size.
+This was made managable by the work of Glauber, who showed that any quantum state could be written as a superposition of coherent states.
+For optical systems, close to harmonic oscillators, this was a very efficient way to represent states.
+Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.
+These have solved many problems, but are now limited by intractable convergence problems.
+
+In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using superpositions of coherent states as the approximant.
+This is an appealing approach.
+Every quantum state can be represented arbitrarily well as the number of components in the superposition becomes large.
+The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.
+There have been several attempts to apply this numerically.
+In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
 
@@
 This is a discretisation of the expansion introduced by Glauber, a mapping $G$ from square-integrable functions to Fock space, where
-$$Gf={1\over π}∫g(α*)|α〉\,d²α.\eqno{(1)}$$
+$$Gf={1\over π}\int g(α*)|α〉\,d²α.\eqno{(1)}$$
 This is different from Glauber's use: he uses $f=ge^{½|α|²}$.
@@
 so the squared norm is
-$$\|g\|²={1\over n!}∫₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}∫₀^∞r^{2n+1}e^{-r²}\,dr.$$
+$$\|g\|²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
 According to Wolfram,
-$$∫₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
+$$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
 so
@@
 This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
-$$f(t)=∫F(ω)e^{iωt}\,dω.$$
-This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.  (What is the SVD of the DFT matrix?)
+$$f(t)=\int F(ω)e^{iωt}\,dω.$$
+This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
+(What is the SVD of the DFT matrix?
+It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
 
@@
 Is there a better way?
+
+The accuracy of the pertubation method is shown below.
+
+\centerline{\XeTeXpicfile psfa.pdf width \hsize}
+
+\centerline{\XeTeXpicfile psv.pdf width \hsize}
 

### 2013-07-10 paper.tex
reorganisation

@@
 
-When the first quantum physicists understood that measurements are essentially uncertain, they naturally wondered what conditions would produce the normal law of probability, and discovered gaussian wave packets.
-Once they got sick of the free particle and turned to the harmonic oscillator, they would have know to look for its ground state, and how it oscillated in the classical limit.
-Gaussian wave packets were the answer to both questions.
-Now, with their width fixed so that the action of a cycle was $\hbar$, they became coherent states.
+%% Background
 
-When quantum mechanics was applied to more complicated systems, the equations of motion could not be solved in closed form.
-Several approaches were used to approximate them: one of the most versatile was the variational method, in which a class of approximants was specified that could be treated exactly, and it was determined which of them most nearly solved the problem.
-The most obvious application was in finding ground states, where the best solution was the one with the lowest energy.
-Variational methods were also applied to quantum dynamics, by seeking approximants for which the difference between the two sides of Schrödinger's equation had the least norm in Hilbert space.
-Gaussian wavepackets were an obvious choice for the class of approximants.
-They are distinguished by their central position and momentum, and their scale: useful things to know about the solution to any problem.
-Also, the gaussian packets and their derivatives are easy to manipulate algebraically.
+When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.
+They soon understood the free particle and turned to the harmonic oscillator, investigating its ground state and how its excited states generate sinusoidal motion in the classical limit.
+These problems were also solved by the gaussian wave packet.
+With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.
+
+The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but this luck doesn't last.
+Most quantum problems must be solved approximately.
+This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.
+To test how well a putative solution approximates the dynamics of a system, it can be substituted into each side of Schrödinger's equation, and the difference between them evaluated by the usual norm in Hilbert space.
+It remains to choose a class of approximants, and the gaussian wave packets are an common choice.
+They differ in their scale and their central position and momentum, useful properties to determine for any quantum state.
 
@@
 This is an appealing approach.
-Every quantum state can be represented arbitrarily well as the number of components in the superposition becomes large.
+Every quantum state can be approximated arbitrarily well as the number of components in the superposition becomes large.
 The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.
-There have been several attempts to apply this numerically.
-In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
+
+%% Problem
+
+Since Heller suggested it, there have been several attempts to apply this numerically.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
+%% Consequence
+These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.
+The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.
+A systematic approach to regularisation is required in order to approach these problems variationally.
+The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.
 
@@
 
-Variational dynamics seeks an expansion $g$ to approximate a quantum state $|ψ〉≈Gg$.
-Given the form of Equation~1, this requires solving a Fredholm integral equation of the first kind.
-That is the canonical example of an ill-posed problem: rounding and discretisation errors in $|ψ〉$ are magnified enormously in the solution $g$.
-To solve these problems, it helps to understand the singular values and vectors of the operator $G$.
+%% Phase space expansion as a linear operator
 
-The singular value expansion of the linear operator $G$ has the form
-$$Gf = ∑_{j=1}^∞ |g_j〉σ_j〈v_j,f〉,$$
-where the left singular vectors $|g_j〉$ are orthonormal kets, the singular values $σ_j$ are positive real numbers, the right singular vectors $v_j$ are normalised entire functions, and presumably $〈,〉$ is the Bargmann inner product.
-This can be written
-$$G = |G〉SV†,$$
-where $|G〉$ is a row vector of the left singular kets, $S$ is a matrix with the singular values along the diagonal, and $V$ maps complex coefficients to linear combinations of the functions $v_j$.
-
-Because the expansions are underdetermined, the set $|g_j〉$ will be complete, but the set $v_j$ will not be.
-Alternatively, some of the $v_j$ have zero singular values and no corresponding left singular vector.
-One can write a full SVE, where $V†$ is a full expansion, but $S$ truncates the vector of coefficients.
-
-The singular value expansion is useful for two reasons.
-Firstly, the discrete version, the singular value decomposition, is the most stable way to solve badly-posed least squares problems.
-Secondly, the sizes of the singular values, and the components of $H|ψ〉$ along the singular directions, determine whether the problem can be solved at all.
-
-The most important issue is the Picard condition.
-This restates a very basic observation: for the problem $G\dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $f$ over an orthonormal basis must have a finite norm.
-If the basis is chosen as the right singular vectors $v_j$, this means that the sum of the squared moduli of the components must be finite.
-For $\dot f$ to satisfy the equation, the component of $v_j$ is $〈g_j|H|ψ〉/σ_j$.
-So the Picard condition is that, for the equation to have a solution, the series
-$$∑_{j=1}^∞{\bigl|〈g_j|H|ψ〉\bigr|²\over σ_j²}$$
-must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.
+%% SVE lemma
 
 Hypothesis: The singular values of $G$ are 0 and $\sqrt π$, and the space belonging to $\sqrt π$ comprises the functions $f$ for which $fe^{½|α|²}$ is analytic.
+
+There are two ways one could set about proving the Hypothesis.
+Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
+Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
+ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
+
+This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
+My numerical experiment suggests otherwise; let's ignore that possibility for now.
+
+Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
+$$f(z)=∑c_n{zⁿ\over (n!)^½},$$
+for some normalised vector $c$.
+
+  Our expansion function is 
+$$g(α)=f(α)e^{-½|α|²}.$$
+If $f$ has the Taylor expansion
+$$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
+then 
+$$Gg = ∑_n c_n|n〉.$$
+The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
+$$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
+the Glauber preimage of $|n〉$.  We have 
+$$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
+so the squared norm is
+$$\|g\|²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
+According to Wolfram,
+$$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
+so
+$$\|g\|=π^½.$$
+
+The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
+
+Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
+Or at least I think so.
+I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
+
+%% Consequences of lemma
+
+To sum up, this analysis shows that $G$ is a projector.
+Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.
+This might depend somewhat on how evenly the amplitudes are spaced: I could start with a regular grid, and add Gaussian noise to see what changed.
 
@@
 This must be entirely an artefact of the discretisation.
+
+The theorems on how quickly singular values of convolution operators decay don't apply here.
+They apply to convolutions of real functions, which can only oscillate at a fixed rate.
+Complex functions have space in the Argand plane to oscillate arbitrarily rapidy; considered as wavefunctions, the coherent states do: the wavenumber is proportional to the imaginary part of $α$.
+
+It turns out that complex convolutions are not like real convolutions: the imaginary part of $t-t'$ can control the rate at which the kernel oscillates.
+Each line of constant imaginary part in the convolution is determined by a limited range of frequencies from the function, but that frequency range can vary with the imaginary part, so that all frequency components of the function contribute to the convolution somewhere.
+Is the Fourier transform a complex convolution?
+
+If the Hypothesis is right, the Picard condition is very obviously satisfied.
+The operator $G$ is a projection.
+The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm.
+Functions in the null space of $G$ don't correspond to left singular vectors, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
+
+%% Discrete phase space
+
+The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.
+The weights of the components don't simple sample $f(α)$: they also adjust for how close the sample points are, and how much the $|α_i〉$ overlap.
+The operator $G$ is probably being disceretised by some kind of Galerkin formula; I don't know what the basis functions are in phase space and Hilbert space.
+
+It would be useful to find a quadrature rule, with weights $w_i$, such that 
+$$∑_i w_if(α_i)|α_i〉≈\int f(α)|α〉d²α.$$
+Suppose we aim to be exactly right, in the case that $f(α)$ is the normalised entire function representing the left hand side, i.e.,
+$$f(β)=∑_i w_if(α_i)〈β|α_i〉.$$
+(I'm assuming we put the factor $1/π$ in $G$.)  This doesn't have a solution for general superpositions of coherent states.  Surely there's some quadrature rule for a sampled phase-space function?
+
+Using the Galerkin criterion, there doesn't seem to be one where the coefficient of $|α_i〉$ only depends on $f(α_i)$.
+However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.
+
+There are two types of ill-posed problems, rank-deficient and discrete.
+They are distingished by how the singular values decay to zero.
+In a rank-deficient problem, there is a sharp cut between large and small singular values.
+The large ones are large enough that components in their singular directions do not contribute excessively to the solution, the small ones would contribute huge components, but they are small enough to round to zero, and ignore the component along that singular direction.
+In discrete problems, the singular values decay continuously.
+The size of the components of the solution varies continuously: there is no point where they become huge, but a continuous minimum and a gradual rise as the singular values get smaller.
+
+I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.
+However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.
+This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
+$$f(t)=\int F(ω)e^{iωt}\,dω.$$
+This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
+(What is the SVD of the DFT matrix?
+It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
+
+%% Truncated Hilbert space
+
+I still haven't thought in detail about the results with truncated Fock spaces.
+Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
+Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
+So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
+And in fact, by solving the inverse problem, we can always get a nice solution in any case.
+Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
+
+The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
+Numerical experiments in the script {\tt sve.m} support this.
+The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
+Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
+The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
+
+\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
+
+The following three figures show the right singular vectors, sampled versions of phase space functions.
+The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
+These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
+In the continuous limit there is only one singular value, and the vectors are degenerate.
+For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
+As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
+
+\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
+\XeTeXpicfile rsv3.pdf width 0.5\hsize}
+\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+
+Reducing the dimension of Fock space so that the problem is underdetermined means that, when the problem is solved by SVD, we only take components along $N+1$ directions in Fock space, and map them to $N+1$ directions in parameter space.
+The remaining $2R-N-1$ components of the solution are zero.
+As we add extra dimensions to Fock space, we are taking components in more Fock space directions, which include more oscillating components of the wavefunction.
+This is like sampling a function on a finer grid.
+With these directions, we can form some of the smaller singular vectors of the full problem.
+
+Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
+(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
+
+As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
+Presumably these converge to limits as $N$ increases.
+There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
+
+%% Why grids don't work, and variation necessary
 
@@
 
-I still haven't thought in detail about the results with truncated Fock spaces.
-Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
-Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
-So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
-And in fact, by solving the inverse problem, we can always get a nice solution in any case.
-Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
-
-The theorems on how quickly singular values of convolution operators decay don't apply here.
-They apply to convolutions of real functions, which can only oscillate at a fixed rate.
-Complex functions have space in the Argand plane to oscillate arbitrarily rapidy; considered as wavefunctions, the coherent states do: the wavenumber is proportional to the imaginary part of $α$.
-
-It turns out that complex convolutions are not like real convolutions: the imaginary part of $t-t'$ can control the rate at which the kernel oscillates.
-Each line of constant imaginary part in the convolution is determined by a limited range of frequencies from the function, but that frequency range can vary with the imaginary part, so that all frequency components of the function contribute to the convolution somewhere.
-Is the Fourier transform a complex convolution?
-
-Not a
-
-If the Hypothesis is right, the Picard condition is very obviously satisfied.
-The operator $G$ is a projection.
-The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm.
-Functions in the null space of $G$ don't correspond to left singular vectors, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
-
-There are two ways one could set about proving the Hypothesis.
-Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
-Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
-ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
-
-This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
-My numerical experiment suggests otherwise; let's ignore that possibility for now.
-
-Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
-$$f(z)=∑c_n{zⁿ\over (n!)^½},$$
-for some normalised vector $c$.
-
-  Our expansion function is 
-$$g(α)=f(α)e^{-½|α|²}.$$
-If $f$ has the Taylor expansion
-$$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
-then 
-$$Gg = ∑_n c_n|n〉.$$
-The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
-$$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
-the Glauber preimage of $|n〉$.  We have 
-$$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
-so the squared norm is
-$$\|g\|²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
-According to Wolfram,
-$$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
-so
-$$\|g\|=π^½.$$
-
-The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
-Numerical experiments in the script {\tt sve.m} support this.
-The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
-Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
-The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
-
-\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
-
-The following three figures show the right singular vectors, sampled versions of phase space functions.
-The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
-These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
-In the continuous limit there is only one singular value, and the vectors are degenerate.
-For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
-As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
-
-\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
-\XeTeXpicfile rsv3.pdf width 0.5\hsize}
-\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
-
-Note that the gaussian entire functions are a vector subspace of the square integrable functions.
-
-Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
-Or at least I think so.
-I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
-
-To sum up, this analysis shows that $G$ is a projector.
-Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.
-This might depend somewhat on how evenly the amplitudes are spaced: I could start with a regular grid, and add Gaussian noise to see what changed.
-
-There are two types of ill-posed problems, rank-deficient and discrete.
-They are distingished by how the singular values decay to zero.
-In a rank-deficient problem, there is a sharp cut between large and small singular values.
-The large ones are large enough that components in their singular directions do not contribute excessively to the solution, the small ones would contribute huge components, but they are small enough to round to zero, and ignore the component along that singular direction.
-In discrete problems, the singular values decay continuously.
-The size of the components of the solution varies continuously: there is no point where they become huge, but a continuous minimum and a gradual rise as the singular values get smaller.
-
-I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.
-However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.
-This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
-$$f(t)=\int F(ω)e^{iωt}\,dω.$$
-This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
-(What is the SVD of the DFT matrix?
-It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
-
-The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.
-I was working in a truncated Fock space; I wonder what happens as the truncation increases, while the grid stays a constant size?
-Or alternately, if the grid shrinks while the Fock space stays constant.
-
-I'm getting a picture of how the final algorithm will look.
-At each timestep, the directions belonging to large singular values of $|Dψ〉$ will be used for variational dynamics.
-This can be done by solving at Tychonov conditioned problem, for example.
-At the same time, the directions with small singular values will be used to keep the discretisation stable.
-I think this requires two things: that $|Dψ〉$ is reasonably well conditioned, and that the angle between $H|ψ〉$ and the span of $|Dψ〉$ is small.
-This raises a question: are there singular directions that have little effect on $|ψ〉$, but a large effect on $|Dψ〉$?
-This seems likely; for example, shuffling the amplitudes to move two components further apart in phase space, then adjusting the weights so that $|ψ〉$ stays the same.
-This could be done by a kind of Tychonov anti-regularisation: we improve the stability as much as we can, while constraining the change in $|ψ〉$.
-I might need to constrain the norm of $dz$ too, so that linear approximations hold.
-
-I've already figured out how to find the change in the condition number by pertubation theory.
-That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.
-How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?
-This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.
-Is there a better way?
-
-The accuracy of the pertubation method is shown below.
-
-\centerline{\XeTeXpicfile psfa.pdf width \hsize}
-
-\centerline{\XeTeXpicfile psv.pdf width \hsize}
-
-The final algorithm would propagate $|ψ〉$ in Hilbert space by a semi-implicit method, solving a variational problem to represent each state.
-This is a slight generalisation, because you need to project $H|ψ_{n+½}〉$ in to the span of $|Dψ_n〉$, and the ensembles at the two times are slightly different.
-It might help to propagate from $n$ to $n+½$ by changing only weights, leaving the same ensemble.
-On the other hand, the change is only that the $α*$ become $β*$, so it should be managable.
-
-There shouldn't be an need for fancy methods to fit the initial state: one can take a guess $z₀$, and variationally solve the equation $|Dψ(z)〉dz=|ψ₀〉-|ψ(z)〉$.
-One could even change the conditioning rules as the state propagates, so that $|Dψ〉$ is aligned with $|ψ₀〉-|ψ(z)〉$ to start with, then with $H|ψ(z)〉$ later on.
-That way, you start the actual simulation with a well conditioned approximant.
+%% Fock space is cheating
 
@@
 
-The following graphs show a coherent state, with real amplitude 2, propagating in a quartic oscillator potential.
-The propagation was done exactly, in a Fock space truncated at $|19〉$; at times $π/10, π/5, 3π/10,\ldots,π$, a 10 component superposition was fit to the state by the routine {\tt cohfit5}.
-The SVD of $|Dψ〉$ was computed for each of these superpositions, and the Picard condition investigated.
-It appears that this discretisation is inherently ill-posed.
+To do this, we need some bracket matrices.  We have
+$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
+The brackets with the number states are
+$$〈m|H/\hbar|n〉=n(n-1)δ_{mn}$$
+and
+$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
+as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
+$$〈n|H/\hbar|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
 
-\centerline{\XeTeXpicfile spsns.pdf width \hsize}
-Amplitudes of components
+%% Singular value decompositions
 
-\centerline{\XeTeXpicfile traj.pdf width \hsize}
-Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted)
+Variational dynamics seeks an expansion $g$ to approximate a quantum state $|ψ〉≈Gg$.
+Given the form of Equation~1, this requires solving a Fredholm integral equation of the first kind.
+That is the canonical example of an ill-posed problem: rounding and discretisation errors in $|ψ〉$ are magnified enormously in the solution $g$.
+To solve these problems, it helps to understand the singular values and vectors of the operator $G$.
 
-\centerline{\XeTeXpicfile picard.pdf width \hsize}
-Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied.
+The singular value expansion of the linear operator $G$ has the form
+$$Gf = ∑_{j=1}^∞ |g_j〉σ_j〈v_j,f〉,$$
+where the left singular vectors $|g_j〉$ are orthonormal kets, the singular values $σ_j$ are positive real numbers, the right singular vectors $v_j$ are normalised entire functions, and presumably $〈,〉$ is the Bargmann inner product.
+This can be written
+$$G = |G〉SV†,$$
+where $|G〉$ is a row vector of the left singular kets, $S$ is a matrix with the singular values along the diagonal, and $V$ maps complex coefficients to linear combinations of the functions $v_j$.
 
-\centerline{\XeTeXpicfile svnum.pdf width \hsize}
-Expected particle number for left singular vectors.
-The svs with small sws have large particle number.
-When $H$ is quartic, these have large components in $H|ψ〉$.
-In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$.
+Because the expansions are underdetermined, the set $|g_j〉$ will be complete, but the set $v_j$ will not be.
+Alternatively, some of the $v_j$ have zero singular values and no corresponding left singular vector.
+One can write a full SVE, where $V†$ is a full expansion, but $S$ truncates the vector of coefficients.
 
-\centerline{\XeTeXpicfile flwc0n10.pdf width \hsize}
+The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
+If a singular value decomposition $|Dψ〉=|U〉SV†$ is taken, the singularity is indicated by small singular values on the diagonal of $S$.
+It is often measured by the condition number of $|Dψ〉$, the ratio of the largest to the smallest singular value.
+Let $C(z)$ be the condition number of $|Dψ(z)〉$.
+As part of the regularisation process, it would be useful to avoid changing $z$ in directions that increase $C(z)$.
+For example, if two coherent states had real amplitudes, and the real part of the expected amplitude needed to increase, it would be done by shifting the right hand one away from the left hand one, not the left hand one closer to the right hand.
+Is there a way to make a computer do that?
 
-\centerline{\XeTeXpicfile flwc0n5.pdf width \hsize}
+The least squares problem can be solved using the singular value decomposition: to approximate a state $|h〉≈|Dψ(z)〉dz$, we set $dz=VS^{-1}〈U|h〉$.
+The first problem is that some of the $σ_i$ on the diagonal of $S$ will be small, because the $|u_i〉$ that correspond to them change different components of the superposition $|ψ(z)〉$ in ways that nearly cancel out.
+So a error in the discrete representation of $|h〉$ in the direction $|u_i〉$ will cause a very large error in the least squares $dz$ in the direction $v_i$.
+The way to avoid such large errors is to shift the small singular values of $S$ up to a lower bound $ε$, and thus reduce the large singular values of $S^{-1}$ below an upper bound $ε^{-1}$.
+There are many and varied ways of doing that.
 
-The above plots show my first attempt at this.
-The initial state was coherent, with real amplitude 2.
-A superposition of 5 coherent states was initialised with amplitudes normally distributed around 2, the norm of the superposition being 1, and the components having equal norms.
-This superposition was fit to the coherent state by minimising their distance in Hilbert space; for stability, the weights were constrained so that no component had a norm less than 0.1.
-This was done by the Matlab routine {\tt fmincon}.
-The initial state was truncated to Fock states $|0〉$ to $|10〉$, and propagated exactly in a Hamiltonian $\hbar a^{2\dagger}a²$ with timestep 0.03.
-After each step, the fitting of the superposition to the exact state was repeated.
+%% Picard plots
 
-The results are shown in the first Figure.
-The left hand plot shows the complex amplitude of the exact solution in black, and the fitted solution in blue.
-The right hand plot shows the residual of the fit as a solid line, and the condition number of $|Dψ(z)〉$ at the fitted $z$ as a dotted line.
-The program stopped after a short time, when the fitted amplitudes caused overflow when expanded over Fock coefficients.
+The singular value expansion is useful for two reasons.
+Firstly, the discrete version, the singular value decomposition, is the most stable way to solve badly-posed least squares problems.
+Secondly, the sizes of the singular values, and the components of $H|ψ〉$ along the singular directions, determine whether the problem can be solved at all.
 
-The second Figure shows the same simulation, with the states expanded over Fock states $|0〉$ to $|5〉$.
-This has a significant effect on the truncated exact solution, adding an extra loop to the amplitude.
-The fitting works a bit longer, but still fails at a very short time.
+The most important issue is the Picard condition.
+This restates a very basic observation: for the problem $G\dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $f$ over an orthonormal basis must have a finite norm.
+If the basis is chosen as the right singular vectors $v_j$, this means that the sum of the squared moduli of the components must be finite.
+For $\dot f$ to satisfy the equation, the component of $v_j$ is $〈g_j|H|ψ〉/σ_j$.
+So the Picard condition is that, for the equation to have a solution, the series
+$$∑_{j=1}^∞{\bigl|〈g_j|H|ψ〉\bigr|²\over σ_j²}$$
+must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.
 
-\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
+The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.
+I was working in a truncated Fock space; I wonder what happens as the truncation increases, while the grid stays a constant size?
+Or alternately, if the grid shrinks while the Fock space stays constant.
 
-\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
+%% Adapting the sample points to maintain condition
 
-The next two plots show a different fitting strategy.
-Instead of constraining the weights, a cost term was added in quadrature to the Hilbert space distance, proportional to the logarithm of the condition number of $|Dψ(z)〉$ for the fitted $z$.
-The constant of proportionality was $10^{-5}$; this was adjusted until a good fit was obtained.
-The most apparent thing is that the fit is much more stable: this is being done with timestep 0.1 instead of 0.3, but the fit stays stable for the whole period.
-(I had to redraw the initial amplitudes a few times before that happened.)  It isn't a very good fit through the middle of the period: the residual increases to order 1.
-The really interesting thing is what happened when, by accident, I truncated the Fock basis at $N=5$.
-As before, this perturbs the trajectory in Hilbert space.
-However, the perturbed trajectory can be fit very well: the residual is between $10^{-2}$ and $10^{-5}$, and this is done with quite well conditioned superpositions.
+I'm getting a picture of how the final algorithm will look.
+At each timestep, the directions belonging to large singular values of $|Dψ〉$ will be used for variational dynamics.
+This can be done by solving at Tychonov conditioned problem, for example.
+At the same time, the directions with small singular values will be used to keep the discretisation stable.
+I think this requires two things: that $|Dψ〉$ is reasonably well conditioned, and that the angle between $H|ψ〉$ and the span of $|Dψ〉$ is small.
+This raises a question: are there singular directions that have little effect on $|ψ〉$, but a large effect on $|Dψ〉$?
+This seems likely; for example, shuffling the amplitudes to move two components further apart in phase space, then adjusting the weights so that $|ψ〉$ stays the same.
+This could be done by a kind of Tychonov anti-regularisation: we improve the stability as much as we can, while constraining the change in $|ψ〉$.
+I might need to constrain the norm of $dz$ too, so that linear approximations hold.
 
-So even without stiffness in a discrete time propagation, there is something in the high energy components of $H|ψ〉$ that makes fitting the state with coherent states unstable.
-This isn't too surprising: these high energy components are rapidly oscillating, and will by magnified when the inverse problem is solved with fairly smooth, low amplitude coherent states.
-To fit the oscillating components, we need coherent states with large amplitudes, and low weights.
-But the low weights will increase the condition number of $|Dψ〉$, so the regularisation scheme will avoid these, unless there are many components in the superposition and all their weights are low.
-Let's investigate what happens as the number of components and the Fock truncation are varied together.
+To solving our differential equation, we need to solve an iterated least squares problem.
+We need to do so in a way that preserves the condition of $|Dψ〉$.
+When we shift $z$ in the direction $v_i$ as $dz=λv_i$, how does this condition number change?
+Without loss of generality, it suffices to consider the effect on the smallest and largest singular values, that determine $C(z)$.
+Note that $v_1$ and $v_R$ are eigenvectors of the matrix ${\cal V}(z)=〈Dψ(z)|Dψ(z)〉$, belonging to the eigenvalues $σ₁²$ and $σ_R²$; this is a familiar problem in pertubation theory.
+Assume for now that we have found $D{\cal V}(z)$—this won't be analytic, so we will need to use real and imaginary parts, or Wirtinger calculus.
+The change in $\cal V$ is of the form $d{\cal V}=∂_i{\cal V}(z)λ+∂*_i{\cal V}(z)λ*$, where the partial derivatives are taken in the directions $v_i$.
 
-The things I've noticed so far are:
-\item{1.} When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.
-It isn't accurate, though.
+As is well known in pertubation theory, the change in the eigenvalue $σ₁²$ corresponding to a small change $d{\cal V}$ is $v₁†d{\cal V}v₁$, and similarly for $σ_R²$.
+Therefore the change in $C(z)²$ is
+$$σ_R²v₁†d{\cal V}v₁-σ₁²v_R†d{\cal V}v_R \over σ_R².$$
+Given a proposed change $dz$, we can find the effect on the condition number.
 
-\item{2.} Measuring state differences in a truncated fock space makes the fit stable and accurate.
-Of course, the truncated state is different from the original one, so the original state is not fit precisely.
-When this was done, the least squares problem involved 6 fock coefficients, and the amplitude vector has 10 elements.
-So almost all the truncated problems were underdetermined: some singular values of the Jacobian are exactly zero.
-Matlab claims that backslash solves underdetermined problems by giving a solution with as many zero elements as possible.
-The documentation doesn't say how it chooses which elements to make zero.
-That's different from the other regularisation methods we've tried.
-Of course, the problems are being solved inside Matlab's optimisation routine, and I don't know what it's doing.
+The question is how to apply this to condition the process of finding $dz$.
+Given $z$, we can find out which directions of $dz$ correspond to large changes in the condition number.
+However,  until we know $dz$, we don't know if these changes help or hurt.
+The process might run like this: first, truncate the singular values.
+Next, find the components $〈U_i|h〉$, and the corresponding components $v_iσ_i〈U_i|h〉$ of $dz$, given the truncation.
+Then, truncate the components that will cause $C(z)$ to increase to values that limits the increase to an acceptable amount.
 
-Reducing the dimension of Fock space so that the problem is underdetermined means that, when the problem is solved by SVD, we only take components along $N+1$ directions in Fock space, and map them to $N+1$ directions in parameter space.
-The remaining $2R-N-1$ components of the solution are zero.
-As we add extra dimensions to Fock space, we are taking components in more Fock space directions, which include more oscillating components of the wavefunction.
-This is like sampling a function on a finer grid.
-With these directions, we can form some of the smaller singular vectors of the full problem.
+I've figured out how to find the change in the condition number by pertubation theory.
+That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.
+How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?
+This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.
+Is there a better way?
 
-Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
-(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
+The accuracy of the pertubation method is shown below.
+The state is coherent, with ampltiude 2.
+It was sampled at 5 amplitudes, shown on the left hand plot.
+The weights were assigned by inverting the discretised $G$ operator on these sample points, with small singular values truncated by Matlab's {\tt pinv} function.
+The the moduli of the resulting weights are shown in the right hand plot.
+This method of fitting usually gives a Hilbert space residual around 0.01.
 
-As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
-Presumably these converge to limits as $N$ increases.
-There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
+The lower plot shows how the smallest singular value of the discrete $G$ changes, as the sample points are shifted.
+The black lines show changes in the real parts of the amplitudes, the red lines in the imaginary parts.
+The dotted black line shows the derivatives for the real parts calculated by pertubation theory, as follows.
+The derivatives of the matrix $〈N|G|A〉$  with respect to each real part were computed by symmetric difference, with $h=0.01$.
+Here, $|Α〉=(|α₁〉 |α₂〉 \cdots |α_R〉)$.
+A matrix
+$$M=\pmatrix{0 & 〈N|G|A〉\cr 〈A|G|N〉 & 0}$$
+was constructed; by substituting $〈N|G|A〉=USV†$, it is clear that the vector
+$$x={1\over √2}\pmatrix{u_R \cr v_R}$$
+is an eigenvector, with eigenvalue $σ_R$, and norm 1.  (Provided $N+1>R$.)  From first-order pertubation theory, the derivative of this eigenvalue with respect to amplitude $i$ is
+$$∂_i σ_i=x†∂_iMx.$$
+As shown in the plots, lines of this slope fit the singular value very well.
 
-\item{3.} The difference between the truncated state and the original one fails to satisfy the Picard condition when with respect to $|Dψ〉$.  What about $H|ψ〉$?
+\centerline{\XeTeXpicfile psfa.pdf width \hsize}
 
-\item{4.} When the least squares problem is solved with a different distance function, which sets distances along the large $N$ directions to zero, the fit fails on the first step.
-Note that this isn't really a distance.
-Also, I did it by solving $P|h〉=P|Dψ〉dz$; when $P$ truncates the Fock space, the operator $P|Dψ〉$ is singular.
-Doing a weighted least squares fit, with the many particle directions having low weights, would be near-singular.
+\centerline{\XeTeXpicfile psv.pdf width \hsize}
+
+Below is a plot of the condition number of $|Dψ(z)〉$ for a superposition, as the value of $z$ varies by $0.1$ in every direction.
+The top line shows the effects of changing $φ$, the bottom line $α$, components are shown left to right.
+Changes to real parts are black, imaginary parts red.
+There are a few things to note.
+The condition is very bad, with number $1.4×10^{11}$.
+This is entirely due to the similar amplitudes of components 1 and 3: those graphs are on a different scale to the others, and it's logarithmic.
+As they move closer and become identical, the condition number increases $10^{14}$.
+No doubt it would go higher if we sampled the graphs more finely, but this is huge—reciprocal floating point epsilon is $5×10^{15}$.
+If the separation between these components is doubled, the condition number improves by an order of magnitude.
+
+\centerline{\XeTeXpicfile j1.pdf width \hsize}
+
+\centerline{\XeTeXpicfile j2.pdf width \hsize}
+
+The top line indicates that the condition number is independent of the phases of the components.
+I didn't forsee that, but it isn't too surprising: we can change the coefficients of the superposition in any direction we like, regardless of their current values, and the derivative is only sensitive to changes.
+
+%% Shape of Hilbert space trajectories
+
+If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.
+The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.
+Different eigenstates circle the origin with different frequencies.
+
+The semi-implicit formula has a special property: when the trajectory of $u(t)$ is a circle, the $vⁿ$ lie exactly on that circle, and the discretisation error is equivalent to rescaling time by a factor that depends on the frequency and the discrete timestep.
+In Hilbert space, this means that the projections of the state onto eigenplanes still circle the origin exactly: only the frequencies are altered by discretisation.
+So the effect of discretising quantum dynamics using the semi-implicit formula is merely to shift the energies of the eigenstates.
+Of course, this can be a significant effect: it's the only difference between the harmonic and quartic oscillators.
+
+%% Discrete time and stiffness
 
@@
 $$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$
-If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.
-The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.
-Different eigenstates circle the origin with different frequencies.
-
-The semi-implicit formula has a special property: when the trajectory of $u(t)$ is a circle, the $vⁿ$ lie exactly on that circle, and the discretisation error is equivalent to rescaling time by a factor that depends on the frequency and the discrete timestep.
-In Hilbert space, this means that the projections of the state onto eigenplanes still circle the origin exactly: only the frequencies are altered by discretisation.
-So the effect of discretising quantum dynamics using the semi-implicit formula is merely to shift the energies of the eigenstates.
-Of course, this can be a significant effect: it's the only difference between the harmonic and quartic oscillators.
 
@@
 
-The initial state was the ground state of the Morse oscillator potential, given various momentum kicks.
-The number of components in the superposition representing this was varied.
-The initial superposition was found by the AMPL optimisation program, with the constraint that no component would have a weight less that ?.
+The following graphs show a coherent state, with real amplitude 2, propagating in a quartic oscillator potential.
+The propagation was done exactly, in a Fock space truncated at $|19〉$; at times $π/10, π/5, 3π/10,\ldots,π$, a 10 component superposition was fit to the state by the routine {\tt cohfit5}.
+The SVD of $|Dψ〉$ was computed for each of these superpositions, and the Picard condition investigated.
+It appears that this discretisation is inherently ill-posed.
 
-I want to measure the numerical stability of the variational dynamics derived from Schrödinger's equation.
-I can do that by measuring how the discretisation error converges as the timestep is reduced.
-The second hypothesis is that the instability is due to stiffness, which can be tested by discretising time in ways with a variety of stability regions, and seeing whether the stability differs.
+\centerline{\XeTeXpicfile spsns.pdf width \hsize}
+Amplitudes of components
+
+\centerline{\XeTeXpicfile traj.pdf width \hsize}
+Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted)
+
+\centerline{\XeTeXpicfile picard.pdf width \hsize}
+Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied.
+
+\centerline{\XeTeXpicfile svnum.pdf width \hsize}
+Expected particle number for left singular vectors.
+The svs with small sws have large particle number.
+When $H$ is quartic, these have large components in $H|ψ〉$.
+In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$.
 
@@
 
-To do this, we need some bracket matrices.  We have
-$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
-The brackets with the number states are
-$$〈m|H/\hbar|n〉=n(n-1)δ_{mn}$$
-and
-$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
-as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
-$$〈n|H/\hbar|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
+\centerline{\XeTeXpicfile flwc0n10.pdf width \hsize}
 
-The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
-If a singular value decomposition $|Dψ〉=|U〉SV†$ is taken, the singularity is indicated by small singular values on the diagonal of $S$.
-It is often measured by the condition number of $|Dψ〉$, the ratio of the largest to the smallest singular value.
-Let $C(z)$ be the condition number of $|Dψ(z)〉$.
-As part of the regularisation process, it would be useful to avoid changing $z$ in directions that increase $C(z)$.
-For example, if two coherent states had real amplitudes, and the real part of the expected amplitude needed to increase, it would be done by shifting the right hand one away from the left hand one, not the left hand one closer to the right hand.
-Is there a way to make a computer do that?
+\centerline{\XeTeXpicfile flwc0n5.pdf width \hsize}
 
-The least squares problem can be solved using the singular value decomposition: to approximate a state $|h〉≈|Dψ(z)〉dz$, we set $dz=VS^{-1}〈U|h〉$.
-The first problem is that some of the $σ_i$ on the diagonal of $S$ will be small, because the $|u_i〉$ that correspond to them change different components of the superposition $|ψ(z)〉$ in ways that nearly cancel out.
-So a error in the discrete representation of $|h〉$ in the direction $|u_i〉$ will cause a very large error in the least squares $dz$ in the direction $v_i$.
-The way to avoid such large errors is to shift the small singular values of $S$ up to a lower bound $ε$, and thus reduce the large singular values of $S^{-1}$ below an upper bound $ε^{-1}$.
-There are many and varied ways of doing that.
+The above plots show my first attempt at this.
+The initial state was coherent, with real amplitude 2.
+A superposition of 5 coherent states was initialised with amplitudes normally distributed around 2, the norm of the superposition being 1, and the components having equal norms.
+This superposition was fit to the coherent state by minimising their distance in Hilbert space; for stability, the weights were constrained so that no component had a norm less than 0.1.
+This was done by the Matlab routine {\tt fmincon}.
+The initial state was truncated to Fock states $|0〉$ to $|10〉$, and propagated exactly in a Hamiltonian $\hbar a^{2\dagger}a²$ with timestep 0.03.
+After each step, the fitting of the superposition to the exact state was repeated.
 
-To solving our differential equation, we need to solve an iterated least squares problem.
-We need to do so in a way that preserves the condition of $|Dψ〉$.
-When we shift $z$ in the direction $v_i$ as $dz=λv_i$, how does this condition number change?
-Without loss of generality, it suffices to consider the effect on the smallest and largest singular values, that determine $C(z)$.
-Note that $v_1$ and $v_R$ are eigenvectors of the matrix ${\cal V}(z)=〈Dψ(z)|Dψ(z)〉$, belonging to the eigenvalues $σ₁²$ and $σ_R²$; this is a familiar problem in pertubation theory.
-Assume for now that we have found $D{\cal V}(z)$—this won't be analytic, so we will need to use real and imaginary parts, or Wirtinger calculus.
-The change in $\cal V$ is of the form $d{\cal V}=∂_i{\cal V}(z)λ+∂*_i{\cal V}(z)λ*$, where the partial derivatives are taken in the directions $v_i$.
+The results are shown in the first Figure.
+The left hand plot shows the complex amplitude of the exact solution in black, and the fitted solution in blue.
+The right hand plot shows the residual of the fit as a solid line, and the condition number of $|Dψ(z)〉$ at the fitted $z$ as a dotted line.
+The program stopped after a short time, when the fitted amplitudes caused overflow when expanded over Fock coefficients.
 
-As is well known in pertubation theory, the change in the eigenvalue $σ₁²$ corresponding to a small change $d{\cal V}$ is $v₁†d{\cal V}v₁$, and similarly for $σ_R²$.
-Therefore the change in $C(z)²$ is
-$$σ_R²v₁†d{\cal V}v₁-σ₁²v_R†d{\cal V}v_R \over σ_R².$$
-Given a proposed change $dz$, we can find the effect on the condition number.
+The second Figure shows the same simulation, with the states expanded over Fock states $|0〉$ to $|5〉$.
+This has a significant effect on the truncated exact solution, adding an extra loop to the amplitude.
+The fitting works a bit longer, but still fails at a very short time.
 
-The question is how to apply this to condition the process of finding $dz$.
-Given $z$, we can find out which directions of $dz$ correspond to large changes in the condition number.
-However,  until we know $dz$, we don't know if these changes help or hurt.
-The process might run like this: first, truncate the singular values.
-Next, find the components $〈U_i|h〉$, and the corresponding components $v_iσ_i〈U_i|h〉$ of $dz$, given the truncation.
-Then, truncate the components that will cause $C(z)$ to increase to values that limits the increase to an acceptable amount.
+\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
 
-Coding this will be quite involved, and the result might not be very efficient.  However, if it allows physically chosen bases to be used in  variational simulations, it is likely to be worthwhile.
+\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
 
-\centerline{\XeTeXpicfile j1.pdf width \hsize}
+The next two plots show a different fitting strategy.
+Instead of constraining the weights, a cost term was added in quadrature to the Hilbert space distance, proportional to the logarithm of the condition number of $|Dψ(z)〉$ for the fitted $z$.
+The constant of proportionality was $10^{-5}$; this was adjusted until a good fit was obtained.
+The most apparent thing is that the fit is much more stable: this is being done with timestep 0.1 instead of 0.3, but the fit stays stable for the whole period.
+(I had to redraw the initial amplitudes a few times before that happened.)  It isn't a very good fit through the middle of the period: the residual increases to order 1.
+The really interesting thing is what happened when, by accident, I truncated the Fock basis at $N=5$.
+As before, this perturbs the trajectory in Hilbert space.
+However, the perturbed trajectory can be fit very well: the residual is between $10^{-2}$ and $10^{-5}$, and this is done with quite well conditioned superpositions.
 
-To test this idea, a five component superposition was fit to an amplitude 2 coherent state.
-The Hilbert space residual was $1.3×10^{-3}$.
-The amplitudes and norms of the components are shown above; the fitting routine constrained the norm of each component to be at least 0.1.
-Note that components 1 and 3 ended up with very similar amplitudes.
+So even without stiffness in a discrete time propagation, there is something in the high energy components of $H|ψ〉$ that makes fitting the state with coherent states unstable.
+This isn't too surprising: these high energy components are rapidly oscillating, and will by magnified when the inverse problem is solved with fairly smooth, low amplitude coherent states.
+To fit the oscillating components, we need coherent states with large amplitudes, and low weights.
+But the low weights will increase the condition number of $|Dψ〉$, so the regularisation scheme will avoid these, unless there are many components in the superposition and all their weights are low.
+Let's investigate what happens as the number of components and the Fock truncation are varied together.
 
-Below is a plot of the condition number of $|Dψ(z)〉$ for this superposition, as the value of $z$ varies by $0.1$ in every direction.
-The top line shows the effects of changing $φ$, the bottom line $α$, components are shown left to right.
-Changes to real parts are black, imaginary parts red.
-There are a few things to note.
-The condition is very bad, with number $1.4×10^{11}$.
-This is entirely due to the similar amplitudes of components 1 and 3: those graphs are on a different scale to the others, and it's logarithmic.
-As they move closer and become identical, the condition number increases $10^{14}$.
-No doubt it would go higher if we sampled the graphs more finely, but this is huge—reciprocal floating point epsilon is $5×10^{15}$.
-If the separation between these components is doubled, the condition number improves by an order of magnitude.
+The things I've noticed so far are:
+\item{1.} When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.
+It isn't accurate, though.
 
-\centerline{\XeTeXpicfile j2.pdf width \hsize}
+Measuring state differences in a truncated fock space makes the fit stable and accurate.
+Of course, the truncated state is different from the original one, so the original state is not fit precisely.
+When this was done, the least squares problem involved 6 fock coefficients, and the amplitude vector has 10 elements.
+So almost all the truncated problems were underdetermined: some singular values of the Jacobian are exactly zero.
+Matlab claims that backslash solves underdetermined problems by giving a solution with as many zero elements as possible.
+The documentation doesn't say how it chooses which elements to make zero.
+That's different from the other regularisation methods we've tried.
+Of course, the problems are being solved inside Matlab's optimisation routine, and I don't know what it's doing.
 
-The top line indicates that the condition number is independent of the phases of the components.
-I didn't forsee that, but it isn't too surprising: we can change the coefficients of the superposition in any direction we like, regardless of their current values, and the derivative is only sensitive to changes.
+%% The complete algorithm
 
-The experiment with following the exact solution by nonlinear optimisation was repeated.
-Instead of constraining the weights of the components, the logarithm of the condition number of the derivative was added in quadrature to the residual of the fit to give a cost.
-This worked much better, getting stably to $t=0.4$ instead of under 0.1.
-At that point, the residuals increased to order 1, and the superposition stopped following the exact solution.
-This cold mean that 5 components weren't enough to represent those states, except that the same thing happened with 10 components.
+The final algorithm would propagate $|ψ〉$ in Hilbert space by a semi-implicit method, solving a variational problem to represent each state.
+This is a slight generalisation, because you need to project $H|ψ_{n+½}〉$ in to the span of $|Dψ_n〉$, and the ensembles at the two times are slightly different.
+It might help to propagate from $n$ to $n+½$ by changing only weights, leaving the same ensemble.
+On the other hand, the change is only that the $α*$ become $β*$, so it should be managable.
 
-Note that the optimisation is a nonlinear least squares problem, and special solvers exist for this case.
+There shouldn't be an need for fancy methods to fit the initial state: one can take a guess $z₀$, and variationally solve the equation $|Dψ(z)〉dz=|ψ₀〉-|ψ(z)〉$.
+One could even change the conditioning rules as the state propagates, so that $|Dψ〉$ is aligned with $|ψ₀〉-|ψ(z)〉$ to start with, then with $H|ψ(z)〉$ later on.
+That way, you start the actual simulation with a well conditioned approximant.
 
-Some furt
+%% Numerical experiments
+
+I want to measure the numerical stability of the variational dynamics derived from Schrödinger's equation.
+I can do that by measuring how the discretisation error converges as the timestep is reduced.
+The second hypothesis is that the instability is due to stiffness, which can be tested by discretising time in ways with a variety of stability regions, and seeing whether the stability differs.
+
+%%%%
+
+\item{4.} When the least squares problem is solved with a different distance function, which sets distances along the large $N$ directions to zero, the fit fails on the first step.
+Note that this isn't really a distance.
+Also, I did it by solving $P|h〉=P|Dψ〉dz$; when $P$ truncates the Fock space, the operator $P|Dψ〉$ is singular.
+Doing a weighted least squares fit, with the many particle directions having low weights, would be near-singular.
 

### 2013-07-10 paper.tex
notes

@@
 was constructed; by substituting $〈N|G|A〉=USV†$, it is clear that the vector
-$$x={1\over √2}\pmatrix{u_R \cr v_R}$$
+$$x={1\over \sqrt2}\pmatrix{u_R \cr v_R}$$
 is an eigenvector, with eigenvalue $σ_R$, and norm 1.  (Provided $N+1>R$.)  From first-order pertubation theory, the derivative of this eigenvalue with respect to amplitude $i$ is
@@
 \centerline{\XeTeXpicfile psv.pdf width \hsize}
+
+Given the gradient $Dσ_R$, we wish to find a direction where $σ_R$ increases most rapidly for a given change in $|ψ〉$.
+Given a small change $|dψ〉$ lying in the tangent plane, the corresponding change in $z$ is given by $dz=〈Dψ^{+}|dψ〉$, where $〈Dψ^{+}|$ is the pseudoinverse of $|Dψ〉$.
+The change in $σ_R$ is $dσ_R=Dσ_Rdz=Dσ_R〈Dψ^{+}|dψ〉$.
+So the gradient direction in Hilbert space is $|Dψ^{+}〉Dσ_R†$; the corresponding direction in $z$ space is $〈Dψ^{+}|Dψ^{+}〉Dσ_R†$.
+In terms of the SVD $|Dψ〉=|U〉SV†$, this is $〈U|S^{-2}|U〉Dσ_R†$.
+Here, $|ψ〉$ is considered a function of $4R$ real variables instead of $2R$ complex ones.
 

### 2013-07-11 paper.tex
writing

@@
 When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.
-They soon understood the free particle and turned to the harmonic oscillator, investigating its ground state and how its excited states generate sinusoidal motion in the classical limit.
+They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.
 These problems were also solved by the gaussian wave packet.
@@
 
-The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but this luck doesn't last.
-Most quantum problems must be solved approximately.
+The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.
 This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.
-To test how well a putative solution approximates the dynamics of a system, it can be substituted into each side of Schrödinger's equation, and the difference between them evaluated by the usual norm in Hilbert space.
-It remains to choose a class of approximants, and the gaussian wave packets are an common choice.
-They differ in their scale and their central position and momentum, useful properties to determine for any quantum state.
+To test how well a wave function approximates the dynamics of a system, the difference between the two sides of Schrödinger's equation can be measured by the Hilbert space norm.
+It remains to choose a class of approximants.
+The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.
 
-The laser was invented, and physicists needed a quantum theory of intense radiation.
-This obeyed simple laws of dynamics, very similar to harmonic oscillators.
-However, It involved up to $10^{18}$ photons, spread across thousands of modes, so the difficulty of the quantum problem came from its sheer size.
-This was made managable by the work of Glauber, who showed that any quantum state could be written as a superposition of coherent states.
-For optical systems, close to harmonic oscillators, this was a very efficient way to represent states.
+The invention of the laser called for a quantum theory of intense radiation.
+Laser beams are difficult because their quantum descriptions are so large: they involve up to $10^{18}$ photons, distributed across thousands of modes.
+The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.
+Laser light is very coherent, so these expansions often represent it efficiently.
 Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.
-These have solved many problems, but are now limited by intractable convergence problems.
+When atomic Bose-Einstein condensates were made, and very coherent matter became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.
+They have solved many problems, but are limited by intractable convergence issues.
 
-In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using superpositions of coherent states as the approximant.
+In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.
 This is an appealing approach.
-Every quantum state can be approximated arbitrarily well as the number of components in the superposition becomes large.
-The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.
+Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.
+The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.
 
@@
 
-Since Heller suggested it, there have been several attempts to apply this numerically.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
+Since Heller suggested it, there have been several attempts to apply this numerically, but none has entirely succeeded.
+In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
 %% Consequence

### 2013-07-14 paper.tex
writing

@@
 
+We need a way to discretise the function $g$.
+There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
+The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$.
+Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
+Here, $w_r$ is a real weight, intended to compensate for the irregularity of the sample set, so that the sampled vector $\vec g=\bigl(w₁g(α₁),…,w_Rg(α_R)\bigr)^{\rm T}$ has a similar norm to the function $g$.
+
+Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.
+The obvious choice is $g_r(α*)=(〈α_r|G^+)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
+The condition becomes
+$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
+	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
+	& & \ddots & \cr
+	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
+\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
+= \pmatrix{π\cr π\cr\vdots\cr π}.
+$$
+Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
+It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
+The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.
+This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
+
 The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.
@@
 It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
+
+A numerical example is shown below.
+The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.
+A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $|g(a*)|²$, where $g=G⁺|ψ〉$.
+These are plotted below, as circles in phase space.
+For smaller ensembles, the results were not consistent between draws.
+
+\centerline{\XeTeXpicfile cats.pdf width \hsize}
+
+The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉f=〈A|ψ〉$, where $f$ is a vector of weights such that $|A〉f$ approximates $|ψ〉$.
+Below it is a plot of the expansions over number states of the singular kets $〈n|u_r〉$, where $|A〉=|U〉SV$.
+For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.
+The components of the cat have coherent amplitude 2, so their largest number component is $n=4$, and the components drop quickly as $n$ increases further.
+Above $r=15$, the singular ket $|u_r〉$ has very small components on these number states, so it hardly overlaps $|ψ〉$.
+At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.
+In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied.
+Seeing that $G$ is perfectly conditioned, this is a very bad discretisation.
+
+The final graph is an L curve when the discrete problem is Tychonov regularised.
+It shows the expected shape, with minimum residual around $2×10^{-8}$.
+Six regularised solutions are shown above; the modulus of $f_r$ is the disk's saturation, and the phase determines its hue.
+Note that least norm regularisation has the desirable property that adjacent $α_r$ have similar $f_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.
+
+\centerline{\XeTeXpicfile catpicard.pdf width \hsize}
+
+\centerline{\XeTeXpicfile catsws.pdf width \hsize}
+
+\centerline{\XeTeXpicfile catl.pdf width \hsize}
 

### 2013-07-15 paper.tex
morning writing

@@
 There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
-The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$.
+The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr).
 Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
@@
 
-The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
+The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.  This can also be seen from the expansions of coherent states.  The normalised entire expansion of $|α〉$ is
+$$g(β*)=〈β|α〉=e^{-½|α|²-½|β|²+β*α},$$
+whence
+$$\|g\|²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
+The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.
+
+The lemma gives a very simple singular value expansion of $G$.
+Because there is only one nonzero singular value, any orthonormal basis of ket space will do as the left singular vectors.
+Let's choose the number states for familiarity.
+Then
+$$Gg=∑_n |n〉π\int ν_n*(α)g(α)\,d²α=|N〉π\int ν*(α)g(α),$$
+where $|N〉=\bigl(|0〉\ |1〉\ |2〉\ …\bigr)$ and $ν(α)=\bigl(ν₀(α)\ ν₁(α)\ …\bigr)$ could be thought of as infinite matrices whose columns are the left singular vectors $|n〉$ and their normalised entire expansions, the right singular vectors $v_n$.
+The pesudoinverse,
+$$G⁺|ψ〉=νπ〈N|ψ〉= ∑_n ν_nπ〈n|ψ〉,$$
+maps $|ψ〉$ to its normalised entire expansion.
+Expanding over number states is useful because the maximum number in a system is usually limited by conservation of energy.
+In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.
+
 Numerical experiments in the script {\tt sve.m} support this.

### 2013-07-15 paper.tex
writing

@@
 
+There are now three goals for optimisation of $\dot z$:
+\item{1.} The ket $|ψ(z)〉$ solves Schrödinger's equation.  I.e., $|Dψ(z)〉\dot z=H|ψ(z)〉$.
+\item{2.} Singular vectors of $|Dψ〉$ belonging to small singular values do not dominate the solution.  This is controlled by a Tychonov condition, $\dot z=0$, or perhaps by truncation.
+\item{3.} The condition of $|A(z)〉$ improves.
+This can be set up as a least squares condition $Dσ_R(z)\dot z+Cσ_R(z)\dot z*=|σ_R||\dot z|_{\rm opt}$, given a desired rate of change of $z$.
+No doubt there is a better way.
+
+Putting these conditions together, with regularisation parameters $λ$ and $ε$, gives the problem
+$$\pmatrix{|Dψ(z)〉& 0 \cr λI & 0 \cr Dσ_R(z) & Cσ_R(z)}
+	\pmatrix{\dot z\cr \dot z*} =
+	\pmatrix{H|ψ(z)〉\cr 0\cr |σ_R||\dot z|_{\rm opt}}.$$
+This raises a technical problem, because we need to constrain $\dot z*$ to be the conjugate of $\dot z$.
+Alternatively, if we write $z=x+iy$, we need to solve a complex least squares problem under the constraint that the solution is real.
+Neither of those sounds trivial, but it shouldn't be impossible.
+
 Below is a plot of the condition number of $|Dψ(z)〉$ for a superposition, as the value of $z$ varies by $0.1$ in every direction.

### 2013-07-16 paper.tex
writing

@@
 There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
-The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr).
+The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr)$.
 Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
@@
 
-Putting these conditions together, with regularisation parameters $λ$ and $ε$, gives the problem
+Putting these conditions together, with regularisation parameters $λ$ and $ε$, gives the least squares problem
 $$\pmatrix{|Dψ(z)〉& 0 \cr λI & 0 \cr Dσ_R(z) & Cσ_R(z)}
-	\pmatrix{\dot z\cr \dot z*} =
+	\pmatrix{\dot z\cr \dot z*} ≈
 	\pmatrix{H|ψ(z)〉\cr 0\cr |σ_R||\dot z|_{\rm opt}}.$$
 This raises a technical problem, because we need to constrain $\dot z*$ to be the conjugate of $\dot z$.
-Alternatively, if we write $z=x+iy$, we need to solve a complex least squares problem under the constraint that the solution is real.
-Neither of those sounds trivial, but it shouldn't be impossible.
+In the space of complex numbers, this constraint is not linear, because $(cz)*≠cz*$.
+However, that is true if $c$.
+Moreover, if $z=x+iy$, then $|z-w|²=|x-\Re w|²+|y-\Im w|²$, so a least squares problem in $z$ has the same solution as the equivalent least squares problem in $x$ and $y$.
+
+Consider the problem 
+$$ \pmatrix{A & B}\pmatrix{z\cr z*}≈c,$$
+with $z∈C^m$, $c∈C^n$, and $A,B∈C^{m×n}$.  If $z=x+iy$, the left hand side can be rearranged
+$$\pmatrix{A & B}\pmatrix{z\cr z*}
+	=\pmatrix{A & B}\pmatrix{I&iI\cr I&-iI}\pmatrix{x\cr y}
+	=\pmatrix{A+B & iA-iB}\pmatrix{x\cr y}.$$
+Separating the rows into real and imaginary parts gives
+$$\pmatrix{\Re (A+B) & \Re(iA-iB)\cr\Im(A+B) & \Im(iA-iB)}\pmatrix{x\cr y}
+	=\pmatrix{\Re A+\Re B & \Im B-\Im A\cr\Im A+\Im B & \Re A-\Re B)}
+		\pmatrix{x\cr y}≈\pmatrix{\Re c\cr \Im c}.$$
 

### 2013-07-16 zls.m
complex conjugate least squares solver

@@ -1,0 +1,13 @@
+function [ z ] = zls( X, b )
+%zls Solve a least squares problem for a complex number and its conjugate
+%   The problem solved is X*[z; conj(z)] = b
+
+n = 0.5*numel(b);
+Xl = X(:,1:n); Xr = X(:,n+1:end);
+A = [real(Xl+Xr) real(1i*Xl-1i*Xr);...
+    imag(Xl+Xr) imag(1i*Xl-1i*Xr)];
+xy = A \ [real(b); imag(b)];
+z = xy(1:n)+1i*xy(n+1:end);
+
+end
+

### 2013-07-16 paper.tex
derivations

@@
 $$σ_R²v₁†d{\cal V}v₁-σ₁²v_R†d{\cal V}v_R \over σ_R².$$
-Given a proposed change $dz$, we can find the effect on the condition number.
+Given a proposed change $dz$, we can find the effect on the condition number.  However, it would be simpler to constrain the smallest singular value, and likely just as effective.
 
-The question is how to apply this to condition the process of finding $dz$.
-Given $z$, we can find out which directions of $dz$ correspond to large changes in the condition number.
-However,  until we know $dz$, we don't know if these changes help or hurt.
-The process might run like this: first, truncate the singular values.
-Next, find the components $〈U_i|h〉$, and the corresponding components $v_iσ_i〈U_i|h〉$ of $dz$, given the truncation.
-Then, truncate the components that will cause $C(z)$ to increase to values that limits the increase to an acceptable amount.
+Let 
+$$|A(α)〉=(|α₁〉 |α₂〉 \cdots |α_R〉)=(\cdots e^{-½|α_r|²+α_ra†}\cdots)|0〉.$$
+This is not an analytic function, so there is no Cauchy derivative such that $|A(z+h)〉=|A(z)〉+|DA(z)〉h+O(h²)$.
+There are, however, Wirtinger derivatives such that $|A(z+h)〉=|A(z)〉+|DA(z)〉h+|CA(z)〉h†$.
+As a point of notation, let $〈A(z+h)|=〈A(z)|+h†〈CA(z)|+h〈DA(z)|$, so that $〈CA(z)|$ is the conjugate bra to $|DA(z)〉$.
+Let $|DA(z)〉=(|∂₁A(z)〉\cdots |∂_RA(z)〉)$, and $|CA(z)〉=(|∂*₁A(z)〉\cdots |∂*_RA(z)〉)$.
+Because $|A〉$ is a matrix valued function, these are also matrices, with $|DA〉$ and $|CA〉$ being higher rank tensors.
+I hope the product notation is not too confusing.
+The Wirtinger derivatives are
+$$|∂_rA(α)〉=\pmatrix{0 & … & 0 & (-{α*_r\over2}+a†)|α_r〉& 0 & … & 0},$$
+and
+$$|∂*_rA(α)〉=\pmatrix{0 & … & 0 & -{α*_r\over2}|α_r〉& 0 & … & 0}.$$
 
-I've figured out how to find the change in the condition number by pertubation theory.
-That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.
-How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?
-This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.
-Is there a better way?
+To find the derivative of the smallest singular value $σ_R(α)$, consider the matrix
+$$M(α)=\pmatrix{0 & |A(α)〉\cr 〈A(α)| & 0}.$$
+Here, $|A(α)〉$ maps weight vectors in $C^R$ to kets, while $〈A(α)|$ maps kets back to column vectors of inner products.
+The operation of the zeros is obvious; formally, they need to be a row of null kets and a column of null bras.
+The matrix operates on vectors of a ket followed by $R$ complex numbers.
+Let $|A(α)〉=|U〉SV†$ be a singular value decomposition, whence $〈A(α)|=VS〈U|$.
+The vector
+$$x={1\over \sqrt2}\pmatrix{|u_R〉 \cr v_R}$$
+is a normalised eigenvector  of $M$, belonging to the eigenvalue $σ_R(α)$.  From first-order pertubation theory, the derivative of this eigenvalue with respect to $α_r$ is
+$$∂_r σ_R=x†∂_rMx\qquad ∂*_r σ_R=x†∂*_rMx.$$
+Then 
+$$\eqalign{∂_r σ_R(α)&=x†∂_rM(α)x\cr
+	&=x†\pmatrix{0 & |∂_rA(α)〉\cr 〈∂_rA(α)| & 0}x \cr
+	&=½\pmatrix{〈u_R| & v†_r}
+		\pmatrix{0 & \matrix{0& (-{α*_r\over2}+a†)|α_r〉&0}\cr 
+			\matrix{0\cr -{α_r\over2}〈α_r|\cr 0}& }
+		\pmatrix{|u_R〉 \cr v_R} \cr
+	&=-½\Re(α*_rv_R^r〈u_R|α_r〉)+½v_R^r〈u_R|a†|α_r〉,
+}$$
+and similarly
+$$∂*_r σ_R(α)=-½\Re(v_R^{r\ast}α_r〈α_r|u_R〉)+½v_R^{r\ast}〈α_r|a|u_R〉.$$
+More simply, if $α=x+iy$,
+$${∂σ_R\over ∂x} = ∂_r σ_R+∂*_r σ_R=\Re(v_R^r〈u_R|(a†-α*_r)|α_r〉),$$
+$${∂σ_R\over ∂y} = i∂_r σ_R-i∂*_r σ_R=-\Im(v_R^r〈u_R|a†|α_r〉).$$
 
@@
 The black lines show changes in the real parts of the amplitudes, the red lines in the imaginary parts.
-The dotted black line shows the derivatives for the real parts calculated by pertubation theory, as follows.
-The derivatives of the matrix $〈N|G|A〉$  with respect to each real part were computed by symmetric difference, with $h=0.01$.
-Here, $|Α〉=(|α₁〉 |α₂〉 \cdots |α_R〉)$.
-A matrix
-$$M=\pmatrix{0 & 〈N|G|A〉\cr 〈A|G|N〉 & 0}$$
-was constructed; by substituting $〈N|G|A〉=USV†$, it is clear that the vector
-$$x={1\over \sqrt2}\pmatrix{u_R \cr v_R}$$
-is an eigenvector, with eigenvalue $σ_R$, and norm 1.  (Provided $N+1>R$.)  From first-order pertubation theory, the derivative of this eigenvalue with respect to amplitude $i$ is
-$$∂_i σ_i=x†∂_iMx.$$
-As shown in the plots, lines of this slope fit the singular value very well.
+The dotted black line shows the derivatives for the real parts calculated by pertubation theory, finding the derivatives of $〈N|∂_rA〉$ by symmetric difference, with $h=0.01$.
+Lines of this slope fit the singular value very well.
 
@@
 No doubt there is a better way.
+\item{4.} There might need to be fourth condition, to prevent amplitudes with small weights from wandering off to infinity.
 

### 2013-07-17 paper.tex
writing

@@
 \title{Stability of variational quantum dynamics with Gaussian approximants}
+
+\def\G{|Γ〉}
+\def\cite#1{$[\hbox{\tt #1}]$}
 
@@
 The invention of the laser called for a quantum theory of intense radiation.
-Laser beams are difficult because their quantum descriptions are so large: they involve up to $10^{18}$ photons, distributed across thousands of modes.
+It is difficult to analyse laser beams because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.
 The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.
@@
 
-When used on computers, the variational methods expand quantum states over finitely many coherent states.
-This is a discretisation of the expansion introduced by Glauber, a mapping $G$ from square-integrable functions to Fock space, where
-$$Gf={1\over π}\int g(α*)|α〉\,d²α.\eqno{(1)}$$
-This is different from Glauber's use: he uses $f=ge^{½|α|²}$.
-Is it easier to work Glauber's way, with the $e^{-½|α|²}$ in the mapping, so that the singular vectors are entire functions?
-You'd have to use a Bargmann inner product to normalise them; the theory of that is well developed.
+%% Phase space expansion as a linear operator
 
-%% Phase space expansion as a linear operator
+The expansion operator introduced by Glauber\cite{prx-131-2766} is
+$$\G f={1\over π}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
+The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.
+The function $f$ is a function of a complex variable, such that $e^{-½|α|²}f(α)$ is square-integrable.
+Bargmann\cite{} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; we will call it the space of phase space distributions, or just distributions.
+In this paper, the application of a linear function is written as a product, and functions whose range is ket space are denoted by kets, usually with a greek letter.
+So $\G f$ is the ket that results from applying the linear operator $\G$ to the phase space distribution $f$.
+
+The state $|α〉$ is a coherent state with complex amplitude $α$.
+These can be defined in a number of ways, perhaps most simply in second quantisation, as eigenstates of a boson annihilation operator, with $a|α〉=α|α〉$.
+The annihilation operator is not hermitian, so the eigenvalues $α$ are complex numbers.
+In first quantisation, such an operator is the complex amplitude
+$$a=?(λx+ip/λ),\qquad\hbox{Get constants from Cahill}$$
+where $x$ and $p$ are the position and momentum operators; the bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency ?.
+
+Coherent states are not orthogonal, but their brackets follow the law
+$$〈β|α〉=e^{-½|β|²-½|α|²+β*α},$$
+whence
+$$|〈β|α〉|²=e^{-|β-α|²}.$$
+Their projector follows
+$$\int |α〉〈α|\,d²α=π.$$
+For an orthonormal basis, the right hand side would be 1 instead of $π$.  It follows that the operator $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=e^{½|α|²}〈α|ψ〉$.  Then
+$$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
+Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$, so $f₀$ is an entire function.
+Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.
+Further properties of the coherent states will be stated as required.
+Glauber's summary of them is paraphrased in every quantum optics textbook ever written.
+
+Obviously, $\G$ is not one to one, because the images of the entire functions already fill ket space.
 

### 2013-07-18 paper.tex
derivative of least SV figured out

@@
 $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
-Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$, so $f₀$ is an entire function.
+Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.
 Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.
@@
 
-To solving our differential equation, we need to solve an iterated least squares problem.
-We need to do so in a way that preserves the condition of $|Dψ〉$.
-When we shift $z$ in the direction $v_i$ as $dz=λv_i$, how does this condition number change?
-Without loss of generality, it suffices to consider the effect on the smallest and largest singular values, that determine $C(z)$.
-Note that $v_1$ and $v_R$ are eigenvectors of the matrix ${\cal V}(z)=〈Dψ(z)|Dψ(z)〉$, belonging to the eigenvalues $σ₁²$ and $σ_R²$; this is a familiar problem in pertubation theory.
-Assume for now that we have found $D{\cal V}(z)$—this won't be analytic, so we will need to use real and imaginary parts, or Wirtinger calculus.
-The change in $\cal V$ is of the form $d{\cal V}=∂_i{\cal V}(z)λ+∂*_i{\cal V}(z)λ*$, where the partial derivatives are taken in the directions $v_i$.
-
-As is well known in pertubation theory, the change in the eigenvalue $σ₁²$ corresponding to a small change $d{\cal V}$ is $v₁†d{\cal V}v₁$, and similarly for $σ_R²$.
-Therefore the change in $C(z)²$ is
-$$σ_R²v₁†d{\cal V}v₁-σ₁²v_R†d{\cal V}v_R \over σ_R².$$
-Given a proposed change $dz$, we can find the effect on the condition number.  However, it would be simpler to constrain the smallest singular value, and likely just as effective.
+To solve our differential equation, we need to solve an iterated least squares problem.
+We need to do so in a way that preserves the condition of $|A(z)〉$.
+When we shift $z$ in the direction $\dot z$, how does the condition change?
+The issue is the small singular values, so it makes sense to constrain the smallest singular value $σ_R$.
+This is a function of $z$.
+We proceed to calculate its derivative by first order pertubation theory.
 
 Let 
-$$|A(α)〉=(|α₁〉 |α₂〉 \cdots |α_R〉)=(\cdots e^{-½|α_r|²+α_ra†}\cdots)|0〉.$$
+$$|A(α)〉=\pmatrix{|α₁〉& \cdots &|α_R〉}
+	=\pmatrix{\cdots &e^{-½|α_r|²+α_ra†}&\cdots}|0〉.$$
 This is not an analytic function, so there is no Cauchy derivative such that $|A(z+h)〉=|A(z)〉+|DA(z)〉h+O(h²)$.
-There are, however, Wirtinger derivatives such that $|A(z+h)〉=|A(z)〉+|DA(z)〉h+|CA(z)〉h†$.
-As a point of notation, let $〈A(z+h)|=〈A(z)|+h†〈CA(z)|+h〈DA(z)|$, so that $〈CA(z)|$ is the conjugate bra to $|DA(z)〉$.
-Let $|DA(z)〉=(|∂₁A(z)〉\cdots |∂_RA(z)〉)$, and $|CA(z)〉=(|∂*₁A(z)〉\cdots |∂*_RA(z)〉)$.
-Because $|A〉$ is a matrix valued function, these are also matrices, with $|DA〉$ and $|CA〉$ being higher rank tensors.
-I hope the product notation is not too confusing.
-The Wirtinger derivatives are
-$$|∂_rA(α)〉=\pmatrix{0 & … & 0 & (-{α*_r\over2}+a†)|α_r〉& 0 & … & 0},$$
-and
-$$|∂*_rA(α)〉=\pmatrix{0 & … & 0 & -{α*_r\over2}|α_r〉& 0 & … & 0}.$$
+There are Wirtinger derivatives such that $|A(z+h)〉=|A(z)〉+|DA(z)〉h+|CA(z)〉h†$.
+However, in this case it is easier to work with real and imaginary parts, defining
+$$|A'(x,y)〉=\pmatrix{|x₁+iy₁〉&\cdots &|x_R+iy_R〉}
+	=\pmatrix{\cdots &e^{-½x_r²-½y_r²+(x_r+iy_r)a†}&\cdots}|0〉.$$
+The partial derivatives are easily determined to be
+$$|∂_rA'(x,y)〉=\cases{
+	\pmatrix{0 & (-x_r+a†)|x_r+iy_r〉 & 0} & $1≤ r≤ R$, \cr
+	\pmatrix{0 & (-y_r+ia†)|x_r+iy_r〉 & 0} & $R<r≤ 2R$.}$$
+Because $|A〉$ matrix-valued, its total derivative is some kind of third rank tensor, and its partial derivatives are simpler to work with.
+The total derivative $Dσ_R$ is just a row vector, and will be used often.
 
-To find the derivative of the smallest singular value $σ_R(α)$, consider the matrix
-$$M(α)=\pmatrix{0 & |A(α)〉\cr 〈A(α)| & 0}.$$
-Here, $|A(α)〉$ maps weight vectors in $C^R$ to kets, while $〈A(α)|$ maps kets back to column vectors of inner products.
+Let the singular value decomposition of $|A(α)〉$ be $|A(α)〉=|U〉SV†$.  The standard ways to compute singular values use the matrix
+$$M(x,y)=\pmatrix{0 & |A'(x,y)〉\cr 〈A'(x,y)| & 0}.$$
+Here, $|A(x,y)〉$ maps weight vectors in $C^R$ to kets, while $〈A(x,y)|=VS〈U|$ maps kets back to column vectors of inner products.
 The operation of the zeros is obvious; formally, they need to be a row of null kets and a column of null bras.
-The matrix operates on vectors of a ket followed by $R$ complex numbers.
-Let $|A(α)〉=|U〉SV†$ be a singular value decomposition, whence $〈A(α)|=VS〈U|$.
-The vector
+The matrix $M(x,y)$ operates on a vector comprising a ket above $R$ complex numbers.
+Now, the vector
 $$x={1\over \sqrt2}\pmatrix{|u_R〉 \cr v_R}$$
-is a normalised eigenvector  of $M$, belonging to the eigenvalue $σ_R(α)$.  From first-order pertubation theory, the derivative of this eigenvalue with respect to $α_r$ is
-$$∂_r σ_R=x†∂_rMx\qquad ∂*_r σ_R=x†∂*_rMx.$$
-Then 
-$$\eqalign{∂_r σ_R(α)&=x†∂_rM(α)x\cr
-	&=x†\pmatrix{0 & |∂_rA(α)〉\cr 〈∂_rA(α)| & 0}x \cr
-	&=½\pmatrix{〈u_R| & v†_r}
-		\pmatrix{0 & \matrix{0& (-{α*_r\over2}+a†)|α_r〉&0}\cr 
-			\matrix{0\cr -{α_r\over2}〈α_r|\cr 0}& }
-		\pmatrix{|u_R〉 \cr v_R} \cr
-	&=-½\Re(α*_rv_R^r〈u_R|α_r〉)+½v_R^r〈u_R|a†|α_r〉,
-}$$
-and similarly
-$$∂*_r σ_R(α)=-½\Re(v_R^{r\ast}α_r〈α_r|u_R〉)+½v_R^{r\ast}〈α_r|a|u_R〉.$$
-More simply, if $α=x+iy$,
-$${∂σ_R\over ∂x} = ∂_r σ_R+∂*_r σ_R=\Re(v_R^r〈u_R|(a†-α*_r)|α_r〉),$$
-$${∂σ_R\over ∂y} = i∂_r σ_R-i∂*_r σ_R=-\Im(v_R^r〈u_R|a†|α_r〉).$$
+is a normalised eigenvector  of $M$, belonging to the eigenvalue $σ_R(x,y)$.
+As is well known in first-order pertubation theory, the derivative of this eigenvalue with respect to $x$ and $y$ is given by
+$$∂_rσ_R=x†(∂_rM)x
+	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA〉v_R\right).$$
+Substituting the partial derivatives gives
+$$∂_rσ_R=\cases{\Re\left(〈u_R|(-x_r+a†)|α_r〉v_R^r\right) &  $1≤ r≤ R$, \cr
+	\Re\left(〈u_R|(-y_r+ia†)|α_r〉v_R^r\right) & $R<r≤ 2R$.}$$
 
 The accuracy of the pertubation method is shown below.
-The state is coherent, with ampltiude 2.
-It was sampled at 5 amplitudes, shown on the left hand plot.
-The weights were assigned by inverting the discretised $G$ operator on these sample points, with small singular values truncated by Matlab's {\tt pinv} function.
-The the moduli of the resulting weights are shown in the right hand plot.
-This method of fitting usually gives a Hilbert space residual around 0.01.
-
-The lower plot shows how the smallest singular value of the discrete $G$ changes, as the sample points are shifted.
-The black lines show changes in the real parts of the amplitudes, the red lines in the imaginary parts.
-The dotted black line shows the derivatives for the real parts calculated by pertubation theory, finding the derivatives of $〈N|∂_rA〉$ by symmetric difference, with $h=0.01$.
-Lines of this slope fit the singular value very well.
+A vector of 5 amplitudes was drawn from a normal distribution around amplitude 2, shown in the upper plot.
+The lower plots show how the smallest singular value of $|A〉$ varies as the amplitudes shift.
+The five columns show changes in the five amplitudes.
+The black lines show how $σ_R(x,y)$ varies with the real part $x$, the red lines how it varies with $y$.
+The dotted lines show the linear approximation derived above by pertubation theory, and the second row of plots show the error in the approximation.
+The linear approximation is good; presumably it holds for changes small compared to the separation of the amplitudes.
 

### 2013-07-18 svgrad.m

@@ -1,0 +1,60 @@
+% Sample a state by inverting G on random amplitudes, then look for
+% unimportant singular directions of |Dpsi> along which the least singular
+% value of G varies rapidly.
+
+global N; N = 15; brackets
+R = 5;
+hs = -0.1:0.02:0.1;
+c0 = nq*evan(-2,2);
+
+% Random sample amplitudes
+a = 2+sqrt(0.5)*randn(R,2)*[1; 1i];
+G = nq*evan(-0.5*abs(a).^2,a);
+ef = pinv(G)*c0;
+f = -0.5*abs(a).^2 + log(ef);
+
+figure
+plot(real(a), imag(a), '*k');
+set(gca, 'DataAspectRatio',[1 1 1])
+for r = 1:R
+    text(real(a(r)), imag(a(r)), sprintf('  %d',r))
+end
+title('Complex amplitudes')
+
+[Ug,Sg,Vg] = svd(G, 'econ');
+u = Ug(:,end); v = Vg(:,end); ev = [u; v];
+sigma = Sg(end,end);
+
+cns = zeros(2*R, numel(hs)); cns(:) = nan;
+basis = [eye(R) 1i*eye(R)];
+for p = 1:2*R
+    for q = 1:numel(hs)
+        da = hs(q)*basis(:,p);
+        [U,S,V] = svd(nq*evan(-0.5*abs(a+da).^2, a+da), 'econ');
+        cns(p,q)=S(end,end);
+    end
+end
+
+gsym = zeros(1,2*R)
+for p = 1:R
+    x = real(a(p))*eye(N+1);  y = imag(a(p))*eye(N+1);
+    gsym(p) = real(v(p)*u'*(aop'-x)*G(:,p));
+    gsym(p+R) = real(v(p)*u'*(1i*aop'-y)*G(:,p));
+end
+   
+figure
+for p = 1:R
+    subplot(2,R,p)
+    plot(hs, cns(p,:), '-k', hs, cns(p+R,:), '-r', ...
+        hs, sigma+gsym(p)*hs, ':k', hs, sigma+gsym(p+R)*hs, ':r')
+    if p==1
+        ylabel('\sigma_r')
+    elseif p==3
+        title('Variation of \sigma_r with x_r (black) and y_r (red)')
+    end
+    subplot(2,R,p+R)
+    plot(hs, cns(p,:)-sigma-gsym(p)*hs, '-k', ...
[…]

### 2013-07-18 paper.tex
writing

@@
 
+%% convergence to the initial state
+
+As discussed below, the time-dependent Schrödinger equation for a quartic oscillator is numerically difficult, no matter how it is discretised.
+The following problem aimed to solve an intrinsically simple differential equation,
+$$ {d\over ds}|ψ(z(s))〉=|ψ₀〉-|ψ(z(s))〉.$$
+Here, $|ψ₀〉$ was the half cycle cat state; $|ψ(z)〉$ had 15 components, and was initialised by sampling random, and linearly fitting weights.
+The result is shown on the left below.
+The clusters of complementary colours indicate components that nearly cancel each other out; these should not be present when the condition of $|A〉$ is constrained, but are expected from random sampling.
+They might also indicate too small a Tychonov parameter: large parameters will tend to give nearby points equal weights.
+
+\centerline{\XeTeXpicfile expens.pdf width \hsize}
+
+The equation was solved, in the script {\tt hzero.m}.
+The results are shown below, and the final superposition is plotted on the right above.
+The second graph below isthe smallest singular value of $|A〉$.
+Many draws of initial samples give the exponential decay one would expect.
+This one gave two surprises: initially, the distance between $|ψ〉$ and $|ψ₀〉$ increases with time.
+This is entirely wrong, according to the differential equation.
+The initial state was a good approximation to $|ψ₀〉$, so there might not be much scope to improve the approximant.
+However, it should certainly not move away from the target.
+
+The variational method was the least distance in Hilbert space, with Tychonov regularisation.  Time was discretised by Euler's formula, which should solve exponential decay very stably.
+
+The second surprise is at large $s$, where the approximant improves, but so does the condition of $|A〉$.
+In most draws of initial samples, the approximant improves by clustering the amplitudes towards the states $|2i〉$ and $|-2i〉$, which makes the conditioning worse.
+
+\centerline{\XeTeXpicfile exp.pdf width \hsize}
+
 %% Shape of Hilbert space trajectories

### 2013-07-18 evan.m

@@ -1,15 +1,17 @@
-function X = evan(f,a)
+function [X, f] = evan(f,a)
 % evan: Calculate the parts of Fock space brackets that depend on the
 % amplitudes of the superposition.
 % 
 % X = evan(z)
 % X = evan(f, a)
+% [X, f] = evan(a, 'even')
 %
 % npsi = sum(nq*evan(z),2)
 % ndpsi = [nq*evan(z), ndqr*evan(z)]
 %
 % The amplitudes can be provided as a pair of vectors f and a, or a vector
-% z = [f a].
+% z = [f a].  The 'even' form sets the weights such that each component is
+% normalised, and returns the f vector that does this as a second value.
 
 global N
 
@@ -17,6 +19,10 @@ if nargin==1
     R = 0.5*numel(f);
     a = f(R+1:end); f = f(1:R);
 end
-a = a(:).';
+if strcmp(a, 'even')
+    a = f;
+    f = -0.5*abs(a).^2;
+end
 
+f = f(:);  a = a(:).';
 X = ((ones(N+1,1)*a).^((0:N)'*ones(size(a))))*diag(exp(f));

### 2013-07-19 paper.tex
writing

@@
 For smaller ensembles, the results were not consistent between draws.
+
+A basic question is why regularisation helps.
+Usually, it's applied to the inversion of measured data, where the right hand side of the least squares problem includes noise.
+The aim is to prevent amplified noise from dominating the solution.
+In our problem, the right hand side is calculated instead of measured, and the only noise is rounding errors.
+Those are 16 orders of magnitude smaller than the exact right hand side, so rounding the small singular values up to $10^{-10}$ should be more than adequate.
+We've needed much stronger regularisation than that: therefore, rounding errors aren't the problem.
+
+The only other error I can think of is nonlinearity.
+Tychonov parameters of order $10^{-10}$ do work for the linear expansion problem.
+Perhaps we're not doing regularisation, we're actually establishing a trust region, where the linear least squares problem is a good approximation to the actual change in $|ψ(z)〉$.
+
+If that's the case, it's tempting to make the weights linear instead of logarithmic, so that that part of the problem is exactly linear.
+However, the function $e^φ$ is pretty nearly linear in the range we're shifting $φ$, so that can't help much.
+In fact, the whole problem is quite linear.
+
+It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the l-curve experiments for the expansion problem.
+To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.
+In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around $π^½$.
+And we're back to the other open question, how to relate the norm of a function to the norm of a set of samples at irregular points.
+How can I estimate how much of phase space “belongs” to each amplitude sample, in order to get a volume over which the function has that value?
+
+The obvious way would be to assign amplitudes to the nearest sample.
+That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
+What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
+There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
+Something like $x^{-2}e^{-|x|²} sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
+I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
+What's the best exponent?
+I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
 
@@
 
+There are a few things to note about the initial trajectory.
+The initial weights already satisfy a least squares condition.
+Therefore, the intial change in the superposition should be entirely in the amplitudes: the weights are already optimal.
+If that isn't the case, the regularisation of the expansion problem would have to be inconsistent with that of the variational problem, so that we're accepting variational changes that we previously rejected when we found the initial state.
+
 %% Shape of Hilbert space trajectories

### 2013-07-22 paper.tex
writing

@@
 
-The expansion operator introduced by Glauber\cite{prx-131-2766} is
-$$\G f={1\over π}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
+The expansion introduced by Glauber\cite{prx-131-2766} is performed by the operator
+$$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
 The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.
 The function $f$ is a function of a complex variable, such that $e^{-½|α|²}f(α)$ is square-integrable.
-Bargmann\cite{} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; we will call it the space of phase space distributions, or just distributions.
+Bargmann\cite{} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.
 In this paper, the application of a linear function is written as a product, and functions whose range is ket space are denoted by kets, usually with a greek letter.
@@
 The state $|α〉$ is a coherent state with complex amplitude $α$.
-These can be defined in a number of ways, perhaps most simply in second quantisation, as eigenstates of a boson annihilation operator, with $a|α〉=α|α〉$.
+These can be defined in a number of ways, perhaps most simply in second quantisation, as eigenstates of a boson annihilation operator $a$, with $a|α〉=α|α〉$.
 The annihilation operator is not hermitian, so the eigenvalues $α$ are complex numbers.
@@
 $$|〈β|α〉|²=e^{-|β-α|²}.$$
-Their projector follows
+The projector satisfies the law
 $$\int |α〉〈α|\,d²α=π.$$
-For an orthonormal basis, the right hand side would be 1 instead of $π$.  It follows that the operator $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=e^{½|α|²}〈α|ψ〉$.  Then
+For an orthonormal basis, the right hand side would be 1 instead of $π$.  It follows that the operator $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
 $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
@@
 
-Hypothesis: The singular values of $G$ are 0 and $\sqrt π$, and the space belonging to $\sqrt π$ comprises the functions $f$ for which $fe^{½|α|²}$ is analytic.
+Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.
 
@@
 
-The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.
-
-Assuming that there is no way to expand a state with a smaller norm than a Gaussian entire function, this does it.
+Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.
 Or at least I think so.
@@
 
-To sum up, this analysis shows that $G$ is a projector.
+If the domain and range of an operator are the same, and the operator has singular values 0 and 1, then that operator is an orthogonal projector onto its range.
+The operator $|Γ〉$ is a generalisation of that.
+It is the composition of an orthogonal projection of function space onto the entire functions followed by an isometric mapping of the entire functions onto the kets.
+The orthogonality and isometry are under the Bargmann inner product for phase space functions.
 Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.
@@
 
-The theorems on how quickly singular values of convolution operators decay don't apply here.
-They apply to convolutions of real functions, which can only oscillate at a fixed rate.
-Complex functions have space in the Argand plane to oscillate arbitrarily rapidy; considered as wavefunctions, the coherent states do: the wavenumber is proportional to the imaginary part of $α$.
-
-It turns out that complex convolutions are not like real convolutions: the imaginary part of $t-t'$ can control the rate at which the kernel oscillates.
-Each line of constant imaginary part in the convolution is determined by a limited range of frequencies from the function, but that frequency range can vary with the imaginary part, so that all frequency components of the function contribute to the convolution somewhere.
-Is the Fourier transform a complex convolution?
-
-If the Hypothesis is right, the Picard condition is very obviously satisfied.
-The operator $G$ is a projection.
-The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm.
-Functions in the null space of $G$ don't correspond to left singular vectors, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
+The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
+$$π^{-½}〈β|Γ〉f=π^{-1}∫〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}∫〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
+This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.
+There are many theorems to the effect that singular values of real convolution operators decay rapidly, and the operators are badly conditioned.
+It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.
+I think the loophole is that of real functions can only oscillate at a limited rate, so that their Fourier transforms have finite norm.
+Complex functions have space in the Argand plane to oscillate arbitrarily rapidy.
+And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.
 
@@
 There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
-Something like $x^{-2}e^{-|x|²} sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
+Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
 I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
@@
 I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+
+The Picard condition is very obviously satisfied by the projection operator $|Γ〉$.
+The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm as itself.
+Functions in the null space of $|Γ〉$ don't have components in the left singular directions, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
 

### 2013-07-23 paper.tex
writing

@@
 And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.
+
+%% Finite phase space
+
+The operator $|Γ〉$ acts on functions defined over the whole complex plane.
+For states with high particle number, the smallest preimages have support far from the origin.
+We can not compute with such functions: any discretisation will sample values of $f$ on a finite subset of the plane.
+Unfortunately, the conditioning of these restricted operators will be worse than that of $|Γ〉$.
+We have excluded the small preimages of the states with large particle number, so they must be represented by functions with support close to the origin, but with large norms.
+Hence the restricted operators will have small singular values corresponding to states with many particles.
+
+Define a restricted operator
+$$|Γ_χ〉f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
+For now, suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.
+The most obvious examples are characteristic functions, where $χ$ is 1 inside a circle of radius $b$, and 0 outside.
+However, the formalism allows more gradual truncation as well.
+We can rewrite this
+$$\eqalign{|Γ_χ〉f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
+	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
+	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
+where
+$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
+Presumably, the right singular vectors have the $f_n$ of least norm such that 
+$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$
+
+What is the appropriate norm to use?  In the case that $χ$ is a characteristic function on the domain $D$, it is presumably the truncated Bargmann norm
+$$〈g,f〉=\int_D e^{-|α|²}g*(α)f(α)\,d²α=\int e^{-|α|²}χ(α)g*(α)f(α)\,d²α.$$
+Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
+$$\|f_n\|²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
+given 
+$$2\sqrt π(n!)^{-½}\int_0^b r^{n+1}e^{-r²}f_n(r)dr=1.$$
+If one considers changing $f$ by $Δa$ in an interval around $a$, the Lagrange condition is
+$$re^{-r²}f_n(r)\propto r^{n+1}e^{-r²},$$
+or 
+$$f_n(r)\propto rⁿ.$$
+The analytic functions still win.  
+In this case, the integrals are identical, so, for the preimage of $|n〉$,
+$$\|f_n\|²=\sqrt{πn!},$$
+and, presuming $|n〉$ is the singular vector, the singular value is
+$$σ_n=(πn!)^{-1/4}.$$
+I've done something wrong, because the result doesn't depend on $b$.
+
+Numerical experiments in the script {\tt sve.m} support this.
+The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
+Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
+The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
+
+\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
+
+The following three figures show the right singular vectors, sampled versions of phase space functions.
+The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
+These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
+In the continuous limit there is only one singular value, and the vectors are degenerate.
+For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
+As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
+
+\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
+\XeTeXpicfile rsv3.pdf width 0.5\hsize}
+\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+
+Reducing the dimension of Fock space so that the problem is underdetermined means that, when the problem is solved by SVD, we only take components along $N+1$ directions in Fock space, and map them to $N+1$ directions in parameter space.
+The remaining $2R-N-1$ components of the solution are zero.
+As we add extra dimensions to Fock space, we are taking components in more Fock space directions, which include more oscillating components of the wavefunction.
+This is like sampling a function on a finer grid.
+With these directions, we can form some of the smaller singular vectors of the full problem.
 
@@
 In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.
-
-Numerical experiments in the script {\tt sve.m} support this.
-The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
-Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
-The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
-
-\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
-
-The following three figures show the right singular vectors, sampled versions of phase space functions.
-The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
-These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
-In the continuous limit there is only one singular value, and the vectors are degenerate.
-For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
-As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
-
-\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
-\XeTeXpicfile rsv3.pdf width 0.5\hsize}
-\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
-
-Reducing the dimension of Fock space so that the problem is underdetermined means that, when the problem is solved by SVD, we only take components along $N+1$ directions in Fock space, and map them to $N+1$ directions in parameter space.
-The remaining $2R-N-1$ components of the solution are zero.
-As we add extra dimensions to Fock space, we are taking components in more Fock space directions, which include more oscillating components of the wavefunction.
-This is like sampling a function on a finer grid.
-With these directions, we can form some of the smaller singular vectors of the full problem.
 

### 2013-07-23 paper.tex
configured home laptop

@@
 
-Obviously, $\G$ is not one to one, because the images of the entire functions already fill ket space.
+Prior to Glauber, Bargmann pointed out that the Hilbert space of wavefunctions is isomorphic to the space of entire functions of a complex variable, under the inner product
+$$\langle g,f\rangle=\int g*(z)f(z)\,d²z.$$
+He introduced the isometry
+$$\langle Γ^+|ψ\rangle(α)=π^{-½}\langle α|ψ\rangle,$$
+which Glauber showed was a pseudoinverse of $|Γ\rangle$, in the sense that
+$$|Γ\rangle\langle Γ^+|=1.$$
+The other combination, $\langle Γ^+|Γ\rangle$, is an orthogonal projector from the functions of a complex variable onto the entire functions, under the Bargmann inner product.
+It follows that $f=\langle Γ^+|ψ\rangle$ is the function with least norm that satisfies $|Γ\rangle f=|ψ\rangle$.
+As shown above, $\|f\|=\| |ψ\rangle\|$.
+
+If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ^+|Γ\rangle$.
+Therefore the singularity of $|Γ\rangle$ is easily resolved.
+If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ^+|ψ\rangle$.
+And $\langle Γ^+|$, an isometry, is as nice a function as could possibly be desired.
+So it is curious that the discrete version of this problem causes so much trouble.
 

### 2013-07-24 paper.tex
writing

@@
 It follows that $f=\langle Γ^+|ψ\rangle$ is the function with least norm that satisfies $|Γ\rangle f=|ψ\rangle$.
-As shown above, $\|f\|=\| |ψ\rangle\|$.
+As shown above, $‖f‖=‖ |ψ\rangle‖$.
 
@@
 so the squared norm is
-$$\|g\|²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
+$$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
 According to Wolfram,
@@
 so
-$$\|g\|=π^½.$$
+$$‖g‖=π^½.$$
 
@@
 Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
-$$\|f_n\|²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
+$$‖f_n‖²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
 given 
@@
 $$f_n(r)\propto rⁿ.$$
-The analytic functions still win.  
-In this case, the integrals are identical, so, for the preimage of $|n〉$,
-$$\|f_n\|²=\sqrt{πn!},$$
-and, presuming $|n〉$ is the singular vector, the singular value is
-$$σ_n=(πn!)^{-1/4}.$$
-I've done something wrong, because the result doesn't depend on $b$.
+The analytic functions still win.  Let $f_n(r)=rⁿ$.  Then
+$$‖f_n‖²=2π\int_0^b r^{2n+1}e^{-r²}\,dr$$
+and
+$$‖ |Γ_χ〉f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
+so we have
+$$σ_n={‖ |Γ_χ〉f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
+	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
+According to Wolfram,
+$$\int_0^b r^{2n+1}e^{-r²}dr=½\left(Γ(n+1)-Γ(n+1,b²)\right),$$
+where $Γ$ is Mathematica's version of the incomplete gamma function.  Therefore, recalling that $n!=Γ(n+1)$,
+$$σ_n=\left(1-{Γ(n+1,b²)\over Γ(n+1)}\right)^½.$$
+Unfortunately, Matlab defines the incomplete gamma function in a different way than Mathematica does.
+Fortunately, that Matlab function, {\tt gammainc}, is exactly the square of the right-hand side of the above equation.
+
+This all seems wrong, because the incomplete gamma function varies the wrong way with $n$.  It becomes small as $n→0$.
 
@@
 
-\centerline{\XeTeXpicfile svgrid.pdf width \hsize}
+\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
 
@@
 
-\centerline{\XeTeXpicfile rsv2.pdf width 0.5\hsize
-\XeTeXpicfile rsv3.pdf width 0.5\hsize}
-\centerline{\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+\topinsert \XeTeXpicfile rsv2.pdf width 0.5\hsize
+\XeTeXpicfile rsv3.pdf width 0.5\hsize \endinsert
+\topinsert \XeTeXpicfile rsv4.pdf width 0.5\hsize \endinsert
 
@@
 With these directions, we can form some of the smaller singular vectors of the full problem.
+
+%% Sampling on random grids
+
+\topinsert \XeTeXpicfile gweight.pdf width \hsize 
+The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
+The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
+Note how nearby points tend to cancel each others' samples.
+I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
+\endinsert
 
@@
 whence
-$$\|g\|²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
+$$‖g‖²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
 The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.

### 2013-07-25 paper.tex
writing

@@
 
-When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.
+When they realised that positions and momenta are essentially uncertain, the first quantum physicists were bound to discover the gaussian wave packet by seeking conditions that produce a normal distribution of probability.
 They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.
@@
 Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.
-The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.
+The experience of quantum optics shows that, in many systems, the approximation converges with a reasonable number of components.
+Also, variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.
 
@@
 
-The expansion introduced by Glauber\cite{prx-131-2766} is performed by the operator
+Expansion over coherent states is a linear operator from functions of a complex variable to quantum states,
 $$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
 The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.
-The function $f$ is a function of a complex variable, such that $e^{-½|α|²}f(α)$ is square-integrable.
-Bargmann\cite{} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.
-In this paper, the application of a linear function is written as a product, and functions whose range is ket space are denoted by kets, usually with a greek letter.
-So $\G f$ is the ket that results from applying the linear operator $\G$ to the phase space distribution $f$.
+The function $f$ is such that $e^{-½|α|²}f(α)$ is square-integrable.
+Bargmann\cite{pam-14-187} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.
+When norms of functions are written below, the norm is the one derived from this inner product.
+In this paper, $\G f$ denotes the ket that results from applying the linear operator $\G$ to the phase space distribution $f$; application and composition of linear operators is shown by juxtaposition, and functions whose values are quantum states are written as kets, usually with a greek letter.
 
@@
 whence
-$$|〈β|α〉|²=e^{-|β-α|²}.$$
-The projector satisfies the law
+$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.$$
+The projector satisfies
 $$\int |α〉〈α|\,d²α=π.$$
-For an orthonormal basis, the right hand side would be 1 instead of $π$.  It follows that the operator $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
-$$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
-Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.
+This is similar to the law for an orthonormal basis, except that the right hand side is $π$ instead of 1.
+Further properties of the coherent states will be stated as required.
+Glauber's summary of them\cite{prx-131-2766} is paraphrased in every quantum optics textbook ever written.
+
+% It follows that $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
+% $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
+% Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.
 Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.
-Further properties of the coherent states will be stated as required.
-Glauber's summary of them is paraphrased in every quantum optics textbook ever written.
 
-Prior to Glauber, Bargmann pointed out that the Hilbert space of wavefunctions is isomorphic to the space of entire functions of a complex variable, under the inner product
-$$\langle g,f\rangle=\int g*(z)f(z)\,d²z.$$
-He introduced the isometry
-$$\langle Γ^+|ψ\rangle(α)=π^{-½}\langle α|ψ\rangle,$$
-which Glauber showed was a pseudoinverse of $|Γ\rangle$, in the sense that
-$$|Γ\rangle\langle Γ^+|=1.$$
-The other combination, $\langle Γ^+|Γ\rangle$, is an orthogonal projector from the functions of a complex variable onto the entire functions, under the Bargmann inner product.
-It follows that $f=\langle Γ^+|ψ\rangle$ is the function with least norm that satisfies $|Γ\rangle f=|ψ\rangle$.
-As shown above, $‖f‖=‖ |ψ\rangle‖$.
+Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
+$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
+Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.
+It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.
+The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.
+It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.
 
-If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ^+|Γ\rangle$.
+If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ⁺|Γ\rangle$.
 Therefore the singularity of $|Γ\rangle$ is easily resolved.
-If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ^+|ψ\rangle$.
-And $\langle Γ^+|$, an isometry, is as nice a function as could possibly be desired.
+If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ⁺|ψ\rangle$.
+And $\langle Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
 So it is curious that the discrete version of this problem causes so much trouble.
@@
 
-Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.
-
-There are two ways one could set about proving the Hypothesis.
+% Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.
+% 
+% There are two ways one could set about proving the Hypothesis.
 Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
 Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
-ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
-
-This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
+% ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
+% 
+% This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
 My numerical experiment suggests otherwise; let's ignore that possibility for now.
-
-Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
-$$f(z)=∑c_n{zⁿ\over (n!)^½},$$
-for some normalised vector $c$.
-
-  Our expansion function is 
-$$g(α)=f(α)e^{-½|α|²}.$$
-If $f$ has the Taylor expansion
-$$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
-then 
-$$Gg = ∑_n c_n|n〉.$$
-The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
-$$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
-the Glauber preimage of $|n〉$.  We have 
-$$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
-so the squared norm is
-$$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
-According to Wolfram,
-$$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
-so
-$$‖g‖=π^½.$$
-
-Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.
+% 
+% Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
+% $$f(z)=∑c_n{zⁿ\over (n!)^½},$$
+% for some normalised vector $c$.
+% 
+%   Our expansion function is 
+% $$g(α)=f(α)e^{-½|α|²}.$$
+% If $f$ has the Taylor expansion
+% $$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
+% then 
+% $$Gg = ∑_n c_n|n〉.$$
+% The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
+% $$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
+% the Glauber preimage of $|n〉$.  We have 
+% $$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
+% so the squared norm is
+% $$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
+% According to Wolfram,
+% $$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
+% so
+% $$‖g‖=π^½.$$
+% 
+% Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.
 Or at least I think so.
@@
 %% Consequences of lemma
-
-If the domain and range of an operator are the same, and the operator has singular values 0 and 1, then that operator is an orthogonal projector onto its range.
-The operator $|Γ〉$ is a generalisation of that.
-It is the composition of an orthogonal projection of function space onto the entire functions followed by an isometric mapping of the entire functions onto the kets.
-The orthogonality and isometry are under the Bargmann inner product for phase space functions.
-Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.
-This might depend somewhat on how evenly the amplitudes are spaced: I could start with a regular grid, and add Gaussian noise to see what changed.
-
-So $G$ is a projection: if the inner product is normalised properly, $G$ keeps the component of $f$ in the space of entire functions, and discards the orthogonal component.
-The continuous problem is simply underdetermined, and the inverse problem can be regularised very easily by excluding components with zero singular values from the solution.
-That isn't what I see with discrete versions: they seem to have a steady decay of singular values, not a sudden drop to zero.
-This must be entirely an artefact of the discretisation.
 
@@
 Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.
-The obvious choice is $g_r(α*)=(〈α_r|G^+)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
+The obvious choice is $g_r(α*)=(〈α_r|G⁺)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
 The condition becomes

### 2013-07-26 paper.tex
writing

@@
 \def\G{|Γ〉}
+\def\Gx{|Γ_χ〉}
 \def\cite#1{$[\hbox{\tt #1}]$}
@@
 
-Define a restricted operator
-$$|Γ_χ〉f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
-For now, suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.
-The most obvious examples are characteristic functions, where $χ$ is 1 inside a circle of radius $b$, and 0 outside.
-However, the formalism allows more gradual truncation as well.
-We can rewrite this
-$$\eqalign{|Γ_χ〉f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
+%% The SVE
+
+The above paragraph can be stated quantitatively.  Define a restricted operator
+$$\Gx f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
+Most obviously, $χ$ could be a characteristic function, with $χ(z)$ being 1 for $|z|≤b$ and 0 outside this disk.  The formalism allows gradual truncation too.
+
+The sizes of states and their expansions can be read from the singular value expansion of $\Gx $.
+If the domain and range of $\G$ were the same space, it would have an eigenvalue decomposition; the SVE is a generalisation of that.
+It has the form
+$$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
+The kets $|g_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.
+The $σ_i$ are positive real numbers, called singular values.
+The functions $v_i$ are are orthonormal under the Bargmann inner product, and are called right singular vectors.
+
+In the case where $χ=1$, and $\Gx=\G$, the expansion is quite arbitrary.
+The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire functions for which $\G v_i=|u_i〉$, and every singular value is 1.
+If the domain and range of $\G$ were the same, this would indicate that it was a projector.
+The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component keeps the same expansion coefficients, while the basis $v_i$ is replaced by the $|u_i〉$.
+
+When the operator is restricted, the SVE gives more information.
+If $\Gx$ were one to one, each $v_i$ would be the unique preimage of $|g_i〉$; since $\Gx$ is singular, $v_i$ is the preimage with least norm.
+The singular values are the ratios of $‖\Gx v_i‖$ to $‖v_i‖$.
+If $|u_i〉$ has a low particle number, it can be expanded efficiently: $v_i$ will be close to $〈Γ⁺|u_i〉$, and $σ_i$ will be close to 1.
+If $|u_i〉$ includes components with large particle number, its expansion will have a larger norm than itself.
+In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, and $σ_i$ will be small.
+
+Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
+$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.
+Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.
+In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.
+When the expansion is discretised, however, it is very easy for the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ to disappear in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.
+
+Let's compute some singular values and vectors.  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
+$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
 	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
@@
 and
-$$‖ |Γ_χ〉f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
+$$‖ \Gx f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
 so we have
-$$σ_n={‖ |Γ_χ〉f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
+$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
 	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
@@
 I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
+
+The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
+This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
+Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.
+Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.
+When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.
+This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
 \endinsert
@@
 However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.
-
-There are two types of ill-posed problems, rank-deficient and discrete.
-They are distingished by how the singular values decay to zero.
-In a rank-deficient problem, there is a sharp cut between large and small singular values.
-The large ones are large enough that components in their singular directions do not contribute excessively to the solution, the small ones would contribute huge components, but they are small enough to round to zero, and ignore the component along that singular direction.
-In discrete problems, the singular values decay continuously.
-The size of the components of the solution varies continuously: there is no point where they become huge, but a continuous minimum and a gradual rise as the singular values get smaller.
-
-I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.
-However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.
-This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
-$$f(t)=\int F(ω)e^{iωt}\,dω.$$
-This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
-(What is the SVD of the DFT matrix?
-It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
-
-A numerical example is shown below.
-The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.
-A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $|g(a*)|²$, where $g=G⁺|ψ〉$.
-These are plotted below, as circles in phase space.
-For smaller ensembles, the results were not consistent between draws.
-
-A basic question is why regularisation helps.
-Usually, it's applied to the inversion of measured data, where the right hand side of the least squares problem includes noise.
-The aim is to prevent amplified noise from dominating the solution.
-In our problem, the right hand side is calculated instead of measured, and the only noise is rounding errors.
-Those are 16 orders of magnitude smaller than the exact right hand side, so rounding the small singular values up to $10^{-10}$ should be more than adequate.
-We've needed much stronger regularisation than that: therefore, rounding errors aren't the problem.
-
-The only other error I can think of is nonlinearity.
-Tychonov parameters of order $10^{-10}$ do work for the linear expansion problem.
-Perhaps we're not doing regularisation, we're actually establishing a trust region, where the linear least squares problem is a good approximation to the actual change in $|ψ(z)〉$.
-
-If that's the case, it's tempting to make the weights linear instead of logarithmic, so that that part of the problem is exactly linear.
-However, the function $e^φ$ is pretty nearly linear in the range we're shifting $φ$, so that can't help much.
-In fact, the whole problem is quite linear.
-
-It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the l-curve experiments for the expansion problem.
-To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.
-In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around $π^½$.
-And we're back to the other open question, how to relate the norm of a function to the norm of a set of samples at irregular points.
-How can I estimate how much of phase space “belongs” to each amplitude sample, in order to get a volume over which the function has that value?
-
-The obvious way would be to assign amplitudes to the nearest sample.
-That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
-What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
-There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
-Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
-I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
-What's the best exponent?
-I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
-
-The Picard condition is very obviously satisfied by the projection operator $|Γ〉$.
-The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm as itself.
-Functions in the null space of $|Γ〉$ don't have components in the left singular directions, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
-
-\centerline{\XeTeXpicfile cats.pdf width \hsize}
-
-The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉f=〈A|ψ〉$, where $f$ is a vector of weights such that $|A〉f$ approximates $|ψ〉$.
-Below it is a plot of the expansions over number states of the singular kets $〈n|u_r〉$, where $|A〉=|U〉SV$.
-For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.
-The components of the cat have coherent amplitude 2, so their largest number component is $n=4$, and the components drop quickly as $n$ increases further.
-Above $r=15$, the singular ket $|u_r〉$ has very small components on these number states, so it hardly overlaps $|ψ〉$.
-At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.
-In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied.
-Seeing that $G$ is perfectly conditioned, this is a very bad discretisation.
-
-The final graph is an L curve when the discrete problem is Tychonov regularised.
-It shows the expected shape, with minimum residual around $2×10^{-8}$.
-Six regularised solutions are shown above; the modulus of $f_r$ is the disk's saturation, and the phase determines its hue.
-Note that least norm regularisation has the desirable property that adjacent $α_r$ have similar $f_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.
-
-\centerline{\XeTeXpicfile catpicard.pdf width \hsize}
-
-\centerline{\XeTeXpicfile catsws.pdf width \hsize}
-
-\centerline{\XeTeXpicfile catl.pdf width \hsize}
 
@@
 
-%% Singular value decompositions
-
-Variational dynamics seeks an expansion $g$ to approximate a quantum state $|ψ〉≈Gg$.
-Given the form of Equation~1, this requires solving a Fredholm integral equation of the first kind.
-That is the canonical example of an ill-posed problem: rounding and discretisation errors in $|ψ〉$ are magnified enormously in the solution $g$.
-To solve these problems, it helps to understand the singular values and vectors of the operator $G$.
-
-The singular value expansion of the linear operator $G$ has the form
-$$Gf = ∑_{j=1}^∞ |g_j〉σ_j〈v_j,f〉,$$
-where the left singular vectors $|g_j〉$ are orthonormal kets, the singular values $σ_j$ are positive real numbers, the right singular vectors $v_j$ are normalised entire functions, and presumably $〈,〉$ is the Bargmann inner product.
-This can be written
-$$G = |G〉SV†,$$
-where $|G〉$ is a row vector of the left singular kets, $S$ is a matrix with the singular values along the diagonal, and $V$ maps complex coefficients to linear combinations of the functions $v_j$.
-
-Because the expansions are underdetermined, the set $|g_j〉$ will be complete, but the set $v_j$ will not be.
-Alternatively, some of the $v_j$ have zero singular values and no corresponding left singular vector.
-One can write a full SVE, where $V†$ is a full expansion, but $S$ truncates the vector of coefficients.
-
 The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
@@
 Or alternately, if the grid shrinks while the Fock space stays constant.
+
+%% ill-posedness
+
+As shown above, the restricted expansion operator $\Gx $ has singular values that decay to very small values, whose singular vectors are states with particle numbers corresponding to coherent amplitudes outside the support of $χ$.
+
+There are two types of ill-posed problems, rank-deficient and discrete.
+They are distingished by how the singular values decay to zero.
+In a rank-deficient problem, there is a sharp cut between large and small singular values.
+The large ones are large enough that components in their singular directions do not contribute excessively to the solution, the small ones would contribute huge components, but they are small enough to round to zero, and ignore the component along that singular direction.
+In discrete problems, the singular values decay continuously.
+The size of the components of the solution varies continuously: there is no point where they become huge, but a continuous minimum and a gradual rise as the singular values get smaller.
+
+I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.
+However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.
+This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
+$$f(t)=\int F(ω)e^{iωt}\,dω.$$
+This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
+(What is the SVD of the DFT matrix?
+It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
+
+A numerical example is shown below.
+The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.
+A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $|g(a*)|²$, where $g=G⁺|ψ〉$.
+These are plotted below, as circles in phase space.
+For smaller ensembles, the results were not consistent between draws.
+
+A basic question is why regularisation helps.
+Usually, it's applied to the inversion of measured data, where the right hand side of the least squares problem includes noise.
+The aim is to prevent amplified noise from dominating the solution.
+In our problem, the right hand side is calculated instead of measured, and the only noise is rounding errors.
+Those are 16 orders of magnitude smaller than the exact right hand side, so rounding the small singular values up to $10^{-10}$ should be more than adequate.
+We've needed much stronger regularisation than that: therefore, rounding errors aren't the problem.
+
+The only other error I can think of is nonlinearity.
+Tychonov parameters of order $10^{-10}$ do work for the linear expansion problem.
+Perhaps we're not doing regularisation, we're actually establishing a trust region, where the linear least squares problem is a good approximation to the actual change in $|ψ(z)〉$.
+
+If that's the case, it's tempting to make the weights linear instead of logarithmic, so that that part of the problem is exactly linear.
+However, the function $e^φ$ is pretty nearly linear in the range we're shifting $φ$, so that can't help much.
+In fact, the whole problem is quite linear.
+
+It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the l-curve experiments for the expansion problem.
+To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.
+In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around $π^½$.
+And we're back to the other open question, how to relate the norm of a function to the norm of a set of samples at irregular points.
+How can I estimate how much of phase space “belongs” to each amplitude sample, in order to get a volume over which the function has that value?
+
+The obvious way would be to assign amplitudes to the nearest sample.
+That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
+What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
+There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
+Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
+I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
+What's the best exponent?
+I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+
+The Picard condition is very obviously satisfied by the projection operator $|Γ〉$.
+The left singular vectors can be any orthonormal basis of ket space; the singular values are all 1, so they don't reduce at all, and any ket has a preimage with the same norm as itself.
+Functions in the null space of $|Γ〉$ don't have components in the left singular directions, so, if the problem is solved by expanding over singular vectors, they will not be present in the solution.
+
+\topinsert\XeTeXpicfile cats.pdf width \hsize\endinsert
+
+The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉f=〈A|ψ〉$, where $f$ is a vector of weights such that $|A〉f$ approximates $|ψ〉$.
+Below it is a plot of the expansions over number states of the singular kets $〈n|u_r〉$, where $|A〉=|U〉SV$.
+For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.
+The components of the cat have coherent amplitude 2, so their largest number component is $n=4$, and the components drop quickly as $n$ increases further.
+Above $r=15$, the singular ket $|u_r〉$ has very small components on these number states, so it hardly overlaps $|ψ〉$.
+At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.
+In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied.
+Seeing that $G$ is perfectly conditioned, this is a very bad discretisation.
+
+The final graph is an L curve when the discrete problem is Tychonov regularised.
+It shows the expected shape, with minimum residual around $2×10^{-8}$.
+Six regularised solutions are shown above; the modulus of $f_r$ is the disk's saturation, and the phase determines its hue.
+Note that least norm regularisation has the desirable property that adjacent $α_r$ have similar $f_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.
+
+\topinsert\XeTeXpicfile catpicard.pdf width \hsize\endinsert
+
+\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
+
+\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
 

### 2013-07-26 paper.tex
writing

@@
 
-%% Discrete phase space
-
-We need a way to discretise the function $g$.
-There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
-The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr)$.
-Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
-Here, $w_r$ is a real weight, intended to compensate for the irregularity of the sample set, so that the sampled vector $\vec g=\bigl(w₁g(α₁),…,w_Rg(α_R)\bigr)^{\rm T}$ has a similar norm to the function $g$.
-
-Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.
-The obvious choice is $g_r(α*)=(〈α_r|G⁺)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
-The condition becomes
-$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
-	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
-	& & \ddots & \cr
-	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
-\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
-= \pmatrix{π\cr π\cr\vdots\cr π}.
-$$
-Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
-It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
-The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.
-This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
-
-The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.
-The weights of the components don't simple sample $f(α)$: they also adjust for how close the sample points are, and how much the $|α_i〉$ overlap.
-The operator $G$ is probably being disceretised by some kind of Galerkin formula; I don't know what the basis functions are in phase space and Hilbert space.
-
-It would be useful to find a quadrature rule, with weights $w_i$, such that 
-$$∑_i w_if(α_i)|α_i〉≈\int f(α)|α〉d²α.$$
-Suppose we aim to be exactly right, in the case that $f(α)$ is the normalised entire function representing the left hand side, i.e.,
-$$f(β)=∑_i w_if(α_i)〈β|α_i〉.$$
-(I'm assuming we put the factor $1/π$ in $G$.)  This doesn't have a solution for general superpositions of coherent states.  Surely there's some quadrature rule for a sampled phase-space function?
-
-Using the Galerkin criterion, there doesn't seem to be one where the coefficient of $|α_i〉$ only depends on $f(α_i)$.
-However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.
-
-%% Truncated Hilbert space
-
-I still haven't thought in detail about the results with truncated Fock spaces.
-Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
-Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
-So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
-And in fact, by solving the inverse problem, we can always get a nice solution in any case.
-Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
-
-The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.  This can also be seen from the expansions of coherent states.  The normalised entire expansion of $|α〉$ is
-$$g(β*)=〈β|α〉=e^{-½|α|²-½|β|²+β*α},$$
-whence
-$$‖g‖²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
-The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.
-
-The lemma gives a very simple singular value expansion of $G$.
-Because there is only one nonzero singular value, any orthonormal basis of ket space will do as the left singular vectors.
-Let's choose the number states for familiarity.
-Then
-$$Gg=∑_n |n〉π\int ν_n*(α)g(α)\,d²α=|N〉π\int ν*(α)g(α),$$
-where $|N〉=\bigl(|0〉\ |1〉\ |2〉\ …\bigr)$ and $ν(α)=\bigl(ν₀(α)\ ν₁(α)\ …\bigr)$ could be thought of as infinite matrices whose columns are the left singular vectors $|n〉$ and their normalised entire expansions, the right singular vectors $v_n$.
-The pesudoinverse,
-$$G⁺|ψ〉=νπ〈N|ψ〉= ∑_n ν_nπ〈n|ψ〉,$$
-maps $|ψ〉$ to its normalised entire expansion.
-Expanding over number states is useful because the maximum number in a system is usually limited by conservation of energy.
-In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.
-
-Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
-(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
-
-As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
-Presumably these converge to limits as $N$ increases.
-There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
-
-%% Why grids don't work, and variation necessary
-
-This paper will talk about two related but distinct inverse problems.
-Both start with a given wavefunction $|ψ\rangle$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
-In systems with few modes, these points could be a grid, but in many-mode systems they will have to be more sparse than that: there isn't space to store grid coefficients.
-The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j\rangle$ approximates $|ψ\rangle$.
-This is simply finding a discrete inverse to $G$ sampled on the grid $α_j$; since $G$ is a projection, the pseudoinverse of $G$ should solve it well enough.
-The differential case, where we have an expansion for $|ψ\rangle$ which we want to perturb to represent $|ψ\rangle+|dψ\rangle$, is no harder.
-Because $G$ is linear, we can expand $|dψ\rangle$, and add it to the expansion of $|ψ\rangle$.
-
-The second problem, the variational one, is a generalisation.
-As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ\rangle$.
-Instead of expanding $|ψ\rangle$ over phase space, we want to expand it over the columns of $|Dψ(f,α)\rangle$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
-Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)\rangle$ to the solution of the inverse problem, and those of $|∂fψ(f,a)\rangle$ to zero.
-Is this the smallest norm solution?
-Probably not: if a coherent state was moving in an oscillator, it might be easier to just rotate the $α_j$.
-
-Since we have a totally stable solution to the problem, why would we try anything else?
-Because we're iterating the problem, and we need to preserve our ability to solve it at future steps.
-If the amplitudes form a grid, which covers all the energies the system could possibly access, we're set: we'll always be able to expand over them.
-However, we can't do that for many mode systems: there will always be accessible states of the system whose support in phase space is remote from the current grid.
-The set $|Dψ\rangle$ will have some small singular values, corresponding to directions where the amplitudes and samples are adjusted in a way that preserves the wavefunction.
-We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ\rangle$ in the future.
-
-So what are the critera for adjusting the grid?
-I think there are two obvious ones: firstly, that the inverse problem stays well-posed.
-Otherwise, we're wasting grid points.
-Secondly, when we're expanding a derivative over the tangent space $|Dψ\rangle$, we want to tilt the tangent space closer to the derivative.
-That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.
-The first criterion seems redundant: if we could better represent the goal by shifting two sample points apart, the second criterion will do that automatically.
-However, if we only care about the tangent space and not the derivative map, we might need very large adjustments of the amplitudes and weights to get to the point in the tangent space that we want.
-So a condition number criterion seems sensible.
-There might be other criteria that I've overlooked.
-
-%% Fock space is cheating
-
-Expanding the quartic oscillator over Fock states allows it to be solved very stably, and as precisely as desired by rotating the phases of the expansion coefficients.
-Also, it is easy to find the expansion over Fock states of a superpositon of coherent states.
-This means that the stability of time discretisation can be removed from the problem entirely.
-The state can be propagated exactly, and the coherent states adjusted to fit.
-The most direct way to do so is by nonlinear optimisation at every timestep; Matlab provides the necessary routines in a convenient package.
-Unfortunately, this is not at all stable.
-
-In a sense, expanding over Fock states is cheating.
-If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis such as the Fock states would be unmanagably large.
-In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.
-However, this would raise the question whether those least-square solutions from the normal equations were stable.
-By using an orthonormal basis, we can be more confident of this.
-
-To do this, we need some bracket matrices.  We have
-$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
-The brackets with the number states are
-$$〈m|H/\hbar|n〉=n(n-1)δ_{mn}$$
-and
-$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
-as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
-$$〈n|H/\hbar|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
-
-The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
-If a singular value decomposition $|Dψ〉=|U〉SV†$ is taken, the singularity is indicated by small singular values on the diagonal of $S$.
-It is often measured by the condition number of $|Dψ〉$, the ratio of the largest to the smallest singular value.
-Let $C(z)$ be the condition number of $|Dψ(z)〉$.
-As part of the regularisation process, it would be useful to avoid changing $z$ in directions that increase $C(z)$.
-For example, if two coherent states had real amplitudes, and the real part of the expected amplitude needed to increase, it would be done by shifting the right hand one away from the left hand one, not the left hand one closer to the right hand.
-Is there a way to make a computer do that?
-
-The least squares problem can be solved using the singular value decomposition: to approximate a state $|h〉≈|Dψ(z)〉dz$, we set $dz=VS^{-1}〈U|h〉$.
-The first problem is that some of the $σ_i$ on the diagonal of $S$ will be small, because the $|u_i〉$ that correspond to them change different components of the superposition $|ψ(z)〉$ in ways that nearly cancel out.
-So a error in the discrete representation of $|h〉$ in the direction $|u_i〉$ will cause a very large error in the least squares $dz$ in the direction $v_i$.
-The way to avoid such large errors is to shift the small singular values of $S$ up to a lower bound $ε$, and thus reduce the large singular values of $S^{-1}$ below an upper bound $ε^{-1}$.
-There are many and varied ways of doing that.
-
 %% Picard plots
-
-The singular value expansion is useful for two reasons.
-Firstly, the discrete version, the singular value decomposition, is the most stable way to solve badly-posed least squares problems.
-Secondly, the sizes of the singular values, and the components of $H|ψ〉$ along the singular directions, determine whether the problem can be solved at all.
 
@@
 Or alternately, if the grid shrinks while the Fock space stays constant.
+
+%% Tychonov conditioning
+
+There is a family of methods for inverting operators with small singular values, known as regularisation methods.  A very useful idea is due to Tychonov.  The problem
+$$\Gx f=|ψ〉$$
+can be regarded as a least squares problem, where $|ψ〉$ is chosen to minimise the residual $‖\Gx f-|ψ〉‖²$.
+The exact solution has residual zero, of course.
+The problem is that it is swamped by components with small singular values of $\Gx$.
+Suppose we modify this problem, and seek $|ψ〉$ to minimise $‖\Gx f-|ψ〉‖²+λ‖f‖²$, where $λ$ is a positive number, known as a Lagrange parameter for reasons that will become clear shortly.
+There are two things to note here.
+Firstly, this is still a least squares problem: it is equivalent to minimising
+$$\left‖ \pmatrix{\Gx \cr λ} f-\pmatrix{|ψ〉\cr 0}\right‖².$$
+Therefore, it can be solved stably by the usual least squares methods.
+Secondly, suppose $f_λ$ is the solution for a given value of $λ$.
+If another function, $g$, had the same norm but a smaller residual, then $f$ would not minimise the Tychonov expression.
+The Tychonov method is equivalent to constraining the norm of $f$, and seeking the solution with smallest residual given the constraint.
+Or, equivalently, constraining the residual and seeking the solution with smallest norm.
+This equivalence to a constrained optimisation problem is why $λ$ is a Lagrange parameter.
+
+Solving the Tychonov problem for various values of $λ$ generates a family of solutions.
+These can be plotted on a graph of residual against solution norm, where they mark the edge of the space of possible solutions.
+Above and to the right of the curve, a solution can be found with the given norm and residual: below and to the left, it can't.
+
+\topinsert\XeTeXpicfile lcurve-1.pdf \endinsert
+
+An example for a typical application is shown below.
+The context is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.
+The real state is $x₀$, and it has generated observations $b=Ax₀+e$, where $e$ consists of measurement error.
+The problem $Ax=b$ is solved for $x$, with various Lagrange parameters.
+At the top left, the norm of $x$ is constrained to be zero, so the solution is $x=0$, and the residual is the norm of $b$.
+As the constraint on the norm is relaxed, the solution becomes closer to $x₀$, and the residual decreases.
+At the elbow of the curve, the norm of the solution is constrained to be close to the norm of the real state $x₀$, the solution is close to $x₀$, and the residual is the norm of the measurement error.
+Relaxing the constraint further allows solutions $x$ that fit the measurements better than the real state $x₀$ does.
+Typically, they won't fit much better, so the residual doesn't decrease much below $‖e‖$.
+However, the solution now includes components belonging to small singular values, and the norm of the solution will increase as far as it is allowed to.
 
@@
 I didn't forsee that, but it isn't too surprising: we can change the coefficients of the superposition in any direction we like, regardless of their current values, and the derivative is only sensitive to changes.
+
+%% Tychonov regularisation
+
+%% Discrete phase space
+
+We need a way to discretise the function $g$.
+There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
+The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr)$.
+Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
+Here, $w_r$ is a real weight, intended to compensate for the irregularity of the sample set, so that the sampled vector $\vec g=\bigl(w₁g(α₁),…,w_Rg(α_R)\bigr)^{\rm T}$ has a similar norm to the function $g$.
+
+Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.
+The obvious choice is $g_r(α*)=(〈α_r|G⁺)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
+The condition becomes
+$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
+	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
+	& & \ddots & \cr
+	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
+\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
+= \pmatrix{π\cr π\cr\vdots\cr π}.
+$$
+Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
+It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
+The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.
+This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
+
+The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.
+The weights of the components don't simple sample $f(α)$: they also adjust for how close the sample points are, and how much the $|α_i〉$ overlap.
+The operator $G$ is probably being disceretised by some kind of Galerkin formula; I don't know what the basis functions are in phase space and Hilbert space.
+
+It would be useful to find a quadrature rule, with weights $w_i$, such that 
+$$∑_i w_if(α_i)|α_i〉≈\int f(α)|α〉d²α.$$
+Suppose we aim to be exactly right, in the case that $f(α)$ is the normalised entire function representing the left hand side, i.e.,
+$$f(β)=∑_i w_if(α_i)〈β|α_i〉.$$
+(I'm assuming we put the factor $1/π$ in $G$.)  This doesn't have a solution for general superpositions of coherent states.  Surely there's some quadrature rule for a sampled phase-space function?
+
+Using the Galerkin criterion, there doesn't seem to be one where the coefficient of $|α_i〉$ only depends on $f(α_i)$.
+However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.
+
+%% Truncated Hilbert space
+
+I still haven't thought in detail about the results with truncated Fock spaces.
+Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
+Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
+So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
+And in fact, by solving the inverse problem, we can always get a nice solution in any case.
+Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
+
+The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.  This can also be seen from the expansions of coherent states.  The normalised entire expansion of $|α〉$ is
+$$g(β*)=〈β|α〉=e^{-½|α|²-½|β|²+β*α},$$
+whence
+$$‖g‖²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
+The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.
+
+The lemma gives a very simple singular value expansion of $G$.
+Because there is only one nonzero singular value, any orthonormal basis of ket space will do as the left singular vectors.
+Let's choose the number states for familiarity.
+Then
+$$Gg=∑_n |n〉π\int ν_n*(α)g(α)\,d²α=|N〉π\int ν*(α)g(α),$$
+where $|N〉=\bigl(|0〉\ |1〉\ |2〉\ …\bigr)$ and $ν(α)=\bigl(ν₀(α)\ ν₁(α)\ …\bigr)$ could be thought of as infinite matrices whose columns are the left singular vectors $|n〉$ and their normalised entire expansions, the right singular vectors $v_n$.
+The pesudoinverse,
+$$G⁺|ψ〉=νπ〈N|ψ〉= ∑_n ν_nπ〈n|ψ〉,$$
+maps $|ψ〉$ to its normalised entire expansion.
+Expanding over number states is useful because the maximum number in a system is usually limited by conservation of energy.
+In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.
+
+Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
+(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
+
+As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
+Presumably these converge to limits as $N$ increases.
+There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
+
+%% Why grids don't work, and variation necessary
+
+This paper will talk about two related but distinct inverse problems.
+Both start with a given wavefunction $|ψ\rangle$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
+In systems with few modes, these points could be a grid, but in many-mode systems they will have to be more sparse than that: there isn't space to store grid coefficients.
+The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j\rangle$ approximates $|ψ\rangle$.
+This is simply finding a discrete inverse to $G$ sampled on the grid $α_j$; since $G$ is a projection, the pseudoinverse of $G$ should solve it well enough.
+The differential case, where we have an expansion for $|ψ\rangle$ which we want to perturb to represent $|ψ\rangle+|dψ\rangle$, is no harder.
+Because $G$ is linear, we can expand $|dψ\rangle$, and add it to the expansion of $|ψ\rangle$.
+
+The second problem, the variational one, is a generalisation.
+As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ\rangle$.
+Instead of expanding $|ψ\rangle$ over phase space, we want to expand it over the columns of $|Dψ(f,α)\rangle$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
+Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)\rangle$ to the solution of the inverse problem, and those of $|∂fψ(f,a)\rangle$ to zero.
+Is this the smallest norm solution?
+Probably not: if a coherent state was moving in an oscillator, it might be easier to just rotate the $α_j$.
+
+Since we have a totally stable solution to the problem, why would we try anything else?
+Because we're iterating the problem, and we need to preserve our ability to solve it at future steps.
+If the amplitudes form a grid, which covers all the energies the system could possibly access, we're set: we'll always be able to expand over them.
+However, we can't do that for many mode systems: there will always be accessible states of the system whose support in phase space is remote from the current grid.
+The set $|Dψ\rangle$ will have some small singular values, corresponding to directions where the amplitudes and samples are adjusted in a way that preserves the wavefunction.
+We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ\rangle$ in the future.
+
+So what are the critera for adjusting the grid?
+I think there are two obvious ones: firstly, that the inverse problem stays well-posed.
+Otherwise, we're wasting grid points.
+Secondly, when we're expanding a derivative over the tangent space $|Dψ\rangle$, we want to tilt the tangent space closer to the derivative.
+That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.
+The first criterion seems redundant: if we could better represent the goal by shifting two sample points apart, the second criterion will do that automatically.
+However, if we only care about the tangent space and not the derivative map, we might need very large adjustments of the amplitudes and weights to get to the point in the tangent space that we want.
+So a condition number criterion seems sensible.
+There might be other criteria that I've overlooked.
+
+%% Fock space is cheating
+
+Expanding the quartic oscillator over Fock states allows it to be solved very stably, and as precisely as desired by rotating the phases of the expansion coefficients.
+Also, it is easy to find the expansion over Fock states of a superpositon of coherent states.
+This means that the stability of time discretisation can be removed from the problem entirely.
+The state can be propagated exactly, and the coherent states adjusted to fit.
+The most direct way to do so is by nonlinear optimisation at every timestep; Matlab provides the necessary routines in a convenient package.
+Unfortunately, this is not at all stable.
+
+In a sense, expanding over Fock states is cheating.
+If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis such as the Fock states would be unmanagably large.
+In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.
+However, this would raise the question whether those least-square solutions from the normal equations were stable.
+By using an orthonormal basis, we can be more confident of this.
+
+To do this, we need some bracket matrices.  We have
+$$|Dψ(φ,α)〉=(1,a†)|ψ(φ,α)〉.$$
+The brackets with the number states are
+$$〈m|H/\hbar|n〉=n(n-1)δ_{mn}$$
+and
+$$〈n|Dψ(φ,α)〉=e^φ\pmatrix{{α^n\over \sqrt{n!}} &  α^{n-1}\sqrt{n\over (n-1)!}},$$
+as is the bracket of the quartic oscillator Hamiltonian $a^{2\dagger}a²$
+$$〈n|H/\hbar|ψ(φ,α)〉=e^φα^n\sqrt{n(n-1)\over(n-2)!}.$$
+
+The instability in the dynamics is caused by the near-singularity of $|Dψ〉$.
+If a singular value decomposition $|Dψ〉=|U〉SV†$ is taken, the singularity is indicated by small singular values on the diagonal of $S$.
+It is often measured by the condition number of $|Dψ〉$, the ratio of the largest to the smallest singular value.
+Let $C(z)$ be the condition number of $|Dψ(z)〉$.
+As part of the regularisation process, it would be useful to avoid changing $z$ in directions that increase $C(z)$.
+For example, if two coherent states had real amplitudes, and the real part of the expected amplitude needed to increase, it would be done by shifting the right hand one away from the left hand one, not the left hand one closer to the right hand.
+Is there a way to make a computer do that?
+
+The least squares problem can be solved using the singular value decomposition: to approximate a state $|h〉≈|Dψ(z)〉dz$, we set $dz=VS^{-1}〈U|h〉$.
+The first problem is that some of the $σ_i$ on the diagonal of $S$ will be small, because the $|u_i〉$ that correspond to them change different components of the superposition $|ψ(z)〉$ in ways that nearly cancel out.
+So a error in the discrete representation of $|h〉$ in the direction $|u_i〉$ will cause a very large error in the least squares $dz$ in the direction $v_i$.
+The way to avoid such large errors is to shift the small singular values of $S$ up to a lower bound $ε$, and thus reduce the large singular values of $S^{-1}$ below an upper bound $ε^{-1}$.
+There are many and varied ways of doing that.
 

### 2013-07-27 paper.tex
writing

@@
 With these directions, we can form some of the smaller singular vectors of the full problem.
-
-%% Sampling on random grids
-
-\topinsert \XeTeXpicfile gweight.pdf width \hsize 
-The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
-The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
-Note how nearby points tend to cancel each others' samples.
-I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
-
-The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
-This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
-Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.
-Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.
-When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.
-This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
-\endinsert
 
@@
 
-%% Tychonov regularisation
-
 %% Discrete phase space
 
-We need a way to discretise the function $g$.
-There are two usual ways to do this: sampling on a grid, or taking components along a fixed basis.
-The size of state space rules these out, so we must sample $g$ at a set of irregularly spaced amplitudes $A=\{α₁,…,α_R\}$, corresponding to a row of kets $|A〉=\bigl(|α₁〉\ ⋯\ |α_R〉\bigr)$.
-Equivalently, we take components of $f$ in function space along the set of vectors $χ_r(α)=w_rδ(α-α_r)$.
-Here, $w_r$ is a real weight, intended to compensate for the irregularity of the sample set, so that the sampled vector $\vec g=\bigl(w₁g(α₁),…,w_Rg(α_R)\bigr)^{\rm T}$ has a similar norm to the function $g$.
+To compute in phase space, we need to reduce the function $g$ to a finite vector of numbers.  This is usually done in one of two ways: by a vector
+$$\vec g=\pmatrix{g(z₁) & ⋯ & g(z_R)},$$
+where the points $z_i$ form a regular grid, or by
+$$\vec g=\pmatrix{\langle u₁,g\rangle & ⋯ & \langle u_R,g\rangle},$$
+where the distributions $u_i$ are a fixed basis.
+The huge dimension of state space rules out any regular approach: we must sample $g$ at a set of irregularly spaced amplitudes or on an irregular basis, adapted to the support of $g$.
 
-Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.
-The obvious choice is $g_r(α*)=(〈α_r|G⁺)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.
-The condition becomes
+This makes regularisation tricky.
+The goal is to represent a state $|ψ\rangle$ as a superposition of coherent states, $|ψ\rangle=|A\rangle c$.
+We want the coefficients $c$ to correspond to a phase space function $g$ that is nearly analytic.
+If the amplitudes $|A\rangle$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
+On an irregular grid, we would expect the coefficients to be larger where the grid is sparse, and smaller where it is dense.
+It would be good to know if irregular coefficients are the result of sampling a smooth function on an irregular grid, or of sampling a wildly oscillating function.
+
+The obvious way to attempt this is to factor the coefficients as $|ψ\rangle=|A\rangle Wc'$, where $c'_i$ are samples that approximate $g(α_i)$, and $W$ is a matrix of weights that adapts the samples to the irregular grid.
+In fact, we can do this exactly.
+If we set $W=(\langle A|A\rangle)^{-1}$, then $c'=\langle A|ψ\rangle$.
+We could try seeking coefficients $c$ such that the norm of $\langle A|A\rangle c$ is minimal.
+
+To avoid the complexities of Wirtinger calculus, the problem of relating the weights of gaussians to the values of a sampled function can be set up for a function of a single real variable.
+It would be appealing for the weight matrix $W$ to be diagonal, so that the weight of the gaussian centred at each point was determined by the value of $f$ at that point.
+This can be set up either as a Galerkin condition, or as a collocation condition: they turn out to be equivalent.
+Unfortunately, the weights end up oscillating wildly.
+This approach doesn't seem useful, but is presented for completeness, and because the phenomenon of adjacent weights cancelling keeps occuring, this being the simplest problem in which I've seen it.
+
+As a collocation problem, consider the case where $f(x)=1$, and we wish to approximate $f$ by a function 
+$$f'(x)=∑_i e^{-(x-x_i)²}w_ic_i$$
+that satisfies the collocation condition $f'(x_i)=f(x_i)$.  The goal is for $c_i=1$, so we need ...
+
 $$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
@@
 \pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
-= \pmatrix{π\cr π\cr\vdots\cr π}.
-$$
+= \pmatrix{π\cr π\cr\vdots\cr π}.$$
+
+The Galerkin condition seeks a set of weights $w$ such that, for the normalised functions $u_i=π^{-½}e^{-(x-x_i)²}$, we have
+$$1=\int u_i=∑_j w_j u_i(x_j)=∑_j w_je^{-(x_j-x_i)²}.$$
+This is the same condition as the collocation condition.  The results are shown below.
+
 Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
@@
 
-The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.
-The weights of the components don't simple sample $f(α)$: they also adjust for how close the sample points are, and how much the $|α_i〉$ overlap.
-The operator $G$ is probably being disceretised by some kind of Galerkin formula; I don't know what the basis functions are in phase space and Hilbert space.
+There is another way to derive coefficients of the constant function, which might be more stable.
+Suppose that $D$ is a subset of the real line, large enough that the tails of the gaussians outside it are neglibible.
+Then
+$$E=\|f'-1\|=\int_D\left(∑_ic_iu_i(x)-1)²=∑_{ij}c_ic_j\int_D u_iu_j-2∑_ic_i\int_D u_i+\mu(D),$$
+where $μ$ is the integration measure.  The derivative of this with respect to $c_i$ is
+$${∂E\over ∂c_i}=2∑_{j}c_j\int u_iu_j-2\int u_i,$$
+at which point we can expand the domain $D$ to the entire real line.
+The condition for stationarity, and presumably best fit, is $Ac=b$, where $a_{ij}=\int u_iu_j$, and $b_i=\int u_i$.
+It will be interesting to see what weights come out of this method.
 
-It would be useful to find a quadrature rule, with weights $w_i$, such that 
-$$∑_i w_if(α_i)|α_i〉≈\int f(α)|α〉d²α.$$
-Suppose we aim to be exactly right, in the case that $f(α)$ is the normalised entire function representing the left hand side, i.e.,
-$$f(β)=∑_i w_if(α_i)〈β|α_i〉.$$
-(I'm assuming we put the factor $1/π$ in $G$.)  This doesn't have a solution for general superpositions of coherent states.  Surely there's some quadrature rule for a sampled phase-space function?
+%% Sampling on random grids
 
-Using the Galerkin criterion, there doesn't seem to be one where the coefficient of $|α_i〉$ only depends on $f(α_i)$.
-However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.
+\topinsert \XeTeXpicfile gweight.pdf width \hsize 
+The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
+The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
+Note how nearby points tend to cancel each others' samples.
+I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
+
+The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
+This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
+Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.
+Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.
+When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.
+This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
+\endinsert
 

### 2013-07-28 paper.tex
writing

@@
 
-If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ⁺|Γ\rangle$.
-Therefore the singularity of $|Γ\rangle$ is easily resolved.
-If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ⁺|ψ\rangle$.
-And $\langle Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
+If, following Bargmann, we identify the quantum space with the entire functions, $|Γ〉$ becomes the projector $〈 Γ⁺|Γ〉$.
+Therefore the singularity of $|Γ〉$ is easily resolved.
+If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.
+And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
 So it is curious that the discrete version of this problem causes so much trouble.
@@
 where the points $z_i$ form a regular grid, or by
-$$\vec g=\pmatrix{\langle u₁,g\rangle & ⋯ & \langle u_R,g\rangle},$$
+$$\vec g=\pmatrix{〈 u₁,g〉 & ⋯ & 〈 u_R,g〉},$$
 where the distributions $u_i$ are a fixed basis.
@@
 This makes regularisation tricky.
-The goal is to represent a state $|ψ\rangle$ as a superposition of coherent states, $|ψ\rangle=|A\rangle c$.
+The goal is to represent a state $|ψ〉$ as a superposition of coherent states, $|ψ〉=|A〉 c$.
 We want the coefficients $c$ to correspond to a phase space function $g$ that is nearly analytic.
-If the amplitudes $|A\rangle$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
+If the amplitudes $|A〉$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
 On an irregular grid, we would expect the coefficients to be larger where the grid is sparse, and smaller where it is dense.
@@
 
-The obvious way to attempt this is to factor the coefficients as $|ψ\rangle=|A\rangle Wc'$, where $c'_i$ are samples that approximate $g(α_i)$, and $W$ is a matrix of weights that adapts the samples to the irregular grid.
+The obvious way to attempt this is to factor the coefficients as $|ψ〉=|A〉 Wc'$, where $c'_i$ are samples that approximate $g(α_i)$, and $W$ is a matrix of weights that adapts the samples to the irregular grid.
 In fact, we can do this exactly.
-If we set $W=(\langle A|A\rangle)^{-1}$, then $c'=\langle A|ψ\rangle$.
-We could try seeking coefficients $c$ such that the norm of $\langle A|A\rangle c$ is minimal.
+If we set $W=(〈 A|A〉)^{-1}$, then $c'=〈 A|ψ〉$.
+We could try seeking coefficients $c$ such that the norm of $〈 A|A〉 c$ is minimal.
 
@@
 There is another way to derive coefficients of the constant function, which might be more stable.
-Suppose that $D$ is a subset of the real line, large enough that the tails of the gaussians outside it are neglibible.
-Then
-$$E=\|f'-1\|=\int_D\left(∑_ic_iu_i(x)-1)²=∑_{ij}c_ic_j\int_D u_iu_j-2∑_ic_i\int_D u_i+\mu(D),$$
-where $μ$ is the integration measure.  The derivative of this with respect to $c_i$ is
-$${∂E\over ∂c_i}=2∑_{j}c_j\int u_iu_j-2\int u_i,$$
-at which point we can expand the domain $D$ to the entire real line.
-The condition for stationarity, and presumably best fit, is $Ac=b$, where $a_{ij}=\int u_iu_j$, and $b_i=\int u_i$.
-It will be interesting to see what weights come out of this method.
+Suppose that $D$ is a subset of the real line, large enough that the tails of gaussians are neglibible outside it.
+Let
+$$U=π^{-½}\pmatrix{e^{-(x-x₁)²}&⋯&e^{-(x-x_R)²}},$$
+and $w$ be a column vector of $R$ real weights.  Then define a residual
+$$r(w)=\int_D(Uw-1)².$$
+The derivative of this with respect to the weights is
+$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U^{\rm T}Uw-U.$$
+(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c^{\rm T}ab$.)
+Unfortunately,
+$$\int (U^{\rm T}U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
+	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
+and $\int u_i=1$, so this is the identical problem again.
 
@@
 This paper will talk about two related but distinct inverse problems.
-Both start with a given wavefunction $|ψ\rangle$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
+Both start with a given wavefunction $|ψ〉$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
 In systems with few modes, these points could be a grid, but in many-mode systems they will have to be more sparse than that: there isn't space to store grid coefficients.
-The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j\rangle$ approximates $|ψ\rangle$.
+The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j〉$ approximates $|ψ〉$.
 This is simply finding a discrete inverse to $G$ sampled on the grid $α_j$; since $G$ is a projection, the pseudoinverse of $G$ should solve it well enough.
-The differential case, where we have an expansion for $|ψ\rangle$ which we want to perturb to represent $|ψ\rangle+|dψ\rangle$, is no harder.
-Because $G$ is linear, we can expand $|dψ\rangle$, and add it to the expansion of $|ψ\rangle$.
+The differential case, where we have an expansion for $|ψ〉$ which we want to perturb to represent $|ψ〉+|dψ〉$, is no harder.
+Because $G$ is linear, we can expand $|dψ〉$, and add it to the expansion of $|ψ〉$.
 
 The second problem, the variational one, is a generalisation.
-As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ\rangle$.
-Instead of expanding $|ψ\rangle$ over phase space, we want to expand it over the columns of $|Dψ(f,α)\rangle$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
-Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)\rangle$ to the solution of the inverse problem, and those of $|∂fψ(f,a)\rangle$ to zero.
+As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ〉$.
+Instead of expanding $|ψ〉$ over phase space, we want to expand it over the columns of $|Dψ(f,α)〉$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
+Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)〉$ to the solution of the inverse problem, and those of $|∂fψ(f,a)〉$ to zero.
 Is this the smallest norm solution?
@@
 However, we can't do that for many mode systems: there will always be accessible states of the system whose support in phase space is remote from the current grid.
-The set $|Dψ\rangle$ will have some small singular values, corresponding to directions where the amplitudes and samples are adjusted in a way that preserves the wavefunction.
-We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ\rangle$ in the future.
+The set $|Dψ〉$ will have some small singular values, corresponding to directions where the amplitudes and samples are adjusted in a way that preserves the wavefunction.
+We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ〉$ in the future.
 
@@
 Otherwise, we're wasting grid points.
-Secondly, when we're expanding a derivative over the tangent space $|Dψ\rangle$, we want to tilt the tangent space closer to the derivative.
+Secondly, when we're expanding a derivative over the tangent space $|Dψ〉$, we want to tilt the tangent space closer to the derivative.
 That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.
@@
 so the Fock states are eigenstates, although the spacing of their energies is uneven.
-The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
+The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1〉 & |2〉 & ⋯ & |N〉}$, with a column vector of coefficients $c(t)$, as
 $$|ν(c(t))〉=|N〉c(t).$$
@@
 In this case, the Fock states diagonalise $H$, so they are exact.
-However, in general the initial state can only be approximately expanded over $|N\rangle$.
+However, in general the initial state can only be approximately expanded over $|N〉$.
 Time was discretised with the explicit Euler formula, which replaces the differential equation

### 2013-07-29 paper.tex
cleanup for Hui

@@
 \input xpmath
-\input respnotes
+\input respnotes \tenrm
 
@@
 
-% It follows that $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
-% $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
-% Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.
-Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.
-
 Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
@@
 The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.
-It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.
+(Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.
 
-If, following Bargmann, we identify the quantum space with the entire functions, $|Γ〉$ becomes the projector $〈 Γ⁺|Γ〉$.
-Therefore the singularity of $|Γ〉$ is easily resolved.
-If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.
+If, following Bargmann, we identify state space with the entire functions, $|Γ〉$ is identified with the projector $〈 Γ⁺|Γ〉$.
+Therefore, although $|Γ〉$ is singular, the singularity is easily resolved.
+If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be done simply enough by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.
 And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
-So it is curious that the discrete version of this problem causes so much trouble.
+So it is curious that the discrete version of this problem has caused so much trouble.
 
@@
 The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
-$$π^{-½}〈β|Γ〉f=π^{-1}∫〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}∫〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
+$$π^{-½}〈β|Γ〉f=π^{-1}\int 〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}\int 〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
 This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.
-There are many theorems to the effect that singular values of real convolution operators decay rapidly, and the operators are badly conditioned.
+There are many theorems to the effect that real convolution operators are badly conditioned.
 It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.
@@
 We have excluded the small preimages of the states with large particle number, so they must be represented by functions with support close to the origin, but with large norms.
-Hence the restricted operators will have small singular values corresponding to states with many particles.
+Hence the restricted operators will be badly conditioned.
 
@@
 $$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
-The kets $|g_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.
+The kets $|u_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.
 The $σ_i$ are positive real numbers, called singular values.
@@
 
-In the case where $χ=1$, and $\Gx=\G$, the expansion is quite arbitrary.
-The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire functions for which $\G v_i=|u_i〉$, and every singular value is 1.
+In the case where $χ(z)=1$, and $\Gx=\G$, the expansion is quite arbitrary.
+The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire function for which $\G v_i=|u_i〉$, and every singular value is 1.
 If the domain and range of $\G$ were the same, this would indicate that it was a projector.
-The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component keeps the same expansion coefficients, while the basis $v_i$ is replaced by the $|u_i〉$.
+The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component retains its expansion coefficients as the basis $v_i$ is replaced by $|u_i〉$.
 
@@
 If $\Gx$ were one to one, each $v_i$ would be the unique preimage of $|g_i〉$; since $\Gx$ is singular, $v_i$ is the preimage with least norm.
-The singular values are the ratios of $‖\Gx v_i‖$ to $‖v_i‖$.
+The singular values are the ratios of $‖\Gx v_i‖=‖σ_i|u_i〉‖$ to $‖v_i‖$.
 If $|u_i〉$ has a low particle number, it can be expanded efficiently: $v_i$ will be close to $〈Γ⁺|u_i〉$, and $σ_i$ will be close to 1.
 If $|u_i〉$ includes components with large particle number, its expansion will have a larger norm than itself.
-In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, and $σ_i$ will be small.
+In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, multiplied by a large constant to maintain normalisation, and $σ_i$ will be small.
 
@@
 In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.
-When the expansion is discretised, however, it is very easy for the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ to disappear in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.
+When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contain the important information about $|ψ〉$ are very likely to be lost in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.
 
-Let's compute some singular values and vectors.  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
+Let's compute some singular values and vectors.  (And the the wrong answer—try again.)  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
 $$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
@@
 
-\topinsert \XeTeXpicfile rsv2.pdf width 0.5\hsize
-\XeTeXpicfile rsv3.pdf width 0.5\hsize \endinsert
-\topinsert \XeTeXpicfile rsv4.pdf width 0.5\hsize \endinsert
+\topinsert \hbox{\XeTeXpicfile rsv2.pdf width 0.5\hsize
+\XeTeXpicfile rsv3.pdf width 0.5\hsize}
+\XeTeXpicfile rsv4.pdf width 0.5\hsize\hfil\break
+The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase spaces is truncated to a square about the origin, whose size varies.
+These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert
 
@@
 
-The most important issue is the Picard condition.
-This restates a very basic observation: for the problem $G\dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $f$ over an orthonormal basis must have a finite norm.
-If the basis is chosen as the right singular vectors $v_j$, this means that the sum of the squared moduli of the components must be finite.
-For $\dot f$ to satisfy the equation, the component of $v_j$ is $〈g_j|H|ψ〉/σ_j$.
+The most basic issue is the Picard condition.
+This restates a simple observation: for the problem $\G \dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $\dot f$ over an orthonormal basis must have a finite norm.
+If the basis is chosen as the right singular vectors $v_i$, this means that the sum of the squared moduli of their coefficients must be finite.
+The component of $v_i$ is $〈u_i|H|ψ〉/σ_i$.
 So the Picard condition is that, for the equation to have a solution, the series
-$$∑_{j=1}^∞{\bigl|〈g_j|H|ψ〉\bigr|²\over σ_j²}$$
+$$∑_{i=1}^∞{\bigl|〈g_i|H|ψ〉\bigr|²\over σ_i²}$$
 must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.
@@
 The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.
-I was working in a truncated Fock space; I wonder what happens as the truncation increases, while the grid stays a constant size?
-Or alternately, if the grid shrinks while the Fock space stays constant.
 
@@
 The exact solution has residual zero, of course.
-The problem is that it is swamped by components with small singular values of $\Gx$.
+The problem is that it is dominated by components belonging to singular values of $\Gx$.
 Suppose we modify this problem, and seek $|ψ〉$ to minimise $‖\Gx f-|ψ〉‖²+λ‖f‖²$, where $λ$ is a positive number, known as a Lagrange parameter for reasons that will become clear shortly.
@@
 
-\topinsert\XeTeXpicfile lcurve-1.pdf \endinsert
+\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
 
-An example for a typical application is shown below.
-The context is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.
-The real state is $x₀$, and it has generated observations $b=Ax₀+e$, where $e$ consists of measurement error.
-The problem $Ax=b$ is solved for $x$, with various Lagrange parameters.
+An example for a typical application is shown above.
+The context in which Tychonov regularisation is usually applied is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.
+The real state is $x₀$, and the observed data are $b=Ax₀+e$, where $e$ consists of measurement error.
+The problem $Ax=b$ is solved for $x$, and regularised with various Lagrange parameters.
 At the top left, the norm of $x$ is constrained to be zero, so the solution is $x=0$, and the residual is the norm of $b$.
@@
 At the elbow of the curve, the norm of the solution is constrained to be close to the norm of the real state $x₀$, the solution is close to $x₀$, and the residual is the norm of the measurement error.
-Relaxing the constraint further allows solutions $x$ that fit the measurements better than the real state $x₀$ does.
+Relaxing the constraint further allows spurious solutions $x$ that fit the measurements better than the real state $x₀$ does.
 Typically, they won't fit much better, so the residual doesn't decrease much below $‖e‖$.
@@
 This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.
-(What is the SVD of the DFT matrix?
-It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)
 
@@
 The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.
-A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $|g(a*)|²$, where $g=G⁺|ψ〉$.
+A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $P(α)=|e^{-½|α|²}〈G⁺|ψ〉(α*)|²$.
 These are plotted below, as circles in phase space.
-For smaller ensembles, the results were not consistent between draws.
+For ensembles with fewer components, the results were very inconsistent between draws.
+
+The state $|ψ〉$ was expanded in the form $|ψ〉=|A〉\vec f$, where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $\vec f$ is a column vector of $R$ complex weights.
+This was done with Tychonov regularisation, for various Lagrange parameters: these and the residuals are shown with the graphs.
+The complex weights $c$ are indicated by brightness and color as above.
+Note that least norm regularisation has the desirable property that nearby $α_r$ have similar $c_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.
+
+\topinsert\XeTeXpicfile cats.pdf width \hsize
+\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
 
@@
 
-It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the l-curve experiments for the expansion problem.
+It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the L-curve experiments for the expansion problem.
 To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.
-In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around $π^½$.
-And we're back to the other open question, how to relate the norm of a function to the norm of a set of samples at irregular points.
-How can I estimate how much of phase space “belongs” to each amplitude sample, in order to get a volume over which the function has that value?
-
-The obvious way would be to assign amplitudes to the nearest sample.
-That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
-What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
-There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
-Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
-I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
-What's the best exponent?
-I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around 1.
 
@@
 
-\topinsert\XeTeXpicfile cats.pdf width \hsize\endinsert
+\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
+Picard plot for cat expansion. \endinsert
 
-The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉f=〈A|ψ〉$, where $f$ is a vector of weights such that $|A〉f$ approximates $|ψ〉$.
-Below it is a plot of the expansions over number states of the singular kets $〈n|u_r〉$, where $|A〉=|U〉SV$.
+The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉\vec f=〈A|ψ〉$, where $\hat f$ is a vector of weights such that $|A〉\vec f$ approximates $|ψ〉$.
+The singular value decomposition is $|A〉=|U〉SV$.
 For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.
@@
 At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.
-In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied.
-Seeing that $G$ is perfectly conditioned, this is a very bad discretisation.
+This is shown in the plot of Fock state components of the singular kets.
+In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied: the discrete doesn't have a finite solution.
+Seeing that the continuous operator $\G$ is perfectly conditioned, this is a very bad discretisation.
 
-The final graph is an L curve when the discrete problem is Tychonov regularised.
-It shows the expected shape, with minimum residual around $2×10^{-8}$.
-Six regularised solutions are shown above; the modulus of $f_r$ is the disk's saturation, and the phase determines its hue.
-Note that least norm regularisation has the desirable property that adjacent $α_r$ have similar $f_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.
-
-\topinsert\XeTeXpicfile catpicard.pdf width \hsize\endinsert
+The final graph is an L curve when the discrete problem is Tychonov regularised.  It shows the expected shape, with minimum residual around $2×10^{-8}$.
 
@@
 At the same time, the directions with small singular values will be used to keep the discretisation stable.
-I think this requires two things: that $|Dψ〉$ is reasonably well conditioned, and that the angle between $H|ψ〉$ and the span of $|Dψ〉$ is small.
-This raises a question: are there singular directions that have little effect on $|ψ〉$, but a large effect on $|Dψ〉$?
-This seems likely; for example, shuffling the amplitudes to move two components further apart in phase space, then adjusting the weights so that $|ψ〉$ stays the same.
-This could be done by a kind of Tychonov anti-regularisation: we improve the stability as much as we can, while constraining the change in $|ψ〉$.
-I might need to constrain the norm of $dz$ too, so that linear approximations hold.
 
@@
 
-\centerline{\XeTeXpicfile psfa.pdf width \hsize}
-
-\centerline{\XeTeXpicfile psv.pdf width \hsize}
+\topinsert 
+\vskip -40mm
+\XeTeXpicfile psfa.pdf width \hsize
+\vskip -90mm
+\XeTeXpicfile psv.pdf width \hsize\endinsert
 
@@
 
+An obvious way would be to assign amplitudes to the nearest sample.
+That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
+What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
+There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
+Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
+I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
+What's the best exponent?
+I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+
 %% Sampling on random grids
@@
 
-The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.  This can also be seen from the expansions of coherent states.  The normalised entire expansion of $|α〉$ is
-$$g(β*)=〈β|α〉=e^{-½|α|²-½|β|²+β*α},$$
-whence
-$$‖g‖²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
-The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.
-
-The lemma gives a very simple singular value expansion of $G$.
-Because there is only one nonzero singular value, any orthonormal basis of ket space will do as the left singular vectors.
-Let's choose the number states for familiarity.
-Then
-$$Gg=∑_n |n〉π\int ν_n*(α)g(α)\,d²α=|N〉π\int ν*(α)g(α),$$
-where $|N〉=\bigl(|0〉\ |1〉\ |2〉\ …\bigr)$ and $ν(α)=\bigl(ν₀(α)\ ν₁(α)\ …\bigr)$ could be thought of as infinite matrices whose columns are the left singular vectors $|n〉$ and their normalised entire expansions, the right singular vectors $v_n$.
-The pesudoinverse,
-$$G⁺|ψ〉=νπ〈N|ψ〉= ∑_n ν_nπ〈n|ψ〉,$$
-maps $|ψ〉$ to its normalised entire expansion.
-Expanding over number states is useful because the maximum number in a system is usually limited by conservation of energy.
-In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.
-
 Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
-(And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
+And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
 
@@
 We are free to add these as we like, in order to maintain a grid on which we can expand $|ψ〉$ in the future.
-
-So what are the critera for adjusting the grid?
-I think there are two obvious ones: firstly, that the inverse problem stays well-posed.
-Otherwise, we're wasting grid points.
-Secondly, when we're expanding a derivative over the tangent space $|Dψ〉$, we want to tilt the tangent space closer to the derivative.
-That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.
-The first criterion seems redundant: if we could better represent the goal by shifting two sample points apart, the second criterion will do that automatically.
-However, if we only care about the tangent space and not the derivative map, we might need very large adjustments of the amplitudes and weights to get to the point in the tangent space that we want.
-So a condition number criterion seems sensible.
-There might be other criteria that I've overlooked.
 
@@
 
-\centerline{\XeTeXpicfile expens.pdf width \hsize}
+\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
 
@@
 
-\centerline{\XeTeXpicfile exp.pdf width \hsize}
+\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
 
@@
 
-\centerline{\XeTeXpicfile spsns.pdf width \hsize}
-Amplitudes of components
+\topinsert \XeTeXpicfile spsns.pdf width \hsize
+Amplitudes of components \endinsert
 
-\centerline{\XeTeXpicfile traj.pdf width \hsize}
-Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted)
+\topinsert \XeTeXpicfile traj.pdf width \hsize
+Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
 
-\centerline{\XeTeXpicfile picard.pdf width \hsize}
-Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied.
+\topinsert \XeTeXpicfile picard.pdf width \hsize
+Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
 
-\centerline{\XeTeXpicfile svnum.pdf width \hsize}
+\topinsert \XeTeXpicfile svnum.pdf width \hsize
 Expected particle number for left singular vectors.
@@
 In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$.
+\endinsert
 
@@
 
-\centerline{\XeTeXpicfile flwc0n10.pdf width \hsize}
+\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
 
-\centerline{\XeTeXpicfile flwc0n5.pdf width \hsize}
+\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
 
@@
 The things I've noticed so far are:
-\item{1.} When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.
+When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.
 It isn't accurate, though.
@@
 
-\item{4.} When the least squares problem is solved with a different distance function, which sets distances along the large $N$ directions to zero, the fit fails on the first step.
+When the least squares problem is solved with a different distance function, which sets distances along the large $N$ directions to zero, the fit fails on the first step.
 Note that this isn't really a distance.

### 2013-07-29 paper.tex
writing

@@
 
-Let's compute some singular values and vectors.  (And the the wrong answer—try again.)  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
-$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
-	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
-	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
-where
-$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
-Presumably, the right singular vectors have the $f_n$ of least norm such that 
-$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$
+Let's compute some singular values and vectors.  Suppose that $χ$ is the characteristic function of a disk, with
+$$χ(α*)=\cases{1 & $|α|≤b$ \cr 0 & otherwise.}$$
+From symmetry, let's guess that the singular kets are the number states—we will check this numerically later.
+It can be shown by a Lagrange multiplier argument that the right singular functions are the truncations to the disk of $f_n=〈Γ⁺|n〉$.
+Then the singular values are
+$$σ_n={‖\Gx f_n‖\over ‖f_n‖}.$$
+We take
+$$f_n=π^{-½}〈α|n〉e^{½|α|²}=π^{-½}∑_{n=0}^∞ {α^{n\ast}\over \sqrt{n!}}.$$
 
-What is the appropriate norm to use?  In the case that $χ$ is a characteristic function on the domain $D$, it is presumably the truncated Bargmann norm
-$$〈g,f〉=\int_D e^{-|α|²}g*(α)f(α)\,d²α=\int e^{-|α|²}χ(α)g*(α)f(α)\,d²α.$$
-Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
-$$‖f_n‖²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
-given 
-$$2\sqrt π(n!)^{-½}\int_0^b r^{n+1}e^{-r²}f_n(r)dr=1.$$
-If one considers changing $f$ by $Δa$ in an interval around $a$, the Lagrange condition is
-$$re^{-r²}f_n(r)\propto r^{n+1}e^{-r²},$$
-or 
-$$f_n(r)\propto rⁿ.$$
-The analytic functions still win.  Let $f_n(r)=rⁿ$.  Then
-$$‖f_n‖²=2π\int_0^b r^{2n+1}e^{-r²}\,dr$$
-and
-$$‖ \Gx f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
-so we have
-$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
-	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
-According to Wolfram,
-$$\int_0^b r^{2n+1}e^{-r²}dr=½\left(Γ(n+1)-Γ(n+1,b²)\right),$$
-where $Γ$ is Mathematica's version of the incomplete gamma function.  Therefore, recalling that $n!=Γ(n+1)$,
-$$σ_n=\left(1-{Γ(n+1,b²)\over Γ(n+1)}\right)^½.$$
-Unfortunately, Matlab defines the incomplete gamma function in a different way than Mathematica does.
-Fortunately, that Matlab function, {\tt gammainc}, is exactly the square of the right-hand side of the above equation.
+First evaluate $‖f_n‖$, using the Bargmann norm.  We have
+$$\eqalign{ ‖f_n‖²&={1\over π}\int_{|α|≤b} e^{-|α|²}{|α|^{2n}\over n!}\,d²α \cr
+	&={1\over π}\int_0^b e^{-r²}{r^{2n}\over n!}2πr\,dr \cr
+	&={1\over n!}\int_0^{b²} u^ne^{-u}\,du=Γ(n+1,b²),}$$
+where $Γ(n,x)$ is the incomplete gamma function, as defined in Matlab.  (Everyone defines this a different way.)  The other norm is given by
+$$\eqalign{ \Gx f_n&=π^{-½}\int_{|α|≤b} |α〉e^{-½|α|²}f_n(α*)\,d²α \cr
+	&={1\over π}∑_{m=0}^∞\int e^{-|α|²}{α^{m}\over \sqrt{m!}}|m〉{α^{n\ast}\over \sqrt{n!}}\,d²α \cr
+	&={1\over π}\int_0^b dr re^{-r²}∑_{m=0}^∞{r^{m+n}\over \sqrt{m!n!}}|m〉\int_0^{2π} dφe^{i(m-n)φ} \cr
+	&=\int_0^b dr 2re^{-r²}{r^{2n}\over n!}|n〉 \cr
+	&=Γ(n+1,b²)|n〉.}$$
 
-This all seems wrong, because the incomplete gamma function varies the wrong way with $n$.  It becomes small as $n→0$.
+The singular value comes out to 
+$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{Γ(n+1,b²)}.$$
 
-Numerical experiments in the script {\tt sve.m} support this.
-The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.
-Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.
-The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.
+Numerical experiments in the script {\tt csve.m} support this.
+The operator $\G$ was discretised on a grid in phase space, with spacing $h=0.3$, bounded by the circle about the origin of radius $b$ for $b=3$, 4 and 5.
+Ket space was discretised by taking inner products with the number states $|0〉$ through $|24〉$.
+The graph below shows the singular values of the truncated operator $\Gx$.
 
@@
 
+The contour plot shows how the analytic expression for $σ_n$ depends on $n$ and the truncation radius $b$.
+The value has decayed to 0.7 when $n=b²$.
+The rule of thumb is that we get one large singular value for every unit circle of phase space, with area $π$.
+
+\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
+
 The following three figures show the right singular vectors, sampled versions of phase space functions.
 The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.
-These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.
-In the continuous limit there is only one singular value, and the vectors are degenerate.
-For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.
+These look very much like truncated number states.
 As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.
 
-\topinsert \hbox{\XeTeXpicfile rsv2.pdf width 0.5\hsize
-\XeTeXpicfile rsv3.pdf width 0.5\hsize}
-\XeTeXpicfile rsv4.pdf width 0.5\hsize\hfil\break
-The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase spaces is truncated to a square about the origin, whose size varies.
+\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
+\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
+The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
 These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert

### 2013-08-02 paper.tex
writing

@@
 
+The goal of fiding samples of the phase space distribution $f(α_r)$ is to regularise the expansion and inverse problems.
+The expansion function with the least norm is guaranteed to be a nice function: we don't know much about the finite superposition whose weights have the least norm.
+
+To the the norm of the function $f$, sampled on an irregular grid $α_r$, the obvious method is to interpolate.
+Doing this might help with irregular weights due to irregular sample points, because a least-squares interpolant will average the weights at several adjacent points.
+However, there is probably a better way to adjust the weights for irregular sampling.
+
+We want to interpolate a function of a complex variable, that should be nearly analytic.  To me, the obvious way to do it is a least squares fit to the monomials
+$$U=e^{-½|z|²}\pmatrix{1 & z & z* & z² & zz* & z^{2\ast} & z³ & z²z* & ⋯}.$$
+It helps to think of $z^mz^{n\ast}=r^{m+n}e^{i(m-n)φ}$.
+Note that if these monomials are orthogonalised by Gramm-Schmidt, in the order they are written, the functions $e^{-½|z|²}\pmatrix{1 & z & z² & z³ &  ⋯}$ are retained as part of the orthonormal set: $z^n$ is the first monomial proportional to $e^{inφ}$, so it is orthogonal to all those before it.
+The same is true for $z^{n\ast}$, the first monomial proportional to $e^{-inφ}$.
+So this basis is an obvious one for projecting onto analytic functions in $z$ or $z*$.
+It will have the form
+$$e^{-½|z|²}\pmatrix{1 & re^{iφ} & re^{-iφ} & r²e^{2iφ} & p_{11}(r) & r²e^{-2iφ} & r³e^{3iφ} & p_{12}(r)e^{iφ} & p_{21}(r)e^{-iφ}& ⋯},$$
+where $p_{ij}$ is a polynomial of degree $i+j$.  Someone must have investigated these orthogonal polynomials before, but they don't appear in Courant \& Hilbert or in Abramowitz \& Stegun.
+
+For least squares fitting, orthogonal polynomials are not necessary: the QR factorisation does that work for us.
+To find norms, we do need to know the grammian matrix, whose elements are the Bargmann inner products of the monomials:
+$$\eqalign{a_{mnpq}&=\int (z^{m\ast}zⁿ)*(z^{p\ast}z^q)e^{-|z|²}\,d²z \cr
+	&= 2πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q+1}\,dr \cr
+	&= πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q}\,2r\,dr \cr
+	&= πδ_{m+n,p+q}Γ({m+n+p+q\over 2}+1)=π(n+p)!δ_{m+n,p+q}.}$$
+Some fits of these to expansions of cat and coherent states over random amplitudes are shown below.
+They suffer the usual problem that polynomial fits oscillate more wildly as the degree of the polynomial increases; there are meant to be ways to avoid that, but I'm not an expert in them.
+
 %% Truncated Hilbert space

### 2013-08-06 paper.tex
cancellation of weights explained

@@
 
-%% Phase space expansion as a linear operator
+\beginsection{The phase space expansion operator}
 
@@
 
-%% SVE lemma
-
-% Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.
-% 
-% There are two ways one could set about proving the Hypothesis.
-Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.
-Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
-% ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
-% 
-% This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.
-My numerical experiment suggests otherwise; let's ignore that possibility for now.
-% 
-% Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
-% $$f(z)=∑c_n{zⁿ\over (n!)^½},$$
-% for some normalised vector $c$.
-% 
-%   Our expansion function is 
-% $$g(α)=f(α)e^{-½|α|²}.$$
-% If $f$ has the Taylor expansion
-% $$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
-% then 
-% $$Gg = ∑_n c_n|n〉.$$
-% The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
-% $$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
-% the Glauber preimage of $|n〉$.  We have 
-% $$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
-% so the squared norm is
-% $$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
-% According to Wolfram,
-% $$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
-% so
-% $$‖g‖=π^½.$$
-% 
-% Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.
-Or at least I think so.
-I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.
-
-%% Consequences of lemma
+%% Consequences of Γ being a projection
 
@@
 
-%% Finite phase space
+\beginsection{Truncated phase space}
 
@@
 
+\beginsection{Regularisation of the truncated expansion problem}
 %% Picard plots
@@
 
-It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the L-curve experiments for the expansion problem.
 To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.
@@
 
-%% Adapting the sample points to maintain condition
+A common observation when states are expanded over an irregular grid is that, when two sample amplitudes are close together, the weights tend to be large, and cancel out.
+This is not at all surprising when one considers that the coherent states $|α〉$ and $|α+h〉$ become nearly parallel in Hilbert space as $h→0$.
+The angle between these vectors satisfies
+$$|〈α|α+h〉|=e^{-½|h|²}=‖|α〉‖‖|β〉‖\cos θ=\cos θ,$$
+and equating Taylor series gives
+$$θ=|h|+O(h²).$$
+Locally, phase space has the same geometry as the surface of the unit sphere.
+The figure demonstrates why weights of nearly parallel (or antiparallel) vectors tend to cancel out, and the second figure shows how this looks in phase space.
+
+\topinsert \XeTeXpicfile geometry-1.pdf
+\it A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
+As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.
+This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
+\endinsert
+
+\topinsert \XeTeXpicfile cancel.pdf width \hsize
+\it The vacuum state was expanded over two coherent states, with amplitudes $±h$.
+The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.
+The norm of the expansion vector, and the residual of the expansion, are graphed above as black lines.
+As the amplitudes converge, the residual decreases—the leftmost phase space plot is very nearly exact—while the norm of the expansion stays around 1.
+This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.
+
+The second row of images, and the red lines on the plot, show similar expansions of the normalised state parallel to $|0.1〉-|-0.1〉$.
+This is almost exactly parallel to the derivative of the vacuum state with respect to the real part of its amplitude.
+Again, the leftmost image is nearly an exact expansion.
+This time, the entire expansion does not have support around $α=0$.
+As the expansion amplitudes converge, the residual decreases, but the norm of the expansion vector explodes.
+As explained in the text, the distance between two small amplitudes in phase space is very nearly the angle between the ket vectors for the corresponding coherent states.
+Thus this is exactly the expansion depicted in the last figure.
+
+If the two sample amplitudes were part of an irregular expansion, and there happened to be no other sample amplitudes nearby, a small component along the direction of the second state would dominate the expansion coefficients at these points.
+But, given those sample amplitudes, that would be the only way to represent such a component.
+\endinsert
+
+\beginsection{Adapting the sample amplitudes to preserve condition}
 
@@
 
-%% Discrete phase space
+\beginsection{Discrete phase space}
 
@@
 
-%% Why grids don't work, and variation necessary
+\beginsection{Ways the problem has been simplified for this paper}
 
@@
 
-%% convergence to the initial state
+\beginsection{Regularising the inital fit equation}
 
@@
 
-%% Shape of Hilbert space trajectories
+\beginsection{Quantum dynamics in Hilbert space}
 
@@
 This would indicate that the ODEs are stiff.
+
+\beginsection{Quantum dynamics in phase space}
 

### 2013-08-06 cancel.m

@@ -1,0 +1,43 @@
+% Expand the derivative of a coherent state over two samples, as they
+% approach each other.
+
+global N; N = 20; brackets
+
+[~, f] = evan([0.1 -0.1], 'even');
+f(1) = f(1)+1i*pi;
+c = zeros(N+1,2);
+c(:,1) = nq*evan(0,0);
+c(:,2) = sum(nq*evan([0 1i*pi] ,[0.1 -0.1]), 2);
+for j = 1:2; c(:,j) = c(:,j)/norm(c(:,j)); end
+
+nh = 13;  h = logspace(-1.2,log10(2.5), nh);  
+psaxis = -3:0.1:3;
+[x,y] = meshgrid(psaxis);  grid = (nq*evan(x(:)+1i*y(:), 'even'))';
+
+figure
+nms = zeros(2,nh);  rds = nms;
+for j = 1:2
+for i=1:nh
+    a = 0.5*[-h(i) h(i)];  [M,f0] = evan(a, 'even');
+    d = (nq*M)\c(:,j);
+    ch = sum(nq*evan(f0+log(d), a), 2);
+    nms(j,i) = norm(d);  rds(j,i) = norm(c(:,j)-ch);
+    np = 4*(i-1)/(nh-1);
+    if mod(np, 1)==0    % Matlab's test for an integer
+        subplot(4, 5, 11+5*(j-1)+np)
+        zplot(psaxis, psaxis, grid*ch);
+        hold on; plot([h(i) -h(i)], [0 0], 'ow')
+        set(gca, 'DataAspectRatio', [1 1 1])
+    end
+end
+end
+
+subplot 411
+plot(h, nms(1,:), '-k', h, nms(2,:), '-r')
+ylabel('expansion norms')
+subplot 412
+plot(h, rds(1,:), '-k', h, rds(2,:), '-r')
+ylabel('expansion residuals')
+xlabel('h')
+
+print(gcf, 'cancel.pdf', '-dpdf')

### 2013-08-06 paper.tex
writing

@@
 
+However, in these single mode numerical experiments, it is often useful to expand over a regular grid.
+For example, this allows us to plot phase space functions.
+It is convenient to normalise the samples so that $‖\vec g‖=‖A\vec g‖$.
+If the sample amplitudes $α_r$ form a square gird with spacing $h$, the condition $\vec g_r=cg(α_r)$ is satisfied when
+$$∑_r|\vec g_r|²=c²∑_r|g(α_r)|²=\int |g|²\,d²α≈h²∑_r|g(α_r)|²,$$
+whence $c=h$, and the discretisation is
+$$\vec g_r=hg(α_r).$$
+
 This makes regularisation tricky.

### 2013-08-09 paper.tex
writing

@@
 The invention of the laser called for a quantum theory of intense radiation.
-It is difficult to analyse laser beams because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.
+Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.
 The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.
@@
 Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.
-When atomic Bose-Einstein condensates were made, and very coherent matter became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.
+When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.
 They have solved many problems, but are limited by intractable convergence issues.
@@
 In first quantisation, such an operator is the complex amplitude
-$$a=?(λx+ip/λ),\qquad\hbox{Get constants from Cahill}$$
-where $x$ and $p$ are the position and momentum operators; the bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency ?.
+$$a=(2\hbar)^{-½}(λx+ip/λ),$$
+where $x$ and $p$ are the position and momentum operators for a particle with mass $m$.
+The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
 
@@
 $$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.$$
+When kets are considered as vectors, the angle between coherent states satisfies
+$$\cos θ=‖|α〉‖‖|β〉‖\cos θ=|〈β|α〉|=e^{-½|β-α|²},$$
+and comparing Taylor series gives
+$$θ=|β-α|+O(|β-α|²).$$
+This has a simple geometric analogy where phase space is mapped locally onto the surface of the unit sphere, the coherent states are vectors from the origin to the surface, and $|α-β|$ is a spherical angle.
+The analogy fails when the angles between vectors are no longer small; for example, it is misleading in regard to the relation between $|α〉$ and $∂_α|α〉$.
+
 The projector satisfies
@@
 This is not at all surprising when one considers that the coherent states $|α〉$ and $|α+h〉$ become nearly parallel in Hilbert space as $h→0$.
-The angle between these vectors satisfies
-$$|〈α|α+h〉|=e^{-½|h|²}=‖|α〉‖‖|β〉‖\cos θ=\cos θ,$$
-and equating Taylor series gives
-$$θ=|h|+O(h²).$$
-Locally, phase space has the same geometry as the surface of the unit sphere.
 The figure demonstrates why weights of nearly parallel (or antiparallel) vectors tend to cancel out, and the second figure shows how this looks in phase space.

### 2013-08-15 paper.tex
writing

@@
 
+%% Overview
+
+There appear to be two sorts of instabilities.
+The first, static, sort occurs when a quantum state is expanded over a finite set of coherent states, and those states have irregular amplitudes.
+This is similar to the kind that numerical analysts call ill-posed problems, and the methods they use to analyse and resolve them can be applied here.
+However, the cause of ill-posedness in our problem differs from those usually studied, and the usual methods of resolution don't work as well as they usually do.
+There is one advantage of the expansion problem: because it is iterated at each timestep, the dynamics of the amplitudes can be exploited to so that they posed better expansion problems as the simulation progresses.
+
+The second type of instability is a dynamical one.
+This occurs whenever Schrödinger's equation is integrated with a quartic hamiltonian, over discrete timesteps.
+In the second quantised picture, it occurs because the repulsion between particles causes components where many particles occupy the mode to oscillate very rapidly.
+The simulation becomes stable when these components are removed, but that changes the dynamics.
+Integrating Schrödinger's equation with repulsive particles either requires very short timesteps, or special integration formulae that exploit features of Hamiltonian dynamics.
+Applying these formulae to the variational dynamics of the coherent states, which are not Hamiltonian, is somewhat subtle.
+
 \beginsection{The phase space expansion operator}
@@
 Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
-$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
+$$〈 Γ⁺|ψ〉(α)=π^{-½}e^{½|α|²}〈 α|ψ〉.$$
 Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.
@@
 With these directions, we can form some of the smaller singular vectors of the full problem.
+
+\beginsection{Convergence of discretely sampled expansions}
+
+To get a sense of what residuals to expect from optimised expansions, the expansion of a state as an entire function, $f=〈Γ⁺|ψ〉$, can be sampled at discrete amplitudes, and the integral $|Γ〉f$ evaluated as a sum.
+In the limit of many amplitudes, this should converge to $|Ψ〉$.
+I chose the vacuum state as $|Ψ〉$.
+I evaluated the integral in two ways, supposed to be extreme cases.
+
+The first way used a grid, restricted to the disk $|α|<2.5$.
+The pitch of the grid, $h$, was varied, and $f(α_r)=π^{-½}e^{½|α_r|²}〈α_r|ψ〉$ evaluated at each point.
+The integral was discretised in the obvious way, to give approximations
+$$|ψ_R〉={1\over πh²} ∑_{r=1}^R |α_r〉〈α_r|ψ〉.$$
+A residual $‖|ψ〉-|ψ_R〉‖$ was found for each $R$; these are plotted as the solid line in the figure.
+As the components increase, the residual asymptotes to the phase space truncation error $‖|Γ_χ〉〈Γ⁺|ψ〉-|ψ〉‖$.
+(Calculate that precisely.)  The gaussian function $f$ is very smooth, so this converges rapidly as $R$ increases.
+
+The second way was a type of stochastic integration.  The samples $α_r$ were drawn randomly, from the normal distribution $|〈α|ψ〉|$.  The integral was approximated by 
+$$|ψ'_R〉=∑_{r=1}^R |α_r〉{〈α_r|ψ〉\over |〈α_r|ψ〉|},\qquad |ψ_R〉={|ψ'_R〉\over||ψ'_R〉|}.$$
+The phase of $f$ is integrated by sampling, but the modulus is integrated stochastically.
+The residuals are plotted as the broken line.
+In principle, this will converge exactly as $R$ increases.
+However, the convergence is $R^{-½}$, as seen on the graph, and expected from the laws of large numbers.
+
+These two lines set bounds on how well we can reasonably expect to solve the expansion problem on an irregular grid.
+A least squares method that does no better than sampling the argument of $f(α)$ at random $α$ is worthless: we need to do better than the dashed line.
+On the other hand, it seems unreasonable to expect samples on an irregular grid to fit $|Ψ〉$ better than samples on a square grid.
+If we try to match the solid line, we should expect the expansion coefficients to be very sensitive to the grid, and not to resemble $f(α_r)$.
+If we try to do better than the truncation error, we will have to represent large number components, and the expansion coefficients will become very large.
+
+\topinsert\XeTeXpicfile converge.pdf width \hsize
+Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.
+\endinsert
 

### 2013-08-21 paper.tex
writing

@@
 where $x$ and $p$ are the position and momentum operators for a particle with mass $m$.
-The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
+(The dimensions don't work: $λ²$ is an action, but then $a$ has dimensions.)  The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
 
@@
 In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.
-When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contain the important information about $|ψ〉$ are very likely to be lost in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.
+When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ is likely to be lost in the large component of $〈Γ⁺_χ|ψ〉$ that gives minor details of high particle number components of $|ψ〉$.
 
@@
 \topinsert\XeTeXpicfile converge.pdf width \hsize
-Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.
+Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.
+The solid line shows sampling on a regular grid, the dashed one at random amplitudes.
+Details in text.
+The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.
 \endinsert
@@
 The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.
+
+%% Expansion tension
+
+To capture the concept of a smooth expansion over an irregular set of nonorthogonal vectors, we'll define a quantity called the sampling tension.
+Suppose that $|u〉$ and $|v〉$ are nearly parallel vectors, and the state $|ψ〉$ is expanded over a set that includes them, as $|ψ〉≈c_u|u〉+c_v|v〉+⋯$.
+A relaxed expansion is one where $c_u|u〉$ and $c_v|v〉$ are as similar as possible; this can be ensured, regardless whether the $c$ are complex numbers of whether $|u〉$ and $|v〉$ are parallel or antiparallel, by requiring that the component of $c_u|u〉$ along $|v〉$ is equal to the component of $c_v|v〉$ along $|u〉$.
+The difference between these components is
+$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖²} - {c_u〈v|u〉\over ‖|v〉‖²}\right|.$$
+These differences form a tension matrix.
+In the case that $|u〉$ and $|v〉$ are nearly orthogonal, their components should be independent.
+In this case, the $T_{uv}$ will be small, because of the proportionality to $〈u|v〉$.
+The tension matrix is a reasonable measure of the smoothness of an expansion: its Frobenius norm, the sum in quadrature of the components, would be a good summary.
 

### 2013-08-27 paper.tex
writing

@@
 
-\beginsection{Convergence of discretely sampled expansions}
+\beginsection{Discretely sampled expansions}
+
+On a computer, the expansion in phase space must be discretised.  The obvious way to do so is to replace the integral with a sum, so that
+$$|ψ〉=d₁|α₁〉+⋯+d_R|α_R〉=|A〉d,$$
+where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $d$ is a column vector of the coefficients.
+In this paper, $c$ will usually denote the coefficients of expansion over orthogonal bases, and $d$ will represent coefficients of nonorthogonal expansions.
+If the system has only a few modes, so that $α$ is a low-dimensional vector, the set of sampled amplitudes $A=\{α₁, …,α_R\}$ could be a regular grid; this makes many things easier.
+When there are many modes, and $α$ is a large vector, the storage required for grids becomes exponentially large, and the sample points will have to be irregular.
+
+This expansion is formally equivalent to the continuous one, in that
+$$∑_r d_r|α_r〉=|Γ〉f_A,$$
+where
+$$f_A(β)=π^½∑_rd_re^{½|α_r|²}δ(β-α_r).$$
+I don't know enough about generalised functions to be comfortable working with these forms; in particular, the norm of $f_A$ is a bit mysterious.
+If this could be clarified, it might provide some useful insights, for example how the norm of $f_A$ compares to that of $〈Γ⁺|ψ〉$.
 
@@
 The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.
+The tension is the Frobenius norm of the matrix $T$ defined in the text.
 \endinsert
 
-\beginsection{Regularisation of the truncated expansion problem}
+The problem of expanding a state over coherent states is similar to expanding a vector over an overcomplete basis in the plane.
+The main difference is that large sets of vectors in the plane are linearly dependent, while sets of coherent states are not.
+The two approaches are illustrated below, with an expansion of the vector marked $+$ along the indicated directions.
+The black lines show the expansion
+$$ \vec x = π^{-1}∑_{r=1}^R (\vec u_r\cdot\vec x)\vec u.$$
+As can be seen, this is a very good approximation even with five vectors; in fact, it is hard to visually distinguish the approximate point from $\vec x$ even with 3 directions.
+The red lines show $\vec x$ expanded over the two directions closest to it.
+The interesting thing is the size of the change of the expansion coefficients when $\vec x$ moves perpendicular to $\vec x$.
+
+\topinsert\XeTeXpicfile circle-1.pdf 
+\endinsert
+
+To compute in phase space, we need to reduce the function $g$ to a finite vector of numbers.  This is usually done in one of two ways: by a vector
+$$\vec g=\pmatrix{g(z₁) & ⋯ & g(z_R)},$$
+where the points $z_i$ form a regular grid, or by
+$$\vec g=\pmatrix{〈 u₁,g〉 & ⋯ & 〈 u_R,g〉},$$
+where the distributions $u_i$ are a fixed basis.
+The huge dimension of state space rules out any regular approach: we must sample $g$ at a set of irregularly spaced amplitudes or on an irregular basis, adapted to the support of $g$.
+
+However, in these single mode numerical experiments, it is often useful to expand over a regular grid.
+For example, this allows us to plot phase space functions.
+It is convenient to normalise the samples so that $‖\vec g‖=‖A\vec g‖$.
+If the sample amplitudes $α_r$ form a square gird with spacing $h$, the condition $\vec g_r=cg(α_r)$ is satisfied when
+$$∑_r|\vec g_r|²=c²∑_r|g(α_r)|²=\int |g|²\,d²α≈h²∑_r|g(α_r)|²,$$
+whence $c=h$, and the discretisation is
+$$\vec g_r=hg(α_r).$$
+
+This makes regularisation tricky.
+The goal is to represent a state $|ψ〉$ as a superposition of coherent states, $|ψ〉=|A〉 c$.
+We want the coefficients $c$ to correspond to a phase space function $g$ that is nearly analytic.
+If the amplitudes $|A〉$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
+On an irregular grid, we would expect the coefficients to be larger where the grid is sparse, and smaller where it is dense.
+It would be good to know if irregular coefficients are the result of sampling a smooth function on an irregular grid, or of sampling a wildly oscillating function.
+
+The obvious way to attempt this is to factor the coefficients as $|ψ〉=|A〉 Wc'$, where $c'_i$ are samples that approximate $g(α_i)$, and $W$ is a matrix of weights that adapts the samples to the irregular grid.
+In fact, we can do this exactly.
+If we set $W=(〈 A|A〉)^{-1}$, then $c'=〈 A|ψ〉$.
+We could try seeking coefficients $c$ such that the norm of $〈 A|A〉 c$ is minimal.
+
+To avoid the complexities of Wirtinger calculus, the problem of relating the weights of gaussians to the values of a sampled function can be set up for a function of a single real variable.
+It would be appealing for the weight matrix $W$ to be diagonal, so that the weight of the gaussian centred at each point was determined by the value of $f$ at that point.
+This can be set up either as a Galerkin condition, or as a collocation condition: they turn out to be equivalent.
+Unfortunately, the weights end up oscillating wildly.
+This approach doesn't seem useful, but is presented for completeness, and because the phenomenon of adjacent weights cancelling keeps occuring, this being the simplest problem in which I've seen it.
+
+As a collocation problem, consider the case where $f(x)=1$, and we wish to approximate $f$ by a function 
+$$f'(x)=∑_i e^{-(x-x_i)²}w_ic_i$$
+that satisfies the collocation condition $f'(x_i)=f(x_i)$.  The goal is for $c_i=1$, so we need ...
+
+$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
+	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
+	& & \ddots & \cr
+	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
+\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
+= \pmatrix{π\cr π\cr\vdots\cr π}.$$
+
+The Galerkin condition seeks a set of weights $w$ such that, for the normalised functions $u_i=π^{-½}e^{-(x-x_i)²}$, we have
+$$1=\int u_i=∑_j w_j u_i(x_j)=∑_j w_je^{-(x_j-x_i)²}.$$
+This is the same condition as the collocation condition.  The results are shown below.
+
+Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
+It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
+The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.
+This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
+
+There is another way to derive coefficients of the constant function, which might be more stable.
+Suppose that $D$ is a subset of the real line, large enough that the tails of gaussians are neglibible outside it.
+Let
+$$U=π^{-½}\pmatrix{e^{-(x-x₁)²}&⋯&e^{-(x-x_R)²}},$$
+and $w$ be a column vector of $R$ real weights.  Then define a residual
+$$r(w)=\int_D(Uw-1)².$$
+The derivative of this with respect to the weights is
+$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U^{\rm T}Uw-U.$$
+(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c^{\rm T}ab$.)
+Unfortunately,
+$$\int (U^{\rm T}U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
+	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
+and $\int u_i=1$, so this is the identical problem again.
+
+An obvious way would be to assign amplitudes to the nearest sample.
+That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
+What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
+There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
+Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
+I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
+What's the best exponent?
+I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+
+%% Sampling on random grids
+
+\topinsert \XeTeXpicfile gweight.pdf width \hsize 
+The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
+The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
+Note how nearby points tend to cancel each others' samples.
+I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
+
+The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
+This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
+Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.
+Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.
+When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.
+This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
+\endinsert
+
+The goal of fiding samples of the phase space distribution $f(α_r)$ is to regularise the expansion and inverse problems.
+The expansion function with the least norm is guaranteed to be a nice function: we don't know much about the finite superposition whose weights have the least norm.
+
+To the the norm of the function $f$, sampled on an irregular grid $α_r$, the obvious method is to interpolate.
+Doing this might help with irregular weights due to irregular sample points, because a least-squares interpolant will average the weights at several adjacent points.
+However, there is probably a better way to adjust the weights for irregular sampling.
+
+We want to interpolate a function of a complex variable, that should be nearly analytic.  To me, the obvious way to do it is a least squares fit to the monomials
+$$U=e^{-½|z|²}\pmatrix{1 & z & z* & z² & zz* & z^{2\ast} & z³ & z²z* & ⋯}.$$
+It helps to think of $z^mz^{n\ast}=r^{m+n}e^{i(m-n)φ}$.
+Note that if these monomials are orthogonalised by Gramm-Schmidt, in the order they are written, the functions $e^{-½|z|²}\pmatrix{1 & z & z² & z³ &  ⋯}$ are retained as part of the orthonormal set: $z^n$ is the first monomial proportional to $e^{inφ}$, so it is orthogonal to all those before it.
+The same is true for $z^{n\ast}$, the first monomial proportional to $e^{-inφ}$.
+So this basis is an obvious one for projecting onto analytic functions in $z$ or $z*$.
+It will have the form
+$$e^{-½|z|²}\pmatrix{1 & re^{iφ} & re^{-iφ} & r²e^{2iφ} & p_{11}(r) & r²e^{-2iφ} & r³e^{3iφ} & p_{12}(r)e^{iφ} & p_{21}(r)e^{-iφ}& ⋯},$$
+where $p_{ij}$ is a polynomial of degree $i+j$.  Someone must have investigated these orthogonal polynomials before, but they don't appear in Courant \& Hilbert or in Abramowitz \& Stegun.
+
+For least squares fitting, orthogonal polynomials are not necessary: the QR factorisation does that work for us.
+To find norms, we do need to know the grammian matrix, whose elements are the Bargmann inner products of the monomials:
+$$\eqalign{a_{mnpq}&=\int (z^{m\ast}zⁿ)*(z^{p\ast}z^q)e^{-|z|²}\,d²z \cr
+	&= 2πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q+1}\,dr \cr
+	&= πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q}\,2r\,dr \cr
+	&= πδ_{m+n,p+q}Γ({m+n+p+q\over 2}+1)=π(n+p)!δ_{m+n,p+q}.}$$
+Some fits of these to expansions of cat and coherent states over random amplitudes are shown below.
+They suffer the usual problem that polynomial fits oscillate more wildly as the degree of the polynomial increases; there are meant to be ways to avoid that, but I'm not an expert in them.
+
+\beginsection{Regularisation of the discrete expansion problem}
 %% Picard plots
@@
 The difference between these components is
-$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖²} - {c_u〈v|u〉\over ‖|v〉‖²}\right|.$$
+$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖} - {c_u〈v|u〉\over ‖|v〉‖}\right|.$$
 These differences form a tension matrix.
@@
 The tension matrix is a reasonable measure of the smoothness of an expansion: its Frobenius norm, the sum in quadrature of the components, would be a good summary.
+
+The idea is that a relaxed expansion has all the coefficients pulling in the same direction, but an expansion with large tension is a tug-of-war, with adjacent coefficients pulling the expanded state in opposite ways.
+
+The tensions were calculated for the grid and irregular expansions above, and plotted in the second graph.
+Despite the much worse convergence of random sampling, the weights have very nearly the same tension as the regular grid.
 
@@
 I didn't forsee that, but it isn't too surprising: we can change the coefficients of the superposition in any direction we like, regardless of their current values, and the derivative is only sensitive to changes.
-
-\beginsection{Discrete phase space}
-
-To compute in phase space, we need to reduce the function $g$ to a finite vector of numbers.  This is usually done in one of two ways: by a vector
-$$\vec g=\pmatrix{g(z₁) & ⋯ & g(z_R)},$$
-where the points $z_i$ form a regular grid, or by
-$$\vec g=\pmatrix{〈 u₁,g〉 & ⋯ & 〈 u_R,g〉},$$
-where the distributions $u_i$ are a fixed basis.
-The huge dimension of state space rules out any regular approach: we must sample $g$ at a set of irregularly spaced amplitudes or on an irregular basis, adapted to the support of $g$.
-
-However, in these single mode numerical experiments, it is often useful to expand over a regular grid.
-For example, this allows us to plot phase space functions.
-It is convenient to normalise the samples so that $‖\vec g‖=‖A\vec g‖$.
-If the sample amplitudes $α_r$ form a square gird with spacing $h$, the condition $\vec g_r=cg(α_r)$ is satisfied when
-$$∑_r|\vec g_r|²=c²∑_r|g(α_r)|²=\int |g|²\,d²α≈h²∑_r|g(α_r)|²,$$
-whence $c=h$, and the discretisation is
-$$\vec g_r=hg(α_r).$$
-
-This makes regularisation tricky.
-The goal is to represent a state $|ψ〉$ as a superposition of coherent states, $|ψ〉=|A〉 c$.
-We want the coefficients $c$ to correspond to a phase space function $g$ that is nearly analytic.
-If the amplitudes $|A〉$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
-On an irregular grid, we would expect the coefficients to be larger where the grid is sparse, and smaller where it is dense.
-It would be good to know if irregular coefficients are the result of sampling a smooth function on an irregular grid, or of sampling a wildly oscillating function.
-
-The obvious way to attempt this is to factor the coefficients as $|ψ〉=|A〉 Wc'$, where $c'_i$ are samples that approximate $g(α_i)$, and $W$ is a matrix of weights that adapts the samples to the irregular grid.
-In fact, we can do this exactly.
-If we set $W=(〈 A|A〉)^{-1}$, then $c'=〈 A|ψ〉$.
-We could try seeking coefficients $c$ such that the norm of $〈 A|A〉 c$ is minimal.
-
-To avoid the complexities of Wirtinger calculus, the problem of relating the weights of gaussians to the values of a sampled function can be set up for a function of a single real variable.
-It would be appealing for the weight matrix $W$ to be diagonal, so that the weight of the gaussian centred at each point was determined by the value of $f$ at that point.
-This can be set up either as a Galerkin condition, or as a collocation condition: they turn out to be equivalent.
-Unfortunately, the weights end up oscillating wildly.
-This approach doesn't seem useful, but is presented for completeness, and because the phenomenon of adjacent weights cancelling keeps occuring, this being the simplest problem in which I've seen it.
-
-As a collocation problem, consider the case where $f(x)=1$, and we wish to approximate $f$ by a function 
-$$f'(x)=∑_i e^{-(x-x_i)²}w_ic_i$$
-that satisfies the collocation condition $f'(x_i)=f(x_i)$.  The goal is for $c_i=1$, so we need ...
-
-$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
-	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
-	& & \ddots & \cr
-	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
-\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
-= \pmatrix{π\cr π\cr\vdots\cr π}.$$
-
-The Galerkin condition seeks a set of weights $w$ such that, for the normalised functions $u_i=π^{-½}e^{-(x-x_i)²}$, we have
-$$1=\int u_i=∑_j w_j u_i(x_j)=∑_j w_je^{-(x_j-x_i)²}.$$
-This is the same condition as the collocation condition.  The results are shown below.
-
-Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
-It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
-The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.
-This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
-
-There is another way to derive coefficients of the constant function, which might be more stable.
-Suppose that $D$ is a subset of the real line, large enough that the tails of gaussians are neglibible outside it.
-Let
-$$U=π^{-½}\pmatrix{e^{-(x-x₁)²}&⋯&e^{-(x-x_R)²}},$$
-and $w$ be a column vector of $R$ real weights.  Then define a residual
-$$r(w)=\int_D(Uw-1)².$$
-The derivative of this with respect to the weights is
-$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U^{\rm T}Uw-U.$$
-(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c^{\rm T}ab$.)
-Unfortunately,
-$$\int (U^{\rm T}U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
-	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
-and $\int u_i=1$, so this is the identical problem again.
-
-An obvious way would be to assign amplitudes to the nearest sample.
-That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.
-What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?
-There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.
-Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.
-I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.
-What's the best exponent?
-I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
-
-%% Sampling on random grids
-
-\topinsert \XeTeXpicfile gweight.pdf width \hsize 
-The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
-The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
-Note how nearby points tend to cancel each others' samples.
-I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
-
-The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
-This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
-Similar things happen in the random ensemble, at the edges of a subgrid of closely spaced points.
-Nearly the same thing happens when sampling with coherent states: the extreme amplitudes need extra weight, which must be cancelled by the interior ones.
-When states are sampled on regular grids of coherent states, the outside amplitudes have very little weight in any case, and don't affect the solution much.
-This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
-\endinsert
-
-The goal of fiding samples of the phase space distribution $f(α_r)$ is to regularise the expansion and inverse problems.
-The expansion function with the least norm is guaranteed to be a nice function: we don't know much about the finite superposition whose weights have the least norm.
-
-To the the norm of the function $f$, sampled on an irregular grid $α_r$, the obvious method is to interpolate.
-Doing this might help with irregular weights due to irregular sample points, because a least-squares interpolant will average the weights at several adjacent points.
-However, there is probably a better way to adjust the weights for irregular sampling.
-
-We want to interpolate a function of a complex variable, that should be nearly analytic.  To me, the obvious way to do it is a least squares fit to the monomials
-$$U=e^{-½|z|²}\pmatrix{1 & z & z* & z² & zz* & z^{2\ast} & z³ & z²z* & ⋯}.$$
-It helps to think of $z^mz^{n\ast}=r^{m+n}e^{i(m-n)φ}$.
-Note that if these monomials are orthogonalised by Gramm-Schmidt, in the order they are written, the functions $e^{-½|z|²}\pmatrix{1 & z & z² & z³ &  ⋯}$ are retained as part of the orthonormal set: $z^n$ is the first monomial proportional to $e^{inφ}$, so it is orthogonal to all those before it.
-The same is true for $z^{n\ast}$, the first monomial proportional to $e^{-inφ}$.
-So this basis is an obvious one for projecting onto analytic functions in $z$ or $z*$.
-It will have the form
-$$e^{-½|z|²}\pmatrix{1 & re^{iφ} & re^{-iφ} & r²e^{2iφ} & p_{11}(r) & r²e^{-2iφ} & r³e^{3iφ} & p_{12}(r)e^{iφ} & p_{21}(r)e^{-iφ}& ⋯},$$
-where $p_{ij}$ is a polynomial of degree $i+j$.  Someone must have investigated these orthogonal polynomials before, but they don't appear in Courant \& Hilbert or in Abramowitz \& Stegun.
-
-For least squares fitting, orthogonal polynomials are not necessary: the QR factorisation does that work for us.
-To find norms, we do need to know the grammian matrix, whose elements are the Bargmann inner products of the monomials:
-$$\eqalign{a_{mnpq}&=\int (z^{m\ast}zⁿ)*(z^{p\ast}z^q)e^{-|z|²}\,d²z \cr
-	&= 2πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q+1}\,dr \cr
-	&= πδ_{m+n,p+q}\int_0^∞r^{m+n+p+q}\,2r\,dr \cr
-	&= πδ_{m+n,p+q}Γ({m+n+p+q\over 2}+1)=π(n+p)!δ_{m+n,p+q}.}$$
-Some fits of these to expansions of cat and coherent states over random amplitudes are shown below.
-They suffer the usual problem that polynomial fits oscillate more wildly as the degree of the polynomial increases; there are meant to be ways to avoid that, but I'm not an expert in them.
-
-%% Truncated Hilbert space
-
-I still haven't thought in detail about the results with truncated Fock spaces.
-Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.
-Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.
-So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.
-And in fact, by solving the inverse problem, we can always get a nice solution in any case.
-Numerical experiments suggest that the variational problem is quite well posed for a single step, too.
-
-Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.
-And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.
-
-As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.
-Presumably these converge to limits as $N$ increases.
-There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.
 

### 2013-08-30 paper.tex
writing

@@
 where $x$ and $p$ are the position and momentum operators for a particle with mass $m$.
-(The dimensions don't work: $λ²$ is an action, but then $a$ has dimensions.)  The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
+The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
 

### 2013-09-08 paper.tex
writing

@@
 
-There appear to be two sorts of instabilities.
-The first, static, sort occurs when a quantum state is expanded over a finite set of coherent states, and those states have irregular amplitudes.
-This is similar to the kind that numerical analysts call ill-posed problems, and the methods they use to analyse and resolve them can be applied here.
-However, the cause of ill-posedness in our problem differs from those usually studied, and the usual methods of resolution don't work as well as they usually do.
-There is one advantage of the expansion problem: because it is iterated at each timestep, the dynamics of the amplitudes can be exploited to so that they posed better expansion problems as the simulation progresses.
+The instability appears to have two major causes.
+The first problem occurs when quantum states are expanded over a finite set of coherent states.
+There is a natural way to expand a state as an integral over the coherent states, with a weight function that varies smoothly with the coherent amplitude.
+When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.
+Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number—for instance, it might be a coherent state.
+When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.
+This causes the expansion weights to be large, to not resemble samples of any smooth function, and to change sharply as the state changes.
+All these properties make it difficult to numerically integrate a rate of change that has been expanded by least squares.
+The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.
+
+The behaviour of the discrete expansion weights is similar to that observed in discrete ill-posed problems, and the methods used to analyse and resolve these can be applied to expansions over coherent states.
+However, the cause of ill-posedness in our problem differs from those usually studied—the large number components of the state are real, while the high-frequency components that play the same role in ill-posed problems are spurious ones introduced by errors in measured data.
+Therefore the standard methods of resolution don't work as well as usual.
+The expansion problem is easier in one way: it is iterated at each timestep, so the dynamics of the coherent amplitudes can be modified so that the expansion problem becomes better posed as the simulation progresses.
 
@@
 
+\beginsection{Variational dynamics}
+
+Schrödinger's equation, the quantum law of motion, is
+$$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
+I will discuss many details of mappings from complex numbers to quantum states, so some notation used in this paper is traditionally used more in mathematics than in physics.
+The symbol $|Ψ〉$ denotes a ket-valued function of a real number; $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.
+The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.
+Therefore it is written outside the ket.
+The total derivative $D$ acts on the function, so it is written inside the ket.
+Because $|Ψ〉$ is a function of a single variable, $|DΨ〉$ is simply another ket-valued function, and $|DΨ(t)〉$ is the ket that would traditionally be written ${∂\over ∂t}|Ψ〉\big|_{t=t}$.
+This could alternatively be written as a partial derivative $|DΨ〉=|∂₀Ψ〉$.
+If $|Ψ〉$ were a function of several variables, the total derivative function $|DΨ〉$ would evaluate to a row vector $|DΨ(t)〉=\pmatrix{|∂₀Ψ(t)〉&⋯&|∂_RΨ(t)〉}$.
+One can regard the kets as column vectors and read this as a Jacobian matrix with infinitely many rows and $R$ columns.
+The notation will be elaborated below as needed.
+
+Schrödinger's equation could even be abbreviated to
+$$i\hbar |DΨ〉=H|Ψ〉.$$
+This is a very simple equation.
+The challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand it over a grid grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes merely thousands of particles.
+In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
+$$|Ψ(t)〉=|ψ(z(t))〉,$$
+or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\left(z₁(t),…,z_R(t)\right)$ is a vector of complex parameters.
+This imposes two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.
+
+Some methods of computing dynamics, such as those involving Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.
+The above model does not apply to these.
+There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which can compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store all the parameters at once.
+
+The next sections will discuss the approximant $|ψ(z)〉$ analysed in this paper.
+Before that, I'll outline the method that will be used to compute the dynamical $z(t)$.
+The final part of the paper discusses the details—and there are many—of making this computation numerically stable.
+
+If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
+$$i\hbar|Dψ(z(t))〉Dz(t)=H|ψ(z(t))〉.$$
+On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.
+This is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to have components outside this subspace.
+This is a consequence of the restricted form of the approximation.
+We can, however, look for a vector $\dot z(z)$ that minimises the residual
+$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\eqno{(4)}$$
+and solve the ordinary differential equation
+$${dz\over dt}=\dot z(z),\tag\ode$$
+giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.
+When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.
+
+The Dirac-Frenkel solution is only optimised locally.
+There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkey solution $|ψ(w(t))〉$ does.
+This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.
+When we discuss regularisation below, we will see that this is in fact the usual case.
+However, global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.
+
+The residual above can be defined using any norm one wishes.
+The aim of computing quantum dynamics is usually to find some expectation values that could be compared to observed quantities, and one could consider using a norm that magnified the discrepencies in these observables, so that the solution with the least residual would correspond closely to the exact solution in the directions of interest.
+On the other hand, we are computing quantum dynamics: these are unitary in the usual norm, so that a small error introduced at an early time is guaranteed to stay small when the state is propagated to later times.
+No other norm guarantees this: they all introduce uncontrolled errors regarding how much early errors are magnified by later dynamics.
+
+Equation~\ode\ will be integrated numerically.
+For this to be done stably and efficiently, $\dot z$ must be a reasonably smooth function of $z$.
+This is formally expressed in terms of the eigenvalues of the Jacobian $D\dot z$; different integration formulae place different bounds on the magnitudes and phases of these, but they all have limited regions of stability.
+Moreover, as we will see below, a technically stable integration formula might still not produce a useful solution.
+
+This complicates the solution of the problem in Formula~4.
+It does not suffice to find a $\dot z$ with the smallest residual, for a given value of $z$.
+In order for Equation~5 to be numerically solvable, the properties of $\dot z$ under perturbation of $z$ are equally important.
+Under the usual Hilbert space norm, Formula~4 is a least squares problem; the pertubation properties of such problems are reasonably well understood.
+We will examine ways to modify the problem, by introducing extra least-squares constraints, that might improve its perturbation properties.
+
 \beginsection{The phase space expansion operator}
+
+The approximant studied in this paper has the form
+$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
+	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
+	=|A〉d,$$
+where $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
+In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.
+The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion $|A〉$.
+Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.
+This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.
 
@@
 
-\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
 
@@
 
-\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
-\XeTeXpicfile rsv4.pdf width 0.5\hsize}
-\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
-The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
+%\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
+%\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+%\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
+%The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
 These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert
@@
 
-\topinsert\XeTeXpicfile converge.pdf width \hsize
+%\topinsert\XeTeXpicfile converge.pdf width \hsize
 Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.
@@
 The tension is the Frobenius norm of the matrix $T$ defined in the text.
-\endinsert
+%\endinsert
 
@@
 
-\topinsert\XeTeXpicfile circle-1.pdf 
-\endinsert
+%\topinsert\XeTeXpicfile circle-1.pdf \endinsert
 
@@
 
-\topinsert \XeTeXpicfile gweight.pdf width \hsize 
+%\topinsert \XeTeXpicfile gweight.pdf width \hsize 
 The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
@@
 This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
-\endinsert
+%\endinsert
 
@@
 
-\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
+%\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
 
@@
 
-\topinsert\XeTeXpicfile cats.pdf width \hsize
-\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
+%\topinsert\XeTeXpicfile cats.pdf width \hsize
+%\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
 
@@
 
-\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
-Picard plot for cat expansion. \endinsert
+%\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
+%Picard plot for cat expansion. \endinsert
 
@@
 
-\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
+%\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
 
-\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
+%\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
 
@@
 
-\topinsert \XeTeXpicfile geometry-1.pdf
-\it A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
+%\topinsert \XeTeXpicfile geometry-1.pdf
+A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
 As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.
 This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
-\endinsert
+%\endinsert
 
-\topinsert \XeTeXpicfile cancel.pdf width \hsize
-\it The vacuum state was expanded over two coherent states, with amplitudes $±h$.
+%\topinsert \XeTeXpicfile cancel.pdf width \hsize
+The vacuum state was expanded over two coherent states, with amplitudes $±h$.
 The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.
@@
 But, given those sample amplitudes, that would be the only way to represent such a component.
-\endinsert
+%\endinsert
 
@@
 $$∂_rσ_R=x†(∂_rM)x
-	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA〉v_R\right).$$
+	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA'〉v_R\right).$$
 Substituting the partial derivatives gives
@@
 
-\topinsert 
-\vskip -40mm
-\XeTeXpicfile psfa.pdf width \hsize
-\vskip -90mm
-\XeTeXpicfile psv.pdf width \hsize\endinsert
+%\topinsert 
+%\vskip -40mm
+%\XeTeXpicfile psfa.pdf width \hsize
+%\vskip -90mm
+%\XeTeXpicfile psv.pdf width \hsize\endinsert
 
@@
 
-\centerline{\XeTeXpicfile j1.pdf width \hsize}
+%\centerline{\XeTeXpicfile j1.pdf width \hsize}
 
-\centerline{\XeTeXpicfile j2.pdf width \hsize}
+%\centerline{\XeTeXpicfile j2.pdf width \hsize}
 
@@
 
-\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile spsns.pdf width \hsize
-Amplitudes of components \endinsert
+%\topinsert \XeTeXpicfile spsns.pdf width \hsize
+%Amplitudes of components \endinsert
 
-\topinsert \XeTeXpicfile traj.pdf width \hsize
-Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
+%\topinsert \XeTeXpicfile traj.pdf width \hsize
+%Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
 
-\topinsert \XeTeXpicfile picard.pdf width \hsize
-Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
+%\topinsert \XeTeXpicfile picard.pdf width \hsize
+%Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
 
-\topinsert \XeTeXpicfile svnum.pdf width \hsize
-Expected particle number for left singular vectors.
+%\topinsert \XeTeXpicfile svnum.pdf width \hsize
+%Expected particle number for left singular vectors.
 The svs with small sws have large particle number.
@@
 
-\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
 
-\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
+%\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
 
@@
 
-\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
+%\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
 
-\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
+%\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
 

### 2013-09-09 paper.tex
writing

@@
 	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
-	=|A〉d,$$
+	=|A〉d,\tag\Ad$$
 where $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
@@
 
-On a computer, the expansion in phase space must be discretised.  The obvious way to do so is to replace the integral with a sum, so that
+As mentioned above, the computer must actually manipulate a discrete expansion with the form of Equation~\Ad,
 $$|ψ〉=d₁|α₁〉+⋯+d_R|α_R〉=|A〉d,$$
@@
 If the system has only a few modes, so that $α$ is a low-dimensional vector, the set of sampled amplitudes $A=\{α₁, …,α_R\}$ could be a regular grid; this makes many things easier.
-When there are many modes, and $α$ is a large vector, the storage required for grids becomes exponentially large, and the sample points will have to be irregular.
+When there are many modes, and $α$ is a large vector, the storage required for grids becomes exponentially large, and the sample points will have to be irregular, most likely drawn at random.
 
@@
 $$f_A(β)=π^½∑_rd_re^{½|α_r|²}δ(β-α_r).$$
-I don't know enough about generalised functions to be comfortable working with these forms; in particular, the norm of $f_A$ is a bit mysterious.
+I don't know enough about generalised functions to reliably manipulate these forms; in particular, the norm of $f_A$ is a bit mysterious.
 If this could be clarified, it might provide some useful insights, for example how the norm of $f_A$ compares to that of $〈Γ⁺|ψ〉$.
@@
 
-Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.
+Unfortunately, for random samples $A$, there tends not to be a solution where the weight matrix is diagonal: some of the $w_r²$ would have to be negative.
 It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.
@@
 This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
+So one solution  would be to find the best expansion of $e^{-x²}$, instead of 1, then adjust the weight for $x_i$ by either $e^{-x_i²}$ or $〈e^{-(x-x_i)²},e^{-x²}〉$.
 %\endinsert

### 2013-09-24 paper.tex
Hong Kong

@@
 
-Some methods of computing dynamics, such as those involving Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.
+Some methods of computing dynamics, such as those involving single-particle Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.
 The above model does not apply to these.
@@
 Before that, I'll outline the method that will be used to compute the dynamical $z(t)$.
-The final part of the paper discusses the details—and there are many—of making this computation numerically stable.
+The final part of the paper discusses the many details of making this computation numerically stable.
 
@@
 When we discuss regularisation below, we will see that this is in fact the usual case.
-However, global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.
+Global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.
 
@@
 The bosons in this case are vibrational quanta in a notional harmonic potential, centred at the origin, with frequency $ω=λ²/m$.
+The coherent state $|α〉$ is a gaussian wave packet, with central position \dots and momentum \dots
 
 Coherent states are not orthogonal, but their brackets follow the law
-$$〈β|α〉=e^{-½|β|²-½|α|²+β*α},$$
+$$〈β|α〉=e^{-½|β|²-½|α|²+β*α}=e^{-½|β-α|²+i\Im α*β},\tag\csip$$
 whence
@@
 The projector satisfies
-$$\int |α〉〈α|\,d²α=π.$$
-This is similar to the law for an orthonormal basis, except that the right hand side is $π$ instead of 1.
+$$π^{-1}\int |α〉〈α|\,d²α=1.$$
+This is similar to the law for an orthonormal basis, except that the left hand side must be scaled by $π$.  Therefore, any state can be expanded as a sum over coherent states, in the form
+$$|ψ〉=π^{-1}\int |α〉〈α|ψ〉\,d²α.\tag\csproj$$
+However, this expansion is not unique: a coherent state $|β〉$ has the trivial expansion $|β〉=|α〉$, but the expansion given by Equations~\csproj\ and~\csip\ is a gaussian function of $α$.
+A set of states like this is known as a frame, in distinction from a basis, whose elements are linearly independent, and whose expansions are unique.
 Further properties of the coherent states will be stated as required.
@@
 And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
-So it is curious that the discrete version of this problem has caused so much trouble.
+So it is curious that the discrete version of this problem has caused so much trouble when solved by least squares.
 
@@
 
-The operator $|Γ〉$ acts on functions defined over the whole complex plane.
-For states with high particle number, the smallest preimages have support far from the origin.
-We can not compute with such functions: any discretisation will sample values of $f$ on a finite subset of the plane.
-Unfortunately, the conditioning of these restricted operators will be worse than that of $|Γ〉$.
-We have excluded the small preimages of the states with large particle number, so they must be represented by functions with support close to the origin, but with large norms.
-Hence the restricted operators will be badly conditioned.
+The operator $\G $ represents kets by phase space distributions, functions defined on the whole complex plane.
+When the ket $|ψ〉$ has many particles, its smallest representation, the distribution $〈Γ⁺|ψ〉$, is concentrated far from the origin.
+Such kets will occur when interacting particles are simulated, the interaction energy is proportional to the square of the number of particles, and the Hamiltonian operator amplifies many-particle components of the kets on which it acts.
+In order to represent kets on a computer, the distribution functions must be sampled on a finite region of the complex plane.
+If $〈Γ⁺|ψ〉$ extends beyond the sampled region, the ket can still be represented by another distribution, because $\G $ is not one-to-one.
+However, this distribution will take larger values and vary more rapidly.
+Therefore, if the simulation is to be numerically stable, it will have to be sampled carefully.
 
@@
 
-The above paragraph can be stated quantitatively.  Define a restricted operator
+This can be made quantitative.  Define a restricted operator
 $$\Gx f=π^{-½}\int |α〉e^{-½|α|²}χ(α*)f(α*)\,d²α.$$
-Most obviously, $χ$ could be a characteristic function, with $χ(z)$ being 1 for $|z|≤b$ and 0 outside this disk.  The formalism allows gradual truncation too.
-
-The sizes of states and their expansions can be read from the singular value expansion of $\Gx $.
-If the domain and range of $\G$ were the same space, it would have an eigenvalue decomposition; the SVE is a generalisation of that.
-It has the form
+For example, $χ$ could be a characteristic function, with $χ(z)$ being 1 for $|z|≤b$ and 0 outside this disk.
+For functions $f$ with support outside this disk, part of $f$ will be multiplied by zero, and the norm of $\Gx f$ will be reduced when compared to the norm of $f$.
+This reduction would cause $\Gx$ to have some small eigenvalues, if it mapped functions onto functions or kets onto kets.
+Because its domain and range are different spaces, it is necessary to use a generalisation, the singular values.
+
+The idea of an eigenvalue expansion is to expand vectors over a carefully chosen basis, so that the effect of an operator is merely to rescale the components.
+For the operator $\Gx$, there will need to be two bases, one in function space and one in ket space.
+If it is required that the bases be orthonormal, and that $\Gx$ has the effect of rescaling the coefficients of vectors and replacing the function basis with the ket basis, the basis vectors are unique under similar conditions to those needed for uniqueness of eigenvalues.
+The scaling factors are known as singular values; the singular value expansion of $\Gx $ has the form
 $$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
@@
 
-In the case where $χ(z)=1$, and $\Gx=\G$, the expansion is quite arbitrary.
-The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire function for which $\G v_i=|u_i〉$, and every singular value is 1.
-If the domain and range of $\G$ were the same, this would indicate that it was a projector.
+In the case where $χ(z)=1$, and $\Gx=\G$, all the singular values are 1, and the expansion is arbitrary.
+The left singular kets $|u_i〉$ can be any orthonormal basis, and the right singular vectors are  $v_i=〈Γ⁺|u_i〉$.
+This indicates that $\G$ is an incomplete projector.
 The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component retains its expansion coefficients as the basis $v_i$ is replaced by $|u_i〉$.
@@
 
-As mentioned above, the computer must actually manipulate a discrete expansion with the form of Equation~\Ad,
-$$|ψ〉=d₁|α₁〉+⋯+d_R|α_R〉=|A〉d,$$
-where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $d$ is a column vector of the coefficients.
-In this paper, $c$ will usually denote the coefficients of expansion over orthogonal bases, and $d$ will represent coefficients of nonorthogonal expansions.
-If the system has only a few modes, so that $α$ is a low-dimensional vector, the set of sampled amplitudes $A=\{α₁, …,α_R\}$ could be a regular grid; this makes many things easier.
-When there are many modes, and $α$ is a large vector, the storage required for grids becomes exponentially large, and the sample points will have to be irregular, most likely drawn at random.
+As mentioned above, the computer manipulates a discrete expansion with the form of Equation~\Ad,
+$$|ψ〉=d₁|α₁〉+⋯+d_R|α_R〉=|A〉d.$$
+This kind of expansion has been studied for a long time.
+An early question was how dense the amplitudes $α_r$ must be in order for the set to span all kets, and $|A〉$ to be a frame.
+In the case that the amplidues are a rectangular grid, with $α_r=ma+inb$ Von~Neumann conjectured in 193?
+that the condition is one amplitude per area $π$ of phase space, which corresponds to an area $\hbar$ of the $x-p$ plane.
+This was proved in 19?.
+The conditioning of expansions as the density of sampled amplitudes approaches this limit appears to be an open question.
+Another open question is how much this limit can be relaxed if squeezed states are included in the expansion.
+
+In recent decades, these expansions have been studied intensely by engineers involved in signal processing, who call the set of coherent states a Moyal~frame.
+These researchers regard gaussian wavepackets as a kind of time-limited Fourier transform, with a gaussian window function, following on from the work of Hamming and others at Bell Labs.
+Moyal frames were investigated by the same community that developed wavelets; the latter became of very wide interest, and obscured the earlier work on Moyal frames.
+
+If the system has only a few modes, so that $α$ is a low-dimensional vector, the set of sampled amplitudes $A=\{α₁, …,α_R\}$ could be a regular grid, with $α_r=(m+in)h$, $m$ and $n$ being integers.
+This makes many things easier.
+In particular, it is straightforward to relate the weights $d$ to the values of a continuous distribution $f$ at the points $α_r$.
+It is convenient to normalise so that $‖d‖=‖f‖$.
+If the $α_r$ form a grid with spacing $h$, as specified above, and we set $d_r=cf(α_r)$, $c$ being constant, this normalisation condition is satisfied when
+$$∑_r|d_r|²=c²∑_r|f(α_r)|²=\int |f|²\,d²α≈h²∑_r|f(α_r)|².$$
+So the proportionality constant is $c=h$, and the discretisation is
+$$d_r=hf(α_r).$$
+
+Relating $d$ to a continuous function is harder when the amplitudes $A$ are irregular.
+The amplitudes will not generally be drawn from a uniform distribution—if they were, we would no doubt be better off using a sparse grid—and if amplitudes distributed with probability $P(α)$ are assigned weights proportional to $d_r=f(α_r)$, then, in the limit of many samples, the state $|ψ〉$ will converge to a multiple of $|Γ〉Pf$, with the continuous distribution $f$ scaled by the distribution $P$ of the sample amplitudes.
+This convergence will be slow, presumably with the square root of $R$, and we might hope to speed it up by compensating for the random spacing of the sample amplitudes.
+If many coherent states overlap $|α_r〉$, then values close to $f(α_r)$ will be counted many times in the sum of Equation~\Ad.
+In order for the sampling to remain even, the ratio of $d_r$ to $f(α_r)$ ought to depend on the density of $A$ around $α_r$, with smaller $d_r$ where the amplitude samples turn out to be dense.
+
+The normal way to discretise a linear operator such as $|Γ〉$ is to select a set of functions $u_r$ over which to expand $f$, and a set of kets $|v_r〉$ over which to expand $|ψ〉$.
+Given a set of sample amplitudes $A$, the only obvious choice for $|v_r〉$ is the set $|A〉$.
+There are two apparent choices for $u_r$: delta functions $δ(α-α_r)$, or the gaussian functions $〈Γ⁺|α_r〉$.
+Operators are usually discretised by the Galerkin condition, where $f$ is approximated by $Ub$, the orthogonal projection of $f$ onto the span of the $u_r$, the discrete approximation to $|ψ〉$ is $|V〉d=|V〉Tb$, for some matrix $T$, and it is required that $|V〉d$ is the orthogonal projection of $|Γ〉Ub$ onto the span of the kets $|v_r〉$.
+If we choose delta functions, the projections are trivial: we're sampling $f$ at the points $α_r$, and adding up coherent states with the resultant weights.
+The expansion of $f$ over Gaussians is more complicated: we are effectively setting $|ψ〉$ to the projection of $|Γ〉f$ onto the span of the $|α_r〉$.
+So it seems that orthogonal projections onto sets of Gaussian functions are an essential part of this discretisation.
+
+There are two usual ways to relate a continuous function $f$ to a discrete set of samples $d$.  This is usually done in one of two ways: by a vector
+$$\vec g=\pmatrix{g(z₁) & ⋯ & g(z_R)},$$
+where the points $z_i$ form a regular grid, or by
+$$\vec g=\pmatrix{〈 u₁,g〉 & ⋯ & 〈 u_R,g〉},$$
+where the distributions $u_i$ are a fixed basis.
+The huge dimension of state space rules out any regular approach: we must sample $g$ at a set of irregularly spaced amplitudes or on an irregular basis, adapted to the support of $g$.
+
+\hrule
 
@@
 
-To compute in phase space, we need to reduce the function $g$ to a finite vector of numbers.  This is usually done in one of two ways: by a vector
-$$\vec g=\pmatrix{g(z₁) & ⋯ & g(z_R)},$$
-where the points $z_i$ form a regular grid, or by
-$$\vec g=\pmatrix{〈 u₁,g〉 & ⋯ & 〈 u_R,g〉},$$
-where the distributions $u_i$ are a fixed basis.
-The huge dimension of state space rules out any regular approach: we must sample $g$ at a set of irregularly spaced amplitudes or on an irregular basis, adapted to the support of $g$.
-
-However, in these single mode numerical experiments, it is often useful to expand over a regular grid.
-For example, this allows us to plot phase space functions.
-It is convenient to normalise the samples so that $‖\vec g‖=‖A\vec g‖$.
-If the sample amplitudes $α_r$ form a square gird with spacing $h$, the condition $\vec g_r=cg(α_r)$ is satisfied when
-$$∑_r|\vec g_r|²=c²∑_r|g(α_r)|²=\int |g|²\,d²α≈h²∑_r|g(α_r)|²,$$
-whence $c=h$, and the discretisation is
-$$\vec g_r=hg(α_r).$$
-
-This makes regularisation tricky.
-The goal is to represent a state $|ψ〉$ as a superposition of coherent states, $|ψ〉=|A〉 c$.
-We want the coefficients $c$ to correspond to a phase space function $g$ that is nearly analytic.
-If the amplitudes $|A〉$ formed a regular grid , it would be obvious enough when $c$ was sampling an analytic function.
-On an irregular grid, we would expect the coefficients to be larger where the grid is sparse, and smaller where it is dense.
-It would be good to know if irregular coefficients are the result of sampling a smooth function on an irregular grid, or of sampling a wildly oscillating function.
+\hrule
 
@@
 The Galerkin condition seeks a set of weights $w$ such that, for the normalised functions $u_i=π^{-½}e^{-(x-x_i)²}$, we have
-$$1=\int u_i=∑_j w_j u_i(x_j)=∑_j w_je^{-(x_j-x_i)²}.$$
+$$1=\int f\cdot u_i=∑_j w_j u_i(x_j)=∑_j w_je^{-(x_j-x_i)²}.$$
 This is the same condition as the collocation condition.  The results are shown below.
@@
 I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.
+
+To prevent the weights from oscillating in order to fit the edges of the samples, it is tempting to do a weighted least squares fit, which emphasises fitting the centre of the cluster over fitting the edges.
+There are two ways to do this.
+Firstly, instead of expanding the constant function $f(x)=1$, we can expand a peaked function such as $f(x)=e^{-x²}$, then divide the coefficient of $e^{-(x-x_r)²}$ by $e^{-(x_r)²}$ to get a weight.
+This mimics the expansions we need to do in practice, where the sample points cover the region where the expanded function has large values.
+The residual in this case is
+$${\cal E} = \int\left(e^{-x²}-∑_rd_re^{(x-x_r)²}\right)²;$$
+the only change to the normal equations is that the right hand side $\pmatrix{1&⋯&1}^{\rm T}$ is replaced by a constant multiple of 
+$$b=\pmatrix{e^{-x₁²/2}&⋯&e^{-x_R²/2}}.\tag\rhs$$
+For now, I'm ignoring factors of 2, $\sqrt π $ and so on: how the weights oscillate is more interesting than their size.
+This seems to work: with random sample points, the weights still oscillate, but remain within an order of magnitude or two, where the weights from expanding the constant function vary by a factor of a million.
+
+The other approach is a weighted least squares fit of the constant function, where the residual is 
+$${\cal E} = \int e^{-x²}\left(1-∑_rd_re^{(x-x_r)²}\right)².$$
+The right hand side of the normal equations remains that of Equation~\rhs, and the matrix becomes
+$$a_{ij}=e^{-(x_i²+x_j²-6x_ix_j)/4}.$$
+(I'm still ignoring constants of order 1.)  Surprisingly, this doesn't seem to work any better than the unweighted problem.
 

### 2013-09-24 weights.m

@@ -1,0 +1,14 @@
+% Find weights for the best fit to exp(-x^2), as a sum of gaussians centred
+% on x_r. 
+
+R = 10;
+x = linspace(-1,1,R).';
+[rw, cl] = meshgrid(x);
+A = exp(-0.5*(rw-cl).^2);
+b = exp(-0.5*x.^2);
+w0 = A \ ones(size(x));
+v1 = A \ b;  w1 = v1./b;
+AA = exp(-0.25*(rw.^2+cl.^2-6*rw.*cl));
+bb = exp(-0.5*x.^2);
+w2 = AA \ bb;
+[x w0 w1 w2]

### 2013-09-30 paper.tex
writing

--- 2013-09-08 paper.tex
@@
 
-The instability appears to have two major causes.
-The first problem occurs when quantum states are expanded over a finite set of coherent states.
-There is a natural way to expand a state as an integral over the coherent states, with a weight function that varies smoothly with the coherent amplitude.
-When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.
-Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number—for instance, it might be a coherent state.
-When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.
-This causes the expansion weights to be large, to not resemble samples of any smooth function, and to change sharply as the state changes.
-All these properties make it difficult to numerically integrate a rate of change that has been expanded by least squares.
-The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.
-
-The behaviour of the discrete expansion weights is similar to that observed in discrete ill-posed problems, and the methods used to analyse and resolve these can be applied to expansions over coherent states.
-However, the cause of ill-posedness in our problem differs from those usually studied—the large number components of the state are real, while the high-frequency components that play the same role in ill-posed problems are spurious ones introduced by errors in measured data.
-Therefore the standard methods of resolution don't work as well as usual.
-The expansion problem is easier in one way: it is iterated at each timestep, so the dynamics of the coherent amplitudes can be modified so that the expansion problem becomes better posed as the simulation progresses.
+There appear to be two sorts of instabilities.
+The first, static, sort occurs when a quantum state is expanded over a finite set of coherent states, and those states have irregular amplitudes.
+This is similar to the kind that numerical analysts call ill-posed problems, and the methods they use to analyse and resolve them can be applied here.
+However, the cause of ill-posedness in our problem differs from those usually studied, and the usual methods of resolution don't work as well as they usually do.
+There is one advantage of the expansion problem: because it is iterated at each timestep, the dynamics of the amplitudes can be exploited to so that they posed better expansion problems as the simulation progresses.
 
@@
 
-\beginsection{Variational dynamics}
-
-Schrödinger's equation, the quantum law of motion, is
-$$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
-I will discuss many details of mappings from complex numbers to quantum states, so some notation used in this paper is traditionally used more in mathematics than in physics.
-The symbol $|Ψ〉$ denotes a ket-valued function of a real number; $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.
-The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.
-Therefore it is written outside the ket.
-The total derivative $D$ acts on the function, so it is written inside the ket.
-Because $|Ψ〉$ is a function of a single variable, $|DΨ〉$ is simply another ket-valued function, and $|DΨ(t)〉$ is the ket that would traditionally be written ${∂\over ∂t}|Ψ〉\big|_{t=t}$.
-This could alternatively be written as a partial derivative $|DΨ〉=|∂₀Ψ〉$.
-If $|Ψ〉$ were a function of several variables, the total derivative function $|DΨ〉$ would evaluate to a row vector $|DΨ(t)〉=\pmatrix{|∂₀Ψ(t)〉&⋯&|∂_RΨ(t)〉}$.
-One can regard the kets as column vectors and read this as a Jacobian matrix with infinitely many rows and $R$ columns.
-The notation will be elaborated below as needed.
-
-Schrödinger's equation could even be abbreviated to
-$$i\hbar |DΨ〉=H|Ψ〉.$$
-This is a very simple equation.
-The challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand it over a grid grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes merely thousands of particles.
-In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
-$$|Ψ(t)〉=|ψ(z(t))〉,$$
-or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\left(z₁(t),…,z_R(t)\right)$ is a vector of complex parameters.
-This imposes two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.
-
-Some methods of computing dynamics, such as those involving Green's functions, derive the expectation values of certain observables without storing data from which the quantum state of the system could be derived.
-The above model does not apply to these.
-There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which can compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store all the parameters at once.
-
-The next sections will discuss the approximant $|ψ(z)〉$ analysed in this paper.
-Before that, I'll outline the method that will be used to compute the dynamical $z(t)$.
-The final part of the paper discusses the details—and there are many—of making this computation numerically stable.
-
-If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
-$$i\hbar|Dψ(z(t))〉Dz(t)=H|ψ(z(t))〉.$$
-On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.
-This is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to have components outside this subspace.
-This is a consequence of the restricted form of the approximation.
-We can, however, look for a vector $\dot z(z)$ that minimises the residual
-$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\eqno{(4)}$$
-and solve the ordinary differential equation
-$${dz\over dt}=\dot z(z),\tag\ode$$
-giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.
-When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.
-
-The Dirac-Frenkel solution is only optimised locally.
-There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkey solution $|ψ(w(t))〉$ does.
-This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.
-When we discuss regularisation below, we will see that this is in fact the usual case.
-However, global optimisation of $z$ is not practical: even computing the local approximation stretches computers to their limits.
-
-The residual above can be defined using any norm one wishes.
-The aim of computing quantum dynamics is usually to find some expectation values that could be compared to observed quantities, and one could consider using a norm that magnified the discrepencies in these observables, so that the solution with the least residual would correspond closely to the exact solution in the directions of interest.
-On the other hand, we are computing quantum dynamics: these are unitary in the usual norm, so that a small error introduced at an early time is guaranteed to stay small when the state is propagated to later times.
-No other norm guarantees this: they all introduce uncontrolled errors regarding how much early errors are magnified by later dynamics.
-
-Equation~\ode\ will be integrated numerically.
-For this to be done stably and efficiently, $\dot z$ must be a reasonably smooth function of $z$.
-This is formally expressed in terms of the eigenvalues of the Jacobian $D\dot z$; different integration formulae place different bounds on the magnitudes and phases of these, but they all have limited regions of stability.
-Moreover, as we will see below, a technically stable integration formula might still not produce a useful solution.
-
-This complicates the solution of the problem in Formula~4.
-It does not suffice to find a $\dot z$ with the smallest residual, for a given value of $z$.
-In order for Equation~5 to be numerically solvable, the properties of $\dot z$ under perturbation of $z$ are equally important.
-Under the usual Hilbert space norm, Formula~4 is a least squares problem; the pertubation properties of such problems are reasonably well understood.
-We will examine ways to modify the problem, by introducing extra least-squares constraints, that might improve its perturbation properties.
-
 \beginsection{The phase space expansion operator}
-
-The approximant studied in this paper has the form
-$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
-	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
-	=|A〉d,$$
-where $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
-In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.
-The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion $|A〉$.
-Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.
-This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.
 
@@
 
-%\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
 
@@
 
-%\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
 
@@
 
-%\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
-%\XeTeXpicfile rsv4.pdf width 0.5\hsize}
-%\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
-%The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
+\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
+\XeTeXpicfile rsv4.pdf width 0.5\hsize}
+\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
+The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
 These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert
@@
 With these directions, we can form some of the smaller singular vectors of the full problem.
+
+\beginsection{Gabor frames}
 
@@
 
-%\topinsert\XeTeXpicfile converge.pdf width \hsize
+\topinsert\XeTeXpicfile converge.pdf width \hsize
 Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.
@@
 The tension is the Frobenius norm of the matrix $T$ defined in the text.
-%\endinsert
+\endinsert
 
@@
 
-%\topinsert\XeTeXpicfile circle-1.pdf \endinsert
+\topinsert\XeTeXpicfile circle-1.pdf 
+\endinsert
 
@@
 
-%\topinsert \XeTeXpicfile gweight.pdf width \hsize 
+\topinsert \XeTeXpicfile gweight.pdf width \hsize 
 The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
@@
 This only happens when the cloud of sampling amplitudes extends beyond the support of the distribution, and the outer amplitudes have very small weights.
-%\endinsert
+\endinsert
 
@@
 
-%\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
+\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
 
@@
 
-%\topinsert\XeTeXpicfile cats.pdf width \hsize
-%\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
+\topinsert\XeTeXpicfile cats.pdf width \hsize
+\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
 
@@
 
-%\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
-%Picard plot for cat expansion. \endinsert
+\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
+Picard plot for cat expansion. \endinsert
 
@@
 
-%\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
+\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
 
-%\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
+\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
 
@@
 
-%\topinsert \XeTeXpicfile geometry-1.pdf
-A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
+\topinsert \XeTeXpicfile geometry-1.pdf
+\it A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
 As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.
 This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
-%\endinsert
+\endinsert
 
-%\topinsert \XeTeXpicfile cancel.pdf width \hsize
-The vacuum state was expanded over two coherent states, with amplitudes $±h$.
+\topinsert \XeTeXpicfile cancel.pdf width \hsize
+\it The vacuum state was expanded over two coherent states, with amplitudes $±h$.
 The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.
@@
 But, given those sample amplitudes, that would be the only way to represent such a component.
-%\endinsert
+\endinsert
 
@@
 $$∂_rσ_R=x†(∂_rM)x
-	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA'〉v_R\right).$$
+	=½\pmatrix{〈u_R| & v†_R}\pmatrix{0 & |∂_rA'(x,y)〉\cr 〈∂_rA'(x,y)| & 0}\pmatrix{|u_R〉 \cr v_R}=\Re\left(〈u_R|∂_rA〉v_R\right).$$
 Substituting the partial derivatives gives
@@
 
-%\topinsert 
-%\vskip -40mm
-%\XeTeXpicfile psfa.pdf width \hsize
-%\vskip -90mm
-%\XeTeXpicfile psv.pdf width \hsize\endinsert
+\topinsert 
+\vskip -40mm
+\XeTeXpicfile psfa.pdf width \hsize
+\vskip -90mm
+\XeTeXpicfile psv.pdf width \hsize\endinsert
 
@@
 
-%\centerline{\XeTeXpicfile j1.pdf width \hsize}
+\centerline{\XeTeXpicfile j1.pdf width \hsize}
 
-%\centerline{\XeTeXpicfile j2.pdf width \hsize}
+\centerline{\XeTeXpicfile j2.pdf width \hsize}
 
@@
 
-%\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
 
@@
 
-%\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
 
@@
 
-%\topinsert \XeTeXpicfile spsns.pdf width \hsize
-%Amplitudes of components \endinsert
+\topinsert \XeTeXpicfile spsns.pdf width \hsize
+Amplitudes of components \endinsert
 
-%\topinsert \XeTeXpicfile traj.pdf width \hsize
-%Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
+\topinsert \XeTeXpicfile traj.pdf width \hsize
+Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
 
-%\topinsert \XeTeXpicfile picard.pdf width \hsize
-%Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
+\topinsert \XeTeXpicfile picard.pdf width \hsize
+Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
 
-%\topinsert \XeTeXpicfile svnum.pdf width \hsize
-%Expected particle number for left singular vectors.
+\topinsert \XeTeXpicfile svnum.pdf width \hsize
+Expected particle number for left singular vectors.
 The svs with small sws have large particle number.
@@
 
-%\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
 
-%\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
 
@@
 
-%\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
+\centerline{\XeTeXpicfile flwc2n10.pdf width \hsize}
 
-%\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
+\centerline{\XeTeXpicfile flwc2n5.pdf width \hsize}
 

### 2013-10-03 dual.tex
Started new dual paper

\input a4
\input xpmath
\input respnotes \tenrm

\title{The dual frame of coherent states with random amplitudes}

\midinsert\XeTeXpicfile approx.pdf height \hsize rotated 90 \endinsert

Let $\{α₁,…,α_R\}$ be a set of $R$ complex amplitudes, distributed irregularly.  An example is plotted in the top line overleaf, with $R=15$, and the amplitudes drawn from a gaussian distribution with $|α|_{\rm rms}=2$.  We aim to approximate kets with linear combinations of the coherent states $|α₁〉$ through $|α_R〉$, in the form $|ψ〉≈|A〉d$, where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $d$ is a column of complex weights.  Presumably, not all kets can be approximated this way: it would be asking too much to approximate a Fock state $|n=100〉$ in terms of coherent states with amplitudes 1 and 2.

The methods of approximation we will consider set $d$ to a linear function of $|ψ〉$.  We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.  We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.  Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.

Some dual kets and error operators are plotted overleaf, for a family of related approximation methods.  Each column of plots illustrates one method.  The first line plots the component amplitudes in phase space; the darkness of the marker for each component is proportional to the norm of its dual ket.  The logarithms of these norms are plotted in the second row.  The third row shows $〈m|E|n〉$, the matrix elements of the error operator between Fock states, for $n=0$ to 15.  The brightness of each element is proportional to its modulus; a perfect approximation would have $E=0$, and a black image.  The hue of each element represents its complex argument.  Cyan is positive real, red is negative real, and so on around the color wheel.

The approximation in the leftmost column is essentially $〈A⁺|=10^{-2}〈A|$, which is very close to zero.  Therefore the error operator is close to $E=1$, with ones on the diagonal, and zeros elsewhere.  The first three dual kets $|φ₁〉$ through $|φ₃〉$ are illustrated below, by plotting the inner products $〈β|φ_r〉$, where $β$ ranges over phase space.  The component amplitudes are overlaid as white dots.  In this case, the dual kets are scaled component kets, with $|φ_r〉\propto|α_r〉$.  The amplitudes $α₁$, $α₂$ and $α₃$ are the white dots at the centres of the plotted coherent states.

The illustrated approximation methods use Tychonoff regularised least squares, with different values of the regularisation parameter $λ$.  They assign expansion coefficients $d$ in order that
$$\pmatrix{|A〉\cr λI}d≈\pmatrix{|ψ〉\cr 0},$$
this equation being satisfied in a least-squares sense.  The solution minimises the residual $‖|ψ〉-|A〉d‖²+λ²‖d‖²$.  It is given by
$$d=〈A⁺|ψ〉=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0}|ψ〉,$$
where $()⁺$ denotes the Moore-Penrose pseudoinverse of a matrix.  I.e., the vector $〈A⁺|$ is the leftmost column of the pseudoinverse matrix.

The leftmost column of the figure illustrates a large regularisation parameter.  The norm of $d$ is constrained to be small, so the approximation is very bad.  However, the approximation coefficients are very well behaved: they are brackets of $|ψ〉$ with coherent states.  The rightmost column illustrates a small regularisation parameter, where $d$ is essentially the least squares solution to $|A〉d=|ψ〉$.  This approximation is good, and the plot of the error operator is mostly black.  In fact, it is surprisingly good: the largest component amplitude is around 2, but the approximation is nearly exact up to $n=11$.  There is a price to pay for this.  The dual kets have large norms, and they are close to the number state $|n=14〉$.  This means that approximations of states with small amplitudes not be stable: perturbing such a state by adding a small component of $|n=14〉$ will wreak havoc with its expansion coefficients.  Using the approximation method of the leftmost column, this component would have very little effect on the coefficients.  This type of instability is very common in least squares problems, and the purpose of regularisation methods such as Tychonov's is to avoid it by trading off the ability to represent large number states.  The middle columns show some intermediate values of $λ$.  In the third column, the dual kets are close to 1, and number states up to $n=4$ or 5 are approximated well, as would be expected for an expansion over amplitude 2 coherent states.  This seems to achieve the right tradeoff.

\midinsert\XeTeXpicfile tych.pdf width\hsize \endinsert

The next figure attempts to illustrate the tradeoff between accuracy and stability as the regularisation parameter varies.  The obvious way to measure accuracy is some norm of the error operator.  However, Hilbert space has infinite dimensions, and only finitely many of them are preserved by these approximations, so the usual norms indicate that all of them are infinitely bad.  I've handled this by truncating Fock space, and taking the Frobenius norm of the matrix 
$$\pmatrix{〈0|E|0〉&&〈0|E|20〉\cr &\vdots&\cr 〈20|E|0〉&&〈20|E|20〉}.$$
This is plotted on the vertical axis.  The horizontal axis is the standard deviation of the norms of the dual kets; no doubt there are better ways to measure the stability of the approximation, but this seems good enough to be useful.  The dots on the graph indicate the approximations illustrated in the first figure; the two on either side of the kink are the ones that seemed reasonable.

The lower left part of the graph corresponds to the left hand columns of the first figure, where the approximation is over regularised, and the dual kets are scaled components.  There is clearly some polynomial scaling.  This is confirmed by the leftmost graphs of logarithmic dual ket norms, which are similar, but have different vertical scales.  In the third illustrated approximation, where the performance graph kinks, the relative weights have a property I've been looking for in other contexts: the weight of the dual vector, and thus the coefficient that tends to be assigned to the component, varies inversely with the density of sampled amplitudes.

After this, the performance line flattens off, and more stability must be sacrificed to improve the accuracy further.  At the top right most point, the effect of Tychonov  regularisation is negligible, and the approximation is very close to a least squares solution to $|A〉d=|ψ〉$.

The form of the dual kets for the intermediate approximations has some features that I can't fully explain.  The phase-space expansion of each dual kets has zeros near the amplitudes of the other components.  I.e., $〈φ_r|α_s〉≈0$ when $r≠s$, unless $α_r$ and $α_s$ are close.  Some of these zeros occur where $〈α_r|α_s〉$ is significant, as seen in the first two columns.  This means that kets that are localised in phase space will be expanded over components with localised amplitudes, to a greater extent than would be predicted from the overlaps of the components.  Furthermore, the bracket $〈β|φ_r〉$ is real and negative when $β$ and $α_r$ are on opposite sides of $α_s$.

Where pairs of amplitudes happen to lie very close together, the norms of the dual kets are very similar, as seen from the darkness of the disks illustrating the components.  This suggests that the dual kets might be negatives of each other, so that the components tend to be assigned equal and opposite coefficients, as we've seen occur in dynamic simulations.  This could be tested by coloring the disks according to the phase of $〈φ_r|α_r〉$: pairs of complementary colors would suggest cancellation.  The problem is that the support of $|φ_r〉$ in phase space is remote from $|α_r〉$; it might be better to plot $〈n=15|φ_r〉$ or similar, if one could predict what number state would dominate the dual kets in the unregularised approximation.

\bye

### 2013-10-03 paper.tex
Started new dual paper

--- 2013-09-24 paper.tex
@@
 The projector satisfies
-$$π^{-1}\int |α〉〈α|\,d²α=1.$$
-This is similar to the law for an orthonormal basis, except that the left hand side must be scaled by $π$.  Therefore, any state can be expanded as a sum over coherent states, in the form
+$$\int |α〉〈α|\,d²α=π.$$
+If the right hand side were 1, this would indicate that the coherent states were an orthonormal basis of ket space.
+As it is, any state can be expanded as a combination of coherent states, in the form
 $$|ψ〉=π^{-1}\int |α〉〈α|ψ〉\,d²α.\tag\csproj$$
@@
 A set of states like this is known as a frame, in distinction from a basis, whose elements are linearly independent, and whose expansions are unique.
+Sets such as the coherent states are known as a Gabor frame.
 Further properties of the coherent states will be stated as required.

### 2013-10-05 dual.tex
writing

@@
 this equation being satisfied in a least-squares sense.  The solution minimises the residual $‖|ψ〉-|A〉d‖²+λ²‖d‖²$.  It is given by
-$$d=〈A⁺|ψ〉=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0}|ψ〉,$$
-where $()⁺$ denotes the Moore-Penrose pseudoinverse of a matrix.  I.e., the vector $〈A⁺|$ is the leftmost column of the pseudoinverse matrix.
+$$d=〈A⁺|ψ〉=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0_R}|ψ〉,$$
+where $()⁺$ denotes the Moore-Penrose pseudoinverse of a matrix, $I$ is the $R×R$ unit matrix, and $0_R$ is a $R×1$ zero vector.
+I.e., the vector $〈A⁺|$ is the leftmost column of the pseudoinverse matrix.
+
+Suppose $|A〉$ has the singular-value decomposition $|A〉=|U〉SV†$, so that
+$$\pmatrix{|A〉\cr λI}v_r=\pmatrix{σ_r|u_r〉\cr λv_r}=\sqrt{σ_r²+λ²}w_r.$$
+The orthonormality of the $w_r$ follows from that of the $|u_r〉$ and $v_r$, so we have a singular-value decomposition
+$$\pmatrix{|A〉\cr λI}=WTV†,$$
+where $τ_r=\sqrt{σ_r²+λ²}$.  Then 
+$$〈A⁺|=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0_R}=VT^{-1}W†\pmatrix{1\cr 0_R},$$
+where the last two factors are the first column of $W†$,
+$$W†\pmatrix{1\cr 0_R}=\pmatrix{{σ₁\over\sqrt{σ₁²+λ²}}〈u₁|\cr\vdots\cr 
+	{σ_R\over\sqrt{σ_R²+λ²}}〈u_R|}.$$
+It follows that the right singular vectors of $〈A⁺|$ are the bras $〈u_r|$, and the singular values are 
+$$σ⁺_r={σ_r\over τ_r\sqrt{σ_r²+λ²}}={σ_r\over σ_r²+λ²}.$$
+Tychonov regularisation divides the singular values of $|A〉$ into large ones and small ones, as compared to $λ$.
+For large $σ_r$, the inverse is very close to the unregularised problem, with $σ⁺_r≈1/σ_r$.
+For small $σ_r$, the component in that singular direction is nearly discarded by $|Α⁺〉$, with $σ⁺_r≈σ_r/λ²$.
+Note that, in this limit, $|Α⁺〉=λ^{-2}|A〉$!
+In between, there is a maximum at $σ_r=λ$, with $σ_r=1/2λ$, half the unregularised value.
+
+This has dramatic consequences for the conditioning of the regularised problem, and how that varies with $λ$.
+When $λ<σ_R$, regularisation has very little effect: the condition of $〈Α⁺|$ is the same as that of $|A〉$, which, for dense sets of coherent states, is usually dreadful.
+As $λ$ increases, so that $σ₁<λ<σ_R$, the singular values lie on either side of the $σ_r=λ$ peak; the condition of $〈Α⁺|$ is determined by how far below this peak $σ₁$ and $σ_R$ fall, and the condition number has a sharp minimum when $σ⁺₁=σ⁺_R$.
+When $λ$ dominates the singular values, and $|Α⁺〉=λ^{-2}|A〉$, it is again the case that their condition is the same.
 

### 2013-10-10 dual.tex
writing

--- 2013-10-03 dual.tex
@@
 Presumably, not all kets can be approximated this way: it would be asking too much to approximate a Fock state $|n=100〉$ in terms of coherent states with amplitudes 1 and 2.
-
-The methods of approximation we will consider set $d$ to a linear function of $|ψ〉$.
-We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
-We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
-Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
-such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
 
@@
 The amplitudes $α₁$, $α₂$ and $α₃$ are the white dots at the centres of the plotted coherent states.
-
-The illustrated approximation methods use Tychonoff regularised least squares, with different values of the regularisation parameter $λ$.  They assign expansion coefficients $d$ in order that
-$$\pmatrix{|A〉\cr λI}d≈\pmatrix{|ψ〉\cr 0},$$
-this equation being satisfied in a least-squares sense.  The solution minimises the residual $‖|ψ〉-|A〉d‖²+λ²‖d‖²$.  It is given by
-$$d=〈A⁺|ψ〉=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0}|ψ〉,$$
-where $()⁺$ denotes the Moore-Penrose pseudoinverse of a matrix.  I.e., the vector $〈A⁺|$ is the leftmost column of the pseudoinverse matrix.
-
-The leftmost column of the figure illustrates a large regularisation parameter.
-The norm of $d$ is constrained to be small, so the approximation is very bad.
-However, the approximation coefficients are very well behaved: they are brackets of $|ψ〉$ with coherent states.
-The rightmost column illustrates a small regularisation parameter, where $d$ is essentially the least squares solution to $|A〉d=|ψ〉$.
-This approximation is good, and the plot of the error operator is mostly black.
-In fact, it is surprisingly good: the largest component amplitude is around 2, but the approximation is nearly exact up to $n=11$.
-There is a price to pay for this.
-The dual kets have large norms, and they are close to the number state $|n=14〉$.
-This means that approximations of states with small amplitudes not be stable: perturbing such a state by adding a small component of $|n=14〉$ will wreak havoc with its expansion coefficients.
-Using the approximation method of the leftmost column, this component would have very little effect on the coefficients.
-This type of instability is very common in least squares problems, and the purpose of regularisation methods such as Tychonov's is to avoid it by trading off the ability to represent large number states.
-The middle columns show some intermediate values of $λ$.
-In the third column, the dual kets are close to 1, and number states up to $n=4$ or 5 are approximated well, as would be expected for an expansion over amplitude 2 coherent states.
-This seems to achieve the right tradeoff.
 

### 2013-10-10 paper.tex
writing

@@
 
-The invention of the laser called for a quantum theory of intense radiation.
+The invention of the laser in 1963 called for a quantum theory of intense radiation.
 Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.
@@
 The problems are exacerbated when the amplitudes are spaced irregularly, instead of on a grid.
+
+Expanding a field over Gaussian functions is a rather obvious idea, and physicists are not the only people to have thought of it.
+The mathematicians and signal theorists who developed wavelet expansions have also analysed Gaussians.
+They call this type of expansion a Gabor frame, or a windowed Fourier transform.
+They have prooved several results that are not well known in physics: for example, that discrete sets of coherent states are complete, not just the continuum.
+They have also studied overcomplete sets of vectors in general, under the field of frame theory.
+Many of these general ideas insight into the specific case of coherent state frames.
 
@@
 
-\beginsection{The phase space expansion operator}
+\beginsection{Continuous phase space expansions}
 
@@
 	=|A〉d,\tag\Ad$$
-where $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
+where $A=\pmatrix{α₁&⋯$α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
 In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.
-The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion $|A〉$.
+The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion operator $|A〉$.
 Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.
@@
 
-\beginsection{Discretely sampled expansions}
+The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.
+We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
+We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
+Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
+such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
+
+\beginsection{Discrete phase space expansions}
 
@@
 
-In recent decades, these expansions have been studied intensely by engineers involved in signal processing, who call the set of coherent states a Moyal~frame.
+In recent decades, these expansions have been studied intensely by engineers involved in signal processing, who call the set of coherent states a Gabor~frame.
 These researchers regard gaussian wavepackets as a kind of time-limited Fourier transform, with a gaussian window function, following on from the work of Hamming and others at Bell Labs.
-Moyal frames were investigated by the same community that developed wavelets; the latter became of very wide interest, and obscured the earlier work on Moyal frames.
+Gabor frames were investigated by the same community that developed wavelets; the latter became of very wide interest, and obscured the earlier work.
 
@@
 
-%% Tychonov conditioning
+\beginsection{Tychonov conditioning}
+
+The simplest approximation method is Tychonoff regularised least squares, with different values of the regularisation parameter $λ$.  This assigns expansion coefficients $d$ in order that
+$$\pmatrix{|A〉\cr λI}d≈\pmatrix{|ψ〉\cr 0},$$
+this equation being satisfied in a least-squares sense.  The solution minimises the residual $‖|ψ〉-|A〉d‖²+λ²‖d‖²$.  It is given by
+$$d=〈A⁺|ψ〉=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0_R}|ψ〉,$$
+where $()⁺$ denotes the Moore-Penrose pseudoinverse of a matrix, $I$ is the $R×R$ unit matrix, and $0_R$ is a $R×1$ zero vector.
+I.e., the vector $〈A⁺|$ is the leftmost column of the pseudoinverse matrix.
+
+Suppose $|A〉$ has the singular-value decomposition $|A〉=|U〉SV†$, so that
+$$\pmatrix{|A〉\cr λI}v_r=\pmatrix{σ_r|u_r〉\cr λv_r}=\sqrt{σ_r²+λ²}w_r.$$
+The orthonormality of the $w_r$ follows from that of the $|u_r〉$ and $v_r$, so we have a singular-value decomposition
+$$\pmatrix{|A〉\cr λI}=WTV†,$$
+where $τ_r=\sqrt{σ_r²+λ²}$.  Then 
+$$〈A⁺|=\pmatrix{|A〉\cr λI}⁺\pmatrix{1\cr 0_R}=VT^{-1}W†\pmatrix{1\cr 0_R},$$
+where the last two factors are the first column of $W†$,
+$$W†\pmatrix{1\cr 0_R}=\pmatrix{{σ₁\over\sqrt{σ₁²+λ²}}〈u₁|\cr\vdots\cr 
+	{σ_R\over\sqrt{σ_R²+λ²}}〈u_R|}.$$
+It follows that the right singular vectors of $〈A⁺|$ are the bras $〈u_r|$, and the singular values are 
+$$σ⁺_r={σ_r\over τ_r\sqrt{σ_r²+λ²}}={σ_r\over σ_r²+λ²}.$$
+Tychonov regularisation divides the singular values of $|A〉$ into large ones and small ones, as compared to $λ$.
+For large $σ_r$, the inverse is very close to the unregularised problem, with $σ⁺_r≈1/σ_r$.
+For small $σ_r$, the component in that singular direction is nearly discarded by $|Α⁺〉$, with $σ⁺_r≈σ_r/λ²$.
+Note that, in this limit, $|Α⁺〉=λ^{-2}|A〉$!
+In between, there is a maximum at $σ_r=λ$, with $σ_r=1/2λ$, half the unregularised value.
+
+This has dramatic consequences for the conditioning of the regularised problem, and how that varies with $λ$.
+When $λ<σ_R$, regularisation has very little effect: the condition of $〈Α⁺|$ is the same as that of $|A〉$, which, for dense sets of coherent states, is usually dreadful.
+As $λ$ increases, so that $σ₁<λ<σ_R$, the singular values lie on either side of the $σ_r=λ$ peak; the condition of $〈Α⁺|$ is determined by how far below this peak $σ₁$ and $σ_R$ fall, and the condition number has a sharp minimum when $σ⁺₁=σ⁺_R$.
+When $λ$ dominates the singular values, and $|Α⁺〉=λ^{-2}|A〉$, it is again the case that their condition is the same.
+
+\midinsert \XeTeXpicfile tysv.pdf width \hsize
+Figure 3 \endinsert
+
+This is illustrated in Figure~3, which plots the logarithms of the singular values of $|A〉$ against those of $|A⁺〉$, both scaled to the regularisation parameter $λ$.
+In the right hand plot, $λ$ is small compared the the singular values, so these lie very close to the line $σ⁺=1/σ$.
+In the left hand plot, $λ$ is large, $|A⁺〉$ is nearly $λ^{-2}|A〉$, and the singular values lie on $λσ⁺=σ/λ$.
+The singular values have been inverted, but this does not change the ratio of the smallest to the largest, so in both cases the condition numbers of $|A〉$ and $|A⁺〉$ are the same.
+
+In the middle plot, $λ$ is of the same order as the singular values, which is the usual case in well-regularised problems.
+In the illustrated case, where the singular values span several orders of magnitude, the condition number of $|A⁺〉$ is about the square root of that of $|A〉$.
+If the singular values were more even to begin with, and all fit in the curve at the top of the peak, the improvement would be greater.
+However, in this case it is unlikely that the problem would need to be regularised, so the square root improvement is likely the best that can be expected.
+
+The leftmost column of the figure illustrates a large regularisation parameter.
+The norm of $d$ is constrained to be small, so the approximation is very bad.
+However, the approximation coefficients are very well behaved: they are brackets of $|ψ〉$ with coherent states.
+The rightmost column illustrates a small regularisation parameter, where $d$ is essentially the least squares solution to $|A〉d=|ψ〉$.
+This approximation is good, and the plot of the error operator is mostly black.
+In fact, it is surprisingly good: the largest component amplitude is around 2, but the approximation is nearly exact up to $n=11$.
+There is a price to pay for this.
+The dual kets have large norms, and they are close to the number state $|n=14〉$.
+This means that approximations of states with small amplitudes not be stable: perturbing such a state by adding a small component of $|n=14〉$ will wreak havoc with its expansion coefficients.
+Using the approximation method of the leftmost column, this component would have very little effect on the coefficients.
+This type of instability is very common in least squares problems, and the purpose of regularisation methods such as Tychonov's is to avoid it by trading off the ability to represent large number states.
+The middle columns show some intermediate values of $λ$.
+In the third column, the dual kets are close to 1, and number states up to $n=4$ or 5 are approximated well, as would be expected for an expansion over amplitude 2 coherent states.
+This seems to achieve the right tradeoff.
 
@@
 
-This paper will talk about two related but distinct inverse problems.
-Both start with a given wavefunction $|ψ〉$, and a set of amplitudes in phase space, $\{α_j\}$, over which to expand it.
-In systems with few modes, these points could be a grid, but in many-mode systems they will have to be more sparse than that: there isn't space to store grid coefficients.
-The first problem, let's call it the inverse problem, is to find a set of weights $f_j$ such that $∑f_j|α_j〉$ approximates $|ψ〉$.
-This is simply finding a discrete inverse to $G$ sampled on the grid $α_j$; since $G$ is a projection, the pseudoinverse of $G$ should solve it well enough.
-The differential case, where we have an expansion for $|ψ〉$ which we want to perturb to represent $|ψ〉+|dψ〉$, is no harder.
-Because $G$ is linear, we can expand $|dψ〉$, and add it to the expansion of $|ψ〉$.
-
-The second problem, the variational one, is a generalisation.
-As well as the grid $α_j$, we have a set of weights $f_j$ representing some wavefunction, presumably not $|ψ〉$.
-Instead of expanding $|ψ〉$ over phase space, we want to expand it over the columns of $|Dψ(f,α)〉$, the set of all changes we can make by slightly adjusting the original expansion, and the grid.
-Obviously, we can solve this problem simply by setting the coefficients of $|∂fψ(f,a)〉$ to the solution of the inverse problem, and those of $|∂fψ(f,a)〉$ to zero.
-Is this the smallest norm solution?
-Probably not: if a coherent state was moving in an oscillator, it might be easier to just rotate the $α_j$.
+The problem we have discussed so far is to expand a state $|ψ〉$ over a set of coherent states, in the form $|ψ〉=|A〉d$.
+In order for a this state to be represented on a computer, the coherent amplitudes $A$ and the coefficients $d$ would all have to be stored in memory.
+In a simulation, we will have an expansion $d$ for $|ψ〉$, which we want to perturb to represent $|ψ〉+|δψ〉$.
+The expansion operator $|A〉$ is linear, so we could expand $|δψ〉=|A〉δd$ and add $δd$ to $d$  However, if the simulation is to represent a many-mode state efficiently, the amplitudes will likely have to vary over time as well as the coefficients.
+Also, it might sometimes be more efficient to vary the ampltiudes.
+For example, if a coherent state was rotating in an oscillator, it might be simpler to rotate the $α_j$ than to vary the weights.
+
+We then have a variational problem $|\dot ψ〉=|Dψ(z)〉\dot z$, where
+$$z=\pmatrix{A\cr d},$$
+and the expansion operator $|Dψ(z)〉$ includes the derivatives of the coherent states with respect to their amplitudes (notation?).
+
+These two problems are closely related.  In fact,
+$$|Dψ(z)〉=
+	\lim_{h→0}\pmatrix{|A+½h〉&|A-½h〉}
+	\pmatrix{{I\over 2}&{Id\over h}\cr{I\over 2}&-{Id\over h}}.
+$$
+The variational problem is equivalent to expanding the state over a frame of $2R$ coherent states, clustered in tight pairs, then taking sums and differences of the expansion coefficients.
+In fact, the vector $d$ could be absorbed into the shifts $+½h$ and $-½h$, so that the transformation matrix was even simpler.
+This offers other approaches to regularisation: we could expand with regularised coefficients, and apply the transformation to the result.
+
+When two amplitudes $α_r$ and $α_s$ are close together, three linearly independent $d$ cause $|Dψ(z)〉d$ become parallel.
+These being $|∂_rψ〉- $|∂_sψ〉$, |∂_{R+r}ψ〉$, and $|∂_{R+s}ψ〉$.
+Two nearby points effectively become one, as far as the variational problem is concerned!
 

### 2013-10-17 paper.tex
writing

@@
 Applying these formulae to the variational dynamics of the coherent states, which are not Hamiltonian, is somewhat subtle.
+
+\beginsection{Notational preliminaries}
+
+Numerical analysis and quantum mechanics are both founded in vector spaces and linear algebra.
+In numerical analysis, the vectors and operators have finitely many dimensions, and are usually represented as matrices.
+In quantum mechanics, they usually come from infinite dimensional spaces, and are written as bras an d kets.
+This paper is about computational quantum mechanics, and it combines both notations.
+
+Let's start with ket expanded as a linear combination, 
+$$|ψ〉=c₁|a₁〉+c₂|a₂〉+⋯+c_R|a_R〉.$$
+This can be written $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}^{\rm T}$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.
+There are other ways to picture the object $|A〉$.
+It can be seen as a linear operator from ${\bf C}^R$ into ket space, or as a matrix representation of that operator, whose columns $|a₁〉$ through $|a_R〉$ are infinite dimensional.
+This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}^{\rm T}$, that maps kets to complex column vectors.
+The products
+$$〈A|A〉=\pmatrix{〈a₁|a₁〉&&〈a₁|a_R〉\cr &\ddots&\cr 〈a_R|a₁〉&&〈a_R|a_R〉}$$
+and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ both follow the usual rules of matrix multiplication.
+Matrix products work too: for instance, the kets $|a_r〉$ are linearly independent, then $|A〉$ has a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix.
+In general, things written as kets have infinite dimensional columns and finitely many rows, while bras are the other way around.
+
+Even more general matrices are sometimes useful.  For instance, the operator
+$$M=\pmatrix{0&〈A|\cr |A〉&0}$$
+can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.
+The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.
+The general rule is that the things in each row must have the same column dimension, and visa-versa.
+Things that have finite dimensional rows and columns will be written as matrices, and things that are infinite in both directions as Hilbert space operators.
+These are both just written as letters.
+
+So far, the operator $|A〉$ has been a linear function from vectors into ket space.
+General functions will be written $|A〉$ as well, but their applications will be written $|A(c)〉$ instead of $|A〉c$.
+The derivatives of these functions are explained below.
 
@@
 We can, however, look for a vector $\dot z(z)$ that minimises the residual
-$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\eqno{(4)}$$
+$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\tag\residual$$
 and solve the ordinary differential equation
@@
 
+The kind of ansatz considered in this paper has the form 
+$$|ψ(z₁,…,z_R)〉=|ψ'(z₁)〉+⋯+|ψ'(z_R)〉.$$
+In fact, this form is very general: the DMRG approximants are the only ones I know of that are not a sum of similar terms.
+Here, $z_r$ may be a vector of complex numbers, perhaps a weight and the amplitudes of several modes.
+In this case, the derivative has the form
+$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&…&|Dψ'(z_R)〉}.$$
+
+I will refer to the kets $|ψ'(z_r)〉$ as the components of $|ψ(z)〉$.
+It is interesting to consider the case where these lie in orthogonal subspaces.
+This will very nearly be the case, for instance, if the components are coherent states, with sparsely distributed amplitudes.
+In this case, the component derivatives $|Dψ'(z_r)〉$ also lie in the orthogonal subspaces, and the total derivative has a block diagonal structure,
+$$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&&\cr&\ddots&\cr&&|Dψ'(z_R)〉},$$
+where the empty elements in row $r$ are null kets in the subspace of the component $|ψ'(z_r)〉$.
+The residual ket thus becomes
+$$i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉=
+	\pmatrix{|Dψ'(z₁)〉-H|ψ'(z₁)〉\cr\vdots\cr |Dψ'(z_R)〉-H|ψ'(z_R)〉},$$
+where the rows are mutually orthogonal.  Therefore, the least squares problem can be solved separately for each component, independently minimising 
+$$\left‖|Dψ'(z_r)〉-H|ψ'(z_r)〉\right‖².$$
+If the system is a BEC, and the components are coherent states, this problem is equivalent solving the Gross-Pitaevskii Equation.
+Heller and others have investigated it for general systems with coherent state components, and found that the amplitudes obey something very like classical dynamics.
+Therefore variational dynamics with sparse components is a waste of time: its results should be very similar to truncated Wigner, and, if they differ, there won't be much reason to believe one rather than the other.
+
+It is interesting to compare this to positive P simulations for the quartic oscillator.
+These have predicted revivals, and have been done for large enough systems that the amplitudes of the coherent states must have been sparse.
+I don't know whether both conditions were ever satisfied in the same simulation: it would be interesting to find out.
+
 The residual above can be defined using any norm one wishes.
@@
 
-This complicates the solution of the problem in Formula~4.
+This complicates the solution of the problem in Formula~\residual.
 It does not suffice to find a $\dot z$ with the smallest residual, for a given value of $z$.
-In order for Equation~5 to be numerically solvable, the properties of $\dot z$ under perturbation of $z$ are equally important.
-Under the usual Hilbert space norm, Formula~4 is a least squares problem; the pertubation properties of such problems are reasonably well understood.
+In order for Equation~\ode\ to be numerically solvable, the properties of $\dot z$ under perturbation of $z$ are equally important.
+Under the usual Hilbert space norm, Formula~\residual is a least squares problem; the pertubation properties of such problems are reasonably well understood.
 We will examine ways to modify the problem, by introducing extra least-squares constraints, that might improve its perturbation properties.
@@
 	=|A〉d,\tag\Ad$$
-where $A=\pmatrix{α₁&⋯$α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
+where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
 In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.
@@
 
+\beginsection{Gabor frames}
+
+Many of the above results for the operator $\G$ are true of expansions over frames in general.
+We will treat the finite case, because every frame will eventually have to be represented on a computer.
+So suppose the set of kets $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a frame for some subspace of Hilbert space.
+
+The theory of frames is traditionally presented in terms of the frame operator,
+$$S=∑_{r=1}^R |a_r〉〈a_r|={\rm Tr}(|A〉〈A|).$$
+In the case that the columns of $|A〉$ are a frame, $S$ is invertible.  The set of vectors
+$$|F〉=S^{-1}|A〉=\pmatrix{S^{-1}|a₁〉&⋯&S^{-1}|a_R〉}
+	=\pmatrix{|φ₁〉&⋯&|φ_R〉}
+$$
+is also a frame, and is known as the dual frame of $|A〉$.  It has the property that $|A〉〈F|$ is the identity operator on the space.
+
+For finite frames, expansions using dual kets have some special properties.
+For a start, the dual expansion is the one where the vector of coefficients has the least norm.
+In fact, if $|ψ〉=|A〉d$, and $c=〈F|ψ〉$, then $‖d‖²=‖c‖²+‖d-c‖²$\cite{1989-sir-31-626}.
+This suggests that $d-c$ is forced to be orthogonal to $c$.
+
+The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.
+We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
+We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
+Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
+such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
+
 \beginsection{Truncated phase space}
@@
 With these directions, we can form some of the smaller singular vectors of the full problem.
-
-\beginsection{Gabor frames}
-
-The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.
-We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
-We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
-Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
-such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
 
@@
 When two amplitudes $α_r$ and $α_s$ are close together, three linearly independent $d$ cause $|Dψ(z)〉d$ become parallel.
-These being $|∂_rψ〉- $|∂_sψ〉$, |∂_{R+r}ψ〉$, and $|∂_{R+s}ψ〉$.
+These being $|∂_rψ〉- |∂_sψ〉$, $|∂_{R+r}ψ〉$, and $|∂_{R+s}ψ〉$.
 Two nearby points effectively become one, as far as the variational problem is concerned!

### 2013-10-25 paper.tex
writing

@@
 
-Numerical analysis and quantum mechanics are both founded in vector spaces and linear algebra.
-In numerical analysis, the vectors and operators have finitely many dimensions, and are usually represented as matrices.
-In quantum mechanics, they usually come from infinite dimensional spaces, and are written as bras an d kets.
-This paper is about computational quantum mechanics, and it combines both notations.
-
-Let's start with ket expanded as a linear combination, 
+Quantum mechanics deals with infinite dimensional linear algebra, represented abstractly as bras and kets.
+Numerical analysis is also founded on vectors and operators, but in finite dimensional spaces, expanded as matrices and column vectors.
+The numerical quantum mechanics discussed in this paper is most conveniently described by combining these notations.
+
+Consider a linear combination of kets,
 $$|ψ〉=c₁|a₁〉+c₂|a₂〉+⋯+c_R|a_R〉.$$
-This can be written $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}^{\rm T}$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.
-There are other ways to picture the object $|A〉$.
-It can be seen as a linear operator from ${\bf C}^R$ into ket space, or as a matrix representation of that operator, whose columns $|a₁〉$ through $|a_R〉$ are infinite dimensional.
-This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}^{\rm T}$, that maps kets to complex column vectors.
-The products
+This can be written as a matrix product $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}^{\rm T}$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.
+Alternatively, the kets $|a_r〉$ can be regarded as infinite dimensional columns, and $|A〉$ as a matrix, representing a linear operator from ${\bf C}^R$ into ket space.
+This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}^{\rm T}$, that maps kets back to column vectors.
+Under the rules of matrix multiplication, the product
 $$〈A|A〉=\pmatrix{〈a₁|a₁〉&&〈a₁|a_R〉\cr &\ddots&\cr 〈a_R|a₁〉&&〈a_R|a_R〉}$$
-and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ both follow the usual rules of matrix multiplication.
-Matrix products work too: for instance, the kets $|a_r〉$ are linearly independent, then $|A〉$ has a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix.
-In general, things written as kets have infinite dimensional columns and finitely many rows, while bras are the other way around.
+is an operator on ${\bf C}^R$, and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ is a Hilbert space operator.
+Other matrix products work too; provided the $|a_r〉$ are linearly independent, $|A〉$ will have a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix%
+\footnote{${}^1$}{In fact, $|A〉$ has an SVD regardless, but $|U〉$ and $S$ are smaller.
+When the $|a_r〉$ are linearly dependent, they span a space of dimension less than $R$, and there is no way to choose in which directions it should be extended to give $R$ orthonormal kets.
+In this case, $|U〉$ will be a row of ${\rm rank}|A〉$ kets, and $S$ will be ${\rm rank}|A〉×R$.}.
+In this paper, things written as kets have a finite number of infinite dimensional columns, and bras are the other way round.
 
@@
 So far, the operator $|A〉$ has been a linear function from vectors into ket space.
-General functions will be written $|A〉$ as well, but their applications will be written $|A(c)〉$ instead of $|A〉c$.
-The derivatives of these functions are explained below.
+General functions will be written $|f〉$ as well, but their applications will be written $|f(c)〉$ instead of $|A〉c$.
+If this function is analytic, it has partial derivatives $|∂_if(c)〉$ with respect to the components of $c$.
+The derivative $∂_i$ acts on the function, not in Hilbert space, so it is written inside the ket.
+These can be assembled into a Jacobian matrix $|Df(c)〉=\pmatrix{|∂₁f(t)〉&⋯&|∂_Rf(t)〉}$, such that $|f(c+h)〉=|f(c)〉+|Df(c)〉h+o(h²)$.
+The derivatives $|∂_if〉$ are ket-valued functions like $|f〉$.
 
@@
 
-Schrödinger's equation, the quantum law of motion, is
+In the notation defined above, Schrödinger's equation is
 $$i\hbar |DΨ(t)〉=H|Ψ(t)〉.$$
-I will discuss many details of mappings from complex numbers to quantum states, so some notation used in this paper is traditionally used more in mathematics than in physics.
-The symbol $|Ψ〉$ denotes a ket-valued function of a real number; $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.
+Here, $|Ψ〉$ denotes a ket-valued function of time, and $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.
 The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.
 Therefore it is written outside the ket.
-The total derivative $D$ acts on the function, so it is written inside the ket.
-Because $|Ψ〉$ is a function of a single variable, $|DΨ〉$ is simply another ket-valued function, and $|DΨ(t)〉$ is the ket that would traditionally be written ${∂\over ∂t}|Ψ〉\big|_{t=t}$.
-This could alternatively be written as a partial derivative $|DΨ〉=|∂₀Ψ〉$.
-If $|Ψ〉$ were a function of several variables, the total derivative function $|DΨ〉$ would evaluate to a row vector $|DΨ(t)〉=\pmatrix{|∂₀Ψ(t)〉&⋯&|∂_RΨ(t)〉}$.
-One can regard the kets as column vectors and read this as a Jacobian matrix with infinitely many rows and $R$ columns.
-The notation will be elaborated below as needed.
-
-Schrödinger's equation could even be abbreviated to
+This could be abbreviated to
 $$i\hbar |DΨ〉=H|Ψ〉.$$
+
 This is a very simple equation.

### 2013-10-31 paper.tex

@@
 Even more general matrices are sometimes useful.  For instance, the operator
-$$M=\pmatrix{0&〈A|\cr |A〉&0}$$
+$$M=\pmatrix{0&|A〉\cr 〈A|&0}$$
 can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.
@@
 Here, $|Ψ〉$ denotes a ket-valued function of time, and $|Ψ(t)〉$ is the ket that this function evaluates to at time $t$.
-The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$.
-Therefore it is written outside the ket.
-This could be abbreviated to
-$$i\hbar |DΨ〉=H|Ψ〉.$$
-
-This is a very simple equation.
-The challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand it over a grid grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes merely thousands of particles.
+The Hamiltonian $H$, a Hilbert space operator, acts on this ket, not on the function $|Ψ〉$, and is written outside the ket.
+Schrödinger's equation could be abbreviated to
+$$i\hbar |DΨ〉=H|Ψ〉,$$
+equating functions instead of their values.
+
+Schrödinger's equation is very simple.
+The main challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand this over a regular basis grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes mere thousands of particles.
 In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
 $$|Ψ(t)〉=|ψ(z(t))〉,$$
-or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\left(z₁(t),…,z_R(t)\right)$ is a vector of complex parameters.
-This imposes two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.
+or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}^{\rm T}$ is a vector of complex parameters.
+This poses two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.
 
@@
 The above model does not apply to these.
-There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which can compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store all the parameters at once.
+There are also methods, such as phase-space Monte-Carlo ones, that have the above form in principle, but which compute different parameters $z_r(t)$ independently, and find expectation values of observables without having to store them all at once.
 
@@
 If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
-$$i\hbar|Dψ(z(t))〉Dz(t)=H|ψ(z(t))〉.$$
+$$i\hbar|Dψ(z)〉Dz=H|ψ(z)〉.$$
 On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.
-This is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to have components outside this subspace.
-This is a consequence of the restricted form of the approximation.
-We can, however, look for a vector $\dot z(z)$ that minimises the residual
-$$\left\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\right\|²,\tag\residual$$
+This equation is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.
+In other words, the restricted form of the approximation does not include the solution to Schrödinger's equation.
+We can, however, look for a least-squares solution, a vector $\dot z(z)$ that minimises the residual
+$$\bigl\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\bigr\|²,\tag\residual$$
 and solve the ordinary differential equation
-$${dz\over dt}=\dot z(z),\tag\ode$$
+$$Dz(t)=\dot z(z(t)),\tag\ode$$
 giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.
@@
 The Dirac-Frenkel solution is only optimised locally.
-There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkey solution $|ψ(w(t))〉$ does.
+There might be some $w(t)$ such that, at large times, $|ψ(w(t))〉$ approximates $|Ψ(t)〉$ better that the Dirac-Frenkel solution $|ψ(w(t))〉$ does.
 This would require that, at early times, $i\hbar|Dψ(z(t))〉$ is closer to $H|ψ(z(t))〉$ than $i\hbar|Dψ(w(t))〉$ is to $H|ψ(w(t))〉$, but that $z(t)$ loses the ability to track the dynamics at later times.
@@
 $$|ψ(z₁,…,z_R)〉=|ψ'(z₁)〉+⋯+|ψ'(z_R)〉.$$
+Here, $z_r$ may be a vector of complex numbers, perhaps a weight and the amplitudes of several modes.
 In fact, this form is very general: the DMRG approximants are the only ones I know of that are not a sum of similar terms.
-Here, $z_r$ may be a vector of complex numbers, perhaps a weight and the amplitudes of several modes.
 In this case, the derivative has the form
@@
 $$|Dψ(z₁,…,z_R)〉=\pmatrix{|Dψ'(z₁)〉&&\cr&\ddots&\cr&&|Dψ'(z_R)〉},$$
-where the empty elements in row $r$ are null kets in the subspace of the component $|ψ'(z_r)〉$.
-The residual ket thus becomes
+where the empty elements in row $r$ are null kets in the subspace of the component $|ψ'(z_r)〉$.  The ket from Equation~\residual\ thus becomes
 $$i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉=
-	\pmatrix{|Dψ'(z₁)〉-H|ψ'(z₁)〉\cr\vdots\cr |Dψ'(z_R)〉-H|ψ'(z_R)〉},$$
+	\pmatrix{|Dψ'(z₁)〉{\dot z}₁(z)-H|ψ'(z₁)〉\cr\vdots\cr |Dψ'(z_R)〉{\dot z}_R(z)-H|ψ'(z_R)〉},$$
 where the rows are mutually orthogonal.  Therefore, the least squares problem can be solved separately for each component, independently minimising 
-$$\left‖|Dψ'(z_r)〉-H|ψ'(z_r)〉\right‖².$$
-If the system is a BEC, and the components are coherent states, this problem is equivalent solving the Gross-Pitaevskii Equation.
-Heller and others have investigated it for general systems with coherent state components, and found that the amplitudes obey something very like classical dynamics.
-Therefore variational dynamics with sparse components is a waste of time: its results should be very similar to truncated Wigner, and, if they differ, there won't be much reason to believe one rather than the other.
+$$\bigl‖|Dψ'(z_r)〉{\dot z}_r(z)-H|ψ'(z_r)〉\bigr‖².$$
+If the system is a BEC, and the components $|ψ'(z_r)〉$ are coherent states with amplitudes $z_r$, this problem reduces to the Gross-Pitaevskii Equation.
+Heller and others have investigated the problem for general systems with coherent state components, and found that the amplitudes obey something very like classical dynamics.
+Therefore variational dynamics with sparse components is a waste of time: its results should be similar to those of the truncated Wigner method, and, if they differ, there won't be much reason to believe one rather than the other.
 
@@
 I don't know whether both conditions were ever satisfied in the same simulation: it would be interesting to find out.
+
+\beginsection{Modifications to Dirac-Frenkel}
 
@@
 where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
-In the limit of large $R$, where the sum approximates an integral over the amplitude $α$, any state can be approximated arbitrarily well by this form.
-The continuous phase-space expansion is the topic of the next section, followed by a discussion of the discrete expansion operator $|A〉$.
+The form on the left is more complicated than the $|A〉d$ one, but it is an analytic function of the parameters, which is convenient for derivatives and optimisation.
+Wavelet theorists have studyed the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.
+This will be discussed later.
+
+In the limit of large $R$, this sum approximates an integral over the amplitude $α$, and any state can be approximated arbitrarily well by this form.
+The continuous phase-space expansion is the topic of this and the next section.
 Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.

### 2013-10-31 paper.tex
writing

@@
 In this case, $|U〉$ will be a row of ${\rm rank}|A〉$ kets, and $S$ will be ${\rm rank}|A〉×R$.}.
-In this paper, things written as kets have a finite number of infinite dimensional columns, and bras are the other way round.
+In this paper, things written as kets are linear mappings from other things to quantum states.
+The other things might be finite vectors of complex numbers, in which case the ket can be thought of as a matrix, or they might be functions from an infinite dimensional space, in which case this is less useful.
+Bras are the other way round: they map states to things other than states.
+Ordinary bras and kets are a special case.
 
@@
 The form on the left is more complicated than the $|A〉d$ one, but it is an analytic function of the parameters, which is convenient for derivatives and optimisation.
-Wavelet theorists have studyed the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.
+Wavelet theorists have studied the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.
 This will be discussed later.
 
-In the limit of large $R$, this sum approximates an integral over the amplitude $α$, and any state can be approximated arbitrarily well by this form.
+In the limit of large $R$, this sum approximates an integral over the amplitude $α$; any state can be approximated arbitrarily well by this form.
 The continuous phase-space expansion is the topic of this and the next section.
@@
 
-Expansion over coherent states is a linear operator from functions of a complex variable to quantum states,
+Expansion over coherent states is a linear operator from functions $f$ of a complex variable to quantum states $\G f$,
 $$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
@@
 When norms of functions are written below, the norm is the one derived from this inner product.
-In this paper, $\G f$ denotes the ket that results from applying the linear operator $\G$ to the phase space distribution $f$; application and composition of linear operators is shown by juxtaposition, and functions whose values are quantum states are written as kets, usually with a greek letter.
 
@@
 
-\midinsert \XeTeXpicfile tysv.pdf width \hsize
+\midinsert %\XeTeXpicfile tysv.pdf width \hsize
 Figure 3 \endinsert

### 2013-11-07 paper.tex
writing

--- 2013-10-31 paper.tex
@@
 \def\cite#1{$[\hbox{\tt #1}]$}
+\def\T{^{\rm T}}
 
@@
 This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.
-To test how well a wave function approximates the dynamics of a system, the difference between the two sides of Schrödinger's equation can be measured by the Hilbert space norm.
+To measure how well a wave function $|ψ(t)〉$ approximates the dynamics of a system, we will compare the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$; these are both vectors in Hilbert space, and the norm defines a distance between them.
 It remains to choose a class of approximants.
@@
 %% Overview
+
+Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem.
+The distinction between them is a bit subtle.
+A problem is ill-conditioned if the solution is very sensitive to the data with which the problem is stated: the canonical example is computing $\sin(x)$ for $x=10⁶$, whence a relative error of $π×10^{-6}$ in $x$ causes a 100\% error in $\sin(x)$.
+The key thing to note is that this has nothing to do with how $\sin(x)$ is computed—the error in $x$ could be due to measurement uncertainty, in which case there would be no way to calculate $\sin(x)$: we'd need to measure it directly.
+Condition is a property of problems.
+Stability is a property of solutions.
+For example, if we asked a computer to calculate $1-\cos(x)$ in the obvious way, with $x$ between 0 and $10^{-10}$, the result might vary randomly around 0, in a range determined by the precision with which the computer stored numbers.
+But this problem is not intrinsically hard: at any reasonable level of precision, it amounts to finding $x²/2$.
+The method of subtracting a numerical value of $\cos(x)$ from 1, which generates random numbers, is unstable.
 
@@
 They call this type of expansion a Gabor frame, or a windowed Fourier transform.
-They have prooved several results that are not well known in physics: for example, that discrete sets of coherent states are complete, not just the continuum.
+They have proved several results that are not well known in physics: for example, that discrete sets of coherent states are complete, not just the continuum.
 They have also studied overcomplete sets of vectors in general, under the field of frame theory.
@@
 $$|ψ〉=c₁|a₁〉+c₂|a₂〉+⋯+c_R|a_R〉.$$
-This can be written as a matrix product $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}^{\rm T}$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.
+This can be written as a matrix product $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}\T$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.
 Alternatively, the kets $|a_r〉$ can be regarded as infinite dimensional columns, and $|A〉$ as a matrix, representing a linear operator from ${\bf C}^R$ into ket space.
-This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}^{\rm T}$, that maps kets back to column vectors.
+This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}\T $, that maps kets back to column vectors.
 Under the rules of matrix multiplication, the product
 $$〈A|A〉=\pmatrix{〈a₁|a₁〉&&〈a₁|a_R〉\cr &\ddots&\cr 〈a_R|a₁〉&&〈a_R|a_R〉}$$
-is an operator on ${\bf C}^R$, and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ is a Hilbert space operator.
+operates on ${\bf C}^R$, and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ operates on Hilbert space.
 Other matrix products work too; provided the $|a_r〉$ are linearly independent, $|A〉$ will have a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix%
@@
 $$M=\pmatrix{0&|A〉\cr 〈A|&0}$$
-can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.
+acts on columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}\T $ to give columns of the same form.
 The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.
@@
 $$|Ψ(t)〉=|ψ(z(t))〉,$$
-or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}^{\rm T}$ is a vector of complex parameters.
+or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}\T $ is a vector of complex parameters.
 This poses two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.
@@
 $$i\hbar|Dψ(z)〉Dz=H|ψ(z)〉.$$
-On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.
-This equation is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.
-In other words, the restricted form of the approximation does not include the solution to Schrödinger's equation.
+On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}\T $.
+This equation is unlikely to have an exact solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.
+In other words, the restricted form of the approximation does not include the exact solution to Schrödinger's equation.
 We can, however, look for a least-squares solution, a vector $\dot z(z)$ that minimises the residual
@@
 	=|A〉d,\tag\Ad$$
-where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
-The form on the left is more complicated than the $|A〉d$ one, but it is an analytic function of the parameters, which is convenient for derivatives and optimisation.
-Wavelet theorists have studyed the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.
+where $A=\pmatrix{α₁&⋯&α_R}\T $ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.
+The form on the left is more complicated than the $|A〉d$ form, but it is an analytic function of $φ$ and $α$, which is convenient for derivatives and optimisation.
+Wavelet theorists have studied the conditions on the set of amplitudes $A$ that allow any state to be expanded in this form, and the numerical stability of those expansions as the density of ampltiudes decreases towards the minimum for completeness.
 This will be discussed later.
@@
 We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
-Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
+Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}\T ,$
 such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
@@
 The derivative of this with respect to the weights is
-$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U^{\rm T}Uw-U.$$
-(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c^{\rm T}ab$.)
+$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U\T Uw-U.$$
+(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c\T ab$.)
 Unfortunately,
-$$\int (U^{\rm T}U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
+$$\int (U\T U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
 	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
@@
 $${\cal E} = \int\left(e^{-x²}-∑_rd_re^{(x-x_r)²}\right)²;$$
-the only change to the normal equations is that the right hand side $\pmatrix{1&⋯&1}^{\rm T}$ is replaced by a constant multiple of 
+the only change to the normal equations is that the right hand side $\pmatrix{1&⋯&1}\T $ is replaced by a constant multiple of 
 $$b=\pmatrix{e^{-x₁²/2}&⋯&e^{-x_R²/2}}.\tag\rhs$$

### 2013-11-07 dual.tex
writing

--- 2013-10-05 dual.tex
@@
 
-\midinsert\XeTeXpicfile approx.pdf height \hsize rotated 90 \endinsert
+%\midinsert\XeTeXpicfile approx.pdf height \hsize rotated 90 \endinsert
 
@@
 
-\midinsert\XeTeXpicfile tych.pdf width\hsize \endinsert
+%\midinsert\XeTeXpicfile tych.pdf width\hsize \endinsert
 

### 2013-11-07 paper.tex

@@
 
-In the limit of large $R$, this sum approximates an integral over the amplitude $α$, and any state can be approximated arbitrarily well by this form.
+In the limit of large $R$, this sum approximates an integral over the amplitude $α$; any state can be approximated arbitrarily well by this form.
 The continuous phase-space expansion is the topic of this and the next section.

### 2013-11-19 paper.tex
writing

@@
 \def\G{|Γ〉}
+\def\A{|A〉}
 \def\Gx{|Γ_χ〉}
@@
 
-When they realised that positions and momenta are essentially uncertain, the first quantum physicists were bound to discover the gaussian wave packet by seeking conditions that produce a normal distribution of probability.
-They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.
-These problems were also solved by the gaussian wave packet.
-With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.
-
+The Gaussian wave packet is a ubiquitous idea in quantum mechanics.
+Its discovery was inevitable: people were provoked by understanding that observables are essentially uncertain to study physical states in which they are normally distributed.
+The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillation of classical mechanics.
+In this potential, there is a natural width for the packet, which ensures its shape remains constant during its motion.
+This width can also be set by requiring equipartition of energy and the uncertainty principle.
+Wavepackets with this natural width are coherent states.
+
+It has also been a productive tool.
 The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.
 This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.
-To measure how well a wave function $|ψ(t)〉$ approximates the dynamics of a system, we will compare the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$; these are both vectors in Hilbert space, and the norm defines a distance between them.
-It remains to choose a class of approximants.
-The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.
+If the problem is to find a wave function $|ψ(t)〉$ that evolves according to a Hamiltonian $H$, the quality of approximation can be measured by comparing the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$.
+These are both vectors in Hilbert space, and can be compared by the norm of their difference.
+The gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.
 
@@
 Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.
-When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics, and have since been applied to other forms of condensed matter.
+When very coherent matter, in the form of atomic Bose-Einstein condensates, became easy to access in the lab, phase space methods were sucessfully applied to calculate its dynamics.
+They have since been applied to other forms of condensed matter.
 They have solved many problems, but are limited by intractable convergence issues.
 
-In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.
+In the 1970s, Heller put variational mechanics and phase space expansions together.
+He proposed to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.
 This is an appealing approach.
-Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.
-The experience of quantum optics shows that, in many systems, the approximation converges with a reasonable number of components.
+Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases, and the finite superposition converges to a continuous phase space expansion.
+The experience of quantum optics shows that, in many systems, the expansion converges when the superposition has a reasonable number of terms.
 Also, variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.
@@
 
-Since Heller suggested it, there have been several attempts to apply this numerically, but none has entirely succeeded.
+There have been several attempts to implement Heller's proposal, but none has entirely succeeded.
 In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
 %% Consequence
-These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.
-The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.
-A systematic approach to regularisation is required in order to approach these problems variationally.
-The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.
+These attempts have aimed to solve chemistry problems, involving dozens of atoms.
+Their failure to converge limited their precision, but useful results were obtained from superpositions with a few components, and ad-hoc methods to improve their stability.
+The problems that people attempt with phase space methods come from atom optics or condensed matter, and involve thousands or millions of atoms.
+In these systems, the instability would prevent any useful information being obtained.
+Therefore a systematic approach to regularisation is required in order to approach these problems variationally.
+The prospective rewards are significant: many-particle problems include high temperature superconductivity and the dynamics of the early universe.
 
@@
 
-Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem.
-The distinction between them is a bit subtle.
+Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem, and the distinction between them is a bit subtle.
 A problem is ill-conditioned if the solution is very sensitive to the data with which the problem is stated: the canonical example is computing $\sin(x)$ for $x=10⁶$, whence a relative error of $π×10^{-6}$ in $x$ causes a 100\% error in $\sin(x)$.
@@
 The method of subtracting a numerical value of $\cos(x)$ from 1, which generates random numbers, is unstable.
+Experiments with many-particle systems suggest that phenomena such as interference and superconductivity are robust, and changing the system slightly does not greatly change the phenomena that result.
+So computing the properties of these systems should be a well-conditioned problem.
+However, methods of solving them variationally, using coherent states, have been very unstable.
 
@@
 The first problem occurs when quantum states are expanded over a finite set of coherent states.
-There is a natural way to expand a state as an integral over the coherent states, with a weight function that varies smoothly with the coherent amplitude.
+There is a natural way to expand a state as an integral over the phase space of coherent states, with a weight function that varies smoothly with the coherent amplitude.
 When this function is sampled at a discrete set of coherent amplitudes and the integral approximated by a sum, the state is approximated to some precision.
-Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number—for instance, it might be a coherent state.
+Part of the error in this appoximation is due to the finite set of coherent states having bounded amplitude, wheras the state being approximated might have components with arbitrarily large particle number.
 When expansion weights for the finite set of coherent states are set by least squares, components with large particle number are approximated by adjacent amplitudes having large weights, with opposite signs so that components with low particle numbers cancel.
@@
 
-\beginsection{Gabor frames}
+\beginsection{Frames}
+
+Because we are after numerical methods, our goal is for expansions such as $\G f$ and $\A d$ to be numerically stable.
+This means, for one thing, that a small change in $f$ should cause a small change in $\G f$.
+The coherent states that compose $\G$ are overcomplete, so fixing the value of $\G f$ does not fix the value of $f$.
+However, in order to compute with this expansion, we need to find a way of solving for $f$ such that a small change in $\G f$ causes a small change in the  $f$ that is the result of our method.
+
+There is a precise way to state these conditions.
+For the expansion $\G$ to be stable, its columns must form a frame\cite{1992-Daubechies-Ten}.
+This means, in the continuous case, that there are constants $A$ and $B$ such that
+$$ A‖|ψ〉‖²≤\int |〈γ|ψ〉|²\,dγ≤B‖|ψ〉‖²,$$
+for all kets $|ψ〉$.
+In the discrete case, the integral is replaced by a sum.
+Because the operator $\G$ is linear, the ket $|ψ〉$ could be a difference $|φ₁〉-|φ₂〉$.
+In this case, the right hand inequality guarantees that a small change in a ket will produce a small change in the components $〈γ|ψ〉$, which the left hand inequality says tht a small change in the components will produce a small change in the expanded ket.
+
+In the particular case of the coherent states, we have
+$$ \int |〈γ|ψ〉|²\,dγ=〈ψ|\int |γ〉〈γ|\,d²γ\,|ψ〉=π,$$
+so the frame condition is satisfied, with $A=B=π$.
+A set of this kind, where the upper and lower frame bounds are the same, is known as a tight frame.
+Mathematicians have known this for a long time, and refer to the coherent states, and other frames that result from translating and modulating functions that aren't necessarily Gaussian, as Gabor frames.
+
+(Talk about dual frames, minimal norm expansions.)
+
+In the numerical case, the number of frame kets is finite.
+Unless $\A$ is singular, the frame bounds always exist, with $A$ and $B$ something like the largest and smallest singular values of $\A$.
+So we need to play a bit loose with the mathematical definitions.
+For numerical purposes, a bound of the order $10^{15}$ might as well be infinite, and one of the order $10^{-15}$ might as well be zero.
+Also, a finite set of states can not  be a frame for the whole of Hilbert space: there will always be directions orthogonal to all the components, for which $∑_r|〈α_r|ψ〉|²=0$.
+So we need to think carefully about what space $|ψ〉$ can lie in.
+The obvious choice would be the range of $\A$, the span of the $|α_r〉$.
+
+But this causes problems with the lower frame bound.
+The amplitudes of the coherent states will be bounded: suppose $|α_r|≤a$.
+A finite set of coherent states with different amplitudes are linearly independent, so the simultaneous equations $〈n|A〉d=0$, where n ranges from $0$ to $R-1$, have a solution.
+This means that $\A d$ is a state of the form $c_R|R〉+c_{R+1}|R+1〉+⋯$.
+Now, the component of $|n〉$ in $|a_r〉$ has modulus no greater than $a^n/\sqrt{n!}$, so $|c_n| ≤ a^n/\sqrt{n!} ‖c‖₁$.
+This means that, in the limit of dense amplitudes, where $R$ increases while $a$ stays constant, we can find vectors $d$ such that $‖\A d‖$ tends to zero.
+(Fix that to argue the right way round.)  There are states in the span of $\A$ with large particle number, that barely overlap any of the states $|α_r〉$, and have very small components along the frame directions.
+So, from a numerical point of view, dense, bounded sets of coherent states fail to be frames.
+If the $α_r$ extended over the whole of phase space, the set would include some larger amplitudes for which $〈α_r|ψ〉$ was not small.
+
+The obvious solution is to not attempt to expand the entire span of $\A$, but first project onto some subspace for which the $|α_r〉$ are a frame with numerically reasonable bounds.
+This is what regularisation methods do.
+In particular, singular vectors of $\A$ with large singular values span a space for which the states are a frame.
+In practice, these correspond to states with limited particle number.
+So this might be why projecting onto smal l particle number makes simulations more stable: the ensemble of coherent states is a frame for that space!
+It also suggest that, for accurate and stable expansions, we want coherent amplitudes large enough that the components don't have a great overlap with the state being expanded.
+
+It is interesting to consider the squeezed states.
+Mathematicians have thought about two types of frames: Gabor frames, generated by translation and momentum kicks, and wavelet frames, generated by translation and limited types of squeezing.
+It is unclear whether anyone has studied if the squeezed states are a frame.
+If they are not, it might explain why numerical expansions over squeezed states have proven to be so unstable.
 

### 2014-04-22 paper.tex
big pre-backup push

@@
 \def\cite#1{$[\hbox{\tt #1}]$}
+\def\define#1{#1}
 \def\T{^{\rm T}}
@@
 Its discovery was inevitable: people were provoked by understanding that observables are essentially uncertain to study physical states in which they are normally distributed.
-The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillation of classical mechanics.
-In this potential, there is a natural width for the packet, which ensures its shape remains constant during its motion.
-This width can also be set by requiring equipartition of energy and the uncertainty principle.
-Wavepackets with this natural width are coherent states.
-
-It has also been a productive tool.
+The Gaussian wavepacket also crops up as the quantum state which, placed in a quadratic potential, best approximates the harmonic oscillations of classical mechanics.
+The potential sets a natural width for the packet, which causes its shape to stay constant as it moves.
+This width is also characterised by equipartition of energy and the uncertainty principle.
+Wavepackets with this natural width are called coherent states.
+
+Wave packets have been a productive tool.
 The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.
@@
 These are both vectors in Hilbert space, and can be compared by the norm of their difference.
-The gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.
+Gaussian wave packets have often been used as variational approximants, because they are defined by their width and their central position and momentum, useful properties to determine for any quantum state.
 
 The invention of the laser in 1963 called for a quantum theory of intense radiation.
-Laser beams are hard to analyse because their quantum descriptions are so large, comprising up to $10^{18}$ photons distributed across thousands of modes.
-The theory was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.
-Laser light is very coherent, so these expansions often represent it efficiently.
+The difficulty of analysing a laser beam stems from the size of its quantum description, up to $10^{18}$ photons distributed across thousands of modes.
+This was made managable by Glauber and Sudarshan, who showed that any quantum state can be expanded as a superposition of coherent states.
+Laser light is very coherent, so these expansions often represent it efficiently, converging with a reasonable number of states in the superposition.
 Moyal(?) developed an efficient way to calculate dynamics, leading to a family of phase space Monte-Carlo methods.
@@
 
-\beginsection{Continuous phase space expansions}
+One modification would be to use a norm derived from lower order moments of the creation and annihilation operators\cite{Drummond-private}.
+In this case, there would not be a direct link between the variational parameters and a state in Hilbert space.
+Instead, the superposition of coherent states would be a way to record moments of the system, which would be propagated so that the rates of change of the moments had the expected values.
+It is unclear why recording the moments indirectly, by an ensemble of coherent amplitudes, would be better than recording the moments themselves.
+Several methods already exist for propagating the moments of a quantum optical system
+: these usually assume that some higher-order moments factor in terms of lower-order ones, in order to obtain a system of ordinary differential equations for a finite set of moments.
+This assumption is at least explicit, and some equivalent assumption would have to be made implicitly if the moments were derived from an ensemble of coherent amplitudes.
+It is unclear why this would be a better approximation to quantum mechanics than the differential equations obtained by assuming the moments factor.
+
+\beginsection{Coherent states}
+
+TODO: Define “phase space” at some point.
 
@@
 
-In the limit of large $R$, this sum approximates an integral over the amplitude $α$; any state can be approximated arbitrarily well by this form.
-The continuous phase-space expansion is the topic of this and the next section.
-Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.
-This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.
-
-Expansion over coherent states is a linear operator from functions $f$ of a complex variable to quantum states $\G f$,
-$$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
-The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.
-The function $f$ is such that $e^{-½|α|²}f(α)$ is square-integrable.
-Bargmann\cite{pam-14-187} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.
-When norms of functions are written below, the norm is the one derived from this inner product.
-
 The state $|α〉$ is a coherent state with complex amplitude $α$.
@@
 whence
-$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.$$
+$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.\tag\csolap$$
 When kets are considered as vectors, the angle between coherent states satisfies
@@
 $$θ=|β-α|+O(|β-α|²).$$
-This has a simple geometric analogy where phase space is mapped locally onto the surface of the unit sphere, the coherent states are vectors from the origin to the surface, and $|α-β|$ is a spherical angle.
-The analogy fails when the angles between vectors are no longer small; for example, it is misleading in regard to the relation between $|α〉$ and $∂_α|α〉$.
+This has a geometric analogy.
+The normalised kets form a hypersphere in ket space.
+A two-dimensional manifold of normalised kets, such as the coherent states, must be locally similar to the unit sphere.
+It turns out that phase space is mapped into ket space without local distortion: the relation between a coherent ket and its neighbours is as if phase space were the surface of a sphere.
+This is only the case when the angles between vectors are small; for example, it is misleading in regard to the relation between $|α〉$ and $∂_α|α〉$.
 
@@
 
-Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
-$$〈 Γ⁺|ψ〉(α)=π^{-½}e^{½|α|²}〈 α|ψ〉.$$
-Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.
-It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.
-The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.
-(Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.
-
-If, following Bargmann, we identify state space with the entire functions, $|Γ〉$ is identified with the projector $〈 Γ⁺|Γ〉$.
-Therefore, although $|Γ〉$ is singular, the singularity is easily resolved.
-If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be done simply enough by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.
-And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
-So it is curious that the discrete version of this problem has caused so much trouble when solved by least squares.
-
-%% Consequences of Γ being a projection
-
-The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
-$$π^{-½}〈β|Γ〉f=π^{-1}\int 〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}\int 〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
-This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.
-There are many theorems to the effect that real convolution operators are badly conditioned.
-It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.
-I think the loophole is that of real functions can only oscillate at a limited rate, so that their Fourier transforms have finite norm.
-Complex functions have space in the Argand plane to oscillate arbitrarily rapidy.
-And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.
+Equation~\csip\ has a consequence for expansions.
+Suppose a coherent state $|β〉$ is expanded in the form in Equation~\Ad, with components such that $|β-α_r|>δ$, and error $‖|β〉-|ψ〉‖<ε$.
+Then $|〈β|ψ〉|>1-ε$, but $|〈β|ψ〉|=|∑〈β|α_r〉d_r|≤e^{-{δ²\over 2}}∑|d_r|$.
+Therefore, as the expanded state moves away from the components, the coefficients must grow as $e^{δ²\over 2}$.
+The error condition requires that $‖|ψ〉‖<1+ε$, so these components must almost entirely cancel each other out.
+This bodes badly for the numerical stability of such expansions: the details are explored below.
 
@@
 
-Because we are after numerical methods, our goal is for expansions such as $\G f$ and $\A d$ to be numerically stable.
-This means, for one thing, that a small change in $f$ should cause a small change in $\G f$.
-The coherent states that compose $\G$ are overcomplete, so fixing the value of $\G f$ does not fix the value of $f$.
-However, in order to compute with this expansion, we need to find a way of solving for $f$ such that a small change in $\G f$ causes a small change in the  $f$ that is the result of our method.
-
-There is a precise way to state these conditions.
-For the expansion $\G$ to be stable, its columns must form a frame\cite{1992-Daubechies-Ten}.
-This means, in the continuous case, that there are constants $A$ and $B$ such that
-$$ A‖|ψ〉‖²≤\int |〈γ|ψ〉|²\,dγ≤B‖|ψ〉‖²,$$
+TODO: say that the default norm in ${\bf C}^n$ is the 2-norm.
+
+In order to support numerical methods, the expansion $|ψ〉=\A d$ needs to be numerically stable.
+This means, most obviously, that changing $d$ by a small amount should cause a small change in $\A d$.
+This is clearly the case.
+The coherent states that compose $\A$ are normalised, so clearly $‖\A d‖≤R‖d‖$, whence $‖\A‖≤R$.
+
+TODO: tighten this bound (can we get $‖\A‖≤\sqrt{R}$?)
+
+The opposite condition, that a small change in $|ψ〉$ should cause a small change in $d$, is trickier.
+Finite sets of coherent states are linearly independent in the mathematical sense, so the equation $\A d₀=0$ has no solution, and the operator $\A$ is invertible.
+This is not true in the numerical sense: given a typical set of amplitudes $\{α_r\}$, there will be vectors $d₀$ such that $\A d₀$ lies within rounding error of zero.
+We can add any multiple of $d₀$ to $d$ without changing the computed $|ψ〉$ at all, so the problem of finding $d$ is utterly unstable.
+The best we can hope for is to find a specific expansion that is numerically stable.
+
+As a first step, we can require that the components $〈A|ψ〉$ are stable.  The conventional way to formulate that is to require there are finite constants $P$ and $Q$ such that
+$$ P‖|ψ〉‖²≤‖〈A|ψ〉‖²≤Q‖|ψ〉‖²,$$
 for all kets $|ψ〉$.
-In the discrete case, the integral is replaced by a sum.
-Because the operator $\G$ is linear, the ket $|ψ〉$ could be a difference $|φ₁〉-|φ₂〉$.
-In this case, the right hand inequality guarantees that a small change in a ket will produce a small change in the components $〈γ|ψ〉$, which the left hand inequality says tht a small change in the components will produce a small change in the expanded ket.
-
-In the particular case of the coherent states, we have
-$$ \int |〈γ|ψ〉|²\,dγ=〈ψ|\int |γ〉〈γ|\,d²γ\,|ψ〉=π,$$
-so the frame condition is satisfied, with $A=B=π$.
-A set of this kind, where the upper and lower frame bounds are the same, is known as a tight frame.
-Mathematicians have known this for a long time, and refer to the coherent states, and other frames that result from translating and modulating functions that aren't necessarily Gaussian, as Gabor frames.
-
-(Talk about dual frames, minimal norm expansions.)
+In this case, the set comprising kets $|α_r〉$ is known as a \define{frame}.
+Strictly, the mapping $〈A|$ from kets to components is called the frame operator; this paper will often refer to $\A$ as a frame, meaning that its columns form a frame.
+The constant $P$ is called the lower frame bound, and $Q$ the upper frame bound.
+It will turn out that, in order to form stable expansions, these bounds ought to be close.
+The ideal case, where they are equal, is known as a tight frame.
+
+There is a complication that we'll ignore for now.
+The state $|ψ〉$ lies in an infinite dimensional Hilbert space, in which there are many directions that are orthogonal to every $|α_r〉$.
+When $|ψ〉$ points along such a direction, $〈A|ψ〉=0$.
+So no finite set of kets is a frame for Hilbert space, but only for some subspace.
+We will discuss this in detail below.
+
+It can be shown that if $\A$ is a frame, the operator $\A〈A|$ is invertible\cite{1992-Daubechies-Ten}.  In this case, the set of kets
+$$|F〉=(\A〈A|)^{-1}|A〉=\pmatrix{(\A〈A|)^{-1}|a₁〉&⋯&(\A〈A|)^{-1}|a_R〉}
+	=\pmatrix{|φ₁〉&⋯&|φ_R〉}
+$$
+is also a frame, called the dual frame of $|A〉$.
+These kets have the property that $|A〉〈F|=|F〉〈A|=1$, the identity operator on the space spanned by the frame.
+In other words, the components of $|ψ〉$ along the directions of the dual frame $|F〉$ are the coefficients of an expansion over the frame $|A〉$, and visa-versa.
+If $|α_r〉$ formed an orthonormal basis, we would have $|F〉=|A〉$, and $|A〉〈A|=1$.
+In this sense, frames are a generalisation of orthonormal bases.
+For a tight frame, $|F〉={1\over P}|A〉$, and the analogy is especially close.
+
+For finite frames, expansions using dual kets have some special properties.
+For a start, the dual expansion is the one where the vector of coefficients has the least norm.
+In fact, if $|ψ〉=|A〉d$, and $c=〈F|ψ〉$, then $‖d‖²=‖c‖²+‖d-c‖²$\cite{1989-sir-31-626}.
+This suggests that $d-c$ is forced to be orthogonal to $c$.
+
+The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.
+We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
+We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
+Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}\T ,$
+such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
 
@@
 
-Many of the above results for the operator $\G$ are true of expansions over frames in general.
-We will treat the finite case, because every frame will eventually have to be represented on a computer.
-So suppose the set of kets $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a frame for some subspace of Hilbert space.
-
-The theory of frames is traditionally presented in terms of the frame operator,
-$$S=∑_{r=1}^R |a_r〉〈a_r|={\rm Tr}(|A〉〈A|).$$
-In the case that the columns of $|A〉$ are a frame, $S$ is invertible.  The set of vectors
-$$|F〉=S^{-1}|A〉=\pmatrix{S^{-1}|a₁〉&⋯&S^{-1}|a_R〉}
-	=\pmatrix{|φ₁〉&⋯&|φ_R〉}
-$$
-is also a frame, and is known as the dual frame of $|A〉$.  It has the property that $|A〉〈F|$ is the identity operator on the space.
-
-For finite frames, expansions using dual kets have some special properties.
-For a start, the dual expansion is the one where the vector of coefficients has the least norm.
-In fact, if $|ψ〉=|A〉d$, and $c=〈F|ψ〉$, then $‖d‖²=‖c‖²+‖d-c‖²$\cite{1989-sir-31-626}.
-This suggests that $d-c$ is forced to be orthogonal to $c$.
-
-The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.
-We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.
-We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.
-Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}\T ,$
-such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.
+\beginsection{Phase space as a Gabor frame}
+
+In the limit of large $R$, this sum approximates an integral over the amplitude $α$; any state can be approximated arbitrarily well by this form.
+The continuous phase-space expansion is the topic of this and the next section.
+Many problems in quantum optics have been solved by a family of Monte-Carlo methods, using similar expansions, where the amplitudes $α_r$ are propagated independently.
+This gives some grounds to hope that the states of interesting many-particle systems can be expanded with a reasonable number of components $R$.
+
+Expansion over coherent states is a linear operator from functions $f$ of a complex variable to quantum states $\G f$,
+$$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
+The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.
+The function $f$ is such that $e^{-½|α|²}f(α)$ is square-integrable.
+Bargmann\cite{pam-14-187} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.
+When norms of functions are written below, the norm is the one derived from this inner product.
+
+In the particular case of the coherent states, we have
+$$ \int |〈γ|ψ〉|²\,dγ=〈ψ|\int |γ〉〈γ|\,d²γ\,|ψ〉=π,$$
+so the frame condition is satisfied, with $A=B=π$.
+A set of this kind, where the upper and lower frame bounds are the same, is known as a tight frame.
+Mathematicians have known this for a long time, and refer to the coherent states, and other frames that result from translating and modulating functions that aren't necessarily Gaussian, as Gabor frames.
+
+Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
+$$〈 Γ⁺|ψ〉(α)=π^{-½}e^{½|α|²}〈 α|ψ〉.$$
+Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.
+It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.
+The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.
+(Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.
+
+If, following Bargmann, we identify state space with the entire functions, $|Γ〉$ is identified with the projector $〈 Γ⁺|Γ〉$.
+Therefore, although $|Γ〉$ is singular, the singularity is easily resolved.
+If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be done simply enough by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.
+And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.
+So it is curious that the discrete version of this problem has caused so much trouble when solved by least squares.
+
+%% Consequences of Γ being a projection
+
+The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
+$$π^{-½}〈β|Γ〉f=π^{-1}\int 〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}\int 〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
+This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.
+There are many theorems to the effect that real convolution operators are badly conditioned.
+It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.
+I think the loophole is that of real functions can only oscillate at a limited rate, so that their Fourier transforms have finite norm.
+Complex functions have space in the Argand plane to oscillate arbitrarily rapidy.
+And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.
 
@@
 
-%\topinsert\XeTeXpicfile circle-1.pdf \endinsert
+\topinsert\XeTeXpicfile circle-1.pdf \endinsert
 
@@
 So one solution  would be to find the best expansion of $e^{-x²}$, instead of 1, then adjust the weight for $x_i$ by either $e^{-x_i²}$ or $〈e^{-(x-x_i)²},e^{-x²}〉$.
-%\endinsert
+\endinsert
 
@@
 
-\beginsection{Regularisation of the discrete expansion problem}
+\beginsection{The discrete variational frame}
+
+In the variational problem, the derivative $-iH|ψ(z)〉$ is expanded over the frame $|Dψ(z)〉$.  For the coherent state approximant of Equation~?,
+$$ |Dψ(z)〉=\pmatrix{1 & a†}|ψ(z)〉.$$
+It follows that
+$$ 〈∂₁ψ|∂₁ψ〉=〈ψ|ψ〉=e^{2\Re φ+|α|}
+	\qquad 〈∂₂ψ|∂₂ψ〉=(1+|α|²)〈ψ|ψ〉,$$
+and
+$$\qquad 〈∂₁ψ|∂₂ψ〉=e^{2\Re φ+|α|}〈α|a†|α〉=α*〈ψ|ψ〉.$$
+The angle $θ$ between $|∂₁ψ〉$ and $|∂₂ψ〉$, such that
+$$ |〈∂₁ψ|∂₂ψ〉|=‖|∂₁ψ〉‖‖|∂₂ψ〉‖\cos θ,$$
+satisfies $\cot θ=|α|$.  The partial derivatives are orthogonal for the vacuum state $α=0$, but become nearly parallel as $|α|$ increases.  This will increase the condition number of $|Dψ〉$.
+
+\beginsection{Regularisation of the discrete problems}
 %% Picard plots
@@
 
-\midinsert %\XeTeXpicfile tysv.pdf width \hsize
+\midinsert \XeTeXpicfile tysv.pdf width \hsize
 Figure 3 \endinsert
@@
 
-%\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
-
-%\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
+\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
+
+\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
 
@@
 
-%\topinsert \XeTeXpicfile geometry-1.pdf
+\topinsert \XeTeXpicfile geometry-1.pdf
 A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
@@
 This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
-%\endinsert
-
-%\topinsert \XeTeXpicfile cancel.pdf width \hsize
+\endinsert
+
+\topinsert \XeTeXpicfile cancel.pdf width \hsize
 The vacuum state was expanded over two coherent states, with amplitudes $±h$.
@@
 
-%\topinsert \XeTeXpicfile spsns.pdf width \hsize
-%Amplitudes of components \endinsert
-
-%\topinsert \XeTeXpicfile traj.pdf width \hsize
-%Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
-
-%\topinsert \XeTeXpicfile picard.pdf width \hsize
-%Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
-
-%\topinsert \XeTeXpicfile svnum.pdf width \hsize
-%Expected particle number for left singular vectors.
+\topinsert \XeTeXpicfile spsns.pdf width \hsize
+Amplitudes of components \endinsert
+
+\topinsert \XeTeXpicfile traj.pdf width \hsize
+Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
+
+\topinsert \XeTeXpicfile picard.pdf width \hsize
+Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
+
+\topinsert \XeTeXpicfile svnum.pdf width \hsize
+Expected particle number for left singular vectors.
 The svs with small sws have large particle number.
@@
 
-%\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
-
-%\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
+\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
+
+\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
 

### 2014-04-22 avsda.m

@@ -1,0 +1,48 @@
+% Compare approximation by 2*R components to R components and
+% their derivatives
+
+global N;  N = 30;  brackets
+R = 20;  a = 2*sqrt(0.5)*[1 1i]*randn(2,R);  gabor
+mesh = -5:0.2:5;  phasespace
+
+strides = [4 2 2 1];
+expansions = {DA(:,1:4:end) A(:,1:2:end) DA(:,1:2:end) A};
+
+figure
+ens = a;
+plot(real(ens), imag(ens), 'ok');  hold on
+ens = a(1:2:end);
+plot(real(ens), imag(ens), 'ok', 'MarkerFaceColor', [0.7 0.7 0.7])
+ens = a(1:4:end);
+plot(real(ens), imag(ens), 'ok', 'MarkerFaceColor', 'k')
+set(gca, 'DataAspectRatio', [1 1 1])
+
+
+figure
+for i=1:4
+    subplot(2,2,i)
+    M = expansions{i};  zplot(0:N, 0:N, eye(N+1)-M*pinv(M))
+    switch i
+        case 1
+            title('Expansion')
+            ylabel([num2str(0.5*R) ' vectors'])
+        case 2
+            title('Variation')
+        case 3
+            ylabel([num2str(R) ' vectors'])
+    end
+end
+
+figure
+for i=1:4
+    [U,S,V] = svd(expansions{i}, 'econ');
+    subplot(4,R+1,R*i+i-R)
+    ns = numel(diag(S));
+    semilogy(0:ns-1, diag(S), '+k')
+    for j=1:ns
+        subplot(4,R+1,R*i+i-R+j)
+        zplot(mesh, mesh, Aps'*U(:,j)); hold on
+        plot(real(a(1:strides(i):end)), imag(a(1:strides(i):end)), '.w')
+        set(gca,'DataAspectRatio',[1 1 1])
+    end
+end

### 2014-04-22 cnsv.m

@@ -1,0 +1,58 @@
+% Sample a state by inverting G on random amplitudes, then look for
+% unimportant singular directions of |Dpsi> along which the least singular
+% value of G varies rapidly.
+
+global N; N = 15; brackets
+R = 5;
+h = 0.01;
+c0 = nq*evan(-2,2);
+
+% Random sample amplitudes
+a = 2+sqrt(0.5)*randn(R,2)*[1; 1i];
+G = nq*evan(-0.5*abs(a).^2,a);
+ef = pinv(G)*c0;
+f = -0.5*abs(a).^2 + log(ef);
+
+figure; subplot(1,2,1)
+plot(real(a), imag(a), '*k');
+set(gca, 'DataAspectRatio',[1 1 1])
+for r = 1:R
+    text(real(a(r)), imag(a(r)), sprintf('  %d',r))
+end
+title('Amplitudes')
+
+[Ud,Sd,Vd] = svd([nq*evan(f,a) ndqr*evan(f,a)], 'econ');
+[Ug,Sg,Vg] = svd(nq*evan(-0.5*abs(a).^2,a), 'econ');
+
+ssdist = zeros(1,R);
+for j = 1:R
+    ssdist(j) = norm(Ud(:,1:j)*Ud(:,1:j)' - Ug(:,1:j)*Ug(:,1:j)');
+end
+
+figure; plot(1:R, ssdist, '+k')
+xlabel('dimension of subspaces')
+ylabel('operator distance between projectors')
+title('Distances between ranges of low-rank approximations to G and D\psi')
+    
+
+ngs = zeros(1,2*R);     % Norms of grad sv
+for p = 1:2*R
+    da = h*Vd(R+1:2*R,p);
+    Gp = nq*evan(-0.5*abs(a+da).^2, a+da);
+    Gm = nq*evan(-0.5*abs(a-da).^2, a-da);
+    A = [zeros(N+1), Gp-Gm; (Gp-Gm)', zeros(R)];
+    ev = [Ug(:,end); Vg(:,end)];
+    dsr = 0.5*ev'*A*ev;
+    da = 1i*da;
+    Gp = nq*evan(-0.5*abs(a+da).^2, a+da);
+    Gm = nq*evan(-0.5*abs(a-da).^2, a-da);
+    A = [zeros(N+1), Gp-Gm; (Gp-Gm)', zeros(R)];
+    ev = [Ug(:,end); Vg(:,end)];
[…]

### 2014-04-22 cohfit2.m

@@ -1,0 +1,27 @@
+function [z, rsdl] = cohfit2(cs, z0, wgt)
+% fit a superposition of coherent states to the fock expansion cs,
+% starting from z.  A term w*log(cond(dq(z))) is added to the cost
+% function in quadrature; there are no hard constraints.
+
+global nq ndqr nwgt
+
+R = 0.5*numel(z0);
+options = optimset('Display','notify');
+w0 = [real(z0), imag(z0)];
+w = fminunc(@(w)cost(w),w0,options);
+z = (w(1:2*R)+1j*w(2*R+1:4*R)).';
+f = z(1:R).'; a = z(R+1:2*R).';
+rsdl = norm(sum([nq*evan(f,a),-cs],2));
+                
+                
+function y = cost(w)
+% number weighted distance of z from state cs
+
+f = w(1:R)+1j*w(2*R+1:3*R); a = w(R+1:2*R)+1j*w(3*R+1:4*R);
+l = nq*evan(f,a);
+ndq = [l, ndqr*evan(f,a)];
+y = norm(nwgt*sum([l,-cs],2))^2 + wgt^2*log(cond(ndq));
+
+end
+
+end

### 2014-04-22 cohfit3.m

@@ -1,0 +1,28 @@
+function [z, rsdl] = cohfit3(cs, z0, wgt)
+% fit a superposition of coherent states to the fock expansion cs,
+% starting from z.  A term w*log(cond(dq(z))) is added to the cost
+% function in quadrature; there are no hard constraints.  This version uses
+% the nonlinear least square optimiser.
+
+global nq ndqr
+
+R = 0.5*numel(z0);
+options = optimset('Display','off');
+w0 = [real(z0), imag(z0)];
+w = lsqnonlin(@cost,w0,[],[],options);
+z = (w(1:2*R)+1j*w(2*R+1:4*R)).';
+rsdl = norm(sum([nq*evan(f,a),-cs],2));
+                
+                
+function y = cost(w)
+% Distance of z from state cs
+
+f = w(1:R)+1j*w(2*R+1:3*R); a = w(R+1:2*R)+1j*w(3*R+1:4*R);
+l = nq*evan(f,a);
+ndq = [l, ndqr*evan(f,a)];
+dcs = sum([l,-cs],2);
+y = [real(dcs); imag(dcs); wgt*log(cond(ndq))];
+
+end
+
+end

### 2014-04-22 cohfit4.m

@@ -1,0 +1,30 @@
+function [z, rsdl] = cohfit4(cs, z0, wgt)
+% fit a superposition of coherent states to the fock expansion cs,
+% starting from z.  A term w*log(cond(dq(z))) is added to the cost
+% function in quadrature; there are no hard constraints.  Cohfit4 uses the
+% global optimisation methods.
+
+global nq ndqr
+
+R = 0.5*numel(z0);
+options = optimset('Display','off');
+w0 = [real(z0), imag(z0)];
+problem = createOptimProblem('fmincon', 'objective', @cost, 'x0', w0);
+gs = GlobalSearch;
+w = run(gs, problem);
+z = (w(1:2*R)+1j*w(2*R+1:4*R)).';
+f = z(1:R).'; a = z(R+1:2*R).';
+rsdl = norm(sum([nq*evan(f,a),-cs],2));
+                
+                
+function y = cost(w)
+% Distance of z from state cs
+
+f = w(1:R)+1j*w(2*R+1:3*R); a = w(R+1:2*R)+1j*w(3*R+1:4*R);
+l = nq*evan(f,a);
+ndq = [l, ndqr*evan(f,a)];
+y = norm(sum([l,-cs],2))^2 + wgt^2*log(cond(ndq));
+
+end
+
+end

### 2014-04-22 cohfit5.m

@@ -1,0 +1,39 @@
+function [z, rsdl] = cohfit5(cs, R, wgt)
+% fit a superposition of coherent states to the fock expansion cs,
+% starting from a sampling of the glauber analytic expansion.
+% A term w*log(cond(dq(z))) is added to the cost
+% function in quadrature; there are no hard constraints.
+
+global nq ndqr
+
+N = numel(cs)-1;  Nbar = (0:N)*abs(cs).^2;
+
+% initial guess by rejection sampling
+a = [];  f = [];
+while numel(a)<R
+    amp = sqrt(Nbar)*[4 4i]*(rand(2,1)-0.5);
+    gmp = exp(-0.5*abs(amp)^2)*conj(amp).^(0:N)./sqrt(factorial(0:N))*cs;
+    if rand()>abs(gmp); continue; end
+    a = [a amp];  f = [f -0.5*abs(amp)^2+1i*angle(gmp)];
+end
+z0 = [f a];
+
+options = optimset('Display','notify');
+w0 = [real(z0), imag(z0)];
+w = fminunc(@(w)cost(w),w0,options);
+z = (w(1:2*R)+1j*w(2*R+1:4*R)).';
+f = z(1:R).'; a = z(R+1:2*R).';
+rsdl = norm(sum([nq*evan(f,a),-cs],2));
+                
+                
+function y = cost(w)
+% number weighted distance of z from state cs
+
+f = w(1:R)+1j*w(2*R+1:3*R); a = w(R+1:2*R)+1j*w(3*R+1:4*R);
+l = nq*evan(f,a);
+ndq = [l, ndqr*evan(f,a)];
+y = norm(sum([l,-cs],2))^2 + wgt^2*log(cond(ndq));
+
+end
+
+end

### 2014-04-22 cohproj.py

@@ -1,0 +1,39 @@
+"""This script compares a direct integration of the quartic oscillator by Euler's formula with one projected onto a superposition of coherent states at every step.
+"""
+
+import brackets; brackets.init(10); from brackets import *
+from numpy import *
+from numpy.linalg import lstsq, norm
+from scipy.special import cbrt
+import matplotlib.pyplot as plt
+
+T = pi
+c0 = nq*evan([-0.5*2**2], [2])
+rcs = logspace(-4, 0, 20)
+errs = zeros_like(rcs)
+ndzs = zeros_like(rcs)
+
+# set up initial guess
+a0 = array([2, 1.5, 2-0.5j, 2.5, 2+0.5j])
+f0 = -0.5*a0**2
+c = sum(nq*evan(f0,a0), axis=1)
+c = c/norm(c)
+f0 = f0 - log(norm(c))
+z0 = hstack((f0, a0))
+dq = hstack((nq*evan(f0,a0), ndqr*evan(f0,a0)))
+
+for i in range(rcs.size):
+	dz = array(lstsq(dq, c0-c, rcond=rcs[i])[0]).flatten()
+	ndzs[i] = norm(dz)
+	z = z0 + dz
+	d = sum(nq*evan(z[0:5],z[5:10]), axis=1)
+	errs[i] = norm(d-c)
+	
+plt.figure(1)
+plt.loglog(ndzs, errs)
+plt.xlabel(r"$|\delta z|$")
+plt.ylabel(r"$||\psi(z)\rangle-|\alpha\rangle|$")
+
+plt.figure(2)
+plt.loglog(rcs, errs)
+plt.show()

### 2014-04-22 condw.m

@@ -1,0 +1,56 @@
+% Fit 5 components to amplitude 2, with condition number weighting and a
+% variety of weights.
+
+global N
+N = 10; brackets
+ws = logspace(-3.5,0,20);
+
+% amplitude 2 coherent state
+c0 = nq*evan(-2,2);
+
+% 5-component approximation, evenly spread
+a0 = 2+[0.5*1i.^(0:3), 0]; R = numel(a0);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+
+cns = zeros(2,numel(ws));
+rsdls = zeros(2,numel(ws));
+
+figure(1)
+for i = 1:numel(ws)
+    z = cohfit3(c0, [f0,a0], ws(i));
+    f = z(1:R).'; a = z(R+1:2*R).';
+    l = nq*evan(f,a);
+    rsdls(1,i) = norm(sum([l,-c0],2));
+    ndq = [l, ndqr*evan(f,a)];
+    cns(1,i) = cond(ndq);
+    subplot(5,4,i)
+    plot(real(a), imag(a), '*k')
+end
+
+% 5-component approximation, randomly spread
+a0 = 2+0.3*randn(1,5)+0.3i*randn(1,5); R = numel(a0);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+
+figure(2)
+for i = 1:numel(ws)
+    z = cohfit3(c0, [f0,a0], ws(i));
+    f = z(1:R).'; a = z(R+1:2*R).';
+    l = nq*evan(f,a);
+    rsdls(2,i) = norm(sum([l,-c0],2));
+    ndq = [l, ndqr*evan(f,a)];
+    cns(2,i) = cond(ndq);
+    subplot(5,4,i)
+    plot(real(a), imag(a), '*k')
+end
+
+
+figure(3); subplot(2,1,1)
+loglog(cns(1,:), rsdls(1,:), '-k', cns(2,:), rsdls(2,:), '-g')
+ylabel('residual')
+xlabel('condition number')
+subplot(2,1,2)
+loglog(cns(1,:), ws, '-k', cns(2,:), ws, '-g')
+xlabel('condition number')
+ylabel('w')

### 2014-04-22 csve.m

@@ -1,0 +1,45 @@
+% Compute the the singular values and vectors of the G operator truncated
+% to a disk, and compare to the calculated ones. 
+
+global N; N = 24; brackets
+mkrs = {'vk' 'sk' 'pk'};  mkr = 1;
+l = zeros(size(mkrs));
+
+h = 0.3;
+for b=3:5;
+    msh = -b:h:b;
+    [x,y] = meshgrid(msh);  x = x(:);  y = y(:);
+    a = x+1i*y;  in = abs(a)<=b; a = a(in);
+    [U,S,V] = svd(h*pi^-0.5*nq*evan(a,'even'));
+
+    figure(1)
+    ll = plot(0:N, diag(S), mkrs{mkr}, ...
+        0:N, sqrt(gammainc(b^2,(0:N)+1)), '-k');
+    l(mkr) = ll(1);
+    mkr = mkr+1;
+    hold on
+
+    smpl = zeros(numel(msh));
+    for i=1:N+1
+        smpl(in) = V(:,i);
+        figure(b); subplot(5,5,i)
+        zplot([-b b], [-b b], smpl)
+        set(gca, 'DataAspectRatio', [1 1 1])
+    end
+end
+
+figure(1)
+set(gca, 'FontSize', 14)
+title('Convergence of |\Gamma_\chi> as phase space boundary expands')
+xlabel('n')
+ylabel('\sigma_n')
+legend(l, 'b=3', '4', '5', 'Location', 'SouthWest')
+
+figure(2)
+set(gca, 'FontSize', 14)
+v = 10.^-[6 4 2 1];  v = [v 0.1:0.2:0.9 1-v];
+[n,m] = meshgrid(linspace(0, 24, 50));
+[C,h] = contour(n,m,sqrt(gammainc(m,n+1)),v,'-k');
+ylabel('b^2'); xlabel n
+title('Singular values of |\Gamma_\chi>')
+clabel(C,h)

### 2014-04-22 expand.m

@@ -1,0 +1,50 @@
+% Sample the vacuum state on regular and random sample amplitudes, and see
+% how it converges.
+
+global N; N = 30; brackets
+
+c = nq*evan(0, 'even');
+b = 2.5;  % radius of regular grid
+nr = 10;  rs = logspace(1,3.5,nr);
+
+% The truncated state the regular expansion should converge to
+
+ct = c.*sqrt(gammainc(b^2, 1:N+1)).';
+
+% Regular grid
+
+rds = zeros(3, numel(rs));  hs = zeros(size(rs));
+tsn = zeros(2, numel(rs));
+for i=1:nr
+    hs(i) = sqrt(pi/rs(i))*b;
+    [x,y] = meshgrid(-b:hs(i):b);
+    z = x(:)+1i*y(:);  z = z(abs(z) <= b);
+    M = nq*evan(z, 'even');
+    d = pi^-0.5*hs(i)*M'*c;
+    T = M'*M*diag(d);   % kets are normalised
+    tsn(1,i) = norm(T-T', 'fro');
+    rds(1,i) = norm(c-pi^-0.5*hs(i)*M*d);
+    rds(3,i) = norm(ct-pi^-0.5*hs(i)*M*d);
+end
+
+% Random grid, even weights, phases of entire expansion
+
+tic; a = gample(c, round(rs(end))); toc
+for i=1:nr
+    r = round(rs(i));
+    [M, f0] = evan(a(1:r), 'even');  M = nq*M;
+    d = M'*c;  d = d/norm(d);
+    f = f0 + log(d);  cfit = sum(nq*evan(f,a(1:r)), 2);
+    T = M'*M*diag(d);
+    tsn(2,i) = norm(T-T', 'fro');
+    rds(2,i) = norm(c - cfit/norm(cfit));
+end
+
+figure
+subplot 211
+loglog(rs, rds(1,:), '-k', rs, rds(2,:), '--k', rs, rds(3,:), ':k')
+xlabel('components');  ylabel('residual')
+subplot 212
+semilogx(rs, tsn(1,:), '-k', rs, tsn(2,:), '--k');
+xlabel('components');  ylabel('tension')
+print(gcf, 'converge.pdf', '-dpdf')

### 2014-04-22 fittest.m

@@ -1,0 +1,15 @@
+% Experiment with fitting gaussian polynomials in z and z* to samples
+
+D = 10;
+x = -5:0.2:5; y = x;  [xs,ys] = meshgrid(x,y);  zs = xs(:)+1i*ys(:);
+fs = sin(2*xs);
+
+figure; subplot 241
+V = diag(exp(-0.5*abs(zs).^2))*zander(zs, D);
+for i = 0:D
+    W = V(:,1:0.5*(i+1)*(i+2));
+    c = W\fs(:);
+    subplot(3,4,i+1); zplot(x, y, W*c);
+    set(gca, 'DataAspectRatio', [1 1 1])
+end
+    

### 2014-04-22 follow.m

@@ -1,0 +1,61 @@
+% Follow the exact solution by nonlinear fitting
+% now uses cohfit2 with w=1e-2, and includes number weighting
+
+global N nwgt
+N = 10; brackets
+T = pi;
+dts = 0.1;
+R = 5;
+w = 1e-5;
+
+% amplitude 2 coherent state and 5-component approximation
+c0 = nq*evan(-2,2);
+% nwgt = diag(abs(c0).^2);
+nwgt = eye(N+1);
+nwgt(7:end,7:end) = 0;  % truncate
+
+% 5-component approximation, randomly spread
+a0 = 2+0.3*randn(1,R)+0.3i*randn(1,R);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+
+% exact solutions
+rs = 0.5*pi*(1+nthroot(linspace(-1, 1, 100), 3));
+crs = diag(c0)*exp(-1i*nhn*rs);
+ars = diag(crs'*aop*crs);
+figure(10)
+plot(real(ars), imag(ars), '-k')
+
+% propagation
+for p=1:numel(dts)
+    dt = dts(p);
+    ts = (0:T/dt+1)*dt;
+    as = zeros(size(ts)); as(:) = nan;
+    rs = zeros(size(ts)); rs(:) = nan;
+    cns = zeros(size(ts)); cns(:) = nan;
+
+
+    f = f0; a = a0;
+    for q=1:numel(ts)
+        [z, rs(q)] = cohfit2(exp(-1i*nhn*ts(q)).*c0, [f,a], w);
+        f = z(1:R).'; a = z(R+1:2*R).';
+        ndq = [nq*evan(f,a), ndqr*evan(f,a)];
+        c = sum(ndq(:,1:R),2);
+        as(q) = c'*aop*c;
+        if any(isnan(c)) || abs(as(q)) > 4
+            break
+        end
+        cns(q) = cond(ndq);
+        figure(3); subplot(5,7,q)
+        plot(real(a), imag(a), '*k');
+        figure(4); subplot(5,7,q)
+        bar(sqrt(abs(sum([nq*evan(f,a),-exp(-1i*nhn*ts(q)).*c0],2)).^2));
+    end
+    figure(1)
+    subplot(1,2,1)
+    plot(real(ars), imag(ars), '-k', real(as), imag(as), '-b');
[…]

### 2014-04-22 follow2.m

@@ -1,0 +1,73 @@
+% Follow the exact solution by nonlinear fitting
+% now uses cohfit5
+
+% Parameters that seem to work:
+% R=5; w=3e-5
+% R = 20; w = 1e-4;     % convergence!
+
+global N
+N = 19; brackets
+ts = 0.5*pi*[0:0.2:1];
+R = 15;
+w = 1e-4;
+nsv = min(N+1,2*R);
+hs = linspace(-0.1,0.1,5);
+
+
+% amplitude 2 coherent state and 5-component approximation
+c0 = nq*evan(-2,2);
+
+% exact solutions
+rs = 0.5*pi*(1+nthroot(linspace(-1, 1, 100), 3));
+crs = diag(c0)*exp(-1i*nhn*rs);
+ars = diag(crs'*aop*crs);
+
+% propagation
+as = zeros(size(ts)); as(:) = nan;
+rs = zeros(size(ts)); rs(:) = nan;
+cns = zeros(size(ts)); cns(:) = nan;
+
+for q=1:numel(ts)
+    [z, rs(q)] = cohfit5(exp(-1i*nhn*ts(q)).*c0, R, w);
+    f = z(1:R).'; a = z(R+1:2*R).';
+    ndq = [nq*evan(f,a), ndqr*evan(f,a)];
+    c = sum(ndq(:,1:R),2);
+    as(q) = c'*aop*c;
+    cns(q) = cond(ndq);
+    [U,S,V] = svd(ndq, 'econ');
+    nbars = sum(((0:N).'*ones(1,nsv)).*abs(U).^2);
+    
+    figure(3); subplot(3,4,q)
+    plot(real(a), imag(a), '*k')
+    set(gca, 'DataAspectRatio',[1 1 1])
+    figure(4); subplot(3,4,q)
+    bar(sqrt(abs(sum([nq*evan(f,a),-exp(-1i*nhn*ts(q)).*c0],2)).^2));
+    figure(5);  subplot(3,4,q)
+    semilogy(1:nsv, abs(U'*(nhn.*c)), '+k', 1:nsv, diag(S), 'ok', ...
+        1:nsv, abs(S\U'*(nhn.*c)), '*r')
+    figure(6);  subplot(3,4,q)
+    plot(1:nsv, nbars, 'xk')
+    
+    % plot changes of condition number
+    figure(9+q)
+    for p = 1:nsv
+        cnhs = zeros(2, numel(hs)); cnhs(:) = nan;
[…]

### 2014-04-22 frame.m

@@ -1,0 +1,135 @@
+% Draw a normal ensemble of coherent states, and plot their frame
+% projectors for a range of Tychonov regularisation parameters.
+
+global N; N = 40; brackets
+R = 15;
+nl = 20;  tych = logspace(-5,1,nl);
+pl = [20 18 15 13 10 1];
+titles = {'components', 'dual ket', 'error', 'first three', '', '';
+    '', 'weights', 'operator', 'dual kets', '', ''};
+
+a = 2*sqrt(0.5)*[1 1i]*randn(2,R);
+A = nq*evan(a,'even');
+
+mesh = -5:0.2:5;
+[x,y] = meshgrid(mesh);  z = x(:)+1i*y(:);
+Aps = nq*evan(z,'even');
+
+apops = zeros(nl, R, N+1);
+enorms = zeros(1,nl);
+wgts = zeros(nl, R);
+conds = zeros(1, nl);
+for i=1:nl
+    M = pinv([A; tych(i)*eye(R)]);
+    M = M(:,1:N+1);  apops(i,:,:) = M;
+    wgts(i,:) = sqrt(abs(diag(M*M')));
+    I = A*squeeze(apops(i,:,:));
+    enorms(i) = sqrt(21) - norm(eye(21)-I(1:21,1:21), 'fro');
+    conds(i) = cond(M);
+end
+
+% Labels
+
+figure
+for j=1:6
+    subplot(6,7,7*j-6); axis off
+    text(1, 0.5,titles{1,j}, ...
+        'HorizontalAlignment', 'right', 'Units', 'Normalized')
+    text(1, 0.3,titles{2,j}, ...
+        'HorizontalAlignment', 'right', 'Units', 'Normalized')
+end
+
+% Plot ensembles and operators
+
+for i=1:6
+    l = pl(i);
+    ty = squeeze(apops(l,:,:));
+    
+    subplot(6,7,i+1); hold on
+    w = wgts(l,:)/max(wgts(l,:));
+    for j=1:R
+        plot(real(a(j)),imag(a(j)), 'ok', ...
+            'MarkerFaceColor', hsv2rgb([0 0 1-w(j)]));
+    end
+    axis equal image;  if i>1; axis off; end
+    
+    subplot(6,7,7+i+1)    
+    semilogy(1:R,wgts(l,:),'+k')
+    
+    subplot(6,7,2*7+i+1)    
+    I = A*ty;
+    zplot(0:15,0:15,eye(16)-I(1:16,1:16))
[…]

### 2014-04-22 gabor.m

@@ -1,0 +1,5 @@
+% Set up Gabor frame
+M = evan(a, 'even');
+A = nq*M;
+DA = [A ndqr*M];
+clear M

### 2014-04-22 galerkin.m

@@ -1,0 +1,19 @@
+% Investigate Galerkin discretisation of G
+
+global N; N = 20; brackets
+
+a0 = 2;
+R = 3;
+t = pi/2;
+c0 = sum(nq*evan(-0.5*abs(a0).^2, a0), 2);
+c0 = (exp(-1i*t*nhn).*c0)/norm(c0);
+a = gample(c0, R);
+
+plot(real(a), imag(a), '*k')
+set(gca, 'DataAspectRatio', [1 1 1])
+
+[ar, as] = meshgrid(a,a);
+M = exp(-abs(ar-as).^2);
+w = M \ (pi*ones(size(a)));
+
+c = norm(exp(-0.5*abs(a(2))-0.5*abs(a)+conj(a)*a(2)));

### 2014-04-22 gample.m

@@ -1,0 +1,25 @@
+function as = gample( c, R )
+%GAMPLE Sample amplitudes from the normalized entire distribution
+%   cs is a vector of number state weights, R is the number of coherent
+%   amplitudes to sample.  as is a column vector of coherent amplitudes.
+%   This relies on a potentially dodgy assumption about how rapidy the
+%   normalised entire expansion decays with amplitude.  This can be tested
+%   with geck(cs).
+
+global N nq
+
+v = 5;  % assumes that |f(a)| <= exp(-|a|^2/v+0.5)
+na = 5*R;    % Number of samples to take at once
+
+as = zeros(R,1);  r = 0;
+while r<R
+    % [r R na]
+    a = sqrt(v)*randn(na,2)*[1; 1i];
+    p0 = exp(-0.5*abs(a).^2/v+0.5);
+    ca = nq*evan(a, 'even');
+    p = abs(ca'*c);
+    a = a(rand(na,1) < p./p0);
+    if numel(a) > R-r; a = a(1:R-r); end
+    as(r+1:r+numel(a)) = a;
+    r = r + numel(a);
+end

### 2014-04-22 gquad.m

@@ -1,0 +1,94 @@
+% Fit a superposition of coherent states to a cat, plot the Picard 
+% and L curves, 
+% interpolate phase space functions by gaussian polynomials, and find
+% its norm.
+
+global N; N = 40; brackets
+
+cat = sum(nq*evan([2i -2i], 'even'), 2);
+R = 40;  a = gample(cat, R);
+ls = logspace(-14,0,51);   % Lagrange parameters
+deg = 4;    % degree of polynomial interpolants
+
+[X, f0] = evan(a,'even');  A = nq*X;  [U,S,V]=svd(A, 'econ');
+[Y, m, n] = zander(a, deg);  B = diag(exp(f0))*Y;
+m = repmat(m,numel(m),1);  n = repmat(n,numel(n),1); 
+G = pi*(m+n'==m'+n).*factorial(n+m');
+
+
+ns = zeros(size(ls));  ms = zeros(size(ls));
+rs = zeros(size(ls));  gs = zeros(size(ls));
+ds = zeros(numel(zander(1,deg)), numel(ls));
+fs = zeros(numel(a), 6);
+for i=1:numel(ls)
+    c = [A; ls(i)*eye(size(A))]\[cat; zeros(N+1,1)];
+    ds(:,i) = B\c;  ms(i) = abs(ds(:,i)'*G*ds(:,i));
+    ns(i) = norm(c);
+    p = abs(c/ns(i)).^2;  gs(i) = exp(-sum(p.*log(p)));
+    f = f0 + log(c);
+    rs(i) = norm(cat - sum(nq*evan(f, a), 2));
+    if any(i==(1:10:51)); fs(:,7-(i+9)/10) = f; end
+end
+
+
+figure
+[x,y] = meshgrid(-2:0.3:2, -4:0.3:4);  z = x+1i*y;
+[X, f1] = evan(z(:),'even');  bras = (nq*X)';
+subplot 278
+buf = zeros(size(z));  buf(:) = bras*cat;
+zplot(-2:0.3:2, -4:0.3:4, buf);  set(gca, 'DataAspectRatio', [1 1 1])
+for i = 1:6
+	subplot(2,7,i+1); hold on
+	supplot([fs(:,i); a])
+	set(gca, 'DataAspectRatio', [1 1 1])
+	title({sprintf('log(\\lambda) = %.1f', log10(ls(61-10*i))); ...
+        sprintf('log(R) = %.1f', log10(rs(61-10*i)))})
+    
[…]

### 2014-04-22 gridverge.m

@@ -1,0 +1,31 @@
+% Fit cat states on a regular grid, and find the lagrange parameter at
+% which the solution departs from an entire function, as the grid spacing
+% varies.
+
+global N; N = 20; brackets
+c = sum(nq*evan([2i -2i], 'even'), 2);
+hs = [0.3 1];
+% hs = logspace(-1,0,3);
+ls = logspace(-15,4,15);
+
+nms = zeros(numel(hs), numel(ls));  rds = nms;
+for i=1:numel(hs)
+    figure
+    [x,y] = meshgrid(-4:hs(i):4);  z = x(:)+1i*y(:);
+    r = abs(z) <= 4;
+    [M,f0] = evan(z(r), 'even');
+    A = hs(i)*nq*M;
+    for j=1:numel(ls)
+        d = [A; ls(j)*eye(size(A))] \ [c; zeros(size(c))];
+        nms(i,j) = norm(d);  rds(i,j) = norm(c-A*d);
+        subplot(4,4,j)
+        supplot([f0+log(d); z(r)])
+    end
+end
+
+figure
+for i=1:numel(hs)
+    subplot(3,2,i);
+    plot(rds(i,:), nms(i,:), '-k');
+    title(num2str(hs(i)));
+end

### 2014-04-22 gweight.m

@@ -1,0 +1,47 @@
+% Draw normally distributed points, and derive a set of weights such that
+% the gaussian centred on each point can be integrated by a weighted sum of
+% samples.
+
+set(0, 'DefaultAxesFontSize', 14)
+
+R = 15;  x = randn(R,1);  y = linspace(-2,2,R).';
+[xr, xc] = meshgrid(x);  [yr, yc] = meshgrid(y);
+Ax = exp(-(xr-xc).^2);  Ay = exp(-(yr-yc).^2);
+rhs = ones(size(x))*sqrt(pi);
+w = Ax\rhs;
+v = Ay\rhs;
+
+figure
+subplot 211
+plot(x, zeros(size(x)), 'ok', x, w, '+k')
+title('Normally distributed abscissae')
+xlabel('x_i')
+ylabel('w_i')
+subplot 212
+plot(y, zeros(size(y)), 'ok', y, v, '+k')
+title('Evenly spaced abscissae')
+xlabel('x_i')
+ylabel('w_i')
+
+l = logspace(-10,1,20);      % reg. parameters
+ry = zeros(size(l));  ny = zeros(size(l));
+rx = zeros(size(l));  nx = zeros(size(l));
+for i=1:numel(l)
+    w = [Ax; l(i)*eye(R)] \ [rhs; zeros(size(x))];
+    nx(i) = norm(w);  rx(i) = norm(Ax*w-rhs);
+    v = [Ay; l(i)*eye(R)] \ [rhs; zeros(size(y))];
+    ny(i) = norm(v);  ry(i) = norm(Ay*v-rhs);
+end
+
+figure
+subplot 311
+loglog(nx, rx, ':k', ny, ry, '-k');
+legend('random', 'regular', 'Location', 'NorthEast')
+xlabel('solution norm'); ylabel('residual')
+title('Tychonov curves')
+subplot 312
+loglog(l, rx, ':k', l, ry, '-k');
+xlabel('parameter'); ylabel('residual')
+subplot 313
+loglog(l, nx, ':k', l, ny, '-k');
+xlabel('parameter'); ylabel('solution norm')

### 2014-04-22 hzero.m

@@ -1,0 +1,48 @@
+% Solve the equation d|psi>/dt = |psi_0>-|psi>, where |psi_0> is a 
+% quartic oscillator state, and |psi> is initially |psi_0> expanded over
+% sampled amplitudes.  Monitor the convergence to |psi_0> and the condition
+% of the Glauber operator.
+
+global N; N = 15; brackets
+
+t = pi/2;  c0 = exp(-1i*nhn*t).*(nq*evan(2, 'even'));
+R = 15;  a = gample(c0, R);
+le = 1e-9;  % Tychonov parameter for initial expansion
+lv = 1e-4;  % variational problem
+[A, f] = evan(a, 'even');  A = nq*A;
+f = f + log([A'*A; le*eye(R)]\[A'*c0; zeros(R,1)]);
+z0 = [f; a];
+
+smax = 4;  ds = 0.03;  s = 0:ds:smax;  z = z0;
+clear scapt zcapt
+rls = zeros(size(s));
+sings = zeros(size(s));
+for i=1:numel(s)
+    c = sum(nq*evan(z), 2);
+    rls(i) = norm(c0-c);
+    if i>2 && rls(i)>rls(i-1) && rls(i-2)>=rls(i-1)
+        scapt = s(i); zcapt = z;
+    end
+    [U,S,V] = svd(evan(z(R+1:2*R), 'even'), 'econ');
+    sings(i) = S(end:end);
+    dq = [nq*evan(z) ndqr*evan(z)];
+    zdot = [dq; lv*eye(size(dq))] \ [c0-c; 0*c];
+    z = z+zdot*ds;
+end
+
+figure(1)
+subplot 121
+supplot(z0)
+set(gca, 'DataAspectRatio', [1 1 1])
+subplot 122
+supplot(z)
+set(gca, 'DataAspectRatio', [1 1 1])
+
+figure(2)
+subplot(2,1,1)
+plot(s, rls, '-k')
+ylabel('|c-c0|')
+subplot(2,1,2)
+plot(s, sings, '-k')
+xlabel('s')
+ylabel('lsv of |A>')

### 2014-04-22 jaccond.m

@@ -1,0 +1,71 @@
+% Fit a 5 component superposition to a coherent state, then investigate how
+% the condition number of its derivative varies for small changes.
+
+global N; N = 10; brackets
+R = 5;
+hs = linspace(-0.1,0.1,10);
+w = 1.5e-3;
+c0 = nq*evan(-2,2);
+
+% Random sample amplitudes
+a = 2+sqrt(0.5)*[1 1i]*randn(2,R);
+G = nq*evan(-0.5*abs(a).^2,a);
+ef = pinv(G)*c0;
+f = -0.5*abs(a).^2 + log(ef).';
+
+% amplitude 2 coherent state and 5-component approximation
+disp(norm(sum([nq*evan(f,a) -c0], 2)))
+
+figure; subplot(1,2,1)
+plot(real(a), imag(a), '*k');
+set(gca, 'DataAspectRatio',[1 1 1])
+for r = 1:R
+    text(real(a(r)), imag(a(r)), sprintf('  %d',r))
+end
+title('Amplitudes')
+
+subplot(1,2,2)
+bar(sqrt(sum(abs(nq*evan(f,a)).^2, 1)))
+title('Weights')
+
+basis = eye(R);
+figure
+for p = 1:R
+    cns = zeros(2, numel(hs)); cns(:) = nan;
+    for q = 1:numel(hs)
+        b = a + hs(q)*basis(p,:);
+        [U,S,V] = svd(nq*evan(-0.5*abs(b).^2,b), 'econ');
+        cns(1,q)=S(end,end);
+        b = a + 1i*hs(q)*basis(p,:);
+        [U,S,V] = svd(nq*evan(-0.5*abs(b).^2,b), 'econ');
+        cns(2,q)=S(end,end);
+    end
+    [U,S,V] = svd(nq*evan(-0.5*abs(a).^2,a), 'econ');
+    b = a + 0.01*basis(p,:);  Gp = nq*evan(-0.5*abs(b).^2,b);
+    b = a - 0.01*basis(p,:);  Gm = nq*evan(-0.5*abs(b).^2,b);
+    A = [zeros(N+1), Gp-Gm; (Gp-Gm)', zeros(R)];
+    ev = [U(:,end); V(:,end)];
+    s = S(end,end);  ds = 0.5*ev'*A*ev;
+    subplot(1,5,p)
+    plot(hs, cns(1,:), '-k', [-0.1 0.1], s+[-0.1 0.1]*ds/0.02, ':k', ...
[…]

### 2014-04-22 oops.py

@@ -1,0 +1,9 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+t = np.arange(19)
+plt.ylabel("Military spending, trillions")
+plt.plot(1971+t, 5*np.exp(t/10.), "bp", label="USA")
+plt.plot(1971+t, 19-t, "r*", label="USSR")
+plt.legend( ("USSR", "USA"), 'upper left')
+plt.savefig("oops.png")

### 2014-04-22 pairing.m

@@ -1,0 +1,13 @@
+% Plot the distance from a(|h>-|-h>) to |n>
+
+
+N = 4;
+[n,s,h] = ndgrid(0:N, logspace(-6,0,5), logspace(-4,0,4));
+
+c = 2*exp(-0.5*h.^2).*h.^n./sqrt(factorial(n));
+c(1:2:end,:,:) = 0;
+c = reshape(c, [1 size(c)]);  c = repmat(c, N+1, 1);
+d = reshape(eye(N+1), [N+1 N+1 1 1]);
+d = repmat(d, 1, 1, numel(s), numel(h));
+
+

### 2014-04-22 phasespace.m

@@ -1,0 +1,4 @@
+% Set up phase space grid
+[x,y] = meshgrid(mesh);  z = x(:)+1i*y(:);
+Aps = nq*evan(z,'even');
+DAps = ndqr*evan(z,'even');

### 2014-04-22 picard.m

@@ -1,0 +1,70 @@
+% Expand the exact solutions over a grid, using the generating function.
+% Find Picard plots for expanding their derivatives over the same grid,
+% and for fitting the grid variationally to the derivatives.
+
+global N; N = 25; brackets
+
+set(0, 'DefaultAxesFontSize', 14)
+
+h = 0.3;  alim = 6;
+c0 = nq*evan(-2,2);
+ts = linspace(0, 0.5*pi, 12);
+[x,y] = meshgrid(-alim:h:alim);
+a = x(:)+1i*y(:);
+zs = [-0.5*abs(a).^2; a];
+xpn = h*evan(zs)'*diag(factorial(0:N).^-0.5); % backward map
+
+figure(1)
+semilogy(0:N, abs(c0), 'vk', 0:N, abs(nhn.*c0), 'pk')
+xlabel('Particle number n')
+ylabel('|<n|?>|')
+legend('|\psi>', 'H|\psi>')
+title('Expansions over number states')
+
+smpl = zeros(numel(-alim:h:alim));
+errs = zeros(2,numel(ts));
+for j=1:numel(ts)
+    c = exp(-1j*nhn*ts(j)).*c0;  Hc = nhn.*c;
+    gs = xpn*(exp(-1j*nhn*ts(j)).*c0);
+    f = -0.5*abs(a).^2 + log(h*gs/pi);
+    ndq = [nq*evan(f,a), ndqr*evan(f,a)];  [U,S,V] = svd(ndq, 'econ');
+    
+    figure(2); subplot(4,3,j)
+    smpl(:) = gs;  zplot([-alim alim], [-alim, alim], smpl);
+    set(gca, 'DataAspectRatio', [1 1 1])
+    if j==2; title('State from time 0 to \pi/2'); end
+    
+    figure(3); subplot(4,3,j)
+    smpl(:) = xpn*Hc;  zplot([-alim alim], [-alim, alim], smpl);
+    set(gca, 'DataAspectRatio', [1 1 1])
+    if j==2; title('Derivative from time 0 to \pi/2'); end
+    
+    figure(4); subplot(4,3,j)
+    smpl(:) = xpn*U(:,1);  zplot([-alim alim], [-alim, alim], smpl);
+    set(gca, 'DataAspectRatio', [1 1 1])
+    if j==2; title('Largest singular ket of variational problem'); end   
+    
+    figure(5); subplot(4,3,j)
[…]

### 2014-04-22 qoq.m

@@ -1,0 +1,24 @@
+% plot <a|psi(t)> for the quartic oscillator
+
+global N; N = 30; n = (0:N)'; brackets
+mesh = linspace(-5,5,50);  phasespace
+
+H = n.^2-n;
+c0 = nq*evan(2, 'even');
+t = [0 0.04 0.1 0.16 0.18 0.2 0.24+0.1*rand() 1/2 2/3 0.9 0.96 1];
+nt = numel(t);
+f = 0:0.1:2*pi;
+
+figure
+set(gcf, 'DefaultAxesDataAspectRatio', [1 1 1])
+for j = 1:nt
+    c = exp(-0.5i*pi*t(j)*H).*c0;
+    dc = -1i*H.*c;
+    subplot(3,nt,j)
+    zplot(mesh, mesh, Aps'*c);
+    title(num2str(t(j)))
+    subplot(3,nt,j+nt)
+    zplot(mesh, mesh, Aps'*dc)
+    subplot(3,nt,j+2*nt)
+    zplot(mesh, mesh, DAps'*dc)
+end

### 2014-04-22 qorf.m

@@ -1,0 +1,50 @@
+% Quartic oscillator with resampling
+
+global N;  N = 30;  brackets
+
+a = 2;  [~,f] = evan(a,'even');
+dt = 1e-3;  T = 15*dt;
+t = 0:dt:T;  S = numel(t);
+
+[f, a] = reframe([f; a]);  z = [f; a];
+c = sum(nq*evan(z),2);
+ct = exp(1i*nhn*t).*repmat(c,1,S);
+
+zt = cell(1,S);  cct = zeros(N+1,S);
+for s=1:S
+    zt{s} = z;  cct(:,s) = c;
+    dz = 0;
+    for j=1:4
+        Dq = [nq*evan(z+dz/2), ndqr*evan(z+dz/2)];
+        dz = pinv(Dq)*(1i*nhn.*c)*dt;
+    end
+    [f, a] = reframe(z + dz);  z = [f; a];
+    c = sum(nq*evan(z),2);
+end
+zt{S} = z;  cct(:,S) = c;
+
+% quadratures
+Xt = diag(ct'*aop*ct);
+XXt = diag(cct'*aop*cct);
+
+figure
+subplot 311
+plot(Xt, ':k');  hold on;  plot(XXt, '-k');
+set(gca, 'DataAspectRatio', [1 1 1])
+title('Phase space trajectory')
+subplot 312
+plot(t, real(XXt), '-k', t, imag(XXt), '-r', ...
+    t, real(Xt), ':k', t, imag(Xt), ':r')
+title('Expected quadratures')
+xlabel('t')
+subplot 313
+semilogy(t, diag(cct'*cct), '-k')
+title('squared ket norm')
+
+
+% components
+figure
+for s=1:S
+    subplot(1,S,s)
+    supplot(zt{s});  set(gca, 'DataAspectRatio', [1 1 1])
+end

### 2014-04-22 randdual.m

@@ -1,0 +1,20 @@
+% Compute the frame bounds, and the dual frame, for a random ensemble of
+% coherent states.
+
+global N; N = 30; brackets
+R = 15; a = 2*sqrt(0.5)*[1 1i]*randn(2,R); gabor
+mesh = linspace(-5,5,20);  phasespace
+
+[~, sv, ~] = svd(A);
+S = A*A';
+F = S\A;
+
+figure
+for r=1:R
+    subplot(1,R+1,r+1)
+    zplot(mesh, mesh, Aps'*F(:,r))
+    set(gca, 'DataAspectRatio', [1 1 1])
+end
+subplot(1,R+1,1)
+text(0.5,0.5, ['Frame bounds'; 'A = ' num2str(sv(end,end)); ...
+    'B = ' num2str(sv(1,1))])

### 2014-04-22 ranterp.m

@@ -1,0 +1,17 @@
+% Interpolate samples on a random grid
+
+kmax = 5;   R = 20;
+k = -kmax:kmax;
+x = pi*(2*rand(1,R)-1).'; xg = (-pi:0.1:pi).';
+A = exp(1i*x*k);    Ag = exp(1i*xg*k);
+f = (x+pi).*(pi-x); fg = (xg+pi).*(pi-xg);
+itp = Ag*pinv(A, 0.1);
+
+figure
+plot(x, zeros(size(x)), 'ok', x, f, '+k', xg, fg, ':k', xg, real(itp*f), '-k')
+
+figure
+for r=1:R
+    subplot(4,5,r)
+    plot(xg, real(itp(:,r)))
+end

### 2014-04-22 reframe.m

@@ -1,0 +1,36 @@
+function [f, a] = reframe(z)
+% REFRAME: improve the condition of parameters describing a ket
+%
+% [f, a] = reframe(z)
+
+global nq
+z = z(:);  c = sum(nq*evan(z),2);
+R = numel(z)/2;  f = z(1:R);  a = z(R+1:end);
+figure;  plot(a, 'ok');  hold on;  axis equal
+
+% Cull amplitudes with low weight
+a = a(real(f) > -6);  R = numel(a);
+
+% Add amplitudes between spread out points
+as = repmat(a,1,R);  S = abs(as - as.') + diag(inf*ones(R,1));
+[nndist, nn] = min(S);
+j = nndist > 1 & nndist < 2;
+a = [a; (a(j)+a(nn(j)))/2];  R = numel(a);
+
+% Delete whichever of each pair has the lower index in a
+as = repmat(a,1,R);  S = abs(as - as.') + diag(inf*ones(R,1));
+[nndist, nn] = min(S);
+a = a(nndist > 0.2 | nn < 1:R);  R = numel(a);
+
+% Replace remaining isolated points with clusters
+as = repmat(a,1,R);  S = abs(as - as.') + diag(inf*ones(R,1));
+nndist = min(S);  j = nndist >= 2;
+b = a(j);  h = rand(size(b));
+a = [a(~j); b+0.3*exp(2i*pi*h); ...
+    b+0.3*exp(2i*pi*(h+1/3)); b+0.3*exp(2i*pi*(h-1/3))];
+
+% Resample in new frame
+[M, f] = evan(a,'even');  A = nq*M;
+f = f + log(pinv(A)*c);
+
+plot(a, '+r')

### 2014-04-22 rgrid.m

@@ -1,0 +1,104 @@
+% Examine the conditioning of expansion on regular and random grids,
+% and the effect of parallel component conditioning
+
+global N; N = 100; brackets
+c = nq*evan(0.3+0.4i, 'even');
+b = 2.5;      % radius of circular grid
+R = 80;
+nl = 15;
+lz = logspace(-14,0,nl);
+lw = logspace(-14,2,nl);
+
+ax = -1.5*b:0.2:1.5*b;
+[x,y] = meshgrid(ax);
+zreg = x(:)+1i*y(:);
+
+% set z to a regular grid
+a = sqrt(2*pi/sqrt(3)/R)*b;
+lim = ceil(2*b/sqrt(3)/a);
+[i,j] = meshgrid(a*(-lim:lim));
+z = i(:) + (0.5+0.5i*sqrt(3))*j(:);
+z = z(abs(z)<=b);  nz = numel(z);
+% set w to a random grid
+w = gample(c, nz);
+
+[Uz,Sz,~] = svd(nq*evan(z, 'even'), 'econ');
+[Uw,Sw,~] = svd(nq*evan(w, 'even'), 'econ');
+m = numel(diag(Sz));
+    
+figure(1);
+subplot 121
+zplot(ax, ax, (nq*evan(zreg, 'even'))'*c); hold on
+plot(real(z), imag(z), '.w', 'MarkerSize', 12)
+set(gca, 'DataAspectRatio', [1 1 1])
+title([num2str(nz) ' points'])
+subplot 122
+zplot(ax, ax, (nq*evan(zreg, 'even'))'*c); hold on
+plot(real(w), imag(w), '.w', 'MarkerSize', 12)
+set(gca, 'DataAspectRatio', [1 1 1])
+
+figure(2);
+subplot 121
+semilogy(1:m, diag(Sz), 'ok', 1:m, abs(Uz'*c), '+k', ...
+    1:m, abs(Sz\Uz'*c), '*k')
+title('Picard plot, regular grid'); xlabel n
+legend('\sigma_n', 'u_n cpt', 'v_n cpt', 'Location', 'SouthWest')
+subplot 122
+semilogy(1:m, diag(Sw), 'ok', 1:m, abs(Uw'*c), '+k', ...
+    1:m, abs(Sw\Uw'*c), '*k')
+title('random grid'); xlabel n
+
[…]

### 2014-04-22 rou.m

@@ -1,0 +1,50 @@
+% Plot SVE of the frame operator |A>, where amplitudes are roots of 49.
+
+global N; N = 100; brackets
+mesh = linspace(-7,7,50);  phasespace
+rs = 5:10:55;  nr = length(rs);
+
+id = zeros(nr, N+1, N+1);
+lsvs = zeros(nr, N+1, 5);
+sws = zeros(nr,55);  sws(:) = nan;
+nbar = sws;
+
+for i=1:nr
+    R = rs(i);
+    h = 2*pi/R;  t = 4.5*exp((1:R)*1i*h);
+    A = nq*evan(t, 'even');
+    id(i,:,:) = A*pinv(A);
+    [U,S,V] = svd(A,'econ');
+    lsvs(i,:,:) = U(:,1:2*i-1:end);
+    sws(i,1:R) = diag(S);
+    nbar(i,1:R) = sum(diag(0:N)*abs(U).^2);
+end
+
+figure
+for i=1:nr
+    subplot(nr,1,i)
+    plot(nbar(i,:),sws(i,:), 'ok')
+    axis([0 55 0 3])
+end
+subplot(nr,1,nr), xlabel('n')
+subplot(nr,1,3), ylabel('\sigma_n')
+
+figure
+set(gcf, 'DefaultAxesDataAspectRatio', [1 1 1])
+for i=1:nr
+    subplot(nr,1,i)
+    zplot(0:50, 0:50, id(i,1:51,1:51))
+end
+
+
+figure
+set(gcf, 'DefaultAxesDataAspectRatio', [1 1 1])
+for i=1:nr
+    for j = 1:5
+        subplot(nr,6,6*i+j-5)
+        zplot(mesh, mesh, Aps'*lsvs(i,:,j).')
+    end
+    subplot(nr,6,6*i-5), hold on
+    plot(4.5*exp(2*pi*(1:rs(i))*1i/rs(i)), '.k')
+    axis([-7 7 -7 7])
+end

### 2014-04-22 sampling.m

@@ -1,0 +1,13 @@
+% Investigate the stability of sinc reconstruction as the density reduces
+% to the nyquist limit.
+
+
+x = linspace(-5,5)';  k = 0:6;
+xnq = (-5:pi/max(k):5)';
+c = rand(numel(k),1);
+f = sin(x*k)*c;  fnq = sin(xnq*k)*c;
+f1 = sinc(0.5*max(k)*(x*ones(size(xnq'))-ones(size(x))*xnq'))*fnq;
+
+figure
+plot(x,f,'-k',x,f1,'--k'); hold on;
+plot(xnq,fnq,'.r','MarkerSize',12)

### 2014-04-22 supplot.m

@@ -1,0 +1,19 @@
+function h = supplot( z )
+%supplot Plot a superposition of coherent states
+%   supplot(z) plots the superposition z on the argand plane, with weights
+%   represented by saturation and phase by hue.
+
+R = 0.5*numel(z);  f = z(1:R);  a = z(R+1:2*R);
+ds = exp(f+0.5*abs(a).^2);
+dm = max(abs(ds));
+for i=1:numel(a)
+    h = plot(real(a(i)), imag(a(i)), 'ok', ...
+        'MarkerFaceColor', ...
+        hsv2rgb([0.5+angle(ds(i))/2/pi abs(ds(i))/dm 1]));
+    hold on
+end
+
+if nargout==0; clear h; end
+
+end
+

### 2014-04-22 sve.m

@@ -1,0 +1,36 @@
+% Investigate the singular values and vectors of the G operator.  Phase
+% space is sampled on a grid with spacing 0.5, ket space is projected onto
+% the fock states |0> through |10>.  The singular values are examined as
+% the edges of the grid expand past alpha=sqrt(10).
+
+global N; N = 24; brackets
+mkrs = {'vk' 'sk' 'pk'};  mkr = 1;
+
+h = 0.3;
+for alim=4:6;
+    msh = -alim:h:alim;
+    [x,y] = meshgrid(msh);  x = x(:);  y = y(:);
+    z = [-0.5*(x.^2+y.^2); x+1i*y];
+    [U,S,V] = svd(h*sqrt(pi)*nq*evan(z)/pi);
+
+    figure(1)
+    semilogy(0:N, 1-diag(S), mkrs{mkr}, 0:N, ...
+        1-sqrt(gammainc(1:N+1,alim^2)), '-k')
+    mkr = mkr+1;
+    hold on
+
+    smpl = zeros(numel(msh));
+    for i=1:N+1
+        smpl(:) = V(:,i);
+        figure(alim); subplot(5,5,i)
+        zplot([-alim alim], [-alim alim], smpl)
+        set(gca, 'DataAspectRatio', [1 1 1])
+    end
+end
+
+figure(1)
+set(gca, 'FontSize', 14)
+title('Convergence of G as phase space boundary expands')
+xlabel('n')
+ylabel('1-\sigma_n')
+legend('4x4', '6x6', '8x8', 'Location', 'SouthEast')

### 2014-04-22 tysv.m

@@ -1,0 +1,22 @@
+s = [logspace(-3,-0.5,5) logspace(-0.5,0.5,10) logspace(0.5,3,5)];
+a = exp(log(10)*(2*rand(30,1)-3));
+
+figure
+set(gcf, 'DefaultLineMarkerSize', 12, ...
+    'DefaultAxesPlotBoxAspectRatio', [1.1 1 1], ...
+    'DefaultAxesXTick', 10.^[-3 0 3])
+for i=1:3
+    subplot(1,3,i)
+    loglog(s, s./(s.^2+1), '-k', a, a./(a.^2+1), '.k')
+    axis([1e-3 1e3 1e-3 1])
+    if i==2; xlabel('\sigma/\lambda'); end
+    if i==1
+        title('Too regularised')
+        ylabel('\lambda\sigma^+')
+    end
+    if i==3 title('Not regularised enough'); end
+    if i>1; set(gca, 'YTick', []); end
+    a = 100*a;
+end
+
+print(gcf, '-depsc', 'tysv.eps')

### 2014-04-22 wobble2.m

@@ -1,0 +1,81 @@
+% Evolve a coherent state for t=0.1, then fit a superposition of five
+% coherent states to the result, truncated to an N=5 Fock basis.  Then
+% expand the basis to N=10, and plot the singular values and the 
+% components of vector from the approximate superposition to the state 
+% along the singular directions in both psi and z space.  Finally, 
+% plot distances from the shifted superposition to the target state, as
+% more singular values are included in the shift, and as a smooth
+% truncation parameter decreases.
+
+global N
+R = 5;
+w = 1e-5;
+t = 0.1;
+
+% amplitude 2 coherent state and 5-component approximation
+N = 7; brackets
+c0 = exp(-1i*nhn*t).*(nq*evan(-2,2));
+
+% 5-component approximation, randomly spread
+a0 = 2+0.3*randn(1,R)+0.3i*randn(1,R);
+f0 = -0.5*abs(a0).^2;
+f0 = f0 + log(norm(c0)) - log(norm(sum(nq*evan(f0,a0),2)));
+z = cohfit2(c0, [f0,a0], w);
+f = z(1:R).'; a = z(R+1:2*R).';
+abar = norm(a)/sqrt(R);
+
+
+N = 10; brackets
+c = exp(-1i*nhn*t).*(nq*evan(-2,2));
+c0 = sum(nq*evan(f,a),2);
+nwgt = diag(abs(sum(nq*evan(-0.5*abar^2,abar),2))); % Poisson
+% nwgt = diag(exp(-(0:N)));     
+% nwgt = eye(N+1);
+
+figure(1)
+plot(real(a), imag(a), '*k')
+title(sprintf('residual = %.4f', norm(c-c0)))
+set(gca, 'DataAspectRatio', [1 1 1])
+xlabel('Re(\alpha)'); ylabel('Im(\alpha)')
+
+figure(2)
+[U,S,V] = svd(nwgt*[nq*evan(f,a), ndqr*evan(f,a)], 'econ');
+dqcpt = U'*nwgt*(c-c0);
+dzcpt = S\dqcpt;
+dz = V*dzcpt;
+semilogy(1:2*R, abs(dqcpt), '+b', 1:2*R, diag(S), 'ok', ...
+    1:2*R, abs(dzcpt), '+r')
+legend('\psi', 'S', 'z', 'Location', 'NorthWest')
+title('Expansion of c-c_0')
[…]

### 2014-04-22 wplot.m

@@ -1,0 +1,15 @@
+function h = wplot(x, y, w)
+%zplot Plot a function of a complex variable as a color image
+%   x and y are as for image.  w is a matrix of samples f(x+iy).
+%   The modulus of f is plotted on the Argand plane as saturation, 
+%   its argument as hue.  Colors might be excessively quantised.
+
+dat = floor(64*abs(w)/max(abs(w(:))));
+CV = ind2rgb(dat, gray(64));
+dat = floor(32*(1+angle(w)/pi));
+CHS = ind2rgb(dat, hsv(64));
+h = image(x, y, CV.*(CHS-1)+1);
+if nargout==0; clear h; end
+
+end
+

### 2014-04-22 zander.m

@@ -1,0 +1,20 @@
+function [A, p, q] = zander( z, n )
+%zander Complex conjugate vandermonde
+%   zander(z,n) where z is a complex number returns the row vector
+%       [1 z z* z^2 zz* z^2* ...]
+%   If z is a vector, it returns a matrix with rows of this form.
+%   The extra return values are the powers of z* in p, and of z in q.
+%   These are vectors.
+
+z = z(:);  d = 0.5*(n+1)*(n+2);
+A = zeros(numel(z), d);
+p = zeros(1,d);  q = zeros(1,d);
+p(1) = 0;  q(1) = 0;  A(:,1) = 1;  j = 1;
+for i=1:n
+    p(j+1:j+i+1) = 0:i;
+    q(j+1:j+i+1) = i:-1:0;
+    A(:,j+1:j+i) = diag(z)*A(:,j-i+1:j);
+    j = j+i+1;
+    A(:,j) = conj(z).^i;
+end
+

### 2014-04-22 zplot.m

@@ -1,0 +1,18 @@
+function h = zplot(x, y, w)
+%zplot Plot a function of a complex variable as a color image
+%   zplot(x, y, w)
+%   x and y are as for image.  w is a matrix of samples f(x+iy).
+%   The modulus of f is plotted on the Argand plane as brightness, 
+%   its argument as hue.  Colors might be excessively quantised.
+%
+%   The data w are automatically reshaped to fit x and y.
+
+w = w(:);
+pixels = hsv2rgb([0.5*(1+angle(w)/pi), ones(size(w)), abs(w)/max(abs(w))]);
+pixels = reshape(pixels, numel(x), numel(y), 3);
+h = image(x, y, pixels);
+set(gca, 'YDir', 'normal')
+if nargout==0; clear h; end
+
+end
+

### 2015-04-13 paper.tex
backup before Yosemite upgrade

@@
 
-\topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
+% \topinsert \XeTeXpicfile svgrid.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
+% \topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert
 
@@
 
-\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
-\XeTeXpicfile rsv4.pdf width 0.5\hsize}
-\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
-The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
+% \topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
+% \XeTeXpicfile rsv4.pdf width 0.5\hsize}
+% \XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
+% The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.
 These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert
@@
 
-\topinsert\XeTeXpicfile converge.pdf width \hsize
-Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.
+% \topinsert\XeTeXpicfile converge.pdf width \hsize
+% Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.
 The solid line shows sampling on a regular grid, the dashed one at random amplitudes.
@@
 The tension is the Frobenius norm of the matrix $T$ defined in the text.
-\endinsert
+% \endinsert
 
@@
 
-\topinsert\XeTeXpicfile circle-1.pdf \endinsert
+% \topinsert\XeTeXpicfile circle-1.pdf \endinsert
 
@@
 
-\topinsert \XeTeXpicfile gweight.pdf width \hsize 
-The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
+% \topinsert \XeTeXpicfile gweight.pdf width \hsize 
+% The weights $w_i$ required in order that the integrals of the functions $u_i(x)=e^{-(x-x_i)²}$ are given by the sums $∑_j w_j u_i(x_j)$.
 The points $x_i$ are normally distributed (top), or evenly spaced (bottom).
@@
 I expect experts in Chebyshev sampling would not be surprised by this, but I am not yet such an expert.
-
-The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
+% 
+% The problem with the evenly spaced points is the gaussians at the edge of the grid: only half their support is covered by the grid, so that half needs to be doubly weighted.
 This would overweight the gaussians just inside the grid, so the only solution is an oscillating one.
@@
 So one solution  would be to find the best expansion of $e^{-x²}$, instead of 1, then adjust the weight for $x_i$ by either $e^{-x_i²}$ or $〈e^{-(x-x_i)²},e^{-x²}〉$.
-\endinsert
+% \endinsert
 
@@
 
-\midinsert \XeTeXpicfile tysv.pdf width \hsize
-Figure 3 \endinsert
+% \midinsert \XeTeXpicfile tysv.pdf width \hsize
+% Figure 3 \endinsert
 
@@
 
-\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
+% \topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert
 
@@
 
-\topinsert\XeTeXpicfile cats.pdf width \hsize
-\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
+% \topinsert\XeTeXpicfile cats.pdf width \hsize
+% \vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert
 
@@
 
-\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
-Picard plot for cat expansion. \endinsert
+% \topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
+% Picard plot for cat expansion. \endinsert
 
@@
 
-\topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
-
-\topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
+% \topinsert\XeTeXpicfile catsws.pdf width \hsize\endinsert
+% 
+% \topinsert\XeTeXpicfile catl.pdf width \hsize\endinsert
 
@@
 
-\topinsert \XeTeXpicfile geometry-1.pdf
-A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
+% \topinsert \XeTeXpicfile geometry-1.pdf
+% A 2 dimensional section through Hilbert space, where fixed vector $|Ψ〉$ is expanded as $|Ψ〉=a|α〉+b|β〉$.
 As $|β〉$ becomes more parallel with $|α〉$, the components grow and cancel.
 This occurs for all vectors outside the shaded segments, which shrink to nothing when $|α〉$ becomes parallel to $|β〉$.
-\endinsert
-
-\topinsert \XeTeXpicfile cancel.pdf width \hsize
-The vacuum state was expanded over two coherent states, with amplitudes $±h$.
+% \endinsert
+
+% \topinsert \XeTeXpicfile cancel.pdf width \hsize
+% The vacuum state was expanded over two coherent states, with amplitudes $±h$.
 The amplitudes and the entire expansion of the least-squares fit are plotted in the first row of images.
@@
 This coincides with the amplitudes passing through bright areas of the images, where the entire expansion has support.
-
-The second row of images, and the red lines on the plot, show similar expansions of the normalised state parallel to $|0.1〉-|-0.1〉$.
+% 
+% The second row of images, and the red lines on the plot, show similar expansions of the normalised state parallel to $|0.1〉-|-0.1〉$.
 This is almost exactly parallel to the derivative of the vacuum state with respect to the real part of its amplitude.
@@
 Thus this is exactly the expansion depicted in the last figure.
-
-If the two sample amplitudes were part of an irregular expansion, and there happened to be no other sample amplitudes nearby, a small component along the direction of the second state would dominate the expansion coefficients at these points.
+% 
+% If the two sample amplitudes were part of an irregular expansion, and there happened to be no other sample amplitudes nearby, a small component along the direction of the second state would dominate the expansion coefficients at these points.
 But, given those sample amplitudes, that would be the only way to represent such a component.
-\endinsert
+% \endinsert
 
@@
 
-\topinsert 
-\vskip -40mm
-\XeTeXpicfile psfa.pdf width \hsize
-\vskip -90mm
-\XeTeXpicfile psv.pdf width \hsize\endinsert
+% \topinsert 
+% \vskip -40mm
+% \XeTeXpicfile psfa.pdf width \hsize
+% \vskip -90mm
+% \XeTeXpicfile psv.pdf width \hsize\endinsert
 
@@
 
-\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
+% \topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
+% \topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert
 
@@
 
-\topinsert \XeTeXpicfile spsns.pdf width \hsize
-Amplitudes of components \endinsert
-
-\topinsert \XeTeXpicfile traj.pdf width \hsize
-Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
-
-\topinsert \XeTeXpicfile picard.pdf width \hsize
-Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
-
-\topinsert \XeTeXpicfile svnum.pdf width \hsize
-Expected particle number for left singular vectors.
+% \topinsert \XeTeXpicfile spsns.pdf width \hsize
+% Amplitudes of components \endinsert
+
+% \topinsert \XeTeXpicfile traj.pdf width \hsize
+% Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert
+
+% \topinsert \XeTeXpicfile picard.pdf width \hsize
+% Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert
+
+% \topinsert \XeTeXpicfile svnum.pdf width \hsize
+% Expected particle number for left singular vectors.
 The svs with small sws have large particle number.
@@
 
-\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
-
-\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
+% \topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert
+% 
+% \topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert
 

### 2017-04-24 VCM/VCM5.m
phase space grid investigations

@@ -95,11 +95,14 @@ x = -5:0.2:5;  y = -5:0.2:5;
 Aps = nq*evan(Z,'even');
 
 % plot snapshots in phase space
+% white contour: coherent amplitudes where the component of |alpha> in the span of the frame drops to 0.5.
+% coloured contours: where the norm of the least-squares expansion of |alpha> is 1, 10, and 100 times the length of the largest element in the frame.
 
 for ti = snapshots
 
 	[~,i] = min(abs(t-ti));
 	ensemble = nq*evan(z(:,i));
+
 	% find residuals by taking components in the orthogonal space
 	[U,~,~] = svd(ensemble);  U = U(:,rank(ensemble)+1:end);
 	rsdl = U'*Aps;  rsdl = sqrt(sum(abs(rsdl).^2));  rsdl = reshape(rsdl, size(X));

### 2017-04-24 VCM/norms.m

@@ -1,0 +1,9 @@
+function B = norms(A,dim)
+% like sum, but for the 2-norm
+
+if nargin < 2, dim = 1; end
+B = sqrt(sum(abs(A).^2,dim));
+s = size(B);
+B = reshape(B,[s(1:dim-1) s(dim+1:end)]);
+
+end

### 2017-04-24 VCM/phasegrid.m

@@ -1,0 +1,260 @@
+% VCM5  Coherent state grid solver for the quartic oscillator
+
+global N;  N = 70;  brackets
+
+% Peter's Hamiltonian
+% nhn = -4*(0:N)' gives an interesting sudden divergence
+
+nhn = nhn/2;
+nhn = nhn-4*(0:N)';
+
+a0 = 2;		% coherent amplitude of initial state
+q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
+epsilon=1i*1.e-4;                                 %%stabilize matrix
+
+T=2*pi;  h=0.005;		% time axis
+T = 0.05;
+t = h*(0:ceil(T/h));
+snapshots = h*(0:5);
+
+iters = 4;                                          %%iterations of ODE solver
+
+% set up coherent state grid
+
+l = 1.1;			% frame density, a0 is not a grid point
+[ax,ay] = meshgrid(-4:l:4);  a = ax(:)+1i*ay(:);
+a = a(abs(a) <= 4);
+A = nq*evan(a,'even');
+pinvA = pinv(A);
+
+% initialise storage for data to plot
+
+c = nan(length(a), length(t));
+cc = nan(length(a), length(t), iters);
+
+BUF = nan(1,length(t));
+alpha.o = BUF;  number = BUF;  csize = BUF;  qsize = BUF;
+rsdl = BUF;  rsdln = BUF;
+%urank = BUF;  rrank = BUF;
+%fcondition = BUF;  vcondition = BUF;
+
+% initial condition, the least squares expansion of |a0> over A
+
+c(:,1) = pinvA*q0;
+
+% Exact values for comparison
+
+qe = exp(-1i*nhn*t).*repmat(A*c(:,1),size(t));	% column i is q(t(i))
+alpha.e = sum(conj(qe).*(aop*qe))/norm(A*c(:,1))^2;
+
+% Integration loop
+
+for i = 1:length(t) 
+
+	% collect data to plot
+
+%	ensemble = nq*evan(z(:,i));
+%	D2 = ndqr*evan(z(:,i));  Dq = [ensemble D2];
+	qo = A*c(:,i);
+%	c = sqrt(sum(abs(ensemble).^2));
[…]

### 2017-04-24 VCM/plog.m

@@ -1,0 +1,7 @@
+function plog(x,y,z,a)
+% plot on grid
+
+zplot(x,y,z), axis equal, hold on
+plot(a,'ow')
+
+end

### 2017-04-24 gquad.m
final figures for thesis

@@ -4,6 +4,8 @@
 % its norm.
 
 global N; N = 40; brackets
+mesh = -5:0.2:5;  phasespace
+
 
 cat = sum(nq*evan([2i -2i], 'even'), 2);
 R = 40;  a = gample(cat, R);
@@ -29,6 +31,14 @@ for i=1:numel(ls)
     rs(i) = norm(cat - sum(nq*evan(f, a), 2));
     if any(i==(1:10:51)); fs(:,7-(i+9)/10) = f; end
 end
+
+figure
+zplot(mesh, mesh, Aps'*cat);
+axis image, hold on
+for i=1:numel(a)
+    plot(real(a(i)), imag(a(i)), 'ow')
+end
+saveTightFigure cample.pdf
 
 
 figure
@@ -57,6 +67,7 @@ title(['Picard plot for expanding a cat over ' num2str(R) ...
     ' coherent states'])
 legend('\sigma_r', '|<u_r|\psi>|', '|<u_r|\psi>|/\sigma_r', ...
     'Location', 'SouthWest')
+saveTightFigure capic.pdf
 
 figure
 imagesc(1:nsv, 0:N, abs(U).^2)
@@ -66,11 +77,13 @@ colormap gray; colormap(flipud(colormap))
 xlabel('singular ket |U_r>');  ylabel('fock state |n>')
 
 figure
-subplot 311
 title('L curve for Tychonov regularisation')
 loglog(ns, rs, '-k')
 xlabel('vector norm');  ylabel('residual')
+title 'Cat state L curve'
+saveTightFigure catyc.pdf
 
+figure
 subplot 312
 loglog(ls, rs, '-k')
 set(gca, 'XDir', 'Reverse')

### 2017-04-24 tysv.m

@@ -14,9 +14,9 @@ for i=1:3
         title('Too regularised')
         ylabel('\lambda\sigma^+')
     end
-    if i==3 title('Not regularised enough'); end
+    if i==3 title('Under regularised'); end
     if i>1; set(gca, 'YTick', []); end
     a = 100*a;
 end
 
-print(gcf, '-depsc', 'tysv.eps')
+saveTightFigure tysv.pdf

### 2017-04-24 jcp-0-0.tex
Started paper

\documentclass[aip,jcp,graphicx,draft]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{On the numerical instability of Gaussian wave packets} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{Rodney E. S. Polkinghorne}
\email[]{rpolkinghorne@swin.edu.au}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{Centre for Quantum and Optical Science, Swinburne University of Technology}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
% insert abstract here
\end{abstract}

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands
\section{Introduction}

Since Heller first proposed the method in 1976\cite{jcp-64-63}, there have been several attempts to compute the dynamics of quantum systems by use of coupled Gaussian wave packets.

{\tt Systems treated}

{\tt Where the method worked}

{\tt When instabilities were noticed}

{\tt What people tried to do about them}

{\tt Proposal for quantum fields}

{\tt Wavelets and Gabor frames in the 1980s}

\section{The naive method}

\section{Expansions over frames of wave packets}

\section{Stiff equations of motion}

\section{Local regularisation}

\section{Global regularisation}

\section{Prospects}

%\label{}
\subsection{}
\subsubsection{}

% If in two-column mode, this environment will change to single-column format so that long equations can be displayed. 
% Use only when necessary.
%\begin{widetext}
%$$\mbox{put long equation here}$$
%\end{widetext}

% Figures should be put into the text as floats. 
% Use the graphics or graphicx packages (distributed with LaTeX2e).
% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz, and Frank Mittelbach for examples. 
%
% Here is an example of the general form of a figure:
% Fill in the caption in the braces of the \caption{} command. 
% Put the label that you will use with \ref{} command in the braces of the \label{} command.
%
% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}%
% \end{figure}

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

% If you have acknowledgments, this puts in the proper section head.
%\begin{acknowledgments}
% Put your acknowledgments here.
%\end{acknowledgments}

% Create the reference section using BibTeX:
\bibliography{your-bib-file}

\end{document}
%
% ****** End of file aiptemplate.tex ******

### 2017-04-26 semii.tex
started notes

\input respnotes

\title Failure of the Iterated Semi-Implicit method at large energy

The semi-implicit method is usually used to solve partial differential equations.  However, in this application it is used as a method of lines, and the semi-implicit method is a way to solve the resulting ordinary differential equations.

For the differential equation~${dx\over dt}=f(x,t)$, the SI formula with time step~$h$ is
$$x_{n+1}=x_n+hf_{n+½},\qquad f_{n+½}=f\left({x_n+x_{n+1}\over 2}\right).$$
This has the nice property that it is symplectic in ket space, but the obvious problem is evaluating~$f_{n+½}$.  This is often done iteratively, with
$$\eqalign{
	x_{n+1}⁰&=x_n\qquad f_{n+½}⁰=f(x_n)\cr
	x_{n+1}^{i+1}&=x_n+hf_{n+½}ⁱ\qquad f_{n+½}^{i+1}=f\left({x_n+x_{n+1}ⁱ\over 2}\right)\cr
}$$

Now consider the quantum case, where~$x$ is the expansion of a ket over an orthonormal basis, and~$f$ is a linear Hamiltonian operator~$-iH$.  In this case, the iteration rule becomes
$$x_{n+1}^{i+1}=x_n+hf_{n+½}ⁱ\qquad f_{n+½}^{i+1}=-{i\over2}Hx_n-{i\over2}Hx_{n+1}ⁱ$$
or
$$x_{n+1}^{i+1}=(1-{ih\over2}H)x_n-{ih\over2}Hx_{n+1}ⁱ$$
with solution
$$x_{n+1}ⁱ=Ux_n=\left(1-ihH-{(hH)²\over 2}+{i(hH)³\over 4}+…+2\left({-ihH\over 2}\right)ⁱ\right)x_n.$$
The eigenvalues of the time step operator~$U$ are the same as those of~$H$.  The iteration will converge if{f} the eigenvalues of~$H$ are bounded by~$2/h$, a kind of Nyquist condition.

Actually, this condition only restricts the eigenvalues whose eigenkets have components in~$x_n$.  Consider a state~$x₀$ with moderate expected energy, but with a small component with a large energy, such as a coherent state.  The iteration will appear to converge for a small number of cycles, and the formula will appear to solve the differential equation.  However, the component with large energy will be amplified by the operator~$U$ at each timestep, and eventually this parasitic solution will dominate the real solution.

This is illustrated in a numerical experiment with the Hamiltonian
$$H=½(a†)²a²-4a†a,$$
and the initial coherent state~$|o(2)〉$ such that~$〈H〉=0$.  The state was expanded over set of 41 coherent states, whose amplitudes form the grid in the figure below.  The figure also shows the largest eigenstate of the discrete Hamiltonian on this grid.

\XeTeXpdffile VCM/aa.pdf width 0.7\hsize

The second figure shows an expansion of this eigenstate over number states.  The most energetic eigenstate is a combination of~$|42〉$ and~$|43〉$, which is not unexpected: any set of 41 number states near the origin of phase space will span nearly the same space as~$|0〉$ through~$|40〉$.

\XeTeXpdffile VCM/ab.pdf width 0.7\hsize

The next figure shows the initial state~$|o(2)〉$ propagated with a 4-iteration semi-implicit rule.  When the initial state is discretised, the components of the number states~$|0〉$ through roughly~$|40〉$ decay exponentially, as they should for a coherent state.  The larger number states form a residual component, that lies orthogonal to the span of the coherent state grid and can not be represented in this discretisation.

The phases of the number state components should rotate, but their sizes should remain constant.  All of the iterations are shown, and the jump every 4 steps comes at the start of each timestep, where the converged solution is replaced by a new rough estimate.  The exponential growth of the parasitic mode is obvious.

\XeTeXpdffile VCM/ac.pdf width 0.7\hsize

The final figure shows the eigenvalues of the exact and discretised Hamiltonians, along with the Nyquist energy limit~$2/h$.  It is clear that parasitic modes will be present.

\XeTeXpdffile VCM/ad.pdf width 0.7\hsize

In an expansion over coherent states, constraining the size of the expansion coefficients has the effect of constraining the energy of the expanded state!  So a solution with constrained norm ought to limit the growth of the parasitic modes, and restrict the simulated state to the part of phase space near the grid.  We will implement this by finding a constrained least-squares solution at each iteration.

\bye

### 2017-04-26 VCM/phasegrid.m
About to put size constraints in phasegrid

@@ -9,6 +9,7 @@ nhn = nhn/2;
 nhn = nhn-4*(0:N)';
 
 a0 = 2;		% coherent amplitude of initial state
+% q0 = zeros(N+1,1);  q0(5) = 1;		% number state 2
 q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
 epsilon=1i*1.e-4;                                 %%stabilize matrix
 
@@ -53,20 +54,13 @@ for i = 1:length(t)
 
 	% collect data to plot
 
-%	ensemble = nq*evan(z(:,i));
-%	D2 = ndqr*evan(z(:,i));  Dq = [ensemble D2];
 	qo = A*c(:,i);
-%	c = sqrt(sum(abs(ensemble).^2));
 	rsdl(i) = norm(qe(:,i)-qo);
 	rsdln(i) = sum(abs(qe(:,i)-qo).^2.*(0:N)')/rsdl(i)^2;
 	qsize(i) = norm(qo);
 	alpha.o(i) = qo'*aop*qo/qsize(i)^2;
 	number(i) = sum(abs(qo).^2.*(0:N)')/qsize(i)^2;
 	csize(i) = norm(c(:,i));
-%	urank(i) = rank(Dq);   rrank(i) = rank([Dq; sqrt(epsilon)*eye(2*R)]);
-%	fcondition(i) = cond(ensemble);  vcondition(i) = cond(Dq);
-%	d = sqrt(sum(abs(D2).^2));
-%	component_lengths(:,i) = [min(d) min(c) max(c)  max(d)];
  	
 	if i == length(t), break, end
 
@@ -84,25 +78,11 @@ end
 
 % Eigenvalues of discretised Hamiltonian and step operators
 
-M = pinvA*diag(1i*nhn)*A;  ew = flip(eig(M));
-figure, plot(1:length(a),real(ew),'+k',1:length(a),imag(ew),'.k',1:length(a),nhn(1:41),'or')
-title 'Discretised H eigenvalues'
-
-P = 1+h*M;  ew = flip(eig(P));
-figure, semilogy(1:length(a),real(ew),'+k',1:length(a),imag(ew),'.k')
-title '1st iteration eigenvalues'
-
-P = 1+h*M*P;  ew = flip(eig(P));
-figure, semilogy(1:length(a),real(ew),'+k',1:length(a),imag(ew),'.k')
-title '2nd iteration eigenvalues'
-
-P = 1+h*M*P;  ew = flip(eig(P));
[…]

### 2017-04-28 semii.tex
phase space grids and regularisation

@@
 
+(Insert figures from phasegrid below.)
+
+It turns out that the steps of the integration formula quickly become larger than the state vector.
+This can be solved by Tychonov conditioning.
+Constraining~$dc$ to be orthogonal to~$c$ doesn't help, because~$dc$ is too large for the small angle approximations to work, and this does not keep the norm constant.
+
+The matrix~$〈A⁺|H|A〉$ is not normal.
+However, because~$|A〉〈A⁺|≈1$, the coefficient vectors~$〈A⁺|E_i〉$ are very nearly eigenvectors, with eigenvalues~$E_i$.
+The approximation is very good for low energy, but starts to fail as the energy~$E_i$ approaches the highest energy in the span of $|A〉$.
+Presumably the eigenvectors are those of the projection of~$H$ on to the span of~$|A〉$.
+
+If~$〈A⁺|$ is Tychonov regularised, then the same argument should work on the low energy space where the regularised expansion is a good approximation.
+But it doesn't appear to: eigenvalues of~$H$ are shifted that are low enoug that the corresponding singular vectors of~$A$ should hardly be touched by regularisation.
+
+Todo.  For number states~$|n〉 with energies~$E_n$, see how the size of~$〈A⁺|n〉$ varies with the lagrange multiplier.
+
 \bye

### 2017-04-28 VCM/phasegrid.m

@@ -11,12 +11,18 @@ nhn = nhn-4*(0:N)';
 a0 = 2;		% coherent amplitude of initial state
 % q0 = zeros(N+1,1);  q0(5) = 1;		% number state 2
 q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
-epsilon=1i*1.e-4;                                 %%stabilize matrix
 
 T=2*pi;  h=0.005;		% time axis
-T = 0.05;
+% T = 0.05;
 t = h*(0:ceil(T/h));
-snapshots = h*(0:5);
+snapshots = [h 2*h 0.1 1 2 4];
+
+gfac = 1.5;		% relative coefficient growth limit
+% gfac = inf;
+
+laf = 3e-4;		% Lagrange multiplier for Tyc conditioning.
+% laf = 3e-4;	% value where parasitic solution takes over at t=pi
+% laf = 0;
 
 iters = 4;                                          %%iterations of ODE solver
 
@@ -29,6 +35,7 @@ A = nq*evan(a,'even');
 pinvA = pinv(A);
 
 % initialise storage for data to plot
+% cc(:,i,:) converge to c(:,i)
 
 c = nan(length(a), length(t));
 cc = nan(length(a), length(t), iters);
@@ -36,12 +43,11 @@ cc = nan(length(a), length(t), iters);
 BUF = nan(1,length(t));
 alpha.o = BUF;  number = BUF;  csize = BUF;  qsize = BUF;
 rsdl = BUF;  rsdln = BUF;
-%urank = BUF;  rrank = BUF;
-%fcondition = BUF;  vcondition = BUF;
 
 % initial condition, the least squares expansion of |a0> over A
 
 c(:,1) = pinvA*q0;
+glim = gfac*norm(c(:,1));
 
 % Exact values for comparison
 
@@ -52,7 +58,7 @@ alpha.e = sum(conj(qe).*(aop*qe))/norm(A*c(:,1))^2;
 
 for i = 1:length(t) 
 
-	% collect data to plot
+	% collect state data
 
 	qo = A*c(:,i);
 	rsdl(i) = norm(qe(:,i)-qo);
@@ -70,7 +76,24 @@ for i = 1:length(t)
 	for j = 1:iters
[…]

### 2017-08-16 jcp-0-0.tex
paper push

@@
 
-\title{On the numerical instability of Gaussian wave packets} %Title of paper
+\title{On the numerical stability of static Gaussian wave packets} %Title of paper
 
@@
 
+\section{Ad-hoc regularisation}
+
 {\tt Systems treated}
@@
 
-\section{The naive method}
+\section{Differential equation formulae and stiffness}
 
-\section{Expansions over frames of wave packets}
+\section{Expansions over wave packets}
 
-\section{Stiff equations of motion}
+{\tt Many wave packets, large number states can be represented, discretised $H$ has large evs, discretised equations of motion are stiff.}
+
+{\tt Why do the least norm expansions of number states have large components near the origin of phase space?}
 
@@
 
-\section{Global regularisation}
+{\tt Effects on evs, stabilisation of dynamics.}
 
 \section{Prospects}
+
+{\tt Other things go wrong when amplitudes are varied, need to study further.}
 

### 2017-08-16 semii.tex
paper push

@@
 
+Explanation.
+Suppose the energy eigenstate~$|E_i〉$ has components along singular directions of~$|A〉$ whose singular values are around~$σ$.
+When the matrix~$〈A⁺_λ|H|A〉$ acts on the coefficient vector~$〈A⁺|E_i〉$, the vector is multiplied successively by scalars of size around~$σ$, $E_i$, and~$σ/(σ²+λ²)$, whose product is~$E_i×σ²/(σ²+λ²)$.
+
+Tychonov regularisation effectively turns off the dynamics in the high-energy Hilbert subspace where the singular values of~$|A〉$ are small.  This is why the semi-implicit method becomes stable.
+
 Todo.  For number states~$|n〉 with energies~$E_n$, see how the size of~$〈A⁺|n〉$ varies with the lagrange multiplier.

### 2017-08-16 slnsize.tex
paper push

\input xpmath
\input respnotes

The way to regularise an ill-posed problem is to bound the size of the solution.  This is straightforward when the problem is being solved once, more complicated when the result is a derivative that will be integrated to find the actual solution.  In this case, small changes can add up to give a large solution, most of which comes from over-fitting.  We want to bound the size of the solution, not the size of the changes.

In the variational gaussian problem, the solution is an expansion $|ψ({\bf φ,α})〉=∑_re^{φ_r}|α_r〉$.  The $|α_r〉$ are translationally invariant, so the size of $\bf α$ doesn't seem to matter, and the size of the solution is measured by $M=\|(e^{φ₁},…,e^{φ_R})\|²=∑_re^{2\Re φ_r}$.  The increment of this is
$$dM=∑_r2e^{2\Re φ_r}\,d(\Re φ_r).$$
We want this increment to be mildly negative, say set it to $-ε\,dt$.

The ill-posed problem is the Dirac-Frankel one,
$$|Dψ(z)〉{\dot z}=-i\hbar H|ψ(z)〉.$$
The conditioning uses $\Re φ$, which is a nonlinear function of $φ$ with respect to complex numbers.  So we must detour to convert this problem to a linear one over real numbers.  Let $φ=x+iy$, and $α=u+iv$.  Kets don't have real and imaginary parts, but their computer representations do, and we'll use $\Re|ψ(z)〉$ and $\Im|ψ(z)〉$ to denote these.  The problem becomes
$$\pmatrix{\Re|D₁ψ〉&-\Im|D₁ψ〉&\Re|D₂ψ〉&-\Im|D₂ψ〉\cr
	\Im|D₁ψ〉&\Re|D₁ψ〉&\Im|D₂ψ〉&\Re|D₂ψ〉
	}\cdot
	\pmatrix{\dot x\cr\dot y\cr\dot u\cr\dot v}
	=\hbar\pmatrix{\Im H|ψ〉\cr -\Re H|ψ〉}.
$$
when split into real and imaginary parts.  In this expression, $|ψ(z)〉$ has been abbreviated to $|ψ〉$.

The appropriate norm for $dz$ is $\|φ\|₂+\|α\|_∞$.  An ensemble with more components should not have a bigger norm if the $α$ change by the same amount.

\bye

### 2017-08-16 VCM/norms.m

@@ -3,7 +3,9 @@ function B = norms(A,dim)
 
 if nargin < 2, dim = 1; end
 B = sqrt(sum(abs(A).^2,dim));
-s = size(B);
-B = reshape(B,[s(1:dim-1) s(dim+1:end)]);
+if ndims(B) > 2
+	s = size(B);
+	B = reshape(B,[s(1:dim-1) s(dim+1:end)]);
+end
 
 end

### 2017-08-16 VCM/phasegrid.m

@@ -31,6 +31,7 @@ iters = 4;                                          %%iterations of ODE solver
 l = 1.1;			% frame density, a0 is not a grid point
 [ax,ay] = meshgrid(-4:l:4);  a = ax(:)+1i*ay(:);
 a = a(abs(a) <= 4);
+[~, i] = sort(abs(a));  a = a(i);
 A = nq*evan(a,'even');
 pinvA = pinv(A);
 
@@ -99,31 +100,80 @@ for i = 1:length(t)
 
 end
 
+% Suffix convention: ev, ew, sw, u = lsv, v = rsv
+
+Hew = nhn(1:length(a));
+M = pinv(A)*diag(nhn)*A;
+[ev,ew] = eig(M);  ew = diag(ew);
+% ns = (norms(aop*A*ev)./norms(A*ev)).^2;
+ns = sum(diag(0:N)*abs(A*ev).^2)./norms(A*ev).^2;
+[ns, P] = sort(ns);
+ev = ev(:,P);  ew = ew(P);
+[Au,Asw,Av] = svd(A,'econ');  Asw = diag(Asw);
+% Tychonov regularised H
+Hy = pinv([A; laf*eye(length(a))])*[diag(nhn)*A; zeros(length(a))];
+[Hyev, Hyew] = eig(Hy);  Hyew = diag(Hyew);
+% yns = (norms(aop*A*Hyev)./norms(A*Hyev)).^2;
+yns = sum(diag(0:N)*abs(A*Hyev).^2)./norms(A*Hyev).^2;
+[yns, P] = sort(yns);
+Hyev = Hyev(:,P);  Hyew = Hyew(P);
+
 % Eigenvalues of discretised Hamiltonian and step operators
 
-M = pinv(A)*diag(nhn)*A;
+figure
+zplot(1:length(a), 0:N, Au), axis image
+title 'svs of A'
+
+unit = eye(N+1,length(a));
+Afock = pinv(A)*unit;
+Afock = Afock/diag(norms(Afock));
+
+figure, subplot 221
+zplot(1:length(a), 1:length(a), ev), axis image
+title 'evs of discretised H'
+
+subplot 222
+tmp2 = A*ev;  tmp2 = tmp2/diag(norms(tmp2));
+imagesc(abs(tmp2(1:length(a),:)).^2), colormap gray, axis image
+set(gca, 'YDir', 'normal')
+title 'Fock discretised H'
+
+subplot 223
+zplot(1:length(a), 1:length(a), Afock), axis image
[…]

### 2017-08-16 VCM/plog.m

@@ -1,7 +1,7 @@
 function plog(x,y,z,a)
 % plot on grid
 
-zplot(x,y,z), axis equal, hold on
+zplot(x,y,z), axis image, hold on
 plot(a,'ow')
 
 end

### 2017-08-16 gquad.m

@@ -67,6 +67,7 @@ title(['Picard plot for expanding a cat over ' num2str(R) ...
     ' coherent states'])
 legend('\sigma_r', '|<u_r|\psi>|', '|<u_r|\psi>|/\sigma_r', ...
     'Location', 'SouthWest')
+xlabel 'r', ylabel 'singular value and vector sizes'
 saveTightFigure capic.pdf
 
 figure

### 2017-08-16 phasexamples.m

@@ -2,12 +2,13 @@
 
 global N; N = 24; brackets
 mesh = -5:0.1:5;  phasespace
-ht = 4;
+ht = 3.7;
 
 figure		% vacuum state
 ax = zplot(mesh, mesh, Aps'*nq*evan(0, 'even'));
-ax.XTick = -4:2:4;  ax.YTick = -4:2:4;
+ax.XTick = -4:4:4;  ax.YTick = -4:4:4;
 ax.Units = 'centimeters';  ax.Position = [0.5 0.5 ht ht];
+xlabel 'Re \alpha',  ylabel 'Im \alpha'
 axis image
 saveTightFigure twoa.pdf
 
@@ -27,10 +28,22 @@ ax.Units = 'centimeters';  ax.Position = [0.5 0.5 ht ht];
 axis off, axis image
 saveTightFigure twoc.pdf
 
+r = linspace(0.75,1,10);
+theta = linspace(0, 2*pi, 100);
+[rg, thg] = meshgrid(r,theta);
+[x,y] = pol2cart(thg,rg);
+figure, ax = axes;
+pcolor(x,y,thg);
+colormap(phase(128));
+ax.Units = 'centimeters';  ax.Position = [0.5 0.5 ht ht];
+axis off, axis image
+text(0.85,0,'1'), text(-0.95,0,'-1'), text(0,0.85,'i'), text(0,-0.9,'-i')
+shading flat;
+saveTightFigure twod.pdf
+
 figure, ax = axes;
 image([113:127 0:128 1:15]'), colormap(phase(129))
 ax.YDir = 'normal';  ax.XTick = [];
 ax.YTick = 16+32*(0:4);  ax.YTickLabel = {'1', 'i', '-1', '-i', '1'};
 ax.YAxisLocation = 'right';	% saveTightFigure clips the left side
 ax.Units = 'centimeters';  ax.Position = [0.5 0.5 0.6 ht];
-saveTightFigure twod.pdf

### 2017-08-16 qplot.m

@@ -19,7 +19,15 @@ for i = 1:numel(ts)
 	figure, zplot(mesh, mesh, Aps'*c), hold on, axis image
 	plot(0, 0, 'ow', real(a), imag(a), '+w'), axis off
 	saveTightFigure(sprintf('qrt%02d.pdf', i))
-%	title(sprintf('Q.O. at t = %.2f*2\\pi', t))
+
+	figure, zplot(mesh, mesh, widi(c,z)), hold on, axis image
+	plot(0, 0, 'ow', real(a), imag(a), '+w'), axis off
+	saveTightFigure(sprintf('qrtw%02d.pdf', i))
+
+	z0 = exp(-2i*pi*t*(2*abs(z).^2-1)).*z;
+	figure, zplot(mesh, mesh, exp(-2*abs(z0-3.5).^2))
+	axis image, axis off
+	saveTightFigure(sprintf('qrtt%02d.pdf', i))
 end
 
 tt = 0:0.001:0.25;  m = numel(tt);

### 2017-08-16 zplot.m

@@ -13,7 +13,7 @@ if nargin < 4, cmap = phase(128); end
 ff = linspace(0, 2*pi, size(cmap,1));
 w = w(:);  f = mod(angle(w), 2*pi);
 pixels = repmat(abs(w)/max(abs(w)), 1, 3).*interp1(ff, cmap, f);
-pixels = reshape(pixels, numel(x), numel(y), 3);
+pixels = reshape(pixels, numel(y), numel(x), 3);
 image(x, y, pixels)
 h = gca;
 set(h, 'YDir', 'normal')

### 2017-10-09 ssc-0-0.tex
add material to ssc paper

% SIAM Article Template
\documentclass[review]{siamart1116}

% Packages and macros go here
\usepackage{lipsum}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithmic}
\ifpdf
  \DeclareGraphicsExtensions{.eps,.pdf,.png,.jpg}
\else
  \DeclareGraphicsExtensions{.eps}
\fi

%strongly recommended
\numberwithin{theorem}{section}

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{Stability of the semi-implicit formula for quantum dynamics} 
\newcommand{\TheAuthors}{Rodney E. S. Polkinghorne}

% Sets running headers as well as PDF title and authors
\headers{\TheTitle}{\TheAuthors}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{{\TheTitle}\thanks{Submitted to the editors DATE.
\funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}}

% Authors: full names plus addresses.
\author{
  Rodney E. S. Polkinghorne\thanks{Swinburne University of Technology
    (\email{rpolkinghorne@swin.edu.au}, \url{http://www.imag.com/\string~ddoe/}).}
}

\usepackage{amsopn}
\DeclareMathOperator{\diag}{diag}

%%% Local Variables: 
%%% mode:latex
%%% TeX-master: "ex_article"
%%% End:

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={\TheTitle},
  pdfauthor={\TheAuthors}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

\externaldocument{ex_supplement}

% FundRef data to be entered by SIAM
%<funding-group>
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>

\begin{document}

\maketitle

% REQUIRED
\begin{abstract}
  This is an example SIAM \LaTeX\ article. This can be used as a
  template for new articles.  Abstracts must be able to stand alone
  and so cannot contain citations to the paper's references,
  equations, etc.  An abstract must consist of a single paragraph and
  be concise. Because of online formatting, abstracts must appear as
  plain as possible. Any equations should be inline.
\end{abstract}

% REQUIRED
\begin{keywords}
  example, \LaTeX
\end{keywords}

% REQUIRED
\begin{AMS}
  68Q25, 68R10, 68U05
\end{AMS}

\section{Introduction}

The semi-implicit method \cite{jkp-132-312,jkp-93-144} is often used to solve stochastic partial differential equations that arise when Monte-Carlo methods are used in quantum optics and in atom optics \cite{pra-58-4824}.  This is done as a method of lines, in which the spatial axes are discretised, often by spectral derivatives, to give a system of ordinary differential equations.  The semi-implicit formula is then used to integrate these.

For a system of equations~${dx\over dt}=f(x,t)$, with time step~$h$,
the formula is
\begin{equation}\label{eq:i}
    x_{n+1}=x_n+hf_{n+\frac12},
\end{equation}
where
\begin{equation}\label{eq:ii}
    f_{n+\frac12}=f\left({x_n+x_{n+1}\over 2}\right).
\end{equation}

The formula has the theoretical advantage that it conserves Hilbert space norm when it is applied to a quantum equation of motion of the form
\begin{equation}\label{eq:iii}
	{dx\over dt}=-iHx,
\end{equation}
where $H$ is a hermitian operator.  However, it is mostly used because it has been found to work well in practice.

It is not obvious how to apply this formula, because of the need
to evaluate~$x_{n+1}$ and~$f_{n+\frac12}$ self-consistently.  The
usual way of doing this is a very simple iterative one.  At first,
$x_{n+1}$ is set to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then
Equations~\cref{eq:i,eq:ii} are evaluated in turn, and the values of~$x_n$
and~$f_{n+\frac12}$ updated.  This ought to continue until they
converge, but in practice this is continued for a fixed number of
iterations.  Formally, this process gives a sequence of~$x_{n+1}^j$
and~$f_{n+\frac12}^j$, with
\begin{subequations}
\begin{align}
	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}=f\left({x_n+x_{n+1}^j\over 2}\right),\qquad j\ge0.
\end{align}
\end{subequations}
Any fixed point of this iteration is a self-consistent solution to
Equations~\cref{eq:i,eq:ii}, but the question remains of when the
iteration converges.

It is a generally known problem that the time steps in formulae for
partial differential equations need to be reduced in proportion to
the spatial grid steps, in order for the formulae to remain stable.
To date, no one has made a specific analysis of the iterative
semi-implicit formula, so the stable time step has had to be
determined by trial and error.  This can be a time consuming process
for fields that are being simulated on large grids.

This paper analyses the semi-implicit formula, in the case that it
is used to integrate linear Hamiltonian dynamics as in
Equation~\cref{eq:iii}.  The main result is that the stability
depends on a Nyquist-like condition, and the iterations converge
if and only if the largest eigenvalue of~$H$ is no more than~$2/h$.

\section{Stability analysis}

Suppose that the equation being integrated has the form of
Equation~\cref{eq:iii}, where~$x$ is a vector of coefficients
obtained by expanding a ket~$|\psi\rangle$ over an orthonormal
basis, and~$f$ is the linear operator~$-iH$, where~$H$ is the matrix
for the Hamiltonian operator in this basis.  The
iteration rules of Equation~\cref{eq:v} become
$$x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j$$
or
$$x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j$$
with solution
$$x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.$$
The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$ as the
hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the sums of a geometric series in~$\omega_m$, with the relation between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$\omega_m^j$ of~$U_j$.  The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.  The exact time evolution operator~$U$ is unitary, with unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The sequence of~$U_j$ is a power series, which will converge if and only if the angular frequencies~$\omega_m$ are bounded by~$2/h$, a kind of Nyquist
condition.  If the discretised Hamiltonian has an eigenstate~$x_\omega$
with a frequency~$\omega\gg 2/h$, then~$U$ will magnify the component
of~$x_\omega$ in~$x$ by a factor close to~$(h\omega/2)^i$, and this
component is a parasitic solution growing steadily at the
rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.

The semi-implicit method is rarely used to solve the linear Equation~\cref{eq:iii}, and more often used to solve a nonlinear Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{something}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

\section{Numerical experiments}

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
  \label{fig:i}
\end{figure}

\section{Conclusion}

The semi-implicit formula considered here is a special case, but most ODE formulae will become unstable when frequencies get large compared to time steps.

Stiff formulae stay stable, but reject the physics that is happening on fast timescales.  If we want to simulate physics at frequency~$\omega$, it seems reasonable to require time steps~$?$.

The problem comes when the interesting physics occurs at slow timescales, but the state representation is inadvertently capable of representing high energies that evolve on fast timescales.

In the case here, where we expanded over energy eigenstates, you can simply truncate.  This is contrived: if you know the energy eigenstates, you already know the answer, and you don't need to simulate anything.

In general, you will use a variational ansatz, or some other tricky representation.  In this case, some care must be taken to avoid inadvertently representing high energy states.  Details to come in future papers.

\section*{Acknowledgments}
We would like to acknowledge the assistance of volunteers in putting
together this example manuscript and supplement.

\bibliographystyle{siamplain}
\bibliography{references}
\end{document}

### 2017-10-10 ssc-0-0.tex
writing

@@
 % Declare title and authors, without \thanks
-\newcommand{\TheTitle}{Stability of the semi-implicit formula for quantum dynamics} 
+\newcommand{\TheTitle}{Stability of the iterative semi-implicit formula for quantum dynamics} 
 \newcommand{\TheAuthors}{Rodney E. S. Polkinghorne}
@@
 
-The semi-implicit method \cite{jkp-132-312,jkp-93-144} is often used to solve stochastic partial differential equations that arise when Monte-Carlo methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
-This is done as a method of lines, in which the spatial axes are discretised, often by spectral derivatives, to give a system of ordinary differential equations.
+The semi-implicit formula is a rule for integrating systems of
+ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
+is often used as a method of lines, to solve stochastic partial
+differential equations that occur when phase space Monte-Carlo
+methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
+This is done as a method of lines, in which the spatial axes are
+discretised, often by spectral derivatives, to give a system of
+ordinary differential equations \cite{2000-Trefethen-Spectral}.
 The semi-implicit formula is then used to integrate these.
+
+{\bf Relate Hilbert space expansions to methods of lines.}
 

### 2017-10-12 ssc-0-0.tex
writing

@@
 
+% formula equations
 The semi-implicit formula is a rule for integrating systems of
@@
 methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
-This is done as a method of lines, in which the spatial axes are
-discretised, often by spectral derivatives, to give a system of
-ordinary differential equations \cite{2000-Trefethen-Spectral}.
-The semi-implicit formula is then used to integrate these.
+These equations are related to Schr\"odinger's equation,
+\begin{equation}
+    \frac{\partial\psi}{\partial t} = -\del^2\psi + V\psi,
+\end{equation}
+with additional nonlinear and random terms.  The spatial axes are
+discretised, often using spectral derivative operators, to give a
+system of ordinary or stochastic differential equations
+\cite{2000-Trefethen-Spectral}, and the semi-implicit formula is
+used to integrate these equations.
 
-{\bf Relate Hilbert space expansions to methods of lines.}
+{\bf Relate Hilbert space expansions to methods of lines using sync wave packets.}
 

### 2017-10-17 ssc-0-0.tex
complete a first draft of the semi-implicit paper

@@
 \usepackage{amsopn}
-\DeclareMathOperator{\diag}{diag}
+\DeclareMathOperator{\sinc}{sinc}
 
@@
 These equations are related to Schr\"odinger's equation,
-\begin{equation}
-    \frac{\partial\psi}{\partial t} = -\del^2\psi + V\psi,
+\begin{equation}\label{eq:vi}
+    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
 \end{equation}
@@
 system of ordinary or stochastic differential equations
-\cite{2000-Trefethen-Spectral}, and the semi-implicit formula is
-used to integrate these equations.
+\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
+by the semi-implicit formula.
 
-{\bf Relate Hilbert space expansions to methods of lines using sync wave packets.}
-
-For a system of equations~${dx\over dt}=f(x,t)$, with time step~$h$,
-the formula is
+For a system of equations~${dx\over dt}=f(x,t)$,
+the semi-implicit formula with time step~$h$ is
 \begin{equation}\label{eq:i}
@@
 \begin{equation}\label{eq:ii}
-    f_{n+\frac12}=f\left({x_n+x_{n+1}\over 2}\right).
+    f_{n+\frac12}=f\left({\textstyle\frac{x_n+x_{n+1}}{2}}\right).
 \end{equation}
-
-The formula has the theoretical advantage that it conserves Hilbert space norm when it is applied to a quantum equation of motion of the form
+When this formula is applied to equations of a form similar to~\cref{eq:vi},
 \begin{equation}\label{eq:iii}
@@
 \end{equation}
-where $H$ is a hermitian operator.  However, it is mostly used because it has been found to work well in practice.
+where~$H$ is a hermitian operator acting on the complex vector~$x$,
+there is an advantage in principle because the norm of~$x$ is
+conserved in the discretised solution.  {\bf Expand on this.}  However, the formula is
+used mainly because it has been found to work well in practice.
 
-It is not obvious how to apply this formula, because of the need
-to evaluate~$x_{n+1}$ and~$f_{n+\frac12}$ self-consistently.  The
-usual way of doing this is a very simple iterative one.  At first,
-$x_{n+1}$ is set to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then
-Equations~\cref{eq:i,eq:ii} are evaluated in turn, and the values of~$x_n$
-and~$f_{n+\frac12}$ updated.  This ought to continue until they
-converge, but in practice this is continued for a fixed number of
-iterations.  Formally, this process gives a sequence of~$x_{n+1}^j$
+% iterations
+It is not obvious how to implement this formula, because~$x_{n+1}$
+and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
+usual way is a simple iteration.  To start,
+$x_{n+1}$~is initialised to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then each iteration
+evaluates Equations~\cref{eq:i,eq:ii} in turn, and updates the values
+of~$x_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the iterations generate a sequence of~$x_{n+1}^j$
 and~$f_{n+\frac12}^j$, with
@@
 	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
-	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}=f\left({x_n+x_{n+1}^j\over 2}\right),\qquad j\ge0.
+	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j
+		\qquad f_{n+\frac12}^{j+1}=f\left({\textstyle\frac{x_n+x_{n+1}^j}{2}}\right),\qquad j\ge0.
 \end{align}
 \end{subequations}
-Any fixed point of this iteration is a self-consistent solution to
-Equations~\cref{eq:i,eq:ii}, but the question remains of when the
-iteration converges.
+This continues until the sequences converge, which is usually
+achieved after a small, fixed number of iterations.  Any fixed point
+of the iteration is a self-consistent solution to
+Equations~\cref{eq:i,eq:ii}.  The remaining question is under what
+conditions the iterations converge.
 
-It is a generally known problem that the time steps in formulae for
-partial differential equations need to be reduced in proportion to
-the spatial grid steps, in order for the formulae to remain stable.
-To date, no one has made a specific analysis of the iterative
-semi-implicit formula, so the stable time step has had to be
-determined by trial and error.  This can be a time consuming process
-for fields that are being simulated on large grids.
+% steps stability
+Generally, when a partial differential equation is discretised by
+a given formula, the time step needs to be reduced in proportion
+to the grid spacing in order for the discretised solution to remain
+stable \cite{something}.  No one has analysed the stability of the
+semi-implicit formula with iterations.  The stable time step has
+had to be determined by trial and error, a time consuming process
+for three-dimensional wave functions that are being simulated on
+large grids.
 
@@
 
-Suppose that the equation being integrated has the form of
-Equation~\cref{eq:iii}, where~$x$ is a vector of coefficients
-obtained by expanding a ket~$|\psi\rangle$ over an orthonormal
-basis, and~$f$ is the linear operator~$-iH$, where~$H$ is the matrix
-for the Hamiltonian operator in this basis.  The
+% basis coefficients
+When the semi-implicit rule is applied to \cref{eq:vi}, or a
+stochastic or nonlinear variant, the laplacian operator is usually
+evaluated by a spectral method.  This is equivalent to expanding
+the wave function~$\psi(q)$ over a basis of sinc wave packets
+\begin{equation}
+	\psi(q) = \sum x_i\phi_i(q),
+\end{equation}
+{\bf figure out range of $i$}
+where
+\begin{equation}
+	\phi_i(q) = \frac1{\sqrt{l}} \sinc(\textstyle\frac{q}{l} - i).
+\end{equation}
+This paper uses the convention
+\begin{equation}
+	\sinc{q} = \frac{\sin(\pi q)}{\pi q},
+\end{equation}
+under which the functions~$\phi_i$ are an orthonormal basis for the
+space of functions limited to bandwidth~$1/2l$.  The spectral method
+reduces \cref{eq:vi} to an equation of the form~\cref{eq:iii} for
+the coefficient vector~$x$, where
+\begin{equation}
+	H_{ij} = \int \phi_i(q) (-\nabla^2 + V(q)) \phi_j^\ast(q).
+\end{equation}
+
+At this point, the
 iteration rules of Equation~\cref{eq:v} become
-$$x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
-	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j$$
+\begin{equation}
+x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
+	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j
+\end{equation}
 or
-$$x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j$$
-with solution
-$$x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
-	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.$$
-The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$ as the
-hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the sums of a geometric series in~$\omega_m$, with the relation between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.
+\begin{equation}
+x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j
+\end{equation}
+with solution {\bf check factor of 2}
+\begin{equation}\label{eq:viii}
+x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
+	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.
+\end{equation}
+The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$
+as the hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the
+sums of a geometric series in~$\omega_m$, with the relation
+between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.
 
@@
   \includegraphics[width=\textwidth]{wev.eps}
-  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$\omega_m^j$ of~$U_j$.
+  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$h\omega_m^j$ of~$U_j$.
 The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.
@@
 
-The sequence of~$U_j$ is a power series, which will converge if and only if the angular frequencies~$\omega_m$ are bounded by~$2/h$, a kind of Nyquist
-condition.  If the discretised Hamiltonian has an eigenstate~$x_\omega$
-with a frequency~$\omega\gg 2/h$, then~$U$ will magnify the component
-of~$x_\omega$ in~$x$ by a factor close to~$(h\omega/2)^i$, and this
-component is a parasitic solution growing steadily at the
-rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.
+The sequence of~$U_j$ is a power series, which will converge if and
+only if the angular frequencies~$\omega_m$ are bounded by the reciprocal time step~$2/h$, a
+kind of Nyquist condition.  If the discretised hamiltonian has an
+eigenstate~$x_\omega$ with a frequency~$\omega\gg 2/h$, then~$U$
+will magnify the component of~$x_\omega$ in~$x$ by a factor close
+to~$2(h\omega/2)^i$, and this component is a parasitic solution
+growing steadily at the rate~$\frac{\ln2}h\frac{i}{h}\ln(\frac{h\omega}2)$.
 
-The semi-implicit method is rarely used to solve the linear Equation~\cref{eq:iii}, and more often used to solve a nonlinear Gross-Pitaevskii~equation of the form
+{\bf Rewrite for clarity.}
+The semi-implicit method is rarely used to solve the linear
+Equation~\cref{eq:iii}, and more often used to solve a nonlinear
+Gross-Pitaevskii~equation of the form
 \begin{equation}
@@
 
+As a test of this analysis, a quantum Kerr oscillator was simulated
+using the semi-implicit method.  This system is a single-orbital
+analog of the fields for repulsive particles that are simulated in
+quantum optics.  The states with definite numbers of particles are
+an orthonormal basis for the quantum Hilbert space, so the coefficient
+vector~$x$ has indices~$n=0$, 1, \dots, $N$, where the number of
+particles is limited to~$N$ to make the simulation finite.  (In the
+experiment shown, $N=25$.)  The hamiltonian matrix is then diagonal,
+with
+\begin{equation}
+	H_{nn} = \omega_n = n(n-9)/2.
+\end{equation}
+The eigenvalues range from~$\omega_4=\omega_5=-10$,
+through~$\omega_1=\omega_9=0$, up to~$\omega_{25}=200$.
+
+In terms of wave functions and Schr\"odinger's equation, this could
+be regarded as the hamiltonian for a single particle, where~$x_n$
+was the coefficient of a Lagrange polynomial~$p_n$ (not quite) in
+the wave function.  It would be quite a wierd hamiltonian physically,
+with terms in products of position and momentum.  On the other hand,
+it makes physical sense in second quantisation, where it describes
+a collection of particles occupying the same orbital, with mutual
+repulsion.
+
+The oscillator was initialised in a quantum-optical coherent state
+\cite{glauber}, where the number of particles is Poisson distributed,
+with
+\begin{equation}\label{eq:vii}
+	x_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
+\end{equation}
+The experiment was run with~$\alpha=2$, which provides some small
+initial coefficients with large eigenfrequencies, which are useful
+for demonstrating the growth of the parasitic solutions.
+
 \begin{figure}[htbp]
@@
   \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
-  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
+  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.  (Tidy up fonts.)}
   \label{fig:i}
@@
 
+The results of solving this Hamiltonian with the semi-implicit
+method, with 4~iterations and time step~$h=0.03$, are shown in
+\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|x_n|$
+are constant in the exact solution.  The right hand side of
+\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
+and the values of~$|x_n|$ computed by the semi-implicit formula are
+shown as the rulings between these lines.  The size of~$x_n$ is
+plotted for all four iterations of the formula within each timestep.
+\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$x_n$
+will still be geometric, resulting in straight lines on the logarithmic
+scale of the graph.
+
+Parasitic growth can be
+seen at the left hand side, where~$\omega_n$ is large.  The limit
+of stability is~$\omega_n=2/h\approx 67$, and the dotted line
+indicates~$\omega_17=68$.  The dashed line shows the result of
+growing~$|x_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
+The undulations in the lines~$|x_n(t)|$ correspond to the growth
+rates given by the different lines in \cref{fig:ii}, which apply
+for the fraction of each step ending with a given iteration.
+
 \section{Conclusion}
 
-The semi-implicit formula considered here is a special case, but most ODE formulae will become unstable when frequencies get large compared to time steps.
+The semi-implicit formula discussed here is only one of the formulae
+that can be used to integrate ordinary differential equations, and
+that could be applied to solve \cref{eq:iii} numerically.  However,
+most of these formulae suffer from similar problems, and become
+unstable when the eigenvalues of the hamiltonian matrix are large
+compared to the time step.  The phenomenon of stiffness in discretised
+partial differential equations is well known \cite{something}.
 
-Stiff formulae stay stable, but reject the physics that is happening on fast timescales.  If we want to simulate physics at frequency~$\omega$, it seems reasonable to require time steps~$?$.
+There are special formulae to integrate stiff differential equations,
+which remain stable when the time steps are long compared to the
+eigenvalues of~$H$.  However, these remain stable by ignoring any
+physics that is happening on the fast timescales corresponding to
+the large eigenvalues.  If the goal is to simulate physics at a
+given frequency, then it seems reasonable that the simulation will
+require a time step around the corresponding Nyquist rate.
 
-The problem comes when the interesting physics occurs at slow timescales, but the state representation is inadvertently capable of representing high energies that evolve on fast timescales.
+The situation where stiff integration formulae would be appropriate
+is when the physics being simulated occurs at slow timescales, but
+the method by which the physical state is being represented
+inadvertently allows the state to contain a component with a large
+energy, that evolves on a fast timescale.  In the numerical example,
+this occured due to the coefficient vector~$x$ including coefficients
+with large~$n$, for which~$x_n$ was too small to have any effect
+on the state.  As demonstrated, the formula caused these minute
+components to grow quickly and dominate the actual solution.  It would
+be possible to address this by using an integration formula that
+was adapted to stiff differential equations.  But it might be better
+to address this by changing the representation so that the simulated
+state can not include such components, or to modify the hamiltonian
+dynamics so that they evolved with reasonable frequencies in the
+simulation.
 
-In the case here, where we expanded over energy eigenstates, you can simply truncate.
-This is contrived: if you know the energy eigenstates, you already know the answer, and you don't need to simulate anything.
+In the example here, the representation was simply an expansion
+over energy eigenstates.  This can be modified very simply, by
+reducing~$N$ to exclude the large eigenvalues and frequencies.  In
+the example, an expansion with coefficients~$x_0$ to~$x_{17}$ would
+be stable with the time step used in the example.  However, this
+is a contrived example.  If the energy eigenstates were known, their
+amplitudes could be calculated directly as complex
+exponentials~$e^{i\omega_nt}$, and there would be no need to use a
+potentially unstable integration formula.
 
-In general, you will use a variational ansatz, or some other tricky representation.
-In this case, some care must be taken to avoid inadvertently representing high energy states.
-Details to come in future papers.
+In real simulations, quantum states are represented by expansion
+over a set of states that are not energy eigenstates, by use of a
+variational ansatz, or by similar methods.  These representations
+are not directly related to the hamiltonian, so it is often the
+case that some states with spuriously high energies can be represented,
+and the resulting ordinary differential equations are unnecessarily
+stiff.  In these simulations, some care must be taken to avoid
+inadvertently representing high energy states, or to modify the
+hamiltonian dynamics to prevent them from causing stiffness.  Details
+for one representation, using Gaussian wave packets, will come in
+future papers.
 
 \section*{Acknowledgments}
-We would like to acknowledge the assistance of volunteers in putting
-together this example manuscript and supplement.
+I would like to acknowledge the support of Swinburne University of Technology.
 

### 2017-10-17 VCM/fockgrid.m
add scripts to test the SI formula on an orthonormal basis

@@ -1,0 +1,232 @@
+% see whether the SI method goes parasitic when Fock states are used instead of coherent states
+
+global N;  N = 25;  brackets
+
+% Peter's Hamiltonian
+% nhn = -4*(0:N)' gives an interesting sudden divergence
+
+nhn = nhn/2;
+nhn = nhn-4*(0:N)';
+
+a0 = 2;		% coherent amplitude of initial state
+% q0 = zeros(N+1,1);  q0(5) = 1;		% number state 2
+q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
+
+T=2*pi;  h=0.03;		% time axis
+T = 0.09;
+t = h*(0:ceil(T/h));
+snapshots = [h 2*h 0.1 1 2 4];
+
+iters = 4;                                          %%iterations of ODE solver
+
+% initialise storage for data to plot
+% cc(:,i,:) converge to c(:,i)
+
+c = nan(N+1, length(t));
+cc = nan(N+1, length(t), iters);
+
+BUF = nan(1,length(t));
+alpha.o = BUF;  number = BUF;  csize = BUF;  qsize = BUF;
+rsdl = BUF;  rsdln = BUF;
+
+% initial condition, the least squares expansion of |a0> over A
+
+c(:,1) = q0;
+cc(:,1) = q0;
+
+% Exact values for comparison
+
+qe = exp(-1i*nhn*t).*repmat(q0,size(t));	% column i is q(t(i))
+alpha.e = sum(conj(qe).*(aop*qe))/norm(q0)^2;
+
+% Integration loop
+
+for i = 1:length(t) 
+
+	% collect state data
+
+	qo = c(:,i);
+	rsdl(i) = norm(qe(:,i)-c(:,i));
+	rsdln(i) = sum(abs(qe(:,i)-c(:,i)).^2.*(0:N)')/rsdl(i)^2;
+	qsize(i) = norm(c(:,i));
+	alpha.o(i) = qo'*aop*c(:,i)/qsize(i)^2;
+	number(i) = sum(abs(c(:,i)).^2.*(0:N)')/qsize(i)^2;
+	csize(i) = norm(c(:,i));
+ 	
+	if i == length(t), break, end
+
+	% propagate state
+	
+	dc = 0;
[…]

### 2017-10-17 VCM/hamplot.m

@@ -1,0 +1,77 @@
+% script to plot numerical features of a discretised Hamiltonian matrix
+% assumes the following variables are set: ...
+
+% Suffix convention: ev, ew, sw, u = lsv, v = rsv
+
+Hew = nhn(1:R);
+M = pinv(A)*diag(nhn)*A;
+[ev,ew] = eig(M);  ew = diag(ew);
+% ns = (norms(aop*A*ev)./norms(A*ev)).^2;
+ns = sum(diag(0:N)*abs(A*ev).^2)./norms(A*ev).^2;
+[ns, P] = sort(ns);
+ev = ev(:,P);  ew = ew(P);
+[Au,Asw,Av] = svd(A,'econ');  Asw = diag(Asw);
+% Tychonov regularised H
+Hy = pinv([A; laf*eye(R)])*[diag(nhn)*A; zeros(R)];
+[Hyev, Hyew] = eig(Hy);  Hyew = diag(Hyew);
+% yns = (norms(aop*A*Hyev)./norms(A*Hyev)).^2;
+yns = sum(diag(0:N)*abs(A*Hyev).^2)./norms(A*Hyev).^2;
+[yns, P] = sort(yns);
+Hyev = Hyev(:,P);  Hyew = Hyew(P);
+
+% Eigenvalues of discretised Hamiltonian and step operators
+
+figure
+zplot(1:R, 0:N, Au), axis image
+title 'svs of A'
+
+unit = eye(N+1,R);
+Afock = pinv(A)*unit;
+Afock = Afock/diag(norms(Afock));
+
+figure, subplot 221
+zplot(1:R, 1:R, ev), axis image
+title 'evs of discretised H'
+
+subplot 222
+tmp2 = A*ev;  tmp2 = tmp2/diag(norms(tmp2));
+imagesc(abs(tmp2(1:R,:)).^2), colormap gray, axis image
+set(gca, 'YDir', 'normal')
+title 'Fock discretised H'
+
+subplot 223
+zplot(1:R, 1:R, Afock), axis image
+title 'number states'
+
+subplot 224
+tmp1 = A*Afock;  tmp1 = tmp1/diag(norms(tmp1));
+imagesc(abs(tmp1(1:R,:)).^2), colormap gray, axis image
+set(gca, 'YDir', 'normal')
+title 'Fock number states'
+
+
+figure
+plot(0:R-1, yns, '.r', 0:R-1, ns, '.k', [0 R], [0 R], ':k')
[…]

### 2017-10-17 VCM/ketgrid.m

@@ -1,0 +1,212 @@
+% VCM5  orthonomal basis solver
+
+global N;  N = 70;  brackets
+
+% Peter's Hamiltonian
+% nhn = -4*(0:N)' gives an interesting sudden divergence
+
+nhn = nhn/2;
+nhn = nhn-4*(0:N)';
+
+T=2*pi;  h=0.005;		% time axis
+T = 0.1;
+t = h*(0:ceil(T/h));
+snapshots = h*(0:5);
+
+iters = 4;                                          %%iterations of ODE solver
+
+q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
+
+% initialise storage for data to plot
+
+c = nan(N+1, length(t));
+cc = nan(N+1, length(t), iters);
+
+BUF = nan(1,length(t));
+alpha.o = BUF;  number = BUF;  csize = BUF;  qsize = BUF;
+rsdl = BUF;  rsdln = BUF;
+
+% initial condition
+
+c(:,1) = q0;
+
+% Exact values for comparison
+
+qe = exp(-1i*nhn*t).*repmat(q0,size(t));	% column i is q(t(i))
+alpha.e = sum(conj(qe).*(aop*qe));
+
+% Integration loop
+
+for i = 1:length(t) 
+
+	% collect data to plot
+
+	qo = c(:,i);
+	rsdl(i) = norm(qe(:,i)-qo);
+	rsdln(i) = sum(abs(qe(:,i)-qo).^2.*(0:N)')/rsdl(i)^2;
+	qsize(i) = norm(qo);
+	alpha.o(i) = qo'*aop*qo/qsize(i)^2;
+	number(i) = sum(abs(qo).^2.*(0:N)')/qsize(i)^2;
+	csize(i) = norm(c(:,i));
+ 	
+	if i == length(t), break, end
+
+	% propagate state
+	
+	dc = 0;
+	for j = 1:iters
+		cc(:,i+1,j) = c(:,i) + dc;
+		ch = c(:,i) + dc/2;
+		dc = -1i*h*nhn.*ch;
+	end
+    	c(:,i+1) = c(:,i) + dc;
+
+end
+
+% Eigenvalues of discretised Hamiltonian and step operators
+
+M = diag(1i*nhn);  ew = eig(M);
+figure, plot(1:(N+1),real(ew),'+k',1:(N+1),imag(ew),'.k',1:(N+1),nhn,'or')
[…]

### 2017-10-17 VCM/phasegrid.m

@@ -13,7 +13,7 @@ a0 = 2;		% coherent amplitude of initial state
 q0 = nq*evan(a0,'even');	% expansion of |a0> over number states
 
 T=2*pi;  h=0.005;		% time axis
-% T = 0.05;
+T = 0.05;
 t = h*(0:ceil(T/h));
 snapshots = [h 2*h 0.1 1 2 4];
 

### 2017-10-17 raise.m
add the script that plots the exact quartic oscillator states

@@ -1,0 +1,17 @@
+% plots of quartic oscillator evolution
+% idea: find the volume of phase space occupied by a ring of N oscillators with a Bose-Hubbard hamiltonian
+
+% note spiral of zeroes at first local maximum.  Interference as larger amplitudes lap smaller ones.
+
+% fucking Matlab pdf black-on-black labels
+
+
+global N; N = 35; brackets
+h = 0.3;  mesh = -5:h:5;  phasespace
+
+for a = [0 1 i 2+3i]
+	c = aop'*nq*evan(a, 'even');
+	figure, zplot(mesh, mesh, Aps'*c), hold on, axis image
+end
+%	saveTightFigure(sprintf('qrt%02d.pdf', i))
+%	title(sprintf('Q.O. at t = %.2f*2\\pi', t))

### 2017-10-17 scat.m
add the script that plots quartic oscillator Wigner functions

@@ -1,0 +1,23 @@
+% Q distributions and phase for cat states
+
+global N;  N = 60;  brackets
+mesh = -5:0.2:5;  phasespace
+
+cs = nq*evan(1.65*[-1 1], 'even');
+cp = sum(cs,2);  cm = cs(:,1) - cs(:,2);
+
+figure
+zplot(mesh, mesh, Aps'*cp);
+axis image, ylim([-3 3]), saveTightFigure va.pdf
+
+figure
+zplot(mesh, mesh, abs(Aps'*cp).^2);
+axis image, ylim([-3 3]), saveTightFigure vb.pdf
+
+figure
+zplot(mesh, mesh, Aps'*cm);
+axis image, ylim([-3 3]), saveTightFigure vc.pdf
+
+figure
+zplot(mesh, mesh, abs(Aps'*cm).^2);
+axis image, ylim([-3 3]), saveTightFigure vd.pdf

### 2017-10-17 sqn.m
add script to plot squeezed states

@@ -1,0 +1,24 @@
+function c = sqn(a,x)
+%SQN expand squeezed states over number states
+%
+%SQN(A,X) where a is a coherent amplitude and x a squeezing factor returns a column vector of coefficients
+%SQN(A,X) where a and x are n-vectors returns a matrix with n columns of expansion coefficients (NYI)
+%
+%See pra-13-2226, Equation 3.23
+
+global N, n = (0:N)';
+
+if (numel(a) > 1) || (numel(x) > 1)
+	error 'vectors of squeezing parameters NYI'
+end
+if x==0
+	error 'shortcut for zero NYI'
+end
+
+mu = cosh(x);  nu = sinh(x);
+c = exp(-abs(a)^2/2 + conj(nu)*a^2/2/mu) * (nu/2/mu).^(n/2) ./ sqrt(mu*factorial(n));
+for j = n'
+	c(j+1) = c(j+1) * hermite(j, a/sqrt(2*mu*nu));
+end
+
+end

### 2017-10-17 squish.m

@@ -1,0 +1,33 @@
+% basic squeezed state play
+
+% The small multiples form a half plane: squeezing in the opposite direction has the same effect.
+
+global N; N = 15; brackets
+mesh = -5:0.1:5;  phasespace
+
+figure, zplot(mesh, mesh, Aps(end,:)), axis image
+title 'Maximum number state'
+
+figure, zplot(mesh, mesh, Aps'*nq*evan(0, 'even')), axis image
+title 'Numerically exact vacuum'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,0.1)), axis image
+title 'Minimally squeezed vacuum'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,1)), axis image
+title 'Squeezed vacuum'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,0.7i)), axis image
+title 'Squeezed vacuum, +i direction'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,-0.7i)), axis image
+title 'Squeezed vacuum, -i direction'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,-0.5-0.7i)), axis image
+title 'Squeezed vacuum, -1-i direction'
+
+figure, zplot(mesh, mesh, Aps'*sqn(0,10)), axis image
+title 'Very squeezed vacuum'
+
+figure, zplot(mesh, mesh, Aps'*sqn(exp(1)*2,1)), axis image
+title 'Squeezed state with coherent amplitude 2'

### 2017-10-17 widi.m
add function to compute Wigner distributions

@@ -1,0 +1,22 @@
+function W = widi(c, a)
+%WIDI calculate the Wigner distribution for a pure quantum state
+%
+% WIDI(C, A) returns values of the Wigner distribution at the complex
+% points A.  The vector C holds the number state components of the
+% quantum state whose Wigner distribution is to be found.
+%
+% The algorithm is from pla-378-117.
+
+W = zeros(size(a));
+
+for m = 0:(length(c)-1)
+	X = 2*(-1)^m/pi * exp(-2*abs(a).^2) .* polyval(LaguerreGen(m, 0), 4*abs(a).^2);
+	W = W + abs(c(m+1))^2*X;
+	for n = 0:(m-1)
+		X = 2*(-1)^n/pi / prod(sqrt((n+1):m)) * exp(-2*abs(a).^2) ...
+			.* (2*a).^(m-n) .* polyval(LaguerreGen(n, m-n), 4*abs(a).^2);
+		W = W + 2*real(c(n+1)*conj(c(m+1))*X);
+	end
+end
+
+end

### 2017-10-18 ssc-0-0.tex
revise the draft semi-implicit paper

@@
 
-For a system of equations~${dx\over dt}=f(x,t)$,
+% iterations
+For a system of equations
+\begin{equation}
+	{ds\over dt}=f(s),
+\end{equation}
 the semi-implicit formula with time step~$h$ is
 \begin{equation}\label{eq:i}
-    x_{n+1}=x_n+hf_{n+\frac12},
+    s_{n+1}=s_n+hf_{n+\frac12},
+	\qquad\text{where}\qquad 
+    f_{n+\frac12}=f\left({\textstyle\frac{s_n+s_{n+1}}{2}}\right).
 \end{equation}
-where
-\begin{equation}\label{eq:ii}
-    f_{n+\frac12}=f\left({\textstyle\frac{x_n+x_{n+1}}{2}}\right).
-\end{equation}
-When this formula is applied to equations of a form similar to~\cref{eq:vi},
-\begin{equation}\label{eq:iii}
-	{dx\over dt}=-iHx,
-\end{equation}
-where~$H$ is a hermitian operator acting on the complex vector~$x$,
-there is an advantage in principle because the norm of~$x$ is
-conserved in the discretised solution.  {\bf Expand on this.}  However, the formula is
-used mainly because it has been found to work well in practice.
-
-% iterations
-It is not obvious how to implement this formula, because~$x_{n+1}$
+It is not obvious how to implement this formula, because~$s_{n+1}$
 and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
-usual way is a simple iteration.  To start,
-$x_{n+1}$~is initialised to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then each iteration
-evaluates Equations~\cref{eq:i,eq:ii} in turn, and updates the values
-of~$x_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the iterations generate a sequence of~$x_{n+1}^j$
-and~$f_{n+\frac12}^j$, with
+usual way is a simple iteration \cite{swx-5-12}.  To start, $s_{n+1}$~is initialised
+to~$s_n$, and~$f_{n+\frac12}$ to~$f(s_n)$.  Then each iteration
+evaluates Equation~\cref{eq:i}, and updates the
+values of~$s_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the
+iterations generate a sequence of~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
+with
 \begin{subequations}
 \begin{align}
-	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
-	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j
-		\qquad f_{n+\frac12}^{j+1}=f\left({\textstyle\frac{x_n+x_{n+1}^j}{2}}\right),\qquad j\ge0.
+	\label{eq:iv} s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
+	\label{eq:v} s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
+		\qquad f_{n+\frac12}^{(j+1)}=f\left({\textstyle\frac{s_n+s_{n+1}^{(j)}}{2}}\right)\qquad \text{when $j\ge0$}.
 \end{align}
@@
 achieved after a small, fixed number of iterations.  Any fixed point
-of the iteration is a self-consistent solution to
-Equations~\cref{eq:i,eq:ii}.  The remaining question is under what
+of the iteration satisfies
+Equation~\cref{eq:i}.  The remaining question is under what
 conditions the iterations converge.
@@
 to the grid spacing in order for the discretised solution to remain
-stable \cite{something}.  No one has analysed the stability of the
+stable \cite{2000-Trefethen-Spectral}.  No one has analysed the stability of the
 semi-implicit formula with iterations.  The stable time step has
@@
 depends on a Nyquist-like condition, and the iterations converge
-if and only if the largest eigenvalue of~$H$ is no more than~$2/h$.
+if and only if the eigenvalues of the hamiltonian are bounded by the reciprocal time step~$2/h$.
 
@@
 When the semi-implicit rule is applied to \cref{eq:vi}, or a
-stochastic or nonlinear variant, the laplacian operator is usually
-evaluated by a spectral method.  This is equivalent to expanding
-the wave function~$\psi(q)$ over a basis of sinc wave packets
+stochastic or nonlinear variant of it, the laplacian operator is usually
+evaluated by a spectral method \cite{something}.  This is equivalent to expanding
+the wave function~$\psi(x)$ over a basis of sinc wave packets
 \begin{equation}
-	\psi(q) = \sum x_i\phi_i(q),
+	\psi(x) = \sum_m s^m\phi_m(x),
 \end{equation}
-{\bf figure out range of $i$}
-where
+where, for a spatial grid of step~$l$,
 \begin{equation}
-	\phi_i(q) = \frac1{\sqrt{l}} \sinc(\textstyle\frac{q}{l} - i).
+	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
 \end{equation}
@@
 \begin{equation}
-	\sinc{q} = \frac{\sin(\pi q)}{\pi q},
+	\sinc x = \frac{\sin\pi x}{\pi x},
 \end{equation}
-under which the functions~$\phi_i$ are an orthonormal basis for the
-space of functions limited to bandwidth~$1/2l$.  The spectral method
-reduces \cref{eq:vi} to an equation of the form~\cref{eq:iii} for
-the coefficient vector~$x$, where
+under which the set of functions~$\phi_m$ is an orthonormal basis for the
+space of functions limited to spatial bandwidth~$1/2l$.  The spectral method
+reduces \cref{eq:vi} to an equation for the complex coefficient vector~$s$ of the form
+\begin{equation}\label{eq:iii}
+	{ds\over dt}=-iHs,
+\end{equation}
+where~$H$ is a Hermitian matrix with elements
 \begin{equation}
-	H_{ij} = \int \phi_i(q) (-\nabla^2 + V(q)) \phi_j^\ast(q).
+	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
 \end{equation}
 
-At this point, the
-iteration rules of Equation~\cref{eq:v} become
+If the formula of Equation~\cref{eq:i} could be applied directly
+to Equation~\cref{eq:iii}, the computed solution would be very
+stable.  The coefficient vector~$s_n$ can be expanded over the
+eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
+shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
+projected onto the complex plane spanned by the eigenvector~$v_k$.
+The exact solution that satisfies Equation~\cref{eq:iii}, with
+initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
+and its projection lies on the circle around the origin of
+radius~$|c_k|$.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics{semii.1}
+  \caption{A step of the semi-implicit formula, projected onto the
+  complex plane spanned by the eigenvector~$v_k$ of~$H$.  The
+  computed solution~$s_{n+1}$ lies on the same circle as the
+  exact solution~$s(h)=\exp(-ihH)s_n$, but it rotates with angular frequency lower
+  than~$\omega_k$.  See text for details.}
+  \label{fig:iii}
+\end{figure}
+
+Suppose that the point~$s_{n+1}$ also lies on this circle.
+Then~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
+chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
+brought towards and away from the origin by moving~$s_{n+1}$ around
+the circle, so there is some position for
+which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, and this value of~$s_{n+1}$
+satisfies Equation~\cref{eq:i}.  The only error introduced by the
+formula is that~$c_k$ rotates with an angular frequency less
+than~$\omega_k$, which is bounded by~$\pi/h$ as~$\omega_k$ becomes
+large.
+
+In practice, the iterations of~Equation~\cref{eq:vi} must be used
+instead of~Equation~\cref{eq:i}, and this is less stable.  For
+Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
+iteration rule becomes
 \begin{equation}
-x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
-	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j
+s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
+	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
 \end{equation}
@@
 \begin{equation}
-x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j
+s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
 \end{equation}
-with solution {\bf check factor of 2}
+with solution
 \begin{equation}\label{eq:viii}
-x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
-	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.
+s_{n+1}^{(j)}=U^{(j)}s_n
+	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
 \end{equation}
-The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$
-as the hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the
-sums of a geometric series in~$\omega_m$, with the relation
-between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.
+The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
+eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
+eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
+in~$-ih\omega_k/2$, which agrees to second order in~$\omega_k$ with
+the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
+operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$
+plotted in \cref{fig:ii}, for the first four iterations.
 
@@
   \includegraphics[width=\textwidth]{wev.eps}
-  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$h\omega_m^j$ of~$U_j$.
-The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.
-The exact time evolution operator~$U$ is unitary, with unimodular eigenvalues.}
+  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.
+The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.
+The exact time evolution operator~$U$ has unimodular eigenvalues.}
   \label{fig:ii}
@@
 
-The sequence of~$U_j$ is a power series, which will converge if and
-only if the angular frequencies~$\omega_m$ are bounded by the reciprocal time step~$2/h$, a
-kind of Nyquist condition.  If the discretised hamiltonian has an
-eigenstate~$x_\omega$ with a frequency~$\omega\gg 2/h$, then~$U$
-will magnify the component of~$x_\omega$ in~$x$ by a factor close
-to~$2(h\omega/2)^i$, and this component is a parasitic solution
-growing steadily at the rate~$\frac{\ln2}h\frac{i}{h}\ln(\frac{h\omega}2)$.
+The geometric series for~$U^{(j)}$ will converge if and only if the
+angular frequencies~$\omega_k$ are bounded by the reciprocal time
+step~$\frac2h$, a kind of Nyquist condition.  If the discretised
+hamiltonian has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
+will enlarge the component of~$v_k$ in~$s_n$ by a factor close
+to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
+thus a parasitic solution growing at the
+rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.
 
@@
 This linearisation has been extensively studied, and the operator~$\hat
-H'$ is the hamiltonian for Bogoliubov sound waves \cite{something}.
+H'$ is the hamiltonian for Bogoliubov sound waves \cite{aop-70-67}.
 The stability conditon for the Gross-Pitaevskii~equation is that
@@
 \section*{Acknowledgments}
-I would like to acknowledge the support of Swinburne University of Technology.
+I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for suggestions that improved the presentation.
 

### 2017-10-19 ssc-0-0.tex
revise the semi-implicit paper into a second draft

@@
 ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
-is often used as a method of lines, to solve stochastic partial
+is most often used as a method of lines, to solve stochastic partial
 differential equations that occur when phase space Monte-Carlo
 methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
-These equations are related to Schr\"odinger's equation,
+These equations resemble Schr\"odinger's equation,
 \begin{equation}\label{eq:vi}
@@
 stochastic or nonlinear variant of it, the laplacian operator is usually
-evaluated by a spectral method \cite{something}.  This is equivalent to expanding
+evaluated by a spectral method \cite{2000-Trefethen-Spectral}.  This is equivalent to expanding
 the wave function~$\psi(x)$ over a basis of sinc wave packets
@@
 \end{equation}
-under which the set of functions~$\phi_m$ is an orthonormal basis for the
-space of functions limited to spatial bandwidth~$1/2l$.  The spectral method
-reduces \cref{eq:vi} to an equation for the complex coefficient vector~$s$ of the form
+under which the set of functions~$\phi_m$ is an orthonormal basis
+for the space of functions limited to spatial bandwidth~$1/2l$.
+The spectral method reduces Equation~\cref{eq:vi} to an ordinary
+differential equation for the complex coefficient vector~$s$ of the
+form
 \begin{equation}\label{eq:iii}
@@
 Suppose that the point~$s_{n+1}$ also lies on this circle.
-Then~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
+Then the projection of~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
 chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
@@
 the circle, so there is some position for
-which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, and this value of~$s_{n+1}$
-satisfies Equation~\cref{eq:i}.  The only error introduced by the
-formula is that~$c_k$ rotates with an angular frequency less
-than~$\omega_k$, which is bounded by~$\pi/h$ as~$\omega_k$ becomes
+which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, so that~$s_{n+1}$
+satisfies Equation~\cref{eq:i}.  The only error introduced by discretisation
+is that~$c_k$ rotates with an angular frequency less
+than~$\omega_k$, bounded by~$\pi/h$ as~$\omega_k$ becomes
 large.
@@
 eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
-in~$-ih\omega_k/2$, which agrees to second order in~$\omega_k$ with
+in~$-ih\omega_k/2$, which agrees to second order in~$h\omega_k$ with
 the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
-operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$
+operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$ is
 plotted in \cref{fig:ii}, for the first four iterations.
@@
   \centering
-  \includegraphics[width=\textwidth]{wev.eps}
+  \includegraphics[width=0.7\textwidth]{wev.eps}
   \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.
@@
 angular frequencies~$\omega_k$ are bounded by the reciprocal time
-step~$\frac2h$, a kind of Nyquist condition.  If the discretised
-hamiltonian has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
+step~$\frac2h$, a kind of Nyquist condition.  If the spectral
+hamiltonian matrix has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
 will enlarge the component of~$v_k$ in~$s_n$ by a factor close
@@
 
-{\bf Rewrite for clarity.}
 The semi-implicit method is rarely used to solve the linear
@@
 \begin{equation}
-	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
+	{\partial\psi\over\partial t}=-i(\nabla^2+V+g|\psi|^2)\psi.
 \end{equation}
-However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
-$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
+However, following the usual approach to the stability of integation formulae \cite{1996-Trefethen-finite}, the mapping from~$\psi$ to $\partial\psi\over\partial t$ can be linearised at each time step, as
 \begin{equation}
-	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
+	\left.{\partial\psi\over\partial t}\right|_{\psi=\psi(t)+\delta\psi}
+		\approx -i(\nabla^2+V+g|\psi(t)|^2)\psi(t) - i\hat H'(\psi(t))\delta\psi,
 \end{equation}
-This linearisation has been extensively studied, and the operator~$\hat
-H'$ is the hamiltonian for Bogoliubov sound waves \cite{aop-70-67}.
-The stability conditon for the Gross-Pitaevskii~equation is that
-the highest frequency of the Bogoliubov sound waves in the discretised
-Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
-high-frequency Bogoliubov sound waves have the same frequencies as
-the eigenstates of~$\hat H$.  Most simulations use grids fine enough
-for this to be the case, and the nonlinear term will not affect the
-stability of such simulations.
+where~$\hat H'(\psi(t))$ is an effective hamiltonian operator for
+the field~$\delta\psi$.  This linearisation has been studied
+extensively, and the dynamics generated by the hamiltonian~$\hat
+H'$ are Bogoliubov sound waves \cite{aop-70-67}.  When this equation
+is discretised with the semi-implicit formula, the stability condition
+is that the Bogoliubov sound wave modes on a spatial grid of
+spacing~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
+fine grid, the high-frequency Bogoliubov modes have the same
+frequencies as the eigenvectors of~$H$ \cite{2002-Pethick-Bose}.
+Most simulations use grids fine enough for this to be the case, and
+the nonlinear term will not affect their stability.
 
@@
 As a test of this analysis, a quantum Kerr oscillator was simulated
-using the semi-implicit method.  This system is a single-orbital
-analog of the fields for repulsive particles that are simulated in
-quantum optics.  The states with definite numbers of particles are
-an orthonormal basis for the quantum Hilbert space, so the coefficient
-vector~$x$ has indices~$n=0$, 1, \dots, $N$, where the number of
-particles is limited to~$N$ to make the simulation finite.  (In the
-experiment shown, $N=25$.)  The hamiltonian matrix is then diagonal,
+using the semi-implicit formula.  This system is an analog of the
+fields that are usually simulated using the formula, where all the
+particles are confined to the same orbital.  The states with definite
+numbers of particles are an orthonormal basis for the quantum Hilbert
+space, so the coefficient vector~$s$ has indices~$n=0$, 1, \dots,
+$N$, where the possible number of particles is limited to~$N$.  In
+the example run here, $N=25$.  The hamiltonian matrix is diagonal,
 with
@@
 
-In terms of wave functions and Schr\"odinger's equation, this could
-be regarded as the hamiltonian for a single particle, where~$x_n$
-was the coefficient of a Lagrange polynomial~$p_n$ (not quite) in
-the wave function.  It would be quite a wierd hamiltonian physically,
-with terms in products of position and momentum.  On the other hand,
-it makes physical sense in second quantisation, where it describes
-a collection of particles occupying the same orbital, with mutual
-repulsion.
-
-The oscillator was initialised in a quantum-optical coherent state
-\cite{glauber}, where the number of particles is Poisson distributed,
-with
+The oscillator was initialised in a quantum optical coherent state
+\cite{prx-131-2766} with a Poisson distribution of particle numbers,
 \begin{equation}\label{eq:vii}
-	x_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
+	s_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
 \end{equation}
-The experiment was run with~$\alpha=2$, which provides some small
-initial coefficients with large eigenfrequencies, which are useful
-for demonstrating the growth of the parasitic solutions.
+The experiment was run with the coherent amplitude~$\alpha=2$.  This
+provides some small initial coefficients with large eigenfrequencies,
+which demonstrate the growth of the parasitic solutions.
 
@@
   \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
-  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.  (Tidy up fonts.)}
+  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
   \label{fig:i}
@@
 method, with 4~iterations and time step~$h=0.03$, are shown in
-\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|x_n|$
+\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|s_n|$
 are constant in the exact solution.  The right hand side of
-\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
-and the values of~$|x_n|$ computed by the semi-implicit formula are
-shown as the rulings between these lines.  The size of~$x_n$ is
+Equation~\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
+and the values of~$|s_n|$ computed by the semi-implicit formula are
+shown as the rulings between those lines.  The size~$|s_n|$ is
 plotted for all four iterations of the formula within each timestep.
-\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$x_n$
-will still be geometric, resulting in straight lines on the logarithmic
+\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$s_n$
+from step to step is still geometric, resulting in straight lines on the logarithmic
 scale of the graph.
+
+{\bf Check growth rates.}
 
@@
 of stability is~$\omega_n=2/h\approx 67$, and the dotted line
-indicates~$\omega_17=68$.  The dashed line shows the result of
-growing~$|x_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
-The undulations in the lines~$|x_n(t)|$ correspond to the growth
-rates given by the different lines in \cref{fig:ii}, which apply
-for the fraction of each step ending with a given iteration.
+indicates~$n=17$, with~$\omega_{17}=68$.  The dashed line shows the result of
+growing~$|s_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
+The undulations in the lines~$|s_n(t)|$ correspond to the growth
+rates given by the different lines in \cref{fig:ii}, each of which applies
+for the initial fraction of each step up to the corresponding iteration.
 
@@
 that can be used to integrate ordinary differential equations, and
-that could be applied to solve \cref{eq:iii} numerically.  However,
+that could be applied to solve Equation~\cref{eq:iii} numerically.  However,
 most of these formulae suffer from similar problems, and become
@@
 
-The situation where stiff integration formulae would be appropriate
-is when the physics being simulated occurs at slow timescales, but
-the method by which the physical state is being represented
-inadvertently allows the state to contain a component with a large
-energy, that evolves on a fast timescale.  In the numerical example,
-this occured due to the coefficient vector~$x$ including coefficients
-with large~$n$, for which~$x_n$ was too small to have any effect
-on the state.  As demonstrated, the formula caused these minute
-components to grow quickly and dominate the actual solution.  It would
-be possible to address this by using an integration formula that
-was adapted to stiff differential equations.  But it might be better
-to address this by changing the representation so that the simulated
-state can not include such components, or to modify the hamiltonian
-dynamics so that they evolved with reasonable frequencies in the
-simulation.
+On the other hand, stiff integration formulae would be appropriate
+where the dynamics being simulated is relatively slow, but the
+representation of the quantum state inadvertently contains some
+components with large energies whose phases oscillate rapidly.  This
+was the case in the example shown above, where the coefficient
+vector included coefficients with large~$n$, for which~$s_n$ was
+too small to matter.  However, $\omega_n$ was large enough that the
+formula caused these minute components to grow and dominate the
+actual solution.  This could be addressed by using an integration
+formula that was adapted to stiff differential equations.  But it
+would be better to change the representation to exclude the high
+energy components, or to modify the simulated dynamics so that these
+components oscillate with reasonable frequencies.
 
-In the example here, the representation was simply an expansion
-over energy eigenstates.  This can be modified very simply, by
-reducing~$N$ to exclude the large eigenvalues and frequencies.  In
-the example, an expansion with coefficients~$x_0$ to~$x_{17}$ would
-be stable with the time step used in the example.  However, this
-is a contrived example.  If the energy eigenstates were known, their
-amplitudes could be calculated directly as complex
-exponentials~$e^{i\omega_nt}$, and there would be no need to use a
-potentially unstable integration formula.
+The example above can be modified very simply, by reducing~$N$ to
+exclude the high frequencies.  An expansion with coefficients~$s_0$
+to~$s_{17}$ would be stable with the time step used above, and the
+coefficients~$s_n$ with $n>17$ are small.  However, this is a
+contrived example.  Since the energy eigenstates are known, their
+amplitudes can be calculated directly as complex
+exponentials~$e^{i\omega_nt}$, and there is no need to use any
+integration formula, stable or otherwise.
 
-In real simulations, quantum states are represented by expansion
-over a set of states that are not energy eigenstates, by use of a
-variational ansatz, or by similar methods.  These representations
-are not directly related to the hamiltonian, so it is often the
-case that some states with spuriously high energies can be represented,
-and the resulting ordinary differential equations are unnecessarily
-stiff.  In these simulations, some care must be taken to avoid
+In a real simulation, the energy eigenstates are not known beforehand,
+and the quantum state is represented some other way.  This could
+be an expansion over a set of states that are not energy eigenstates,
+a variational ansatz, or a similar method.  Some of these representations
+will allow states with spuriously high energies to be represented,
+so that the differential equations satisfied by the representation
+are unnecessarily stiff.  In this case, care must be taken to avoid
 inadvertently representing high energy states, or to modify the
-hamiltonian dynamics to prevent them from causing stiffness.  Details
-for one representation, using Gaussian wave packets, will come in
-future papers.
+Hamiltonian dynamics to prevent them from causing stiffness.  Details
+for one such representation, using Gaussian wave packets, will be
+given in a future paper.
 
@@
 \bibliographystyle{siamplain}
-\bibliography{references}
+\bibliography{references,goats}
 \end{document}

### 2017-11-13 VCM/notes.tex
graph pseudospectra along the real axis

\input respnotes
\input xpmath
\input unifonts \tenrm
\input xref

%% Inserts

\newcount\figno
\def\figlbl{\global\advance\figno by 1\bf Figure \the\figno}
\def\nextfig{{\advance\figno by 1 \the\figno}}
\def\figure#1{$$\hbox{\figlbl}\quad\vcenter{#1}$$}
\def\widefigure#1{\midinsert#1
	\smallskip\centerline{\figlbl}\endinsert}

\title Normality of the discretised number operator

The script {\tt phaseham.m} examines the discretised number operator~$M=〈A⁺|\hat n|A〉$, where
$$|A〉=\pmatrix{|α₁〉&…&|α_R〉},$$
and the coherent amplitudes~$α_i$ lie on a square grid bounded by a disk about the origin of phase space.  This makes~$M$ an~$R×R$ matrix.

The pseudospectrum of the exact number operator is pretty well as expected: the eigenstate~$|n〉$ causes the complex number~$λ$ to be in the~$ε$-pseudospectrum of~$\hat n$ whenever~$|λ-n|<ε$.

\figure{\XeTeXpdffile resp021117b.pdf width  0.7\hsize }

Contrast the pseudospectrum of~$M$:

\figure{\XeTeXpdffile resp021117a.pdf width  0.7\hsize }

Pretty well every~$λ$ has some coefficient vector~$c$ such that~$|(〈A⁺|\hat n|A〉-λ)·c|$ is small.  Correspondingly, the eigenvectors of~$M$ are nearly parallel:

\figure{\XeTeXpdffile resp021117c.pdf width  0.7\hsize }

\bye

### 2017-11-13 VCM/phaseham.m

@@ -1,0 +1,180 @@
+%PHASEHAM analyse the Kerr Hamiltionian discretised with wave packets
+
+global N;  N = 70;  brackets
+
+% Peter's Hamiltonian
+nhn = nhn/2;
+nhn = nhn-4*(0:N)';
+
+laf = 3e-4;		% Lagrange multiplier for Tyc conditioning.
+h = 0.005;		% time step
+
+% phase space grid and expansion operator
+
+l = 1.1;			% frame density
+[ax,ay] = meshgrid(-4:l:4);  a = ax(:)+1i*ay(:);
+a = a(abs(a) <= 4);
+R = length(a);
+[~, i] = sort(abs(a));  a = a(i);
+A = nq*evan(a,'even');
+pinvA = pinv(A);
+
+% Discretised number operator and hamiltonian
+
+M = pinvA*diag(0:N)*A;
+G = pinvA*diag(nhn)*A;
+
+% Eigenvalues are stored as vectors
+% Eigenvectors are stored as columns, in order of Fock number
+% Suffix convention: ev, ew, sw, u = lsv, v = rsv
+% The discrete hamiltonian has eigenkets in Hilbert space, 
+% and eigenvectors in C^R.  These are also the eigenkets of the
+% discrete hamiltonian.
+
+[Au,Asw,Av] = svd(A,'econ');  Asw = diag(Asw);
+
+Hew = nhn(1:R);   % Hev is the unit matrix
+Ans = sum(diag(0:N)*abs(Au).^2);
+Avarn = sum(diag((0:N).^2)*abs(Au).^2) - Ans.^2;
+
+[Mev,Mew] = eig(M);  Mew = diag(Mew);
+Mkets = A*Mev;  Mkets = Mkets / diag(norms(Mkets));
+Mns = sum(diag(0:N)*abs(Mkets).^2);
+[Mns, P] = sort(Mns);
+Mew = Mew(P);  Mev = Mev(:,P);  Mkets = Mkets(:,P);
+Mvarn = sum(diag((0:N).^2)*abs(Mkets).^2) - Mns.^2;
+
+% The Mev turn out identical for large n, and numerically better
+% So we'll use those instead.
+[Gev,~] = eig(G);
+Gkets = A*Gev;  Gkets = Gkets / diag(norms(Gkets));
+Gns = sum(diag(0:N)*abs(Gkets).^2);
[…]

### 2017-11-24 pra-0-1.tex
add some initial notes on phase space wave functions, analytical continuation and degeneracy

\title{Wave functions in phase space}

\begin{document}

Most potentials have characteristic length scales

Blurs around classical trajectories, with quantised phase stripes.  (How does the phase correspond to action-angle coordinates?)

Immediately obvious why these are stationary states, and why they are quantised.

More physical insight: Airy function shows incoming and outgoing particles transparently.

``Quantisation is always due to some topological invariant''.  Analytic structure of Bargmann function vs quantum number and hamiltonian.  How many degenerate zeros are there?  Are there ever poles?

Numerical analysis.  How to expand a wave function over Lagrange(?) polynomials.

Can we construct approximate eigenstates by sweeping a coherent state around the classical trajectory, and adjusting the size of the trajectory to satisfy some quantum condition?  Can we then construct a better approximation to |13〉 by exact diagonalisation of H over |10〉⋯|15〉?

Is it possible to construct 〈α|φ_i〉 on a classical trajectory with phase e^{iS} and magnitude by continuity, then extend it Bargmann-analytically to get a wave function?  Is this any different from the fluid dynamics limit of wave mechanics, other than making it obvious why things are quantised?

Degeneracy.  In a double well system, you can construct eigenstates for each well separately, then diagonalise the ones in a certain energy range to account for tunnelling.  There is a sense here of "real degeneracy", where the wave functions have to overlap in phase space, e.g. the cw and ccw travelling waves in a ring, and "energy coincidences", such as exactly symmetric double wells where the bound states of each well happen to have the same energy, but that would change under a small pertubation.

\end{document}

### 2017-11-24 pra-0-1.tex
add some thoughts on phase space wave functions

@@
 
-Numerical analysis.  How to expand a wave function over Lagrange(?) polynomials.
+Numerical analysis.
+How to expand a wave function over Hermite polynomials.
+I.e., is there a matrix A_n and points x₁ through x_n, such that A_n·(f(x₁) ⋯ f(x_n))' gives the components of the Hermite functions H₁ through H_n in f, and is exactly right for H₁ through H_n?
+(And therefor for linear combinations of them?)  This is almost certainly Gauss-Hermite quadrature, because e^{-x²}·e^{-x²}=e^{-2x²}, and the Gauss-Hermite formula is exact for e^{-x²}·P(x) with P a low-order polynomial.
 
@@
 
+Plots of how tunnelling looks in phase space over time.  If we start off with a real wave packet in one side of a double well, how does it get momentum to tunnel to the other side?
+
 \end{document}

### 2017-12-19 ssc-0-0.tex
complete second draft of SI paper

@@
 \begin{abstract}
-  This is an example SIAM \LaTeX\ article. This can be used as a
-  template for new articles.  Abstracts must be able to stand alone
-  and so cannot contain citations to the paper's references,
-  equations, etc.  An abstract must consist of a single paragraph and
-  be concise. Because of online formatting, abstracts must appear as
-  plain as possible. Any equations should be inline.
+The iterative semi-implicit formula, or some variant of it, is often
+used to solve stochastic partial differential equations that arise
+in quantum optics and in atom optics.  When this formula is applied
+to a quantum system with unitary dynamics, stability depends on a
+kind of Nyquist condition, where the angular frequency corresponding
+to the largest eigenvalue of the discretised hamiltonian is at most
+twice the reciprocal time step.  The formula develops parasitic
+solutions very abruptly as this condition is exceeded, which makes
+it troublesome for practical use, but quite useful as a test of the
+stability of discretisation methods.
 \end{abstract}
@@
 system of ordinary or stochastic differential equations
-\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
+\cite{2000-Trefethen-Spectral}.  These equations are integrated
 by the semi-implicit formula.
@@
 and its projection lies on the circle around the origin of
-radius~$|c_k|$.
+radius~$|c_k|$.  Here~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.
 
@@
 
-{\bf Check growth rates.}
-
 Parasitic growth can be
@@
 for the initial fraction of each step up to the corresponding iteration.
+
+\Cref{fig:a} shows how abruptly the semi-implicit formula becomes
+unstable as the frequency~$\omega_k$ increases through the limit
+of stability.  While this might be unfortunate from a practical
+point of view, it makes the formula useful for evaluating new ways
+to discretise the hamiltonian.  The largest~$\omega_k$ can be
+estimated by integrating the dynamics over a few time steps, and
+adjusting the time step to find the limit of stability.  Quantum
+states inspire some very clever methods of discretisation
+\cite{jcp-64-63,jcp-119-1289}, because the state of a quantum system
+with many degrees of freedom lies in a vector space of just as many
+dimensions, a space where any grid will have impossibly many points.
+It is not always straightforward to estimate the eigenvalues of the
+discrete hamiltonians that result from these methods.  In particular,
+the discrete hamiltonian can be a non-normal operator, in which
+case quite delicate arguments about pseudospectra would be required
+to rigorously determine the stability of the discretised dynamics.
+Running those dynamics, and looking out for parasitic solutions,
+is a fast and simple alternative.
 
@@
 
+When the equation being solved is simply Schr\"odinger's equation
+with repulsion, there are ways to modify the semi-implicit method
+and avoid the large energies.  The most commonly used is an interaction
+picture method \cite{2000-CaradocDavies-vortex}.  This is a split
+step method, where a Fourier transform~$\psi(t,k)$ is taken at each
+time step, and the kinetic energy term proportional to~$-i\nabla^2\psi$
+is evaluated exactly as~$\psi(t+h,k)=e^{ik^2h}\psi(t,k)$.  This
+ensures that~$\|\psi(t+h)\|=\|\psi(t)\|$, so that parasitic modes
+can not grow only because of their high kinetic energy.
+
+Interaction picture methods are very effective when they work.
+However, the nonlinear terms can also become large, and this occurs
+quite often in practice.  For example, it is often desirable to
+simulate a cold gas in a harmonic trap, with~$V(x)\propto x^2$.  In
+this case, the energy of the gas oscillates between potential and
+kinetic, and the potential energy of a particle far from the origin
+can be as great as the kinetic energy of a particle with large
+momentum.  This causes the equation to remain just as stiff even
+when the kinetic energy term is solved exactly.
+
 In a real simulation, the energy eigenstates are not known beforehand,
@@
 \section*{Acknowledgments}
-I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for suggestions that improved the presentation.
+I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for useful suggestions.
 

### 2017-12-21 VCM/phaseham.m
preserve phaseham with the quartic hamiltonian

@@ -1,4 +1,4 @@
-%PHASEHAM analyse the Kerr Hamiltionian discretised with wave packets
+%PHASEHAM analyse the wave packet discretisation of the number operator
 
 global N;  N = 70;  brackets
 
@@ -76,7 +76,7 @@ ylabel '<n>',  xlabel 'eigenkets of discrete number operator'
 figure
 errorbar(0:R-1, Ans, sqrt(Avarn), 'k'), hold on
 plot([0 R-1], [0 R-1], ':k')
-ylabel '<n>',  xlabel 'left SVs of expansion operator'
+ylabel 'n',  xlabel 'left singular kets of expansion operator'
 
 
 eyeR = eye(R);
@@ -125,8 +125,6 @@ xlabel 'real \lambda',  legend discrete exact Location SouthEast
 title 'pseudospectra of hamiltonian'
 
 
-return
-
 figure
 zplot(0:R-1, 0:R-1, Mkets'*Gkets), axis image
 title 'brackets of discrete N and H eigenkets'
@@ -152,11 +150,20 @@ for i = [1:5 (R-5):R]
 	title(['Discrete eigenstate number ' num2str(i-1)])
 	text(-9,6, {sprintf('<n> = %.1f', Mns(i)); sprintf('\\Deltan = %.1f', sqrt(Mvarn(i)))}, ...
 		'Color', 'white')
+end
 
+% Singular vectors of A
+
+for i = ixs
 	figure, zplot(x,y,Aps'*Au(:,i)), hold on, axis image
-	title(['Left singular ket number ' num2str(i-1)])
+	title(sprintf('Left singular ket number %d', i))
 	text(-9,6, {sprintf('<n> = %.1f', Ans(i)); sprintf('\\Deltan = %.1f', sqrt(Avarn(i))); ...
 		sprintf('\\sigma = %.1e', Asw(i))}, 'Color', 'white')
+	scale = 10/max(abs(Av(:,i)));
+	for j = 1:length(a)
+		plot(a(j), 'o', 'MarkerSize', scale*abs(Av(j,i)), ...
+			'MarkerEdgeColor', interp1(ff, cmap, mod(angle(Av(j,i)), 2*pi)))
+	end
 end
 
 figure, subplot 311

### 2018-01-08 jcp-0-0.tex
start drafting the static wave packets paper

@@
 \begin{abstract}
-% insert abstract here
+The ability to expand high-energy states over a set of non-orthogonal wave packets causes the wave packet equations of motion to become stiff.
+On the other hand, a dense set of wave packets is required to have stable expansions of wave functions.
+Tychonov regularisation of the expansion operator suppresses the energies of the high-energy states, while retaining all the physics that happens slower than a given time scale.
 \end{abstract}
@@
 Since Heller first proposed the method in 1976\cite{jcp-64-63}, there have been several attempts to compute the dynamics of quantum systems by use of coupled Gaussian wave packets.
+
+Schemes such as \cite{jcp-132-244111,jcp-136-014109} have used moving grids of coherent states, to which all these considerations would apply.
 

### 2018-01-08 VCM/phasegrid.m
neaten up fock state plots

@@ -160,33 +160,21 @@ ixs = [1:4 round(linspace(5,length(a),4))];
 cmap = phase(128);
 ff = linspace(0, 2*pi, size(cmap,1));
 
-% Singular vectors of A
-
-for i = ixs
-	figure, zplot(x,y,Aps'*Au(:,i)), hold on, axis image
-	title(sprintf('Singular vector %d', i))
-	text(-9,8,sprintf('\\sigma = %.1e', Asw(i)), 'Color', 'white')
-	scale = 10/max(abs(Av(:,i)));
-	for j = 1:length(a)
-		plot(a(j), 'o', 'MarkerSize', scale*abs(Av(j,i)), ...
-			'MarkerEdgeColor', interp1(ff, cmap, mod(angle(Av(j,i)), 2*pi)))
-	end
-end
 
 % Eigenvectors of M
 
 [~, P] = sort(abs(ew));
-for i = ixs
-	figure, zplot(x,y,Aps'*A*ev(:,P(i))), hold on, axis image
-	title(sprintf('Discrete H eigenvector %d', i))
-	text(-9,8,sprintf('ew = %.1e', ew(P(i))), 'Color', 'white')
-	text(-9,7,sprintf('|A*ev| = %.1e', norm(A*ev(:,P(i)))), 'Color', 'white')
-	scale = 10/max(abs(ev(:,P(i))));
-	for j = 1:length(a)
-		plot(a(j), 'o', 'MarkerSize', scale*abs(ev(j,P(i))), ...
-			'MarkerEdgeColor', interp1(ff, cmap, mod(angle(ev(j,P(i))), 2*pi)))
-	end
-end
+% for i = ixs
+% 	figure, zplot(x,y,Aps'*A*ev(:,P(i))), hold on, axis image
+% 	title(sprintf('Discrete H eigenvector %d', i))
+% 	text(-9,8,sprintf('ew = %.1e', ew(P(i))), 'Color', 'white')
+% 	text(-9,7,sprintf('|A*ev| = %.1e', norm(A*ev(:,P(i)))), 'Color', 'white')
+% 	scale = 10/max(abs(ev(:,P(i))));
+% 	for j = 1:length(a)
+% 		plot(a(j), 'o', 'MarkerSize', scale*abs(ev(j,P(i))), ...
+% 			'MarkerEdgeColor', interp1(ff, cmap, mod(angle(ev(j,P(i))), 2*pi)))
+% 	end
+% end
 
 % expansions of number states
 

### 2018-01-08 VCM/phaseham.m

@@ -27,6 +27,7 @@ G = pinvA*diag(nhn)*A;
 % Eigenvalues are stored as vectors
 % Eigenvectors are stored as columns, in order of Fock number
 % Suffix convention: ev, ew, sw, u = lsv, v = rsv
+%     c = expansion coefficients
 % The discrete hamiltonian has eigenkets in Hilbert space, 
 % and eigenvectors in C^R.  These are also the eigenkets of the
 % discrete hamiltonian.
@@ -44,13 +45,8 @@ Mns = sum(diag(0:N)*abs(Mkets).^2);
 Mew = Mew(P);  Mev = Mev(:,P);  Mkets = Mkets(:,P);
 Mvarn = sum(diag((0:N).^2)*abs(Mkets).^2) - Mns.^2;
 
-% The Mev turn out identical for large n, and numerically better
-% So we'll use those instead.
-[Gev,~] = eig(G);
-Gkets = A*Gev;  Gkets = Gkets / diag(norms(Gkets));
-Gns = sum(diag(0:N)*abs(Gkets).^2);
-[Gns, P] = sort(Gns);  Gkets = Gkets(:,P);
-Gew = diag(Mev'*G*Mev);
+nr = zeros(N+1,R);  nr(1:R,:) = diag(0:R-1);
+fockc = pinvA*nr;  fockc = fockc / diag(norms(fockc));
 
 % Tychonov regularised H
 Hy = pinv([A; laf*eye(R)])*[diag(nhn)*A; zeros(R)];
@@ -69,81 +65,41 @@ ff = linspace(0, 2*pi, size(cmap,1));
 % Graphs, as described in titles.
 
 figure
-errorbar(0:R-1, Mns, sqrt(Mvarn), 'k'), hold on
-plot([0 R-1], [0 R-1], ':k')
-ylabel '<n>',  xlabel 'eigenkets of discrete number operator'
-
-figure
 errorbar(0:R-1, Ans, sqrt(Avarn), 'k'), hold on
-plot([0 R-1], [0 R-1], ':k')
+plot([0 R-1], [0 R-1], ':k'),  xlim([0 45])
+title 'ranges of Fock kets in expansion singular kets'
 ylabel 'n',  xlabel 'left singular kets of expansion operator'
-
-
-eyeR = eye(R);
-x2 = 0:0.1:R+10;  y2 = -3:0.1:3;
[…]

### 2018-05-17 cohrep.m
add axis labels to phase space plots

@@ -32,6 +32,7 @@ for i = 1:length(a)
 	imagesc(x([1 end]), y([1 end]), rsdl),  colormap gray, axis equal
 	hold on,  plot(a{i}, 'ow')
 	[C,H] = contour(x,y,nrms, [1.5 5 20 100 300],'-r');  clabel(C,H,'Color','red')
+	xlabel 'Re \beta', ylabel 'Im \beta'
 	ax = gca;  ax.CLim = [0 1];
 	saveTightFigure(sprintf('cohrep%d.pdf', i))
 end

### 2018-05-17 gaussian.tex
Gaussian wave packet literature review

\input respnotes

\def\H{{\bf H}}

\title A review of variational Gaussian wave packets

This will summarise the work that has been done from the time of Heller to the present day.  It will take the viewpoint of frames and stability.

This literature is disjointed, e.g. there are 20 page papers with “coherent state” in the title, that never mention Roy Glauber \cite{cpx-304-103}.  Name games: Gaussian wave packets vs.\ coherent states vs.\ Gabor frames vs.\ windowed Fourier transforms.

Schrödinger's equation for a phase space wave function is \cite{cpx-304-103}
$$i\hbar{d\over dt}〈α|ψ(t)〉={1\over π}∫〈α|β〉H_n(α*,β)〈β|ψ(t)〉\,d²β.$$
This is an integro-differential equation, not just a differential equation.

Notation.  A superposition of~$r$ multimode wave packets, each with~$n$ modes, is~$r×n$.  (The~$×n$ can be omitted for a single mode.)  MCTDH matrices use~$⊗$, so a full tensor product of wave packets over 3 modes might be~$2⊗2⊗2$, or~$2³$.  (Evaluating these as integer expressions gives you a count of the parameters.)  A Hilbert space for a single degree of freedom, expanded over~$r$ basis states that are not gaussian wave packets, is written~$r×\H$.  So the normal MCTDH has structure~$(n₁×\H)⊗…⊗(n_f×\H)$.  An arbitrary expansion is~$\H$, so~$\Hⁿ$ is a general state with~$n$ degrees of freedom.

\section Dirac and Frenkel

Dirac-Frenkel dynamics necessarily conserves the norm of the ket and~$〈H〉$.  Dirac-Frenkel variational coefficients conserve the norm, whatever the amplitudes are forced to do \cite{jcp-136-014109}.

Kay explains that there are several forms of the variational principle
\cite{cpx-137-165}.  In all cases, there is a ket~$|ψ(z)〉$ that
sits in a complex vector space, and depends on variational
parameters~$z$.  In principle, we always want to select~$dz$ so
that~$|ψ(z+dz)〉=|ψ(z)〉+|dψ〉$ in a Hilbert-space least squares
sense.  This occurs when the residual~$i\hbar|ψ'(z)〉·dz-H|ψ(z)〉·dt$
is Hilbert-space orthogonal to the column space of~$|ψ'(z)〉$.  The
simplest case is when~$z$ is complex, and~$|ψ〉$ is analytic in
each parameter, and orthogonality reduces
to
$$〈D_iψ(z)|\left(i\hbar|ψ'(z)〉·dz-H|ψ(z)〉·dt\right)=0\eqno{\tag\eqi}$$
for all~$i$.  This is how Frenkel did it \cite{1934-Frenkel-Wave}, and he's OK.

This doesn't work when~$z$ is real, and therefore~$dz$ is real.
Taking real and imaginary parts, there are twice as many Equations~\eqi\
as there are unknowns.  What we should do is take real and imaginary
parts of the residual and the~$|D_iψ(z)〉$ as expanded over some
orthonormal basis, to get a real least squares problem where more
equations than unknowns is business as usual.  This sensible approach is developed by Raab \cite{cpl-319-674}.

What the 
inventors of quantum mechanics did instead was to arbitrarily demand that
we set the real part of Equations~\eqi\ to zero following
the “least-action form” \cite{cpx-137-165[3]}, or their imaginary
parts to zero following McLachlan \cite{cpx-137-165[2]}.  This has
led to lots of point scoring and name calling regarding which is
better.

\section Heller

\section Heller's followers

Much work was done with~$1×1$ structures, because this is easy to compute.  The independent Gaussian approximation does an ensemble of~$1×1$ simulations, with different initial conditions.  The local harmonic approximation truncates the potential around each wave packet to a second-order Taylor series, whence Gaussian wave packets retain their form.

Heather and Metiu \cite{cpl-118-558} treat a Morse oscillator using a~$1×1$ approximant.  They find that the LHA gives a different result than a variational method.  Their approximant includes some real paramters, but they claim to minimised the error, implying they did least squares over real and imaginary parts.

\section MCTDH

MCTDH is \dots  It was introduced bye Meyer, Manthe and Cederbaum \cite{cpl-165-73}.  They note that there is a redundancy if the number of single-particle states in one degree of freedom exceeds the product over the other degrees of freedom, because the relative states can not be linearly independent in that case.

The “stays orthogonal” constraint can be expressed as the single-particle modes in each subspace obey some Hamiltonian dynamics.

The original idea \cite{jcp-111-2927} was to treat the system with normal MCTDH, and the bath with a multimode Gaussian wave packet, with structure~$(n₁×\H)⊗…⊗(n_{f-1}×\H)⊗(n_f×m)$.  Motivated by molecules in solution.  Assume the solvents have a small range of motion, so that their potential is nearly harmonic.  This paper derived the variational equations for a density matrix and master equation.

Worth and Burghardt \cite{cpl-368-502} combined variational wave packets with MCTDH.  The idea was simply to use gaussian wave packets as the orbitals~$φⁿ_i$ in that method.  (Note that this arranges the tensor products differently than in the multimode coherent state version.)  The hope was to use multidimensional wave packets, with each set of parameters covering several degrees of freedom.  This was applied to a Henon-Heiles potential, and reproduced a numerically exact autocorrelation function.  It's unclear why this was numerically stable.

This used structures~$(5×2)²$ to~$(25×2)²$, in a phase space of volume~?.
{\bf Check out who cited it.}

This managed to superpose 625 wavepackets in 4D space.  {\bf How did they deal with the lack of orthogonality?}  Wave packets were deleted when they escaped to large distance.

Noted that tunnelling happens by lots of energy being transferred to a subset of the wave packets.

Another application in \cite{jcp-119-5364}.

\section Static grids

cpl-318-305
jcp-116-2748
jcp-87-5302
jcp-89-4752
jcp-90-7317
jcp-115-1158

\section Frozen trajectories

Habershon \cite{jcp-136-014109} deleted redundant wave packets to maintain full rank, used Eherenfest type trajectories.

Saleshin and Child \cite{cpx-304-103} propagated an ensemble of coherent amplitudes by treating the normally-ordered Hamiltonian as a classical Hamiltonian.  They argue in Section~4 that when~$α(t)$ follows a classical trajectory governed by that Hamiltonian, the wave function has the form~$〈α(t)|ψ(t)〉=C(α(t))e^{iS_α/\hbar}$, where~$S_α$ is the classical action along the trajectory~$α(t)$.  The fuction~$C(α(t))$ obeys a partial differential equation with small and local coupling.

jcp-113-10028
jcp-132-244111

\section Regularisation

MCTDH was orthogonalised from the start \cite{cpl-165-73}, ostensibly to make the parameters change quickly and the expansion functions change slowly, and to simplify the mechanics.  They noted that their expansion matrix could become singular, and used Tychonov regularisation as well as truncating singular values.  They state that most work has used two degrees of freedom, and use a Henon-Heiles potential as a numerical example.

Kay cpx-137-165 thought about SVDs and rank-deficiency, and noted that truncating the small sws and taking a pseudoinverse helped in some cases.  He didn't directly consider gaussian wave packets.  Follow up his refs and cites.  (This is actually a discrete problem, not a rank-deficient one 2010-Hansen-Discrete).

Habershon \cite{jcp-136-014109} propagated an ensemble of wave packets using Eherenfest type trajectories, but forced the variational coefficients of the redundant ones to zero.  He substituted packets in and out as the condition of the expansion operator changed.  This paper treated a double well potential.

The initial amplitudes were sampled from the Wigner distribution for the initial state, no justification for why it wasn't Q.  For a quartic oscillator state, Wigner sampling would generate lots of wave packets that don't overlap the state.  This used a pseudoinverse that truncated small singular values.  Habershon's results diverge in a very similar way to our quartic oscillator results.  He attributes this to conservation of energy failing, but he doesn't think through the stiffness and high energy state ideas.  His stabilisation approach works.  He claims that it works because he is constantly re-expanding the state over a new set of active wave packets.  It's plausible that a parasitic solution in one expansion is well-behaved in the next expansion.

{\bf Discuss section III of jcp-136-014109 in detail.}

\section Index of potentials treated

\section To look at

2014-Grabowski-Review
jcp-83-3009
acr-14-368
1968-Klauder-Fundamentals
2014-Ali-Coherent
cpx-304-103
Herman, Annu. Rev. Phys. Chem. 45, 83.
pcp-9-3210
mop-106-2145

These have yet to be categorised:

https://doi.org/10.1080/00268978600100551
https://doi.org/10.1090/S0025-5718-04-01685-0 
https://link.springer.com/article/10.1007%2Fs00791-006-0019-8?LI=true
ijc-47-75
jcp-110-5526
jcp-111-2927
jcp-112-6113
jcp-117-4738
jcp-119-1961
jcp-119-5364
jcp-121-9247
jcp-128-054102
jcp-129-174104
jcp-130-134113
jcp-136-014109[1:17]
jcp-62-1544
jcp-64-63
jcp-65-4979
jcp-67-2017
jcp-71-3383
jcp-75-246
jcp-75-2923
jcp-80-3123
jcp-83-3009[1:10]
jcp-84-227
jcp-84-3250
jcp-84-6293
jcp-85-4129
jcp-87-5987
jcp-90-3060
jcp-91-170
jcp-93-3919
jcp-94-2723
jcp-96-4266
jpa-19-2559
njp-10-115005
pna-14-187
pra-13-2226
pra-73-033619
prb-73-125112
prx-131-2766
prx-177-1882
rmp-2-221
ssc-31-3027
xxx-physics-0510076
cpl-118-558[1:20]
cpl-368-502[1:7 10 11 12 13:15 21]
jcp-111-2927[12]
cpx-137-165[3]
cpx-137-165[2]

\section Rejected as irrelevant

jcp-117-4738 (only one wave packet, $r=1$)

### 2018-07-03 ssc-0-0.tex
add a conclusion to the draft semi-implicit paper

@@
 % SIAM Article Template
+% This needs to be compiled with LaTeX and dvipdf.  XeTeX doesn't work.
 \documentclass[review]{siamart1116}
@@
 The iterative semi-implicit formula, or some variant of it, is often
-used to solve stochastic partial differential equations that arise
+used to solve stochastic partial differential equations that form part of Monte-Carlo methods used
 in quantum optics and in atom optics.  When this formula is applied
@@
 % iterations
-For a system of equations
+For a system of ordinary differential~equations
 \begin{equation}
@@
 	\qquad\text{where}\qquad 
-    f_{n+\frac12}=f\left({\textstyle\frac{s_n+s_{n+1}}{2}}\right).
+    f_{n+\frac12}=f\left(\frac{s_n+s_{n+1}}{2}\right).
 \end{equation}
@@
 evaluates Equation~\cref{eq:i}, and updates the
-values of~$s_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the
-iterations generate a sequence of~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
-with
+values of~$s_{n+1}$ and~$f_{n+\frac12}$.  This produces sequences~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
+where
 \begin{subequations}
+\label{eq:v}
 \begin{align}
-	\label{eq:iv} s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
-	\label{eq:v} s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
-		\qquad f_{n+\frac12}^{(j+1)}=f\left({\textstyle\frac{s_n+s_{n+1}^{(j)}}{2}}\right)\qquad \text{when $j\ge0$}.
+	s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
+	s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
+		\qquad f_{n+\frac12}^{(j+1)}=f\left(\frac{s_n+s_{n+1}^{(j)}}{2}\right)\qquad \text{when $j\ge0$}.
 \end{align}
@@
 This continues until the sequences converge, which is usually
-achieved after a small, fixed number of iterations.  Any fixed point
+achieved by a small, fixed number of iterations.  Any fixed point
 of the iteration satisfies
-Equation~\cref{eq:i}.  The remaining question is under what
+Equation~\cref{eq:i}.  The question is under what
 conditions the iterations converge.
 
-% steps stability
 Generally, when a partial differential equation is discretised by
@@
 
+{\bf Make this less abrupt.}
 This paper analyses the semi-implicit formula, in the case that it
@@
 depends on a Nyquist-like condition, and the iterations converge
-if and only if the eigenvalues of the hamiltonian are bounded by the reciprocal time step~$2/h$.
+if and only if the eigenvalues of the Hamiltonian are bounded by the reciprocal time step~$2/h$.
 
@@
 
-% basis coefficients
-When the semi-implicit rule is applied to \cref{eq:vi}, or a
-stochastic or nonlinear variant of it, the laplacian operator is usually
+When the semi-implicit formula is applied to Equation~\cref{eq:vi}, or a
+stochastic or nonlinear variant of it, the Laplacian operator is usually
 evaluated by a spectral method \cite{2000-Trefethen-Spectral}.  This is equivalent to expanding
@@
 \end{equation}
+{\bf We're thinking about spectral collocation methods, with a delta function in place of~$\phi_m(x)$.
+In the band-limited case, the collocation value is the same as the sinc component.
+Explain this more fully.}
 
@@
 projected onto the complex plane spanned by the eigenvector~$v_k$.
-The exact solution that satisfies Equation~\cref{eq:iii}, with
-initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
-and its projection lies on the circle around the origin of
-radius~$|c_k|$.  Here~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.
+The exact solution to Equation~\cref{eq:iii}, with
+initial condition~$s_n$ at time~$t_n$, is~$s(t_n+t)=\sum_k c^ke^{-i\omega_k(t_n+t)}v_k$,
+where~$\omega_k$ is the eigenvalue of~$H$ corresponding to the eigenvector~$v_k$.   Its projection lies on the circle of
+radius~$|c_k|$ centred on the origin.
 
@@
   \includegraphics{semii.1}
-  \caption{A step of the semi-implicit formula, projected onto the
-  complex plane spanned by the eigenvector~$v_k$ of~$H$.  The
-  computed solution~$s_{n+1}$ lies on the same circle as the
-  exact solution~$s(h)=\exp(-ihH)s_n$, but it rotates with angular frequency lower
-  than~$\omega_k$.  See text for details.}
+  \caption{The numerical solution of Schr\"odinger's equation over
+  a single time step~$h$, given by the ideal semi-implicit formula
+  of Equation~\cref{eq:i}.  The trajectory of the discretised wave
+  function~$s$ has been projected onto the complex plane spanned
+  by the eigenvector~$v_k$ of the Hamiltonian operator.  The computed
+  solution~$s_{n+1}$ follows the same circle as the exact
+  solution~$s(t)=\exp(-iH(t-t_n))s_n$, but it rotates with a lower angular
+  frequency than the eigenvalue~$\omega_k$.}
   \label{fig:iii}
@@
 
-Suppose that the point~$s_{n+1}$ also lies on this circle.
-Then the projection of~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
-chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
-brought towards and away from the origin by moving~$s_{n+1}$ around
-the circle, so there is some position for
-which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, so that~$s_{n+1}$
-satisfies Equation~\cref{eq:i}.  The only error introduced by discretisation
-is that~$c_k$ rotates with an angular frequency less
-than~$\omega_k$, bounded by~$\pi/h$ as~$\omega_k$ becomes
-large.
+Suppose that the point~$s_{n+1}$ also lies on this circle, forming
+a chord with~$s_n$.  In the projection shown, the point~$s_{n+\frac12}$
+is the midpoint of the chord.  The
+vector~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ points along the
+chord.  Its magnitude is determined by how closely the chord passes
+the origin, a distance that can be reduced at the same time
+that~$|s_{n+1}-s_n|$ is increased by moving~$s_{n+1}$ further around
+the circle.  There is always some position for
+which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, satisfying Equation~\cref{eq:i}.
+The only error introduced by discretisation is that~$c_k$ rotates
+with an angular frequency less than~$\omega_k$, bounded by~$\pi/h$
+as~$\omega_k$ becomes large.
 
-In practice, the iterations of~Equation~\cref{eq:vi} must be used
-instead of~Equation~\cref{eq:i}, and this is less stable.  For
-Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
+In practice, the iterations of~Equation~\cref{eq:v} must be used
+instead of~Equation~\cref{eq:i}, and this is less stable.  With the
+assumption of Hamiltonian dynamics, the
 iteration rule becomes
@@
 s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
-	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
+	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}\label{eq:x}
 \end{equation}
@@
 \end{equation}
-The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
-eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
+The time step matrix~$U^{(j)}$ has the the same
+eigenvectors~$v_k$ as the Hamiltonian matrix~$H$.  Each of its
 eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
@@
   \includegraphics[width=0.7\textwidth]{wev.eps}
-  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.
-The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.
-The exact time evolution operator~$U$ has unimodular eigenvalues.}
+  \caption{The eigenvalue~$\omega_k$ of the Hamiltonian operator is compared to the size of the eigenvalue~$u_k^{(j)}$ of the time step operator~$U^{(j)}$ for $j$~iterations of the semi-implicit formula of Equation~\cref{eq:x}.
+The exact time step operator has unimodular eigenvalues.
+The lightest line shows~$j=1$ and the heavier ones show~$j=2$, 3, and 4.
+The symbol~$h$ represents the time step, not Planck's constant.}
   \label{fig:ii}
@@
 The geometric series for~$U^{(j)}$ will converge if and only if the
-angular frequencies~$\omega_k$ are bounded by the reciprocal time
-step~$\frac2h$, a kind of Nyquist condition.  If the spectral
-hamiltonian matrix has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
-will enlarge the component of~$v_k$ in~$s_n$ by a factor close
-to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
-thus a parasitic solution growing at the
-rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.
+eigenvalues~$\omega_k$ are bounded by the reciprocal time
+step~$\frac2h$, a kind of Nyquist condition.  If the spectral {\bf or discrete}
+Hamiltonian matrix has a large eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
+will grow the component of~$v_k$ in~$s_n$ by a factor close
+to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component becomes a parasitic solution growing at the
+rate
+\begin{equation}
+\lambda_k=\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right),\label{eq:xi}
+\end{equation}
+so the semi-implicit formula is unstable unless the~$\omega_k$ are bounded by~$\frac2{h}$.
 
@@
 	{\partial\psi\over\partial t}=-i(\nabla^2+V+g|\psi|^2)\psi.
+	label{eq:xii}
 \end{equation}
@@
 \end{equation}
-where~$\hat H'(\psi(t))$ is an effective hamiltonian operator for
-the field~$\delta\psi$.  This linearisation has been studied
-extensively, and the dynamics generated by the hamiltonian~$\hat
-H'$ are Bogoliubov sound waves \cite{aop-70-67}.  When this equation
-is discretised with the semi-implicit formula, the stability condition
-is that the Bogoliubov sound wave modes on a spatial grid of
-spacing~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
+where~$\hat H'(\psi(t))$ is an effective Hamiltonian operator for
+the field {\bf wave function?}~$\delta\psi$.  This linearisation has been studied
+extensively: the dynamics generated by the Hamiltonian~$\hat
+H'$ are Bogoliubov sound waves \cite{aop-70-67}.  The stability condition when this equation
+is discretised with the semi-implicit formula
+is that the Bogoliubov sound wave modes on a grid of
+step~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
 fine grid, the high-frequency Bogoliubov modes have the same
@@
   \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
-  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
+  \caption{The growth of parasitic solutions when the dynamics of a quantum Kerr oscillator are computed with 4 iterations of the semi-implicit formula of Equation~\cref{eq:x}.
+The coefficients~$s_n$ of the Hamiltonian eigenvectors should have constant sizes, shown by the parallel curves at~$t=0$ and~0.09.
+The computed values are shown by the rulings between those lines.
+Their sizes are constant for small~$n$, to the right of the dotted line at the stability limit~$\omega_n=2/h$.
+Above the limit, the coefficients grow exponentially.
+The growth rates predicted by Equation~\cref{eq:xi} are shown by the dashed line.}
   \label{fig:i}
@@
 
-\section{Conclusion}
+\section{Discussion}
 
@@
 
+\section{Conclusion}
+
+For a wave function that follows Schr\"odinger's equation, the stability of the iterated semi-implicit formula is determined by the largest eigenvalue~$\omega$ of the Hamiltonian.
+The formula is stable if~$\omega<\frac2{h}$, where~$h$ is the time step.
+Otherwise, the eigenvector corresponding to~$\omega$ is a parasitic solution, whose rate of growth is stated in Equation~\cref{eq:xi} for the case~$\omega\gg\frac2{h}$, and can be read from from \cref{fig:ii} in general.
+
+\Cref{fig:ii} suggests the performance of the formula when~$\omega\approx\frac2{h}$.
+The discrete time step operator~$U^{(j)}$ differs significantly from the exact operator~$U$ for~$\omega>\frac1{h}$.
+If the high-frequency dynamics are physically significant, a time step should be chosen well inside the limit of stability.
+
+When the dynamics include a nonlinear term, as in Equation~\cref{eq:xii}, stability is determined by the largest eigenvalue of the linearised Hamiltonian about the solution.
+In the common case of a Gross-Pitaevskii~equation, with grid spacing small compared to the healing length, the nonlinear term will have no effect on stability.
+
+The stability limit is a kind of Nyquist condition, which makes it especially easy to reason about stability.
+Physically, it is utterly reasonable for a simulation to require a time step on the characteristic time scale of the dynamics being simulated.
+
 \section*{Acknowledgments}
